### 2020-12-18 response-1.tex
add overleaf paper to github

\documentclass{article}
\usepackage{fullpage}
\usepackage[utf8]{inputenc}

\title{Review Response JDS2007-001RA0}

\date{Revision due: 2020-11-24}

\begin{document}

\maketitle

\parindent 0pt We are happy to resubmit our paper and are very thankful for the great reviews we received. Thanks to the review comments, the quality of the paper has significantly improved.

In particular, we addressed the following:

\begin{enumerate}

\item The paper has been checked multiple times by multiple native English speakers.

\item The organization of the paper has been improved while focusing entirely on the deep learning framework. Several sections and images that distracted the reader from the paper's primary focus (which was evident by the careful reviews) were removed.

\item Reviewer 1: The method has been described using a recipe approach without giving the rationality behind it. There are too many predictive models for COVID-19. What are the advantages of the proposed method? The authors should explain their competitiveness, at least from the following two perspectives. We addressed these issues through the next two items.

\item Reviewer 1: First, compared with the very classical SIR/SEIR and their improved models, what are the advantages of the proposed model?

    A new section has been added to discuss the advantages and disadvantages.

\item Reviewer 1: Is the accuracy improved significantly?

    While using the real-time data our accuracy is very good as it, for example accounts for the weekly fluctuations that are not covered by other prediction models. Moreover, the point is not only a single model in time, but that the model adapts itself over time while new data is added. This has been pointed out in the paper.

\item Reviewer 1: Second, it seems that the data used are all very structured, what about regression model framework?

    The beauty of this deep learning algorithm is that it learns the trends and performs well even if the data is not smoothed with regression or a seven-day average. Instead, it even learns the weekly fluctuations which we point out in the revised version of the paper.

\item Reviewer 1: Is the performance competitive?

    The performance is very competitive and an explanation is added to the conclusion.

\item Reviewer 1: The current deep learning framework is lack of interpretation of the investigated risk factors.

    We added the following section:

While this study focuses on if and which risk factors can improve the predictions, we do not focus on why. Such an analysis could provide insights into causal relationships. For example,  information such as health insurance and the number of hospitals and beds are often related to the quality of services that affect early detection and treatment. Furthermore, population parameters such as density can determine the rate of spread and the number of cases. Similarly, distributions of comorbidities in a given population such as physical health, cholesterol screening, diabetes, etc., are found to be insightful in predicting outcomes. Identifying such risk factors and correlating them to the prediction accuracy provides valuable candidates for future studies.

\item Reviewer 2: The manuscript is in a rather preliminary shape, and needs significant work before being in a publishable shape.

    We believe we have addressed this issue by addressing all reviewer's comments and more.

\item Reviewer 2: Section 2.2 does not look sufficient as a subsection.  It should be either expanded to include more details, or combined with Section 2.1.3.

    The sections have been improved/combined, and a section from later on in the paper was added.

\item Page 5, under "Data Source Integration", what does "we include integrate data integration capabilities" mean?

    The sentence has been corrected.

\item Reviewer 2: 4.  Please consider writing the requirements on Pages 5 and 6 into structured paragraphs.

    The requirements have been rewritten as paragraphs. Also, we added a numbered requirement tag to them and kept the high-level name of the requirement for increased readability and structure.

\item Reviewer 2: Item 5., 6., 7., 8., 9., and 10. Have been fixed by excluding the text related to them from the paper as we found they distracted too much from our main contributions.

\item Reviewer 2: 12.  In Section 2.1,  the activation function is the sigmoid function.  In Section 6, RELU was used.

    Thanks for detecting it. We used in both approaches sigmoid only in the last layer and corrected it in the paper

\item Reviewer 2: 13.  Please consider adding labels for the yaxes in Figure 10 and Figure 11.

    All images now have information about the axes.

\item Reviewer 2: 14.  Page 19, please fix the broken link to reference.

    Broken links were fixed.

\item Reviewer 2: 15.  Sections 7 and 8 are rather redundant.  The language does not read clear.  I suggest moving these two sections, and make Section 5 and 6 a more complete story.

    Thanks for this suggestion. To better showcase our arguments we have done the following: (a) most of the theory was moved into the same section in order to reduce redundancy, (b) we introduced a number of Questions that we analyze within this report to provide additional structure and delineate the sections which answer each question, (c) a more comprehensive explanation has been added the later sections. Together they provide a more complete story.

\item Reviewer 2: 16.  Please consider making the code publicly available at GitHub.

    We intend to make the code public after publication. More than 30K lines of underlying infrastructure code and documentation is already available on GitHub and citations have been added.

\end{enumerate}

\end{document}

### 2020-12-18 deprecated/jds-paper-todo.tex
add overleaf paper to github

\section{todo}

We might want to have two documents; one with "everything" -- the other with highlights referencing first

\begin{verbatim}

In conclusions/highlights we could include

a) Data sample

b) simple remarks on data such as correlations between demographics and 
   that case/fatality data is quite rough and this limits insights

c) magic fits illustrate that DL can get good results when input data smooth

d) basic DL with just case and death data

e) DL+Demographics with clear but modest improvement limited (probably) by data quality

f) technical issues in DL

    Comparison with other DL for time series

    Errors and why use sqrt(count)

    Alternative normalization by population

    Method to combine case/death with demographics

    Calculating future values and custom loss function

    Dropout

    Hyper-parameter search

g) Sensitivity and other discussions of demographics 
h) Futures

What is related work?

\end{verbatim}

\subsection{Geoffrey}

...

### 2020-12-18 deprecated/jds-paper-v2.tex
add overleaf paper to github

\documentclass[12pt]{article}

\usepackage{amsfonts,amsmath,amssymb,amsthm}
\usepackage{array}
\usepackage{authblk}
\usepackage{color}
\usepackage[table]{xcolor}
\usepackage{epsfig,fancyhdr}
\usepackage{geometry,graphicx}
\usepackage[]{hyperref} 
\usepackage{indentfirst}
\usepackage{mathtools}
\usepackage{multirow}
\usepackage{listings}
\usepackage{natbib}
\usepackage{rotating}
\usepackage{comment}
\usepackage{booktabs}
\usepackage{todonotes}
\setuptodonotes{inline}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{sectsty,secdot}
\sectionfont{\bf\fontsize{14}{2em}\selectfont}
\subsectionfont{\bf\fontsize{12}{2em}\selectfont}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\geometry{left=2.5cm,right=2cm,top=2.5cm,bottom=1.5cm}
\setlength{\parindent}{0.63cm}

\setcounter{page}{1}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}
\newtheorem{proposition}{Proposition}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}
\newtheorem{remark}{Remark}
\pagestyle{empty}

\setcitestyle{square}

%---------page header---------%

%\pagestyle{fancy}
%fancyhead[RO]{odd number page}
%fancyhead[LE]{even number page}

%-----------------------------%
\newtheorem{question}{Question}

\newtheorem{requirement}{Requirement}

\renewcommand{\_}{%
    \textunderscore\hspace{0pt}%
}
\setlength{\emergencystretch}{3em}

\begin{document}

% \input{response/response}

\clearpage

\title{{\bf\fontsize{14pt}{18pt}\selectfont AICov: An Integrative Deep Learning Framework for COVID-19 Forecasting with Population Covariates}}
\author[1]{{\fontsize{12pt}{0.5em}\selectfont Geoffrey C. Fox}}
\author[1]{{\fontsize{12pt}{18pt}\selectfont Gregor von Laszewski}}
\author[1]{{\fontsize{12pt}{0.5em}\selectfont Fugang Wang}}
\author[2,3]{{\fontsize{12pt}{0.5em}\selectfont Saumyadipta Pyne}}

\affil[1]{{\emph\fontsize{12pt}{0.5em}\selectfont Digital Science Center, Indiana University, Bloomington Indiana, USA}}
\affil[2]{{\emph\fontsize{12pt}{0.5em}\selectfont Public Health Dynamics Lab, and Department of Biostatistics, University of Pittsburgh, Pittsburgh, Pennsylvania, USA}}
\affil[3]{{\emph\fontsize{12pt}{0.5em}\selectfont Health Analytics Network, Pennsylvania, USA}}

\newcommand{\ABSTRACT}{
The COVID-19 pandemic has had profound global consequences on health, economic, social, behavioral, and almost every major aspect of human life. Therefore, it is of great importance to model COVID-19 and other pandemics in terms of the broader social contexts in which they take place. 
We present the architecture of AICov, which provides an integrative deep learning framework for COVID-19 forecasting with population covariates, some of which may serve as putative risk factors. 
We have integrated multiple different strategies into AICov, including the ability to use deep learning strategies based on LSTM and event modeling.
To demonstrate our approach, we have introduced a framework that integrates population covariates from multiple sources. Thus, AICov not only includes data on COVID-19 cases and deaths but, more importantly, the population's socioeconomic, health, and behavioral risk factors at a local level. The compiled data are fed into AICov, and thus we obtain improved prediction by the integration of the data to our model as compared to one that only uses case and death data. As we use deep learning our models adapt over time while learning the model from past data.
}
\date{}

\newcommand{\TODO}[1]{
\todo[color=yellow!20]{#1}
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\renewcommand{\baselinestretch}{1.07}
\renewcommand{\thefootnote}{}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\maketitle
\thispagestyle{empty}

\begin{quotation}
\centerline{\bfseries{\fontsize{14pt}{1em}\selectfont Abstract}}\par
\setlength{\parindent}{3.22cm}
{\fontsize{12pt}{1.07em}\selectfont~~\ABSTRACT}

\vspace{12pt}
\noindent {\bfseries{Key words and phrases:}}
COVID-19, deep learning, prediction, risk factors, comorbidities.
\par
\end{quotation}\par

\def\thefigure{\arabic{figure}}
\def\thetable{\arabic{table}}

\renewcommand{\theequation}{\thesection.\arabic{equation}}

\fontsize{12}{1.07em}\selectfont

\setcounter{section}{0} %***
\setcounter{equation}{0} %-1

\section{Introduction}

The COVID-19 pandemic has caused an enormous health and humanitarian crisis worldwide. It is unlike any other single phenomenon that has occurred in modern history since the end of World War II. 
The pandemic’s effects have spanned over a range that is so vast over space, and yet so condensed over time, that the dual blows of intensity and rapidity have exposed myriad systemic vulnerabilities in many societies around the world.

The first known case was traced back to  17  November  2019. Since the emergence of a cluster of cases in Wuhan, China, on 31 December 2019, COVID-19 has spread rapidly worldwide. Just six months later, as of 23 June 2020, there are 8,993,659 COVID-19 cases and 469,587 deaths globally. On 30 January 2020, the World Health Organization (WHO) \cite{www-who} declared COVID-19 to be a Public Health Emergency of International Concern and subsequently, on 11 March 2020, a pandemic.

In the United States, the first case of COVID-19 was confirmed on 20 January 2020 in Washington State, and as of 23 June 2020, there were over 2.3 million confirmed cases and 121,167 deaths \cite{www-nyt-map}. 
Although initial efforts to reduce the spread took place, recent data shows a resurgence at an increased scale. In the absence of a vaccine or treatment to effectively combat the disease, non-pharmaceutical policy interventions such as social distancing and lockdowns are recommended by health experts to prevent further transmission. To gain insights into the possible impact of such measures on COVID-19 outcomes, we depend upon the ability to accurately forecast the spread of reported cases and confirmed deaths and recoveries. Naturally, the accuracy of forecasting relies on the availability of current, reliable data and historical data to determine estimates of uncertainty.

During outbreaks of epidemics, initially data on cases and deaths could be scarce, and the quality of data annotation, validation, and aggregation might be uncertain. For instance, changes in clinical data entry, such as the addition of a new category {\em clinically diagnosed} to the existing {\em lab-confirmed} category, may likely have been reflected in the reporting \cite{Petropoulos2020-hh}. Initial fears about COVID-19 among the general population with regards to the trajectory of the pandemic could also affect administrative reporting. Diverse media sources and differences in local and federal policies may add to the general uncertainty about disease progression. Nevertheless, a data-driven approach to forecasting can offer valuable insights into the disease dynamics, and thereby an ability to objectively plan for the near future.

Unlike earlier global outbreaks, during COVID-19, the current worldwide digital ecosystem allows for real-time data collection, which in turn is used by AI and deep learning systems to understand healthcare trends, model risk associations, and predict outcomes \cite{Ting2020-lw}. 
In addition to traditional public health surveillance strategies available to most countries, a variety of static and dynamic data types may be integrated to model the scale and dynamics of this pandemic \cite{Pyne2015-ao}. Several organizations, including the Johns Hopkins University’s Center for Systems Science and Engineering, the New York Times, the Atlantic’s COVID-19 tracking project, among others have developed real-time tracking maps for following COVID-19 cases around the world using data from the U.S.~Centers for Disease Control and Prevention (CDC), WHO, and other international health agencies. Notably, the CDC receives forecasts from several modeling groups that use a wide variety of approaches ranging from SEIR (Susceptible-Exposed-Infectious-Recovered) to Agent-based to Bayesian models to understand the non-pharmaceutical policy interventions (or the lack thereof) to predict disease dynamics and its impact on human lives \cite{www-cdc-modeling-forecast}.

For precise and contextually relevant modeling of COVID-19 dynamics for a given population or community, we use data integration to combine community-specific health and behavioral risk factors, demographic and socioeconomic variables that are available at the local levels along with the cases and deaths data for COVID-19 while leveraging the recently developed and increasingly popular deep learning strategies to forecast COVID-19 dynamics using our model. Towards this, we develop a parallel computing platform called AICov standing for  {\em AI-driven Platform for COVID-19} to implement the different components of our framework and obtain the needed resources through multi-cloud interfaces in a robust manner.

In this study, while we incorporate information from community and county-specific covariates to inform our forecasting model for each metropolitan area, we want to avoid the so-called ecological fallacy in drawing inferences about individual disease outcomes based on large area-level aggregated risk factors. Our objective is also to underscore the nuanced roles played by pre-existing socioeconomic and other prevailing conditions of a given community that can act as determinants of its health outcomes and possible disparities,  especially under such sudden stresses to its current systems as those felt during any pandemic.
It is important to note that the underlying models are {\em learned} directly from the data instead of being projected by a theoretical epidemiological model that relies on simulation populations. This allows quick adaptation and self-adaptation in regards to changing behaviors and regional differences.

The paper is structured as follows. In Section \ref{sec:lstm-theory}, we provide a short overview of deep learning-based time series forecasting that we use in AICov. In Section \ref{sec:arch}, we outline our architecture that enables the user to conduct powerful time series forecasting by integrating static and dynamic information on population risk factors with time-series disease data. In Section \ref{sec:data}, we describe the input data for AICov. Next, we demonstrate our analysis involving the population risk factors and identify those that contribute to our prediction accuracy. Finally, we present our conclusion.

\section{LSTM}
\label{sec:lstm-theory}

This section presents a short overview of the recurrent neural network algorithms used in this work.

\begin{comment}
(Section \ref{sec:lstm-background}).
Additionally, we provide a short introduction of the LSTM implementation as used in Keras (Section \ref{sec:keras}) that we used as a foundation of this work.
\end{comment}

% ------------------------------
\subsection{LSTM Background}
\label{sec:lstm-background}
% ------------------------------

In this study, we used a Long Short-Term Memory (LSTM) algorithm for our predictions \cite{www-keras-lstm,Hochreiter1997-dk}. An LSTM is a recurrent neural network (RNN) \cite{Rumelhart1986-li} with feedback connections allowing the use of data input sequences to predict data output sequences. LSTMs have been applied to many application areas from handwriting \cite{Graves2009-qb}, image, feature detection, and time series prediction \cite{Schmidhuber2005-oy}. LSTM's have an internal state that is used to prevent the vanishing gradient problem \cite{Hochreiter1991-mp} during the training of RNNs.

Different variants of LSTM algorithms exist. Ours is based on an LSTM cell depicted in Figure \ref{fig:lstm}. 
It has an input gate, output gate, and a forget gate. The cell is maintaining how a subsequent value is calculated. This includes (a) how input values are influencing the cell via the input gate, (b) how the forget gate influences a memory value within the cell via the forget gate, and (c) how the output gate influences the output activation of the LSTM unit. A cell has a number of inputs and outputs that are weighted. During a training step, these weights are learned to be reused in a prediction process between input and output sequences.

\begin{figure}[h!]
\begin{minipage}[b]{0.6\textwidth}
    \centering
    \includegraphics[width=0.6\columnwidth]{images/lstm.pdf}
    \caption{LSTM Unit Diagram}
    \label{fig:lstm}
\end{minipage}
\ \
\begin{minipage}[b]{0.3\textwidth}
\vspace{-3cm}
\begin{equation*}
\begin{split}
    f_t &= \sigma(W_f x_t + U_f h_{t-1} + b_f) \\
    i_t &= \sigma(W_i x_t + U_i h_{t-1} + b_i) \\
    o_t &= \sigma(W_o x_t + U_o h_{t-1} + b_o)\\
    \tilde{c}_t &= tanh(W_c x_t + U_c h_{t-1} + b_c)\\
    c_t &= f_t \circ c_{t-1} +i_t \circ \tilde{c}_t \\ 
    h_t &= o_t \circ tanh(c_t) 
    \label{eq:lstm_equations}
\end{split}
\end{equation*}
\caption{LSTM Equations}
\label{fig:eq}
\end{minipage}
\end{figure}

\newcommand{\Rh}{\mathbb{R}^{h}}

We have summarized the equations used in LSTM in Figure~\ref{fig:eq}. We denote variables in lower case and matrices in upper case letters. $W_{q}$ and $U_{q}$ contain the weights of the input and recurrent connections, where the subscript denotes either the input gate $i$, the output gate $o$, the forget gate $f$ or the memory cell $c$.  It is important to note that  $c_{t}\in \mathbb {R}^{h}$ is not just one cell of one LSTM unit, but contains $h$ LSTM units' cells, representing a vector. To summarize our variables we use:
$\circ$ is the Hadamard product,
$\sigma$ is the sigmoid function,
$d$ the number of input features,
$h$ the number of hidden units,
$x_{t}\in \mathbb R^d$ the input vector to the LSTM unit,
$f_t\in \Rh$ the forget gate's activation vector,
$i_{t}\in \Rh$ the input/update gate's activation vector,
$o_{t}\in \Rh$ the output gate's activation vector,
$h_{t}\in \Rh$ the hidden state vector (e.g., the output vector of the LSTM unit),
$\tilde{c}_t\in \Rh$ the cell input activation vector,
$c_{t}\in \Rh$ the cell state vector,
$W\in \mathbb{R}^{h\times d}$, $U\in \mathbb{R} ^{h\times h}$, and $b\in \Rh$ are the weight matrices and bias vector parameters which need to be learned during training.

\begin{comment}
% ------------------------------
\subsection{LSTM in Keras}
\label{sec:keras}
% ------------------------------

LSTM has been available in Keras since 2015 \cite{www-keras-lstm,www-keras-api}. However, it does require some level of familiarity with RNNs for the required manipulation of the input data which needs to be adjusted for the application. The lack of complex and general examples involving multiple input and output variables, in conjunction with secondary information, calls for special attention from the users. Most examples focus on very simple usage of LSTM and do not to take into account the integration of other variables, such as variables represented by spacial population risk factors, in addition to time series data as input and output used in our approach. 
A further issue is that Keras depends on and is distributed with Tensorflow \cite{www-tensorflow}. Despite the popularity of Tensorflow as a Deep Learning framework, its adoption by modern operating systems has lagged behind other frameworks such as PyTorch \cite{www-pytorch}. 
Our framework overcomes these important limitations, especially while addressing the integration of spacial variables that could pose additional input for improving the accuracy of the predictions.
\end{comment}

\subsection{Covariate Enhanced LSTM}

We often think of the laws of physics described by operators that
evolve the system given sufficient initial conditions and in this
language, we have shown how to represent Newton’s law operator by a
recurrent network \cite{Kadupitiya2020-zq}. We expect that the time
dependence of many complex systems: Covid pandemics, Southern
California earthquakes, traffic flow, security events can be described
by deep learning operators that both capture the dynamics and allow
predictions.

In this paper, we adopt this approach to describe the
time dependence of COVID-19 infection and fatality counts in different
regions. 
We intend to
process the two time-series (infections, fatalities/deaths) associated with
each city plus the covariates (fixed in time but dependent on region)
to build a model for the time evolution of the data. There is a
nontrivial architecture design choice as to the method of handling the
time-independent covariates. We could combine a recurrent network for
the time series with some fixed network for the covariates. However,
that does not fit with our concept of deriving an evolution operator
for COVID-19 data that naturally depends on the covariates for each
city. Thus instead, we feed the covariates into the recurrent network.

\section{AICov Architecture}
\label{sec:arch}

The motivation for creating this architecture is based on our experience with deep learning toolkits such as Keras and PyTorch. While these systems provide the necessary APIs to integrate time series solutions, they target primarily more general user communities. In particular, they do not consider existing data sets or specific needs and analysis options to evaluate them for the purpose of COVID-19 risk factor data integration. We describe below the different requirements that motivate the design of our architecture framework, AICov, and how we address each of these.

%\newcommand{\Solution}{{\bf Solution: }}
\newcommand{\Solution}{}

\begin{requirement}{\bf Ease of Use.} A major design criterion for AICov is that the framework must be easy to use, and extensions can be made to allow usability of both the API as well as the user interface. \Solution To address these requirements, we are devising a specialized but easy to use time series API for COVID-19 that integrates common tasks such as automated filtering and normalization of the data.
\end{requirement}

\begin{requirement}{\bf Interactive Access.} Due to the experimental nature of analyzing the data, AICov must support the exploration of the data in an interactive fashion. \Solution As many data scientists use Jupyter Notebooks to integrate their work interactively, AICov integrates such notebooks. Jupyter notebooks allow rapid modification of the analysis workflow through the ease of Python as programming language. Additionally, they enable us to formulate easily sophisticated scientific analytics workflows with the help of Python.
\end{requirement}

\begin{requirement}{\bf Expandable API.} Assuming that new models and other analysis algorithms will be developed over time, AICov must allow the integration of such APIs. \Solution To address this requirement, we are developed an abstraction API for data, analysis, and metadata parameter adjustments. External services could be integrated with the help of REST services. The REST services are automatically generated from function specifications. 
\end{requirement}

\begin{requirement}{\bf Flexible Data Source Integration.} The integration of various sources of data is a key part of AICov. \Solution To achieve this goal, we provide a number of abstractions and data manipulation functions to extract needed data from established sources. Furthermore, the data can be combined, and additional data wrangling by external groups can be integrated.
Through such abstractions, it is possible to replace, add, and correct data sets that are part of the analysis. This allows us to compare different results created from different data sets. Moreover, we include data selection directly in our framework to easily and quickly reuse them in a data mashup. An important example is that the data need not be static over the disease's progression but can be continuously updated. This also has a significant impact on our analytics algorithms.
\end{requirement}

\begin{requirement}{\bf Automated Update of the Analytics.} As the data can change, previous analyses may have to be updated. \Solution Our forecast is not just a single script but includes a mechanism to register multiple workflows of continuously integrated (CI) analysis that are automatically rerun when new data are made available to the system. Previous results are maintained. Metadata with these runs can retrieve the versioned data sets used to calculate and store the versioned analytics workflow.
\end{requirement}

\begin{requirement}{\bf Flexible Model definition.} We want to allow experimentation with various models in AICov that may be easily and flexibly definable as workflows. \Solution We give functional abstraction definitions that enables us to define the models ourselves or integrate third party models describing the spread of the disease. 
\end{requirement}

\begin{requirement}{\bf Deep Learning Forecast.} As the forecasting models depend on changing data that are made available daily, it is important for AICov to utilize the newly available data. Many models can be applied to this, including moving averages based models or models that are purely derived from deep learning. \Solution To address this requirement, AICov allows for the integration of both. Furthermore, in our deep learning framework, we can  integrate an automated search for important risk factors, which is the focus of this study.
\end{requirement}

\begin{requirement}{\bf Model Orchestration.} To coordinate the different model predictions and the generation of the deep learning forecasts, we need the ability to orchestrate them while applying a number of parameters. \Solution To address this requirement, the AICov architecture includes the ability to integrate parameter sweeps to, for example, identify hyperparameters, or integrate different data sets as parameters to identify suitable forecast models that can deliver the one which fits the best.
\end{requirement}

\begin{requirement}{\bf Compute Resource Mapping.} For other uses of AICov, we need to integrate a flexible resource utilization framework. \Solution To address this requirement, we leverage from our earlier work and utilize through Cloudmesh and NIST NBDIF definitions cloud services into the architecture \cite{las-19-nist}. This includes containers but also infrastructure services in the cloud. Through these service integrations, we also provide access to  GPUs.
\end{requirement}

\begin{requirement}{\bf Convenient Interfaces.} As we have a wide variety of users, we need to enable interfaces that are used by the various communities. \Solution To address this requirement, we provide a number of APIs, REST services and make them available via Jupyter Notebooks. In addition, we developed some custom widgets for the notebooks that are specifically targeted towards data integration, parameter manipulation, and visualization of the results 
\end{requirement}

Putting all these requirements together results in an architecture as shown in Figure \ref{fig:arch}.\\

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.6\columnwidth]{images/arch.pdf}
    \caption{AICov Architecture}
    \label{fig:arch}
\end{figure}

In summary, the AICov architecture allows the inclusion of new data sources and models. An orchestrator enables modifying parameters and selecting data sources and models used in the analysis. The framework can be accessed via the interface layer that includes APIs, REST, and Jupyter Notebooks. Compute interfaces are provided either by direct access of GPUs from local resources or through the staging of automatically generated REST services via our Cloudmesh REST Service integrator \cite{cloudmesh-openapi-install,cloudmesh-openapi-benchmark,cloudmesh-manual}. Additionally, we can also access through abstractions cloud services that are offered by the various cloud providers and are targeting our application domain. Different models can be easily integrated into our framework to verify our forecast or enhance our forecast based on the users' models and data.

\section{Data for the COVID-19 Analysis}
\label{sec:data}

While the U.S. has seen an overall case fatality rate of 5.14\%, 
some U.S. locations have experienced a disproportionate number of deaths compared to others. This disparity of deaths among different communities could be attributed to a diversity of local risk factors and socioeconomic determinants of health that now constitute a very active area of investigation. Researchers are studying the effects of different community-specific socioeconomic variables, underlying health conditions and comorbidities, infrastructure, and systemic responses.
Given the complexity of such underlying conditions that can determine individual disease outcomes,  where and under what conditions people get infected by the virus, and then recover or die take place not in isolation but rather in an interplay of diverse factors that are characteristic to a local community, which need to be synthesized. In other words, a basic requirement for effective modeling of COVID-19 dynamics is that the information regarding a particular community’s vulnerabilities and strengths pertaining to the disease must be as comprehensive and integrative as possible. 
The regular reporting of accurate Coronavirus outbreaks data from local health departments has been understandably difficult, especially in the areas that are hardest hit. We use data on community-specific health and behavioral risk factors aggregated at the county level from the large-scale longitudinal CDC BRFSS surveys \cite{www-cdc-brfss}, and demographic and socioeconomic data from the U.S. Census Bureau that are known at the local levels along with the cases and deaths data compiled by NCHS \cite{www-cdc-nchs} and integrate them with other data sources.

\subsection{Data Collection and Processing}

Among the different sources of data on daily cases and deaths data collected and aggregated at different spatial levels, especially for the US, many are either freely available repositories such as GitHub or accessible via specifically developed online resources such as the New York Times~\cite{github-nytimes} and the Johns Hopkins Coronavirus Resource Center~\cite{www-jh-covid19}. 
For this study, we generated two data sets based on the latter resource upon our checking for the locations and times at which cumulative counts of COVID-19 cases and deaths on any given day were no less than those on the previous day. This led us to compile daily coronavirus cases and deaths data for 345 US counties from 32 states and the District of Columbia and matched by five-digit FIPS code or county name to dynamic and static variables from additional data sources.

The static data in this repository was collected from the US Census Bureau and the Centers for Disease Control and Prevention (CDC). The different fields cover the domains of behavioral and health risk factors, hospital capacity, and socioeconomic and demographic conditions. Data on health outcomes, prevention measures, and unhealthy behaviors were extracted from the CDC's 500 Cities Local Data for Better Health program \cite{www-cdc-chronic-data} in the form of crude or age-adjusted prevalence of conditions such as respiratory disease, obesity or smoking. These values were aggregated from the census tract to the county level using the first five digits of the eleven-digit tract FIPS code. Because a single county may consist of multiple individual cities, we include the list of all city labels within each aggregate group to represent a greater metropolitan area. 110 of such metropolitan areas that had more than 500 reported cases of COVID-19 by April 15, 2020, were selected for this study.

Notably, a greater metropolitan area such as New York City could be spread across more than one county, which can lead to the complexity of aggregating counts and other variables. Time series data for the 5 individual counties (boroughs) in New York City were not available in the Johns Hopkins CSSE data set. Rather, the total number of cases and deaths for the entire city of New York are reported and assigned the FIPS code of New York County (Manhattan). To ensure consistent geography across the static and dynamic data, we compute the population-weighted sum or median of each covariate over the assigned counties.

Demographic variables were gathered from the Census QuickFacts~\cite{www-census} online resource using an automated web scraping algorithm and cover relevant areas such as age, race, income, and population density. Additional socioeconomic variables include the Gini Index, which measures economic inequality and CDC Social Vulnerability Index (SVI)~\cite{www-cdc-svi}. The SVI was created to guide public health officials and disaster response efforts by identifying the communities across the United States most likely to need support during a crisis. Census tracts and corresponding counties are ranked across 15 social factors, which are grouped into four themes: socioeconomic status, household composition and disability, minority status and language, and housing and transportation.

Lastly, the total number of general acute care, critical access, and military hospitals within each county are included in the data. Such variables included the number of relevant hospitals per county, and the estimated number of beds (total known bed counts added to the number of hospitals in the county with missing data times the average number of beds per hospital in that state) per 1,000 people using the American Hospital Directory~\cite{www-ahd}.

The list of covariates is presented in Table \ref{tab:risk-factors}.

\begin{comment}
We are using the following abbreviations in the paper, and the data with data from \cite{cdc-brfss}, 
\cite{cdc-svi} and \cite{www-500Citie16} of population $>=18$ years of age that have been data fused by University of Pittsburgh.
\end{comment}

\begin{table}[!hptb]
\caption{Covariate Risk factor abbreviations as used in this study.}
\label{tab:risk-factors}
\bigskip
\resizebox{0.90\textwidth}{!}{
\centering
\begin{tabular}{ll}
\toprule
Risk Factor & Description \\
\midrule
 \underline{None} & no risk factor used\\
 ARTHRITIS 	  & Percent of arthritis \\
 BINGE 		  & Percent of binge drinking \\
 BPHIGH 	  & Percent of high blood pressure\\
 BPMED 		  & Percent of taking blood pressure medication\\
 CANCER 	  & Percent of cancer patients\\
 CASTHMA 	  & Percent of current asthma\\
 CHD 		  & Coronary heart disease\\
 CHECKUP 	  & Percent of health checkup\\
 CHOLSCREEN   & Percent of cholesterol screening\\
 COPD 		  & Percent of Chronic obstructive pulmonary disease\\
 CSMOKING 	  & Current smoking in percent\\
 DIABETES 	  & Percent of diabetes\\
 Estbeds 	  & Number of estimated beds in hospitals\\
 HIGHCHOL 	  & Percent of high cholesterol\\
 Insurance 	  & Percent of Insurance\\
 KIDNEY 	  & Percent of kidney disease\\
 LPA 		  & Percent of No leisure-time physical activity \\
 MHLTH 		  & Percent of not good mental health\\
 Nbeds 		  & Number of beds in hospitals \\
 Nbeds\_per1000   & Number of beds per 1000 people\\
 Nhosp 		  & Number of hospitals\\
 OBESITY 	  & Percent of population that is obese\\
 PHLTH        & Percent of Physical Health\\
 PVI 		  & Pandemic vulnerability index \\
 Percentblacks 	  & Percent of Blacks in the population\\
 Percenthispanics & Percent of Hispanics in the population\\
 STROKE 	  & Percent of stroke\\
 black\_percent	  & Percent of blacks in the population\\
 poverty\_percent & Percent of poverty in the population\\
 senior\_percent  & Percent of seniors in the population\\
 svi\_minority 	  & Percent of social vulnerability index in the minority population\\
 svi\_overall 	  & Percent of social vulnerability index in the overall population\\
\bottomrule
\end{tabular}
}
\end{table}

\section{Analysis}

We have conducted a number of analysis efforts to evaluate the validity of our approach. We focus on the following questions:

\begin{question} Can we apply deep learning strategies that self-learn the behavior of a model in order to provide future predictions? See Section \ref{sec:emperical}.
\label{q:1}
\end{question}

\begin{question} Can we use real time data as input and apply deep learning strategies instead of models that self learn and provide future predictions? See Section \ref{sec:lstm-covariate}.
\label{q:2}
\end{question}

\begin{question} Does the inclusion of covariates improve the prediction quality while including geospatial and socio-economic risk factors? See Section \ref{sec:lstm-covariate}.
\label{q:3}
\end{question}

\begin{question} What impact do we see when we introduce many risk factors? See Section \ref{sec:lstm-many}.
\label{q:4}
\end{question}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% COMMENTED OUT
%
\begin{comment}

\subsection{Covariance Analysis}

For each FIPS we have a number of risk factors for which we have values. These values can be correlated with each other through a correlation analysis.

To provide an initial idea how the risk factors are correlated with each other we showcase in Figure \ref{fig:cov-heatmap-1}  and \ref{fig:cov-heatmap-2} in the scale of -1.0 to 1.0. While the first figure shows all variables, we display in the second figure, only the variables related to health measurements and not social-economic factors.

In addition, we show in Figure \ref{fig:cov-pairplot} the more detailed pair plots. However, an online version of this plot is available, as indicated in the caption, as the information for the later is very condensed.

We observe that most social-economic variables are more strongly correlated with each other than the health variables. This is evident from the plot due to its coloring that is darker than the rest of Figure \ref{fig:cov-heatmap-1}. Obviously, the number of available hospitals vs. the number of estimated beds and the population are strongly correlated.

When looking at the health-related risk factors, we, for example, see that diabetes and high blood pressure are correlated with each other. This reflects observations from health care providers that have correlated these two variables. A similar correlation we see in the correlation between obesity and diabetes. Having such detailed information at hand makes it now suitable to integrate such health-related factors to COVID-19 cases and death.  This is conducted through two different comprehensive and sophisticated analysis studies.

In the first, we relate the risk factors based on information we predict for the maximum of the cases and death from the model we used for the given time period. The second is when we use our model predictions and find the best predictions while minimizing the cumulative error for the given prediction time period. To be complete, we have included all risk factors in our analysis rather than to restrict them based on our correlation matrix.

\begin{figure}[p]
    \centering
    \includegraphics[width=0.8\textwidth]{images/cov_analysis/heatmap.pdf}
    \caption{Covariance Analysis Heatmap of Socio-economic and Health Risk Factors}
    \label{fig:cov-heatmap-1}

    \bigskip
    \centering
    \includegraphics[width=0.8\textwidth]{images/cov_analysis/heatmap2.pdf}
    \caption{Covariance Analysis Heatmap of Health Related Risk Factors}
    \label{fig:cov-heatmap-2}
\end{figure}

\begin{figure}[p]
    \centering
    \includegraphics[width=1.0\textwidth]{images/cov_analysis/pairplot.png}
    \vspace{-1cm}
    \caption{Covariance Analysis Pairplot of Socio-economic and Health Risk Factors.
        An online version of the image is available at \url{https://github.com/laszewski/papers/blob/master/vonLaszewski-covid19-cov-factors.pdf} 
    }
    \label{fig:cov-pairplot}
\end{figure}
\end{comment}

\subsection{Deep Learning Predictions of Model-based Empirical Fits} 
\label{sec:emperical}

To establish the feasibility of using deep learning, we have, in our analysis, verified that through deep learning we can recreate the predictions identified by a model-based approach using empirical fits. In this comprehensive analysis, we obtain predictions through deep learning while using data generated from epidemiological models as discussed in \cite{marsland20-covid-paper}. The time series we used was 100 days long, and a multi-layered Long Short-Term Memory (LSTM) recurrent network was used. Our prediction approach differs by learning not only from the demographics (fixed data for each city) and time-dependent data but by integrating the population model for the underlying prediction, as shown in Fig. \ref{fig:magic-1}. Such model predictive integration capability is important in any application with multiple time scales. For example, this allows us to integrate multi-scale time effects into the forecast, which could address the combination of general forecasts and the next time steps of choice, such as days, weeks, months, or longer. For this analysis, we used 37 of the 110 cities with reliable empirical fits (that we derived from an epidemiological model and not deep learning  \cite{marsland20-covid-paper}) for the cases and deaths data up to April 15, 2020. Our analysis identified a sophisticated single deep learning time evolution operator that can describe these 37  separate data sets, and smooth fitted data leads to very accurate deep learning descriptions via LSTM as depicted in both Figure \ref{fig:magic-1} that are smaller than 1\% from the model predictions.

\begin{figure}[!h]
    \centering
    \includegraphics{images/magic-1.png}
    \caption{Cumulative empirical fits can be very accurately predicted with LSTM. The data is taken from 37 US Cities. Our comparison showcases that given a model, we can replicate the prediction of such models using the model data over time. The x-axis represents the days since the first data was fed to the model.}
    \label{fig:magic-1}
\end{figure}

\begin{comment}
Important to note is that such  models can be reused to conduct further studies to identify correlations towards particular risk factors, as shown in Section \ref{sec:max-analysis}.

\newcommand{\mysize}{0.9}

% \begin{comment}
\begin{figure}[p]
    \centering
    \begin{minipage}{0.45\textwidth}   %left column

    \includegraphics[width=\mysize\textwidth]{images/risk/risk-ARTHRITIS.png}
    \includegraphics[width=\mysize\textwidth]{images/risk/risk-BINGE.png}
    \includegraphics[width=\mysize\textwidth]{images/risk/risk-BPHIGH.png}
    \includegraphics[width=\mysize\textwidth]{images/risk/risk-BPMED.png}
    \includegraphics[width=\mysize\textwidth]{images/risk/risk-CANCER.png}
    \includegraphics[width=\mysize\textwidth]{images/risk/risk-CSMOKING.png}
    \includegraphics[width=\mysize\textwidth]{images/risk/risk-Estbeds.png}
    \includegraphics[width=\mysize\textwidth]{images/risk/risk-INSURANCE.png}

    \end{minipage}
    \ \hfill\vline\hfill \
    \begin{minipage}{0.45\textwidth} %right column

    \includegraphics[width=\mysize\textwidth]{images/risk/risk-CASTHMA.png}
    \includegraphics[width=\mysize\textwidth]{images/risk/risk-CHD.png}
    \includegraphics[width=\mysize\textwidth]{images/risk/risk-CHECKUP.png}
    \includegraphics[width=\mysize\textwidth]{images/risk/risk-CHOLSCREEN.png}
    \includegraphics[width=\mysize\textwidth]{images/risk/risk-COPD.png}
    \includegraphics[width=\mysize\textwidth]{images/risk/risk-DIABETES.png}
    \includegraphics[width=\mysize\textwidth]{images/risk/risk-HIGHCHOL.png}
    \includegraphics[width=\mysize\textwidth]{images/risk/risk-OBESITY.png}

    \end{minipage}
    \caption{Bivariate plot of risk factors when looking at maximum of cases and death in the empirical fit}
    \label{fig:risk-1}
\end{figure}

%\includegraphics[width=\mysize\textwidth]{images/risk/risk-Lat.png}

\begin{figure}
    \centering

    \begin{minipage}{0.45\textwidth}   %left column

    \includegraphics[width=\mysize\textwidth]{images/risk/risk-KIDNEY.png}
    \includegraphics[width=\mysize\textwidth]{images/risk/risk-LPA.png}
    \includegraphics[width=\mysize\textwidth]{images/risk/risk-MHLTH.png}
    \includegraphics[width=\mysize\textwidth]{images/risk/risk-Nbeds.png}
    \includegraphics[width=\mysize\textwidth]{images/risk/risk-Nhosp.png}
    \includegraphics[width=\mysize\textwidth]{images/risk/risk-STROKE.png}
    \includegraphics[width=\mysize\textwidth]{images/risk/risk-svi_households.png}

    \end{minipage}
    \ \hfill\vline\hfill \
    \begin{minipage}{0.45\textwidth} %right column

    \includegraphics[width=\mysize\textwidth]{images/risk/risk-Percentblacks.png}
    \includegraphics[width=\mysize\textwidth]{images/risk/risk-Percenthispanics.png}
    \includegraphics[width=\mysize\textwidth]{images/risk/risk-Percentseniors.png}
    \includegraphics[width=\mysize\textwidth]{images/risk/risk-PHLTH.png}
    \includegraphics[width=\mysize\textwidth]{images/risk/risk-Population.png}
    \includegraphics[width=\mysize\textwidth]{images/risk/risk-PVI.png}
    \includegraphics[width=\mysize\textwidth]{images/risk/risk-svi_housing_transport.png}
    \includegraphics[width=\mysize\textwidth]{images/risk/risk-svi_ses.png}

    \end{minipage}
    \caption{Bivariate plot of risk factors when looking at maximum of cases and death  in the empirical fit}
    \label{fig:risk-2}
\end{figure}
\end{comment}

\begin{comment}
\subsection{Correlation to the Maximum}
\label{sec:max-analysis}

One of the questions that arise is what risk factors, in addition to the cases and death, influence the prediction. In this study, we are not focusing on the cause, but like to identify suitable candidates to analyze further. For this reason, we look at the risk factors that are listed in Table \ref{tab:risk-factors}.

To identify a first set of suitable candidates, we used the model as described in Section \ref{sec:emperical} and identified for each risk factor the maximum for death and for cases for each of the 110 FIPS. We display the bivariate plots for each of the risk factors in Figures~\ref{fig:risk-1} and \ref{fig:risk-2}. We can see two types of effects in these bivariate plots.

\begin{description}

\item[Increased numbers with increased population when a risk factor is present.]~\\
When looking at the variables we see for example that the risk factor STROKE has a stronger correlation to the  death per population, then it has for the number of cases. Although we are not commenting on the causality, it is intuitive for this variable to think about that a person with prior history of a STROKE has an increased risk to dying in case of an infection. However, we also see an increased number for the cases. This indicates that STROKE is a potential important RISK factor and should be considered when conducting predictions based on death and cases.

\item[Value dependency categories for risk factors.] ~\\
When looking at the bivariate plots for death and for cases for a particular risk factor, we can identify trends within the plot that cluster some of the 110 FIPS into different categories. In the most simple aspect of this we can find liner trends that define the upper and lower bound of the risk factor bivariate plot. For example, when we look at the encompassing trend lines for cases vs. death by population, we find on the one hand FIPS for 17031,6037, 36119, 12086, representing  Cook, Illinois, US,     Los Angeles, California, US and  Miami-Dade, Florida with the PVI indices 0.582,    0.63,    0.635. On the other hand, we find Wayne, Michigan,     Essex, New Jersey,      Hartford, Connecticut,     Oakland, Michigan,     Macomb, Michigan,      Norfolk, Massachusetts. Which have  a PVI index of 0.477,     0.564,     0.527,     0.471,     0.499,     0.498 respectively. Hence, we can assume that a factor falling into different categories such as high PVI vs. low PVI have an impact on the cases and death. Associating them with the maximum cases and death helps to identify such categories. In our follow up analysis, we identify if this is indeed the case.

\end{description}

\end{comment}

%
%  END COMMENTED OUT
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Real time-Data Covariate LSTM Prediction}
\label{sec:lstm-covariate}

In this  section, we establish that using deep learning can not only be applied to model predictions from  real time data. As we want to evaluate the influence of the risk factors on our prediction framework using not only models but deep learning algorithms, we have devised a parameter sweep framework based on LSTM that conducts the prediction for the 110 cities while also considering the risk factors as listed in Table \ref{tab:risk-factors}. This framework can query any combination of risk factors as well as not including any, in which case it defaults to the traditional LSTM algorithm as introduced in Section \ref{sec:lstm-theory}. Hence, we will answer Questions \ref{q:2} and \ref{q:3}.

To answer these questions, we chose for our next analysis a date range between 2020-02-01 and 2020-05-25. We start with a basic LSTM operator with two layers of LSTM, and initial and final fully connected layers \cite{Kadupitiya2020-zq}. 
For LSTM, we have selected the activation function based on Rectified Linear Units ({\em RELU}). For the last layer {\em sigmoid} is chosen. A drop out rate of 0.2 is used, we use 16 input nodes, and have a batch size of 110. The maximum epoch is chosen to be 200. The number of nodes internally is 32. The maximum total number of Samples is 12540 when using 3 days as input length. We use a two-layer LSTM, as shown in Table \ref{tab:model}. Within our system, we have the ability to set a number of input days. For this analysis, we have chosen three different input lengths, namely 3, 4, and 5 days.

\begin{table}[p]
    \caption{The deep learning parameters used for the model sweep creation in as exported by Keras using 1 risk factor. The activation function is {\em RELU} and the recurrent activation function is {\em sigmoid}.}
    \label{tab:model}
    \bigskip

    \begin{footnotesize}
    \centering
    \resizebox{1.0\textwidth}{!}{ 
    \begin{tabular}{llp{2cm}p{2cm}p{2cm}p{2cm}p{2cm}p{2cm}p{2cm}p{2cm}p{2cm}}
    \toprule 
     Layer & (type)   &
     Output Shape &  Parameters for 0 risk factor &  
     Output Shape &  Parameters for 1 risk factor &
     Output Shape &  Parameters for 33 risk factors &
     Output Shape &  Parameters for 37 risk factors\\
    \midrule
    dense & (Dense)    & (None,1,32) & 96   &  (None,1,32)  & 128  &  (None,5,32) & 1152 & (None,5,32) & 1280 \\
    lstm & (LSTM)      & (None,1,16) & 3136 &  (None,1,16)  & 3136 &  (None,5,16) & 3136 & (None,5,16) & 3136 \\
    lstm\_1 & (LSTM)   & (None,16)  & 2112  &  (None,16)    & 2112 &  (None,16)   & 2112 & (None,16)   & 2112 \\
    dense\_1 & (Dense) & (None,32)  &  544  &  (None,32)    & 544  &  (None,32)   &  544 & (None,32)   &  544 \\
    dense\_2 & (Dense) & (None,30)  &  990  &  (None,30)    & 990  &  (None,30)   &  990 & (None,30)   &  990 \\
    \bottomrule
    \end{tabular}
    }
    ~\\
    The LSTM contains a total of 6,910 trainable parameters in case we include one risk factor. In the case, we do not use any risk factor it is 6878, and when we include 33 risk factors, it is 7934.  
    \end{footnotesize}
\end{table}

\begin{comment}
\begin{verbatim}
    model.add(Dense(InitialMLP, activation=activationvalue, input_shape=(n_timesteps, n_features)))

    model.add(LSTM(n_nodes, unroll=True, recurrent_dropout= recurrent_dropout1, dropout = dropout1,
                   activation= activationvalue , return_sequences=True, recurrent_activation="sigmoid",
                   input_shape=(n_timesteps, nextround)))

    model.add(LSTM(n_nodes, unroll=True, recurrent_dropout= recurrent_dropout2, dropout = dropout2,
               activation= activationvalue, recurrent_activation="sigmoid",
               input_shape=(n_timesteps, n_nodes)))
\end{verbatim}

\begin{verbatim}
Window Size  1  
n_samples: 12540 
n_timesteps: 1 
n_features: 3 
n_outputs: 30 
batch_size: 110 
n_nodes: 16 
epochs: 200
is NaN  77000  
percent  20.47  
not NaN  299200
\end{verbatim}
\end{comment}

We have run the training on all 110 cities from our data that represented each FIPS while not including any risk factors (Question \ref{q:2}) and running them with a single risk factor while iterating over all risk factors (Question \ref{q:3}). We then summed up all absolute errors for the predicted data points over the same time period and repeated each experiment 10 times. We summarized all the results in the form of a box-whisker diagram each for deaths and for cases. We sorted the diagram in such a fashion that the risk factor that has the model with the lowest error appears first. We preceded the graph with the {\em None} experiment that does not use any risk factors to provide a comparison. The graphs are depicted in Figures \ref{fig:box-cases} for cases, and \ref{fig:box-death} for deaths. 
Further, the obtained sorted order of the best prediction models by minimum error based on the risk factors for cases and deaths are as follows:

\begin{description}

\item[Order for cases sorted by minimum model error:] None, pop\_density\_2010, PHLTH, Insurance, Percentblacks, Percenthispanics, Nhosp, INSURANCE, CHOLSCREEN, DIABETES, STROKE, CHD, CHECKUP, Nbeds, svi\_overall, KIDNEY, CASTHMA, 
black\_percent, BPMED, LPA, CSMOKING, ARTHRITIS, poverty\_percent, Estbeds, MHLTH, senior\_percent, COPD, CANCER, Nbeds\_per1000, BPHIGH, HIGHCHOL, BINGE, OBESITY, svi\_minority

\item[Order for deaths sorted by minimum model error:]
None, svi\_overall, CSMOKING, Percenthispanics, pop\_density\_2010, senior\_percent, CHOLSCREEN, CASTHMA, Insurance, INSURANCE, MHLTH, LPA, poverty\_percent, OBESITY, CANCER, black\_percent, Nbeds\_per1000, Percentblacks, DIABETES, ARTHRITIS, Nbeds, Nhosp, BPHIGH, CHECKUP, PHLTH, CHD, STROKE, HIGHCHOL, Estbeds, KIDNEY, BINGE, BPMED, COPD, svi\_minority
\end{description}

We show the errors and the association with a risk factor in Table \ref{tab:top-1}, where cum\_error represents the cumulative error of the prediction. We sorted it by the cumulative error for cases. We see that in our experiment, almost all risk factors lead to an increased model prediction accuracy. As a result, we note that by including factors such as population density, physical health Insurance, population breakdown by ethnicity, number of hospitals, and diabetes, it leads to better overall predictions for cases. We also see that many of these factors perform better on average.

We observed a different scenario for deaths, in which we only identified two risk factors, namely the Social Vulnerability Index (SVI) and chain-smoking, that, when integrated into our deep learning model, leads to better prediction results. However, the accuracy of the overall prediction of deaths is far less precise than that of the cases. To showcase this, we have included Figures \ref{fig:error-case-popdensity} and \ref{fig:error-death-popdensity} that show the respective errors. This is following our input data distribution in which we noted higher fluctuations relative to the overall value. A mitigation to this issue would be using a seven day average over the analyzed period. However, we have not done this on purpose for this analysis as we wanted to identify how a covariate enhanced LSTM behaves that includes risk factors based on the daily fluctuations. This was done to avoid and showcase any needed preproccessing on the data and to identify the capability of the deep learning framework while adapting to the fluctuations without any special activities. Previously, we have already shown in Section \ref{sec:emperical} that for smooth data inputs the prediction is very accurate.
Using fluctuation data allows our framework for an automated ingest of data on a daily basis to enable data fusion of new cases and death information. Hence, we can re-train the model with new incoming data every day. Through repeated training experiments, we found the LSTM hyperparameters such as epochs and dropout values that work well for our experiments.

We conducted one additional analysis and plotted the best risk factor prediction model errors, the five best, and the ten best from our experiments to identify if we have better predictions than by including no risk factor. The best result found with no risk factor is depicted with the black horizontal line. Figures~\ref{fig:place-top1-cases} to  \ref{fig:place-top10-death} show the results sorted by the minimized error. Notably, we identified that a jump occurs when we sort the top 10 model predictions at around 330 models from a total of 3400 automatically generated models. This also underscores the need to run our deep learning model multiple times to obtain best parameter candidates.

While this study focuses on if and which risk factors can improve the predictions, we do not focus on why. Such analysis could provide insights into causal relationships. For example,  information such as health insurance and the number of hospitals and beds are often related to the quality of services that affect early detection and treatment. Furthermore, population parameters such as density can determine the rate of spread and the number of cases. Similarly, distributions of comorbidities in a given population such as physical health, cholesterol screening, diabetes, etc., are found to be insightful in predicting outcomes \cite{Maleki}. Identifying such risk factors and correlating them to the prediction accuracy provides valuable candidates for future studies.

\begin{table}[p]
\caption{Best models for each risk factor and their respective errors. The model that uses no risk factor is highlighted in grey.}
\label{tab:top-1}
\bigskip
\centering
\resizebox{0.7\textwidth}{!}{ 
\begin{tabular}{rrrrrlrl}
\toprule
 rmse\_cases &  rmse\_death &  cum\_error\_cases &  cum\_error\_death &  days in & risk &  place & \\
\midrule
 0.055443 & 0.067617 & 7.94 & 22.74 & 5 & pop\_density\_2010 & 0 \\
 0.054795 & 0.068529 & 8.01 & 20.55 & 3 & PHLTH & 1 \\
 0.055199 & 0.067641 & 8.07 & 20.43 & 4 & Insurance & 2 \\
 0.055028 & 0.067207 & 8.12 & 20.17 & 4 & Percentblacks & 3 \\
 0.056061 & 0.067489 & 8.22 & 19.04 & 5 & Percenthispanics & 4 \\
 0.054621 & 0.068255 & 8.26 & 20.10 & 3 & Nhosp & 5 \\
 0.054557 & 0.068572 & 8.34 & 24.61 & 3 & INSURANCE & 6 \\
 0.054712 & 0.069121 & 8.37 & 21.70 & 3 & CHOLSCREEN & 7 \\
 0.054586 & 0.067636 & 8.40 & 20.50 & 5 & DIABETES & 8 \\
 0.054160 & 0.066906 & 8.45 & 20.16 & 4 & STROKE & 9 \\
 0.054027 & 0.067642 & 8.46 & 21.93 & 5 & CHD & 10 \\
 0.054927 & 0.067652 & 8.51 & 20.42 & 5 & CHECKUP & 11 \\
 0.054146 & 0.067536 & 8.54 & 21.15 & 5 & Nbeds & 12 \\
 0.054795 & 0.067676 & 8.58 & 19.94 & 5 & svi\_overall & 13 \\
 0.054600 & 0.067565 & 8.61 & 21.86 & 5 & KIDNEY & 14 \\
 0.055089 & 0.068248 & 8.63 & 22.58 & 5 & CASTHMA & 15 \\
 0.054547 & 0.068266 & 8.68 & 20.49 & 3 & black\_percent & 16 \\
 0.054685 & 0.068920 & 8.74 & 21.69 & 3 & BPMED & 17 \\
 0.054323 & 0.067149 & 8.75 & 19.54 & 5 & LPA & 18 \\
 0.053716 & 0.066898 & 8.75 & 19.01 & 5 & CSMOKING & 19 \\
 0.054053 & 0.068716 & 8.83 & 20.94 & 3 & ARTHRITIS & 20 \\
 0.055029 & 0.068672 & 8.84 & 22.02 & 3 & poverty\_percent & 21 \\
 0.054545 & 0.067267 & 8.86 & 21.22 & 4 & Estbeds & 22 \\
 0.054864 & 0.068742 & 8.90 & 19.53 & 3 & MHLTH & 23 \\
 0.055079 & 0.069024 & 8.93 & 21.23 & 3 & senior\_percent & 24 \\
 0.055589 & 0.068250 & 8.93 & 23.26 & 4 & COPD & 25 \\
 0.054500 & 0.068363 & 8.99 & 19.70 & 3 & CANCER & 26 \\
 0.054397 & 0.068491 & 9.05 & 21.89 & 3 & Nbeds\_per1000 & 27 \\
 0.054609 & 0.067346 & 9.13 & 21.29 & 4 & BPHIGH & 28 \\
\rowcolor{lightgray}  0.054868 & 0.069432 & 9.26 & 25.88 & 3 & None & 29 \\
 0.056429 & 0.069345 & 9.28 & 20.98 & 3 & HIGHCHOL & 30 \\
 0.054317 & 0.067638 & 9.32 & 21.00 & 5 & BINGE & 31 \\
 0.054469 & 0.068540 & 9.50 & 20.65 & 3 & OBESITY & 32 \\
 0.055094 & 0.067443 & 9.54 & 21.53 & 4 & svi\_minority & 33 \\
\bottomrule
\end{tabular}
}
\end{table}

\begin{figure}[!p]
    \centering
    \includegraphics[width=1.0\textwidth]{images/boxwhisker/boxplot_cases.pdf}
    \vspace{-1cm}
    \caption{Influence of risk factors on the accuracy of the prediction of COVID-19 cases. The y-axis represents the cumulative error over all input data for the cities.}
    \label{fig:box-cases}
    \bigskip

    \includegraphics[width=1.0\textwidth]{images/boxwhisker/boxplot_death.pdf}
    \vspace{-1cm}
    \caption{Influence of risk factors on the accuracy of the prediction of COVID-19 death. The y-axis represents the cumulative error over all input data for the cities.}
    \label{fig:box-death}
\end{figure}

\begin{figure}[!p]
    \begin{minipage}{.45\textwidth}
    \centering
    \includegraphics[width=1.0\textwidth]{images/predict/Length_5-fields_pop_density_2010-Cumulative-Cases.png}
    \vspace{-1cm}
    \caption{Model prediction error for cumulative cases when including one risk factor - pop\_density\_2010 in the features, in addition to the past number of cases and deaths.}
    \label{fig:error-case-popdensity}
    \end{minipage}
    \begin{minipage}{.45\textwidth}
    \ \
    \centering
    \includegraphics[width=1.0\textwidth]{images/predict/Length_5-fields_pop_density_2010-Cumulative-Death.png}
    \vspace{-1cm}
    \caption{Model prediction error for cumulative deaths when including one risk factor - pop\_density\_2010 in the features, in addition to the past number of cases and deaths.}
    \label{fig:error-death-popdensity}
    \end{minipage}
\end{figure}

\begin{figure}[!p]
    \begin{minipage}{.45\textwidth}
        \centering
        \includegraphics[width=1.0\textwidth]{images/predict/PlaceTop1_Cases.pdf}
        \vspace{-1cm}
        \caption{Best predictions for cases over all risk factors.}
        \label{fig:place-top1-cases}

    \end{minipage}
    \ \
    \begin{minipage}{0.45\textwidth}
        \centering
        \includegraphics[width=1.0\textwidth]{images/predict/PlaceTop1_Death.pdf}
        \vspace{-1cm}
        \caption{Best predictions for deaths over all risk factors.}
        \label{fig:place-top1-death}
    \end{minipage}

    \begin{minipage}{.45\textwidth}
        \centering
        \includegraphics[width=1.0\textwidth]{images/predict/PlaceTop5_Cases.pdf}
        \vspace{-1cm}
        \caption{Top 5 predictions from all cases over all risk factors.}
        \label{fig:place-top5-cases}
    \end{minipage}
    \ \
    \begin{minipage}{.45\textwidth}
        \centering
        \includegraphics[width=1.0\textwidth]{images/predict/PlaceTop5_Death.pdf}
        \vspace{-1cm}
        \caption{Top 5 predictions from all deaths  over all risk factors.}
        \label{fig:place-top5-death}
    \end{minipage}

    \begin{minipage}{.45\textwidth}

        \centering
        \includegraphics[width=1.0\textwidth]{images/predict/PlaceTop10_Cases.pdf}
        \vspace{-1cm}
        \caption{Top 10 predictions from all deaths  over all risk factors.}
        \label{fig:place-top10-cases}
    \end{minipage}
    \ \
    \begin{minipage}{.45\textwidth}

        \centering
        \includegraphics[width=1.0\textwidth]{images/predict/PlaceTop10_Death.pdf}
        \vspace{-1cm}
        \caption{Top 10 predictions from all deaths over all risk factors.}
        \label{fig:place-top10-death}
    \end{minipage}
\end{figure}

\begin{comment}
\begin{figure}[!h]
    \centering
    \includegraphics[width=1.0\textwidth]{images/boxwhisker/swarmplot_cases.pdf}
    \caption{Caption}
    \label{fig:my_label}

    \centering
    \includegraphics[width=1.0\textwidth]{images/boxwhisker/swarmplot_death.pdf}
    \caption{Caption}
    \label{fig:my_label}
\end{figure}
\end{comment}

\subsection{Impact and Design of a Multi-Covariate-LSTM}
\label{sec:lstm-many}

In our next comprehensive analysis, we try to answer Question \ref{q:4} if all risk factors impact the analysis. Thus we feed the covariates into the recurrent network as additional features. In addition to cases and deaths, this leads to 35 variables, including 33 unique covariates. We looked at two related ways of doing this:

\begin{description}

\item [(a)] Have covariates fixed in time (static)

\item[(b)] Multiply the covariates by the times series (we used the average of cases and deaths)

\end{description}

We expected the second choice to be the most natural as it gives all
35 input streams similar time dependence. However, in our initial
tests, we found these choices gave very similar results, and so we
focused on the simpler choice (a) in this paper. In the experiments discussed
here, we either used 0, 1, or 33 covariates. Choosing 33 gives the best
description (reducing loss by about 1\%) while using each covariate as
a solitary extra feature allows us a rather sensitive study of the
importance of each individual covariate as we noted in Section \ref{sec:lstm-covariate}.

The final choice was the quantities to predict and train
on. We choose this to be both the cases and deaths values at the
time value following the sequence combined with the same quantities
for the following 14 days. This gives us 30 different output features
for each sequence. We renormalized, so the two basic features
(following day) contributed 50\% of the loss function. We supported
this in TensorFlow with a custom mean square error ($MSE$) loss
function that dropped training values that were not available for time
sequences near the end of the examined period. While one can not
train on future values, we can predict them, as seen in our sample
outputs. We extensively studied both 2 and 30 output features, and they
give similar descriptions of the observed data with the 30 feature
choice enabling 14-day predictions.

We performed an extensive hyper-parameter search and we show the values
in Table \ref{tab:model} that give robust results without
overfitting. Furthermore, we looked at four different activation functions for 
this analysis in the
LSTM and fully connected layers. $Selu$, $Relu$ and $Tanh$ activations gave
very similar results while $Sigmoid$ activation was much worse. However we still use a $Sigmoid$ function in the last stage just as we did in Section \ref{sec:lstm-covariate}. We also looked at the number of hidden units (separately for LSTM and fully
connected layers), the number of LSTM layers and the dropout values.
Only the latter produced a significant change and, for example, removing
dropouts decreased the loss by about 16\% but we retained them to provide
a robust environment. We chose randomly 20\% of the data for
validation and testing and this had similar loss values to the
training data (the other 80\%) with our chosen dropout.

The final technical issue concerns the data normalization where each
feature is transformed by a feature dependent -- but city independent -- way
to lie between 0 and 1. Further all extensive (proportional to size)
covariates were first divided by the city population except for
the population itself where we took the logarithm; intensive covariates
were just rescaled to lie between 0 and 1. The basic COVID-19 daily
data is extensive but compared two approaches with and without
renormalization, but in this case, we took the square root of the data value.
This is natural if there are counting (square root(N) for N
observations) errors as the deep learning loss function is
$(predicted-observed)^{2}$ without the error traditional in
$\chi^{2}$, namely $((predicted-observed)/error)^{2}$. This approach
with standard deep learning $MSE$ appears sound instead of dividing by
the population. That will emphasize the error contribution of
smaller cities but is still interesting to look at due to the presence of many systematic errors.

\section{Direct Deductions from Deep Learning Data Representation}

The deep learning model contains a lot of information much of which we have already analyzed and discussed in Section
\ref{sec:lstm-covariate} and Figures \ref{fig:magic-1} and
\ref{fig:error-case-popdensity}. In addition, we have looked at the
sensitivity of the results as measured by the predicted daily rates
and the value of the loss function. In Table \ref{tab:corr-matrix}, we
present one way of looking at it. We calculate the correlation
coefficients of many of the covariates from an accurate LSTM model fit
to the data. We divide the time series into three regions

\begin{itemize}
\item \textbf{Past} or all days up to 14 days before the final observed date (May 25, 2020)
\item \textbf{Now} or the 14 days from May 12 to May 25, 2020
\item \textbf{Future} or the 14 days after May 25, 2020 predicted by the model
\end{itemize}

These results are different between population normalized or direct
non-normalized data. We have kept the square-root used in the model
but divided the model values by the square root of the population when
computing the correlation. Significant positive correlations are seen
for CASTHMA, HIGHCHOL, DIABETES, OBESITY, CSMOKING, CHD, and CHECKUP
, while negative correlations for PVI and norm\_pop. We also looked at
the impact of these covariates on the evolution function. As this
operates on data at previous time values, this is sensitive to
variations across cities and times. The evolution operator is
intuitively the time derivative of the absolute values of daily rates
seen in the correlation in Table 3. Nhosp and Estbeds have the most
significance in impacting (as a reduction) the values of both the
infection and fatality rates from the evolution operator. The fatality
rates are also significantly impacted by Insurance and norm\_pop
(reduction) and senior\_percent and OBESITY (increase) from the
evolution operator. norm\_pop impacts infection values as an increase.
We can also identify which covariates have the most significant impact
on the loss function. Here Nhosp, CHD, KIDNEY, BINGE, and PHLTH give 5\%
or more reduction in loss function while norm\_pop, svi\_overall,
CASTHMA, DIABETES, and BPHIGH have small (1\%) effects.

\begin{table}[!h]
\caption{Correlation Matrix}
\bigskip
\label{tab:corr-matrix}
\resizebox{1.0\textwidth}{!}{
\begin{tabular}{lrrrrrr}
\toprule
Risk Factor		& Past Case & Now Case & Future Case & Past Death & Now Death & Future Death \\
\midrule
PVI                     & -0.425      & -0.292      & -0.124      & -0.434      & -0.373      & -0.239    \\
Insurance               & 0.018       & 0.017       & 0.012       & 0.018       & -0.002      & 0.001     \\
Nbeds                   & 0.136       & 0.131       & 0.217       & 0.089       & 0.080        & 0.109     \\
Nbeds\_per1000           & 0.138       & 0.121       & 0.209       & 0.092       & 0.069       & 0.088     \\
Nhosp                   & 0.074       & 0.119       & 0.165       & -0.009      & 0.028       & 0.021     \\
Estbeds                 & 0.136       & 0.131       & 0.217       & 0.089       & 0.08        & 0.109     \\
senior\_percent          & 0.104       & 0.028       & 0.033       & 0.152       & 0.164       & 0.156     \\
black\_percent           & 0.209       & 0.099       & 0.208       & 0.216       & 0.085       & 0.116     \\
Percenthispanics        & -0.072      & -0.032      & -0.064      & -0.091      & -0.068      & -0.046    \\
pop\_density\_2010        & -0.302      & -0.21       & -0.175      & -0.272      & -0.231      & -0.216    \\
poverty\_percent         & 0.128       & 0.059       & 0.2         & 0.148       & 0.031       & 0.08      \\
svi\_minority            & 0.029       & 0.069       & 0.081       & 0.01        & 0.006       & 0.087     \\
svi\_overall             & 0.14        & 0.158       & 0.238       & 0.125       & 0.113       & 0.179     \\
CASTHMA                 & 0.238       & 0.271       & 0.307       & 0.205       & 0.288       & 0.309     \\
HIGHCHOL                & 0.202       & 0.185       & 0.131       & 0.192       & 0.243       & 0.224     \\
DIABETES                & 0.200         & 0.200         & 0.178       & 0.206       & 0.232       & 0.23      \\
OBESITY                 & 0.199       & 0.205       & 0.194       & 0.191       & 0.229       & 0.212     \\
CANCER                  & -0.145      & -0.102      & -0.129      & -0.138      & -0.058      & -0.059    \\
STROKE                  & 0.123       & 0.125       & 0.117       & 0.167       & 0.143       & 0.135     \\
MHLTH                   & 0.11        & 0.161       & 0.186       & 0.076       & 0.175       & 0.222     \\
CSMOKING                & 0.189       & 0.258       & 0.249       & 0.16        & 0.289       & 0.323     \\
CHOLSCREEN              & 0.164       & 0.001       & -0.026      & 0.187       & 0.07        & 0.038     \\
INSURANCE               & 0.018       & 0.017       & 0.012       & 0.018       & -0.002      & 0.001     \\
CHD                     & 0.225       & 0.284       & 0.233       & 0.178       & 0.327       & 0.337     \\
CHECKUP                 & 0.206       & 0.221       & 0.28        & 0.201       & 0.272       & 0.343     \\
KIDNEY                  & 0.045       & 0.028       & 0.032       & 0.022       & 0.071       & 0.053     \\
BINGE                   & 0.031       & 0.044       & 0.009       & 0.083       & 0.03        & 0.022     \\
LPA                     & 0.191       & 0.187       & 0.197       & 0.147       & 0.22        & 0.214     \\
ARTHRITIS               & 0.046       & 0.013       & -0.021      & 0.061       & 0.089       & 0.079     \\
BPMED                   & 0.060        & 0.080        & 0.124       & 0.056       & 0.073       & 0.09      \\
PHLTH                   & 0.064       & 0.132       & 0.104       & 0.022       & 0.129       & 0.114     \\
BPHIGH                  & 0.056       & 0.051       & 0.048       & 0.058       & 0.084       & 0.068     \\
COPD                    & 0.163       & 0.163       & 0.153       & 0.175       & 0.22        & 0.24      \\
norm\_pop                & -0.305      & -0.258      & -0.218      & -0.193      & -0.218      & -0.099  \\
\bottomrule
\end{tabular}
}
\end{table}

\section{Conclusion}

A variety of mathematical and computational models have been used for predicting COVID-19 outcomes \cite{Jewell}. It includes compartmental models such as the SIR (Susceptible-Infected-Recovered) and SEIR (Susceptible-Exposed-Infected-Recovered) which are relatively easy to compute, and thus, commonly used in epidemiology. However, their simplistic assumptions, such as homogeneous mixing within populations, might make them less realistic for pandemics such as COVID-19. For instance, such models often assume a closed population that does not change. However, COVID-19 dynamics have prominently included massive migration, deaths due to comorbidities temporarily conferred immunity, and notably, major disparities in terms of socio-economic determinants of health.

In terms of implementation, SIR and SEIR models are fitted with point estimates, which may not adequately reflect the disease dynamics. For instance, it is difficult to account for behavioral changes (e.g., human mobility) or systemic inadequacies (e.g., availability of hospital beds) that may appear in specific subpopulations. Finally, the compartmental model estimates do not allow for quantification of uncertainty, which makes it harder to use them in policy-making. A comparison of modeling approaches concluded that while SEIR model performed better for a couple of states in the U.S., alternatives have performed better for the other states \cite{Bertozzi}. AICov has addressed most of the above modeling issues in its design, as per the requirements of its forecasting objectives.

We present AICov as an integrative deep learning framework for COVID-19 forecasting with the help of population covariates. Thus, its objectives are different from those of traditional compartmental models. One of the important features of the architecture for AICov is that it is by design targeting Cloud and HPC resources to conduct parameter sweeps to leverage sophisticated deep learning toolkits. The architecture allows the integration of various data sources that can update the data from its sources on demand and update its model predictions based on newly introduced data. Parameters can easily be adjusted via Jupyter notebooks that, in turn, call the computational backends on HPC and cloud resources.

In addition to this architecture, we used data collected from multiple public sources and agencies, and integrated the same across spatially contiguous units such as cities or metropolitan areas. In our analysis, we have focused on 110 selected cities of the U.S., but the framework is general and can be used to analyze similar data on pandemics from anywhere in the world.

The comprehensive analysis showcases the feasibility of our approach. Based on the outcomes reported it can lead to an improved prediction once we integrate risk factors in addition to the time-dependent data such as cases and death resulting from COVID-19. While the improvements observed were modest, they do present a concrete way of improving the forecast models. Inclusion of further putative factors from the ongoing worldwide studies on Covid-19 will only strengthen the future applications of our integrative framework.

We have shown that deep learning can return very good results while using smooth data, which we used in our empirical fits. For real data as presented to us for the daily changes, it still produced good results and even was able to predict changes based on weekly fluctuations. This is typically not achieved by other non-data based model approaches. Also, we have used in our data only cases and deaths, but we intend to expand this by using data about recovery and, in the future, immunization.

We have experimented with different hyperparameters and included in this study a selection of hyperparameters that have worked well for this data set.

As we have set up the first version of our AICov software, we have used cloud and high-performance computers. All of our sophisticated analyses were run in a day on at most 16 computers. All deep learning algorithms that were run on GPUs were run on NVIDIA K80s. However, our framework is on purpose generalized so that other compute resources can be integrated and leveraged. This includes local, HPC, and cloud computing resources, as well as different GPUs. Naturally, we also need to take into consideration not only the compute time but also the time it takes for the researcher to derive the models. As our approach is self-learning based on automated data updates, the time to prepare an updated model can be automatized and minimal input is needed. Thus the overall effort is very competitive.

\begin{comment}
Furthermore, we are currently conducting experiments with multiple risk factors at the same time to identify the most significant combinations of them.
As we have shown that the bivariate analysis has a locality dependency, we will explore categorizing locations with similar demographics and conduct further experiments on demographical similar regions. 
\end{comment}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{comment}

%
% SUPLEMENTAL
%

\vskip 14pt
\noindent {\bfseries{\fontsize{14pt}{1em}\selectfont Supplementary Materials}}

Contain
the brief description of the online supplementary materials.
\par

\end{comment}
%
% ACKNOWLEDGMENT
%

\vskip 14pt
\noindent {\bfseries{\fontsize{14pt}{1em}\selectfont Acknowledgements}}

This work is partially supported by the National Science Foundation (NSF) through awards CIF21 DIBBS 1443054, nanoBIO 1720625, Cybertraining 1829704, CINES 1835598 and Global Pervasive Computational Epidemiology 1918626.  We thank Jcs Kadupitiya for several useful discussions on deep learning frameworks.
\par

%
% BIBLIOGRAPHY
%

%\iffalse
\fontsize{11}{1em}\selectfont
\renewcommand\bibname{\bfseries{\fontsize{14pt}{1em}\selectfont References}}
\expandafter\ifx\csname
natexlab\endcsname\relax\def\natexlab#1{#1}\fi
\expandafter\ifx\csname url\endcsname\relax
  \def\url#1{\texttt{#1}}\fi
\expandafter\ifx\csname urlprefix\endcsname\relax\def\urlprefix{URL}\fi
%\fi

%\clearpage

\bibliographystyle{plainurl}
%\bibliographystyle{plain}

\bibliography{bib/data-covid19,bib/lstm,bib/cloudmesh}

\end{document}

### 2020-12-18 deprecated/proposal.tex
add overleaf paper to github

\subsection{Advanced Deep-Learning for COVID-19}

We discuss two objectives that are closely associated with Aim 3. This includes the introduction of Deep learning-based forecasting of COVID-19 as well as the introduction of Generative Adversarial Network (GANs) for generating new model-based data for COVID-19. Combined with our other Aims, we will develop a novel system for analyzing disease spread. Notably, GANs form a cutting edge, new development in deep learning that were introduced in different applications successfully since 2014 [15].

Objective 1. Deep Learning-based forecasting of COVID-19. There is increased recognition of the importance of deep learning in data-driven discovery across a broad range of applications \TODO{[16]–[18]}. Deep learning uses multiple layers of neural networks to consecutively extract multi-level features from the raw input. Due to the advent of fast GPUs, deep learning has become a formidable option in many optimization problems. The advantage of deep learning is the ability to search for features that are inherent in the underlying dataset without explicitly identifying them. This makes them adaptable for time series analysis as applied to COVID-19 and other infectious diseases. They offer us the opportunity to include not only the time-dependent variables but also additional health and socio-economic risk factors that influence the prediction.

In this Aim, we propose to leverage deep learning and integrate risk factors that will ultimately lead to better predictions that obviously could have an impact on scheduling scarce resources. Furthermore, the inclusion of risk factors can be used to identify population sectors that are most vulnerable. The inclusion of micro and macro data is essential to allow for prediction on various time scales and geographically and socio-economically diverse populations. The system will be designed in such a fashion that not only quasi-real time predictions can be made, but that encourages the inclusion of “what-if-scenarios” helping policymakers to decide what actions need to be taken in which scenario. Hence, the overall framework must be easily reusable by the community and is discussed in more detail in Aim 3.

Preliminary investigation. We are well-positioned to develop such a system as our preliminary investigation (manuscript under preparation [19]) has shown that we can create new predictions rapidly through deep learning while integrating many different risk factors too, for example, deal with new trends, under-reporting [20], or data corrections as we have seen in COVID-19 data reporting [21]. In our pilot study, we have shown that deep learning can be applied to complex COVID-19 pandemic data analysis that can both capture the dynamics to create a deep learning operand and allow predictions. Using county-level CDC BRFSS risk factors, Census-based demographic and socioeconomic variables, and February-March 2020 time-series data on COVID-19 data on cases and deaths from 110 US cities, we used an extensive parameter sweep to identify deep learning operators with very accurate predictions for not only the overall prediction of the cases and deaths but also in regards to their geographical distributions. Fig 1. shows the accuracy of our modeling and forecasting results with error estimates summed up over all locations and contrasting them to the real and predicted data. 
Our pilot incorporates (a) health and behavioral risk factors, (b) socioeconomic parameters, (c) an adaptable and sophisticated time series analysis incorporating multivariate and multistep [22] input and outputs (e.g. in our preliminary analysis steps refer to days), and  (d) and allows inclusion of updated, different, and new data sources [23][24][25][26][27][28]. From this and other data sources our Aim 1 will create temporal and spatial useful datasets for our deep learning framework. The suitability of our deep learning framework for utilizing information for covariates was demonstrated in this preliminary work as many risk factors increase the accuracy of our predictions, as shown in Fig 5. Our analysis “learned” from our experiments that the average accuracy of the model increases while incorporating parameters such as county-level availability of hospitals, the percentage of African Americans or seniors, the percentage of insured people, among others.

Fig. 1: Agglomerated deep learning fits of 110 US cities for COVID-19 case and death data in the US from Feb. 1 to May 25, 2020, with predictions 2 weeks out and showing weekly structure (to showcase the sum over the error of all 110 Cities the  axis is normalized and not shown in \# count)

Fig. 2: Distribution of the accuracy of the prediction upon inclusion of health factors by county.

In this project, we will further use our integrative capability to include more datasets describing risk factors in particular but social and environmental determinants of health in general, and determine their combinations that are relevant for our model predictions. This will allow us to model the disease dynamics in a much broader and powerful framework as we not only focus on time-dependent and spatial instances of the disease, but also on the underlying population characteristics. To establish the feasibility of our approach, we have in our second pilot analysis compared them to epidemiological models with empirical fits [29]. The time series were 100 days long and a multi-layered Long Short-Term Memory (LSTM) [30] recurrent network was used. It differed by learning not only from the demographics (fixed data for each city) as well as time-dependent data but also by integrating the population model for the underlying prediction as shown in Fig. 2. Such model predictive integration capability is important in any application with multiple time scales. For example, it will allow us to integrate multiscale time effects into the forecast, which could address the combination of general forecasts and the next time steps of choice such as days, weeks, months, or longer. For this pilot, we used 37 of the 110 cities with reliable empirical (not deep learning) fits for the case and death data up to April 15, 2020 [29]. Our pilot identified a sophisticated single deep learning time evolution operator that can describe these 37  separate datasets and smooth fitted data leads to very accurate deep learning descriptions as depicted in both Fig. 1 and 3. In our planned work, we will expand this surrogate for an empirical fit to a surrogate for a sophisticated epidemiological simulation combining all our aims and giving the community a new novel tool for analysis (Fig. 4). The new deep learning operators thus developed will become an essential part of the model prediction toolkit of an epidemiologist or disease modeler and pave the way for future epidemiologic studies. 
Our next objective will allow the researchers to generate realistic samples that mimic real statistical properties of a given population, in this case, corresponding to COVID-19 cases and deaths. As we will likely identify further and newer parameters and networks based on those parameters, we will also integrate them into a GAN in order to improve their contributions, and go on to identify accurate model predictions when input data may not be available but at the same time, the output sample will continue to be realistic.

Fig. 3: Deep learning fits to empirical COVID-19 data descriptions with 37 separate results shown as summed over cities. The cases and death were learned together on each separate fit (to showcase the sum over the error of all 37 Cities the  axis is normalized and not shown in \verb|#| count) 	 
Fig. 4: AICov Architecture

\section{Dessimination}

Implementation and distribution of data products and open-source software packages 
We will make the software and data APIs and sources freely available to the community. Tutorials for guiding the users of the platform will be made available so that programs can be easily recreated or reproduced. The code will be managed in a GitHub.com and replications will be hosted in the organizations as backups. 
Impact on other disciplines: Our main focus is to make this framework available to others within the NIH community. In 2004, Modeling of Infectious Disease Agent Study (MIDAS) [40] was established by the National Institute of General Medical Sciences (NIGMS) as a collaborative network of research scientists who use computational, statistical and mathematical models to understand infectious disease dynamics and assist the nation to prepare for, detect and respond to infectious disease threats. As a member of the MIDAS Network, Dr. Pyne will share the results and platforms developed by this project with his fellow-members.
We intend to donate the application of COVID-19 data to MLPerf (with representation from major vendors) [41], [42] as a contribution to evaluate deep learning systems. It could serve as an essential contribution to two working groups: Time Series led by Huang and Science Data which was just set up and is led by Fox (PI of this project) and Hey (chief data scientist of UK science instruments) [43]. Hence this work could be reused as a modern and important benchmark for machine learning services and hardware. Furthermore, lessons learned from this effort can also be applied to other applications [44], [45], [46], [43], [47]. 
Preliminary Results: We have rich experience in managing large scale open source projects on GitHub with many contributors as we see in Cloudmesh and other projects we execute at IU. For example, Cloudmesh has been utilized by more than 200 students and researchers and received direct contributions from 70 that have been integrated [38]. IU has rich experience in big data management and has worked on many big data projects. Through the reuse of these activities, we can reduce the overall cost of the proposal, while at the same time adding functionality that would otherwise not be possible. Dr. Fox is leading the effort of MLPerf Time Series benchmark, while Dr von Laszewski is leading the NIST NBDIF Interfaces working group allowing easy dissemination and adoption of Health and other discipline members. Dr. Pyne is a member of the MIDAS Network of NIGMS, and will share the results and modeling frameworks of the project with his fellow-members.
Timeline (Table 1)

### 2020-12-18 jds-paper-v3.tex
add overleaf paper to github

--- 2020-12-18 deprecated/jds-paper-v2.tex
@@
 \setuptodonotes{inline}
+\usepackage{makecell}
+\usepackage{enumitem}
+\usepackage{tcolorbox}
 
@@
 \pagestyle{empty}
-
+\usepackage{lineno}
 \setcitestyle{square}
 
+\newcommand*\rot{\rotatebox{90}}
+
 %---------page header---------%
 
-%\pagestyle{fancy}
-%fancyhead[RO]{odd number page}
-%fancyhead[LE]{even number page}
+%\fancyhead[RO]{odd number page}
+%\fancyhead[LE]{even number page}
+
+\pagestyle{plain} 
+\linenumbers
 
@@
 
+\newtcolorbox{redbox}{colback=red!5!white,colframe=red!75!black,width=3cm}
+\newtcolorbox{yellowbox}{colback=yellow!5!white,colframe=yelloe!75!black,width=3cm}
+
+\newcommand{\DONE}{{\color{green!60!black}\makebox[0pt][l]{$\square$}\raisebox{.15ex}{\hspace{0.1em}$\checkmark$}}~}
+\newcommand{\DOIT}{{\color{red!60!black}\makebox[0pt][l]{$\square$}\raisebox{.15ex}{\hspace{0.1em}$\boxtimes}}~}
+\newcommand{\PROGRESS}{(in progress)~}
+
 \begin{document}
@@
 % \input{response/response}
+
+REVISION DUE: 01.12.2021
+
+\section*{Reviewer 1}
+
+My comments are the following:
+
+The revised manuscript is much smoother than the first version, but I still feel that less than sufficient care was put into the writing.
+The reviewer pointed some in the pdf report.
+I have some additional editorial suggestions:
+
+\begin{enumerate}
+
+\item \PROGRESS Submit the point-by-point response to the reviewer's comments and my comments in a pdf file.
+
+\begin{quote}
+The original submisison web page had a field to enter the comments as text which we used. We will change this and introduce a PDF file instead, which is easier for us also.
+\end{quote}
+
+\item \DONE Needs page number (line number) for ease of referencing in the review.
+
+\begin{quote}
+Page numbers and line numbers were activated
+\end{quote}
+
+\item \DONE Do not italicize in the text.
+
+\begin{quote}
+All italicized text have been removed.
+inline formulas are done by latex italized. 
+\end{quote}
+
+\item \DONE AICov needs to be defined in the abstract.
+
+\begin{quote}
+We added
+
+"We present the architecture of an artificial intelligence enhanced COVID-19 analysis (in short AICov)"
+
+was added to the abstract.
+\end{quote}
+
+\item \PROGRESS Acronyms need to be defined at their first occurrences (e.g., AI in the intro). An acronyme table is included in table 1, we added the following acronyms to the paper
+
+\begin{quote}
+    All acronyms have been defined at their first usage.
+In addition we have added a section Acronyms in which all acronyms are listed to make it easier for the reader to look them up in case they are used later on in the paper.
+\end{quote}
+
+\DOIT capitalization in definition sin the table 1 and others
+
+\DOIT PERCENT, PERCENTBALCK, BLACK PERCENT, also for hispanics and senior
+
+\DOIT use risk factor in table headings
+
+\DOIT PER1000 is /1000
+
+\DOIT make acronyms appear in text
+
+\DOIT Table 1. capitalization,explain why they are different or make them the same
+
+\item \DOIT  Figure 2: please present the equations as equations.
+The notations used in the equation need a full check (e.g., h is used as both the dimension and vector).
+I'd like to see a clear distinction between inputs/outputs and parameters.
+I'd suggest that the inputs be set up before the equations are presented.
+
+\begin{quote}
+    \todo{todo}
+\end{quote}
+
+\item \DOIT  Regenerate the figures into pdf format (the current jpg/png figures blur when zoomed in).
+
+\begin{quote}
+    \todo{todo}
+\end{quote}
+
+\item \DONE  Table 2: consider reformating it; it would not fit the page width in print.
+
+\begin{quote}
+Table 2 has been reformatted. It was previously fit in the column with resize box. While rotating the headers the font is increased. The table still fits in the column as it still uses resizebox.
+\end{quote}
+
+\item \DONE  Table 3: tidy the headers.
+
+\begin{quote}
+    The headers have been cleaned up and linebreaks wer introduced so the resizebox of the table presents the table now larger
+\end{quote}
+
+\item \DOIT  Figure 5-6: can be combined to one figure with two panels (using ggplot2 and save the space of the labels of the horizontal axis).
+
+\begin{quote}
+    \todo{todo}
+\end{quote}
+
+\item \DOIT  Figure 7-8: can be one figure with two panels; pop density 2010 needs to be cleaned.
+
+\begin{quote}
+    \todo{todo}
+\end{quote}
+
+\item \DOIT  Figure 9-14: combine into one figure.
+
+\begin{quote}
+    \todo{todo}
+\end{quote}
+
+\item \DONE  Table 4: keep the number of digits consistent.
+
+\begin{quote}
+    The digits have been padded with the missing 0 to make them all the same length
+\end{quote}
+
+\item \DOIT  Reproducibility: submit data/code/README in a compressed file as supplementary material, or put them in a public repo and give the link under an unnumbered section Supplementary Material.
+
+\begin{quote}
+    \todo{todo}
+\end{quote}
+
+\end{enumerate}
+
+To submit your revision, please log in to EJMS and submit it as a revised file to the original submission.
+Please also include a detailed description of how you addressed all the points raised by the reviewers.
+
+\section*{Reviewer 2}
+
+This paper presents AICov, an integrated deep learning framework for COVID-19 forecasting.
+It takes into account population covariates, and uses LSTM and event modeling to do the forecasting.
+Data integration is done at a local level.
+Integration of such information brings improvement to the prediction accuracy.
+Despite the authors’ effort to address my comments in their revision, the paper is still in rough shape.
+Please see my comments.
+
+\begin{enumerate}
+
+\item \DOIT Page 5, please write the full name for REST (representational state transfer ) when it first appears.
+
+\begin{quote}
+    The term has been defined at first usage, we also checked all other abbreviations.
+
+    \todo{check other abbreviations}
+
+\end{quote}
+
+\item \DOIT Page 5, line 3 under Requirement 3, “we are developed an abstraction API for data” does not read right.
+
+\begin{quote}
+    \todo{todo}
+\end{quote}
+
+\item \DOIT Page 6, line 8, what level is “local level”?
+
+\begin{quote}
+    \todo{todo}
+\end{quote}
+
+\item (partially done) Page 7, Table 1, is “CHD” also a percent?
+If yes please modify the description; Percent- blacks and black percent both correspond to “Percent of blacks in the population”.
+Both seem to have been used according to Table 3, with difference in the four model performance measurements.
+Do the authors have an explanation for this?
+
+    \begin{quote}
+    The table has been updated and it was indicated that it is in percent.
+There were indeed to different information sources that reported different values for percentage of blacks and other populations.
+This however has since than corrected and we have been able to use just one population value.
+
+    \todo{fix the values and only have one population value}
+    \end{quote}
+
+\item \DOIT Page 8, line 5 of Section 5.1, the abbreviation LSTM has already appeared before, and the full name can be removed.
+
+    \begin{quote}
+        \todo{todo}
+    \end{quote}
+
+\item \DOIT What do the two panels in Figure 4 represent? Please give clearer description either in the text or figure caption.
+
+    \begin{quote}
+        \todo{todo}
+    \end{quote}
+
+\item \DOIT Section 5.2, paragraph 2 line 6, no need to capitalize for the word “Samples”. Line 5 from the bottom, no need to capitalize for the word “Insurance”. Please be careful with your writing.
+
+    \begin{quote}
+        \todo{todo}
+    \end{quote}
+
+\item \DOIT Section 5.2, the “order for cases sorted by minimum model order” and “order for deaths sorted by minimum model error” paragraphs are redundant, as the same information is already displayed in the x-legend for Figures 5 and 6.
+
+    \begin{quote}
+        \todo{todo}
+    \end{quote}
+
+\item \DOIT In Table 3, what is the unit for the error terms?
+The errors in Figures 7 and 8 are percentages.
+If errors in Table 3 are also percentages please clearly mark.
+Also, the numbers in Figures 7 and 8 do not align with numbers in row 1 of Table 3.
+Please check your numbers with care.
+
+    \begin{quote}
+        \todo{todo}
+    \end{quote}
+
+\item \DOIT Please adjust the location of your figures so they do not look pages apart from their interpretations.
+
+    \begin{quote}
+        \todo{todo}
+    \end{quote}
+
+\item \DOIT If I am reading it correctly, Figures 9 and 11 are only the “zoom in” versions of partial information from Figure 13.
+Similarly for the right column.
+Why waste space on two more figures when you could make Figures 13 and 14 bigger and explain the findings?
+
+    \begin{quote}
+        \todo{todo}
+    \end{quote}
+
+\item \DOIT The last paragraph of Section 5.2 suits the conclusion/discussion for future work section, not in the middle of a paper.
+
+    \begin{quote}
+        \todo{todo}
+    \end{quote}
+
+\item \DOIT Section 5.3, the authors mentioned two way of constructing covariates.
+Since the second way was more complicated while yielding no better performance the authors did not use it.
+Then it can be removed without damaging the storyline.
+
+    \begin{quote}
+        \todo{todo}
+    \end{quote}
+
+\item \DOIT The description “We renormalized, so the two basic features (following day) contributed 50% of the loss function” is hard to interpret. Please elaborate.
+
+    \begin{quote}
+        \todo{todo}
+    \end{quote}
+
+\item \DOIT Paragraph 2 on page 14, “Selu, Relu and Tanh activations gave very similar results while Sigmoid activation was much worse.
+However we still use a Sigmoid function...” is very counter-intuitive.
+Isn’t the point of trying different activation functions to better fit the data?
+16.
+Paragraph 3 on page 14, “The basic COVID-19 daily data is extensive but compared two approaches with and without renormalization, but in this case, we ...” The sentence does not read correct.
+Please check your sentences with care!
+
+    \begin{quote}
+        \todo{todo}
+    \end{quote}
+
+\item \DOIT How to interpret the “significant positive correlation” in Section 6 from the practical perspec- tive?
+Also, the term “evolution function” made its first appearance in Section 6 without any definition or explanation; I suggest either removing the whole Section 6, or extend it with clearer definition of evolution function.
+18. The first two paragraphs do not read like to be in a “Conclusion” section. Rather, they better fit into the introduction.
+
+    \begin{quote}
+        \todo{todo}
+    \end{quote}
+
+\end{enumerate}
 
@@
 
-\affil[1]{{\emph\fontsize{12pt}{0.5em}\selectfont Digital Science Center, Indiana University, Bloomington Indiana, USA}}
-\affil[2]{{\emph\fontsize{12pt}{0.5em}\selectfont Public Health Dynamics Lab, and Department of Biostatistics, University of Pittsburgh, Pittsburgh, Pennsylvania, USA}}
-\affil[3]{{\emph\fontsize{12pt}{0.5em}\selectfont Health Analytics Network, Pennsylvania, USA}}
+\affil[1]{{\fontsize{12pt}{0.5em}\selectfont Digital Science Center, Indiana University, Bloomington Indiana, USA}}
+\affil[2]{{\fontsize{12pt}{0.5em}\selectfont Public Health Dynamics Lab, and Department of Biostatistics, University of Pittsburgh, Pittsburgh, Pennsylvania, USA}}
+\affil[3]{{\fontsize{12pt}{0.5em}\selectfont Health Analytics Network, Pennsylvania, USA}}
 
@@
 Therefore, it is of great importance to model COVID-19 and other pandemics in terms of the broader social contexts in which they take place.
-We present the architecture of AICov, which provides an integrative deep learning framework for COVID-19 forecasting with population covariates, some of which may serve as putative risk factors. 
+We present the architecture of an artificial intelligence enhanced COVID-19 analysis (in short AICov) , which provides an integrative deep learning framework for COVID-19 forecasting with population covariates, some of which may serve as putative risk factors.
 We have integrated multiple different strategies into AICov, including the ability to use deep learning strategies based on LSTM and event modeling.
@@
 During outbreaks of epidemics, initially data on cases and deaths could be scarce, and the quality of data annotation, validation, and aggregation might be uncertain.
-For instance, changes in clinical data entry, such as the addition of a new category {\em clinically diagnosed} to the existing {\em lab-confirmed} category, may likely have been reflected in the reporting \cite{Petropoulos2020-hh}.
+For instance, changes in clinical data entry, such as the addition of a new category clinically diagnosed to the existing lab-confirmed category, may likely have been reflected in the reporting \cite{Petropoulos2020-hh}.
 Initial fears about COVID-19 among the general population with regards to the trajectory of the pandemic could also affect administrative reporting.
@@
 
-Unlike earlier global outbreaks, during COVID-19, the current worldwide digital ecosystem allows for real-time data collection, which in turn is used by AI and deep learning systems to understand healthcare trends, model risk associations, and predict outcomes \cite{Ting2020-lw}.
+Unlike earlier global outbreaks, during COVID-19, the current worldwide digital ecosystem allows for real-time data collection, which in turn is used by artificial intelligence (AI) and deep learning systems to understand healthcare trends, model risk associations, and predict outcomes \cite{Ting2020-lw}.
 In addition to traditional public health surveillance strategies available to most countries, a variety of static and dynamic data types may be integrated to model the scale and dynamics of this pandemic \cite{Pyne2015-ao}.
@@
 For precise and contextually relevant modeling of COVID-19 dynamics for a given population or community, we use data integration to combine community-specific health and behavioral risk factors, demographic and socioeconomic variables that are available at the local levels along with the cases and deaths data for COVID-19 while leveraging the recently developed and increasingly popular deep learning strategies to forecast COVID-19 dynamics using our model.
-Towards this, we develop a parallel computing platform called AICov standing for  {\em AI-driven Platform for COVID-19} to implement the different components of our framework and obtain the needed resources through multi-cloud interfaces in a robust manner.
+Towards this, we develop a parallel computing platform called AICov standing for  AI-driven Platform for COVID-19 to implement the different components of our framework and obtain the needed resources through multi-cloud interfaces in a robust manner.
 
@@
 Our objective is also to underscore the nuanced roles played by pre-existing socioeconomic and other prevailing conditions of a given community that can act as determinants of its health outcomes and possible disparities,  especially under such sudden stresses to its current systems as those felt during any pandemic.
-It is important to note that the underlying models are {\em learned} directly from the data instead of being projected by a theoretical epidemiological model that relies on simulation populations.
+It is important to note that the underlying models are learned directly from the data instead of being projected by a theoretical epidemiological model that relies on simulation populations.
 This allows quick adaptation and self-adaptation in regards to changing behaviors and regional differences.
@@
 This section presents a short overview of the recurrent neural network algorithms used in this work.
-
-\begin{comment}
-(Section \ref{sec:lstm-background}).
-Additionally, we provide a short introduction of the LSTM implementation as used in Keras (Section \ref{sec:keras}) that we used as a foundation of this work.
-\end{comment}
 
@@
 
-\begin{comment}
-% ------------------------------
-\subsection{LSTM in Keras}
-\label{sec:keras}
-% ------------------------------
-
-LSTM has been available in Keras since 2015 \cite{www-keras-lstm,www-keras-api}.
-However, it does require some level of familiarity with RNNs for the required manipulation of the input data which needs to be adjusted for the application.
-The lack of complex and general examples involving multiple input and output variables, in conjunction with secondary information, calls for special attention from the users.
-Most examples focus on very simple usage of LSTM and do not to take into account the integration of other variables, such as variables represented by spacial population risk factors, in addition to time series data as input and output used in our approach.
-A further issue is that Keras depends on and is distributed with Tensorflow \cite{www-tensorflow}.
-Despite the popularity of Tensorflow as a Deep Learning framework, its adoption by modern operating systems has lagged behind other frameworks such as PyTorch \cite{www-pytorch}.
-Our framework overcomes these important limitations, especially while addressing the integration of spacial variables that could pose additional input for improving the accuracy of the predictions.
-\end{comment}
-
 \subsection{Covariate Enhanced LSTM}
@@
 \Solution To address this requirement, we are developed an abstraction API for data, analysis, and metadata parameter adjustments.
-External services could be integrated with the help of REST services.
+External services could be integrated with the help of REST (REpresentational State Transfer) services.
 The REST services are automatically generated from function specifications.
@@
 
-Among the different sources of data on daily cases and deaths data collected and aggregated at different spatial levels, especially for the US, many are either freely available repositories such as GitHub or accessible via specifically developed online resources such as the New York Times~\cite{github-nytimes} and the Johns Hopkins Coronavirus Resource Center~\cite{www-jh-covid19}.
+Among the different sources of data on daily cases and deaths data collected and aggregated at different spatial levels, especially for the U.S., many are either freely available repositories such as GitHub or accessible via specifically developed online resources such as the New York Times~\cite{github-nytimes} and the Johns Hopkins Coronavirus Resource Center~\cite{www-jh-covid19}.
 For this study, we generated two data sets based on the latter resource upon our checking for the locations and times at which cumulative counts of COVID-19 cases and deaths on any given day were no less than those on the previous day.
-This led us to compile daily coronavirus cases and deaths data for 345 US counties from 32 states and the District of Columbia and matched by five-digit FIPS code or county name to dynamic and static variables from additional data sources.
-
-The static data in this repository was collected from the US Census Bureau and the Centers for Disease Control and Prevention (CDC).
+This led us to compile daily coronavirus cases and deaths data for 345 U.S.
+counties from 32 states and the District of Columbia and matched by five-digit FIPS code or county name to dynamic and static variables from additional data sources.
+
+The static data in this repository was collected from the U.S.
+Census Bureau and the Centers for Disease Control and Prevention (CDC).
 The different fields cover the domains of behavioral and health risk factors, hospital capacity, and socioeconomic and demographic conditions.
@@
 Notably, a greater metropolitan area such as New York City could be spread across more than one county, which can lead to the complexity of aggregating counts and other variables.
-Time series data for the 5 individual counties (boroughs) in New York City were not available in the Johns Hopkins CSSE data set.
+Time series data for the 5 individual counties (boroughs) in New York City were not available in the Johns Hopkins CSSE (Center for Systems Science and Engineering) data set.
 Rather, the total number of cases and deaths for the entire city of New York are reported and assigned the FIPS code of New York County (Manhattan).
@@
 The list of covariates is presented in Table \ref{tab:risk-factors}.
-
-\begin{comment}
-We are using the following abbreviations in the paper, and the data with data from \cite{cdc-brfss}, 
-\cite{cdc-svi} and \cite{www-500Citie16} of population $>=18$ years of age that have been data fused by University of Pittsburgh.
-\end{comment}
 
@@
 \midrule
- \underline{None} & no risk factor used\\
- ARTHRITIS 	  & Percent of arthritis \\
- BINGE 		  & Percent of binge drinking \\
- BPHIGH 	  & Percent of high blood pressure\\
- BPMED 		  & Percent of taking blood pressure medication\\
- CANCER 	  & Percent of cancer patients\\
- CASTHMA 	  & Percent of current asthma\\
- CHD 		  & Coronary heart disease\\
- CHECKUP 	  & Percent of health checkup\\
- CHOLSCREEN   & Percent of cholesterol screening\\
- COPD 		  & Percent of Chronic obstructive pulmonary disease\\
- CSMOKING 	  & Current smoking in percent\\
- DIABETES 	  & Percent of diabetes\\
- Estbeds 	  & Number of estimated beds in hospitals\\
- HIGHCHOL 	  & Percent of high cholesterol\\
- Insurance 	  & Percent of Insurance\\
- KIDNEY 	  & Percent of kidney disease\\
- LPA 		  & Percent of No leisure-time physical activity \\
- MHLTH 		  & Percent of not good mental health\\
- Nbeds 		  & Number of beds in hospitals \\
- Nbeds\_per1000   & Number of beds per 1000 people\\
- Nhosp 		  & Number of hospitals\\
- OBESITY 	  & Percent of population that is obese\\
- PHLTH        & Percent of Physical Health\\
+ NONE & No risk factor used\\
+ ARTHRITIS 	  & Percent reported with arthritis \\
+ BINGE 		  & Percent reported with binge drinking \\
+ BPHIGH 	  & Percent reported with  high blood pressure\\
+ BPMED 		  & Percent reported taking blood pressure medication\\
+ CANCER 	  & Percent reported as cancer patients\\
+ CASTHMA 	  & Percent reported with current asthma\\
+ CHD 		  & Percent reported with coronary heart disease\\
+ CHECKUP 	  & Percent reported with health checkup\\
+ CHOLSCREEN   & Percent reported with cholesterol screening\\
+ COPD 		  & Percent reported with chronic obstructive pulmonary disease\\
+ CSMOKING 	  & Percent reported as currently smoking\\
+ DIABETES 	  & Percent reported with diabetes\\
+ ESTBEDS 	  & Number of estimated beds in hospitals\\
+ HIGHCHOL 	  & Percent reported with high cholesterol\\
+ INSURANCE 	  & Percent reported with insurance\\
+ KIDNEY 	  & Percent reported with kidney disease\\
+ LPA 		  & Percent reported with no leisure-time physical activity \\
+ MHLTH 		  & Percent reported with not good mental health\\
+ NBEDS 		  & Number of beds in hospitals \\
+ NBEDS/1000   & Number of beds per 1000 people\\
+ NHOSP 		  & Number of hospitals\\
+ OBESITY 	  & Percent reported with  population that is obese\\
+ PHLTH        & Percent reported with physical health issues\\
  PVI 		  & Pandemic vulnerability index \\
- Percentblacks 	  & Percent of Blacks in the population\\
- Percenthispanics & Percent of Hispanics in the population\\
- STROKE 	  & Percent of stroke\\
- black\_percent	  & Percent of blacks in the population\\
- poverty\_percent & Percent of poverty in the population\\
- senior\_percent  & Percent of seniors in the population\\
- svi\_minority 	  & Percent of social vulnerability index in the minority population\\
- svi\_overall 	  & Percent of social vulnerability index in the overall population\\
+ PERCENTBLACKS 	  & Percent of blacks in the population\\
+ PERCENTHISPANICS & Percent of hispanics in the population\\
+ STROKE 	  & Percent reported with stroke\\
+ BLACK	  & Percent of Blacks in the population\\
+ POVERTY & Percent of poverty in the population\\
+ SENIOR  & Percent of seniors in the population\\
+ SVI\_MINORITY 	  & Percent of social vulnerability index in the minority population\\
+ SVI\_OVERALL 	  & Percent of social vulnerability index in the overall population\\
 \bottomrule
@@
 \end{question}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-% COMMENTED OUT
-%
-\begin{comment}
-
-\subsection{Covariance Analysis}
-
-For each FIPS we have a number of risk factors for which we have values. These values can be correlated with each other through a correlation analysis.
-
-To provide an initial idea how the risk factors are correlated with each other we showcase in Figure \ref{fig:cov-heatmap-1}  and \ref{fig:cov-heatmap-2} in the scale of -1.0 to 1.0.
-While the first figure shows all variables, we display in the second figure, only the variables related to health measurements and not social-economic factors.
-
-In addition, we show in Figure \ref{fig:cov-pairplot} the more detailed pair plots.
-However, an online version of this plot is available, as indicated in the caption, as the information for the later is very condensed.
-
-We observe that most social-economic variables are more strongly correlated with each other than the health variables.
-This is evident from the plot due to its coloring that is darker than the rest of Figure \ref{fig:cov-heatmap-1}.
-Obviously, the number of available hospitals vs.
-the number of estimated beds and the population are strongly correlated.
-
-When looking at the health-related risk factors, we, for example, see that diabetes and high blood pressure are correlated with each other.
-This reflects observations from health care providers that have correlated these two variables.
-A similar correlation we see in the correlation between obesity and diabetes.
-Having such detailed information at hand makes it now suitable to integrate such health-related factors to COVID-19 cases and death.
-This is conducted through two different comprehensive and sophisticated analysis studies.
-
-In the first, we relate the risk factors based on information we predict for the maximum of the cases and death from the model we used for the given time period.
-The second is when we use our model predictions and find the best predictions while minimizing the cumulative error for the given prediction time period.
-To be complete, we have included all risk factors in our analysis rather than to restrict them based on our correlation matrix.
-
-\begin{figure}[p]
-    \centering
-    \includegraphics[width=0.8\textwidth]{images/cov_analysis/heatmap.pdf}
-    \caption{Covariance Analysis Heatmap of Socio-economic and Health Risk Factors}
-    \label{fig:cov-heatmap-1}
-
-    \bigskip
-    \centering
-    \includegraphics[width=0.8\textwidth]{images/cov_analysis/heatmap2.pdf}
-    \caption{Covariance Analysis Heatmap of Health Related Risk Factors}
-    \label{fig:cov-heatmap-2}
-\end{figure}
-
-\begin{figure}[p]
-    \centering
-    \includegraphics[width=1.0\textwidth]{images/cov_analysis/pairplot.png}
-    \vspace{-1cm}
-    \caption{Covariance Analysis Pairplot of Socio-economic and Health Risk Factors.
-        An online version of the image is available at \url{https://github.com/laszewski/papers/blob/master/vonLaszewski-covid19-cov-factors.pdf} 
-    }
-    \label{fig:cov-pairplot}
-\end{figure}
-\end{comment}
 
@@
     \caption{Cumulative empirical fits can be very accurately predicted with LSTM.
-The data is taken from 37 US Cities.
+The data is taken from 37 U.S.
+Cities.
 Our comparison showcases that given a model, we can replicate the prediction of such models using the model data over time.
@@
 
-\begin{comment}
-Important to note is that such  models can be reused to conduct further studies to identify correlations towards particular risk factors, as shown in Section \ref{sec:max-analysis}.
-
-\newcommand{\mysize}{0.9}
-
-% \begin{comment}
-\begin{figure}[p]
-    \centering
-    \begin{minipage}{0.45\textwidth}   %left column
-
-    \includegraphics[width=\mysize\textwidth]{images/risk/risk-ARTHRITIS.png}
-    \includegraphics[width=\mysize\textwidth]{images/risk/risk-BINGE.png}
-    \includegraphics[width=\mysize\textwidth]{images/risk/risk-BPHIGH.png}
-    \includegraphics[width=\mysize\textwidth]{images/risk/risk-BPMED.png}
-    \includegraphics[width=\mysize\textwidth]{images/risk/risk-CANCER.png}
-    \includegraphics[width=\mysize\textwidth]{images/risk/risk-CSMOKING.png}
-    \includegraphics[width=\mysize\textwidth]{images/risk/risk-Estbeds.png}
-    \includegraphics[width=\mysize\textwidth]{images/risk/risk-INSURANCE.png}
-
-    \end{minipage}
-    \ \hfill\vline\hfill \
-    \begin{minipage}{0.45\textwidth} %right column
-
-    \includegraphics[width=\mysize\textwidth]{images/risk/risk-CASTHMA.png}
-    \includegraphics[width=\mysize\textwidth]{images/risk/risk-CHD.png}
-    \includegraphics[width=\mysize\textwidth]{images/risk/risk-CHECKUP.png}
-    \includegraphics[width=\mysize\textwidth]{images/risk/risk-CHOLSCREEN.png}
-    \includegraphics[width=\mysize\textwidth]{images/risk/risk-COPD.png}
-    \includegraphics[width=\mysize\textwidth]{images/risk/risk-DIABETES.png}
-    \includegraphics[width=\mysize\textwidth]{images/risk/risk-HIGHCHOL.png}
-    \includegraphics[width=\mysize\textwidth]{images/risk/risk-OBESITY.png}
-
-    \end{minipage}
-    \caption{Bivariate plot of risk factors when looking at maximum of cases and death in the empirical fit}
-    \label{fig:risk-1}
-\end{figure}
-
-%\includegraphics[width=\mysize\textwidth]{images/risk/risk-Lat.png}
-
-\begin{figure}
-    \centering
-
-    \begin{minipage}{0.45\textwidth}   %left column
-
-    \includegraphics[width=\mysize\textwidth]{images/risk/risk-KIDNEY.png}
-    \includegraphics[width=\mysize\textwidth]{images/risk/risk-LPA.png}
-    \includegraphics[width=\mysize\textwidth]{images/risk/risk-MHLTH.png}
-    \includegraphics[width=\mysize\textwidth]{images/risk/risk-Nbeds.png}
-    \includegraphics[width=\mysize\textwidth]{images/risk/risk-Nhosp.png}
-    \includegraphics[width=\mysize\textwidth]{images/risk/risk-STROKE.png}
-    \includegraphics[width=\mysize\textwidth]{images/risk/risk-svi_households.png}
-
-    \end{minipage}
-    \ \hfill\vline\hfill \
-    \begin{minipage}{0.45\textwidth} %right column
-
-    \includegraphics[width=\mysize\textwidth]{images/risk/risk-Percentblacks.png}
-    \includegraphics[width=\mysize\textwidth]{images/risk/risk-Percenthispanics.png}
-    \includegraphics[width=\mysize\textwidth]{images/risk/risk-Percentseniors.png}
-    \includegraphics[width=\mysize\textwidth]{images/risk/risk-PHLTH.png}
-    \includegraphics[width=\mysize\textwidth]{images/risk/risk-Population.png}
-    \includegraphics[width=\mysize\textwidth]{images/risk/risk-PVI.png}
-    \includegraphics[width=\mysize\textwidth]{images/risk/risk-svi_housing_transport.png}
-    \includegraphics[width=\mysize\textwidth]{images/risk/risk-svi_ses.png}
-
-    \end{minipage}
-    \caption{Bivariate plot of risk factors when looking at maximum of cases and death  in the empirical fit}
-    \label{fig:risk-2}
-\end{figure}
-\end{comment}
-
-\begin{comment}
-\subsection{Correlation to the Maximum}
-\label{sec:max-analysis}
-
-One of the questions that arise is what risk factors, in addition to the cases and death, influence the prediction.
-In this study, we are not focusing on the cause, but like to identify suitable candidates to analyze further.
-For this reason, we look at the risk factors that are listed in Table \ref{tab:risk-factors}.
-
-To identify a first set of suitable candidates, we used the model as described in Section \ref{sec:emperical} and identified for each risk factor the maximum for death and for cases for each of the 110 FIPS.
-We display the bivariate plots for each of the risk factors in Figures~\ref{fig:risk-1} and \ref{fig:risk-2}.
-We can see two types of effects in these bivariate plots.
-
-\begin{description}
-
-\item[Increased numbers with increased population when a risk factor is present.]~\\
-When looking at the variables we see for example that the risk factor STROKE has a stronger correlation to the  death per population, then it has for the number of cases.
-Although we are not commenting on the causality, it is intuitive for this variable to think about that a person with prior history of a STROKE has an increased risk to dying in case of an infection.
-However, we also see an increased number for the cases.
-This indicates that STROKE is a potential important RISK factor and should be considered when conducting predictions based on death and cases.
-
-\item[Value dependency categories for risk factors.] ~\\
-When looking at the bivariate plots for death and for cases for a particular risk factor, we can identify trends within the plot that cluster some of the 110 FIPS into different categories.
-In the most simple aspect of this we can find liner trends that define the upper and lower bound of the risk factor bivariate plot.
-For example, when we look at the encompassing trend lines for cases vs.
-death by population, we find on the one hand FIPS for 17031,6037, 36119, 12086, representing  Cook, Illinois, US,     Los Angeles, California, US and  Miami-Dade, Florida with the PVI indices 0.582,    0.63,    0.635.
-On the other hand, we find Wayne, Michigan,     Essex, New Jersey,      Hartford, Connecticut,     Oakland, Michigan,     Macomb, Michigan,      Norfolk, Massachusetts.
-Which have  a PVI index of 0.477,     0.564,     0.527,     0.471,     0.499,     0.498 respectively.
-Hence, we can assume that a factor falling into different categories such as high PVI vs.
-low PVI have an impact on the cases and death.
-Associating them with the maximum cases and death helps to identify such categories.
-In our follow up analysis, we identify if this is indeed the case.
-
-\end{description}
-
-\end{comment}
-
-%
-%  END COMMENTED OUT
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \subsection{Real time-Data Covariate LSTM Prediction}
@@
 We start with a basic LSTM operator with two layers of LSTM, and initial and final fully connected layers \cite{Kadupitiya2020-zq}.
-For LSTM, we have selected the activation function based on Rectified Linear Units ({\em RELU}).
-For the last layer {\em sigmoid} is chosen.
+For LSTM, we have selected the activation function based on Rectified Linear Units (RELU).
+For the last layer sigmoid is chosen.
 A drop out rate of 0.2 is used, we use 16 input nodes, and have a batch size of 110.
@@
 
-\begin{table}[p]
+\begin{table}[!h]
     \caption{The deep learning parameters used for the model sweep creation in as exported by Keras using 1 risk factor.
-The activation function is {\em RELU} and the recurrent activation function is {\em sigmoid}.}
+The activation function is RELU and the recurrent activation function is sigmoid.}
     \label{tab:model}
@@
 
+    \settowidth{\rotheadsize}{Parameters\quad}
+
     \begin{footnotesize}
@@
     \resizebox{1.0\textwidth}{!}{ 
-    \begin{tabular}{llp{2cm}p{2cm}p{2cm}p{2cm}p{2cm}p{2cm}p{2cm}p{2cm}p{2cm}}
+    \begin{tabular}{|ll|lll|ll|ll|}
     \toprule 
-     Layer & (type)   &
-     Output Shape &  Parameters for 0 risk factor &  
-     Output Shape &  Parameters for 1 risk factor &
-     Output Shape &  Parameters for 33 risk factors &
-     Output Shape &  Parameters for 37 risk factors\\
+     \rothead{Layer} & \rothead{Type}   &
+     \rothead{Output Shape} &  \rothead{Parameters for 0 risk factor}  
+                  &  \rothead{Parameters for 1 risk factor} &
+     \rothead{Output Shape} &  \rothead{Parameters for 33 risk factors} &
+     \rothead{Output Shape} &  \rothead{Parameters for 37 risk factors}\\
     \midrule
-    dense & (Dense)    & (None,1,32) & 96   &  (None,1,32)  & 128  &  (None,5,32) & 1152 & (None,5,32) & 1280 \\
-    lstm & (LSTM)      & (None,1,16) & 3136 &  (None,1,16)  & 3136 &  (None,5,16) & 3136 & (None,5,16) & 3136 \\
-    lstm\_1 & (LSTM)   & (None,16)  & 2112  &  (None,16)    & 2112 &  (None,16)   & 2112 & (None,16)   & 2112 \\
-    dense\_1 & (Dense) & (None,32)  &  544  &  (None,32)    & 544  &  (None,32)   &  544 & (None,32)   &  544 \\
-    dense\_2 & (Dense) & (None,30)  &  990  &  (None,30)    & 990  &  (None,30)   &  990 & (None,30)   &  990 \\
+    dense & Dense    & (None,1,32) & 96   & 128  &  (None,5,32) & 1152 & (None,5,32) & 1280 \\
+    LSTM & LSTM      & (None,1,16) & 3136 & 3136 &  (None,5,16) & 3136 & (None,5,16) & 3136 \\
+    LSTM 1 & LSTM   & (None,16)  & 2112  & 2112 &  (None,16)   & 2112 & (None,16)   & 2112 \\
+    Dense 1 & Dense & (None,32)  &  544  & 544  &  (None,32)   &  544 & (None,32)   &  544 \\
+    Dense 2 & Dense & (None,30)  &  990  & 990  &  (None,30)   &  990 & (None,30)   &  990 \\
     \bottomrule
@@
 We sorted the diagram in such a fashion that the risk factor that has the model with the lowest error appears first.
-We preceded the graph with the {\em None} experiment that does not use any risk factors to provide a comparison.
+We preceded the graph with the experiment labeled as ``NONE'' that which does not use any risk factors to provide a comparison.
 The graphs are depicted in Figures \ref{fig:box-cases} for cases, and \ref{fig:box-death} for deaths.
+
+\begin{figure}[!p]
+    \centering
+    \includegraphics[width=1.0\textwidth]{images/boxwhisker/boxplot_cases.pdf}
+    \vspace{-1cm}
+    \caption{Influence of risk factors on the accuracy of the prediction of COVID-19 cases. The y-axis represents the cumulative error over all input data for the cities.}
+    \label{fig:box-cases}
+    \bigskip
+
+    \includegraphics[width=1.0\textwidth]{images/boxwhisker/boxplot_death.pdf}
+    \vspace{-1cm}
+    \caption{Influence of risk factors on the accuracy of the prediction of COVID-19 death. The y-axis represents the cumulative error over all input data for the cities.}
+    \label{fig:box-death}
+\end{figure}
+
+%%%%%%
+
 Further, the obtained sorted order of the best prediction models by minimum error based on the risk factors for cases and deaths are as follows:
@@
 
-\item[Order for cases sorted by minimum model error:] None, pop\_density\_2010, PHLTH, Insurance, Percentblacks, Percenthispanics, Nhosp, INSURANCE, CHOLSCREEN, DIABETES, STROKE, CHD, CHECKUP, Nbeds, svi\_overall, KIDNEY, CASTHMA,
-black\_percent, BPMED, LPA, CSMOKING, ARTHRITIS, poverty\_percent, Estbeds, MHLTH, senior\_percent, COPD, CANCER, Nbeds\_per1000, BPHIGH, HIGHCHOL, BINGE, OBESITY, svi\_minority
+\item[Order for cases sorted by minimum model error:] NONE, POP\_DENSITY\_2010, PHLTH, INSURANCE, PERCENTBLACKS, PERCENTHISPANICS, NHOSP, INSURANCE, CHOLSCREEN, DIABETES, STROKE, CHD, CHECKUP, NBEDS, SVI\_OVERALL, KIDNEY, CASTHMA,
+BLACK, BPMED, LPA, CSMOKING, ARTHRITIS, POVERTY, ESTBEDS, MHLTH, SENIOR, COPD, CANCER, NBEDS/1000, BPHIGH, HIGHCHOL, BINGE, OBESITY, SVI\_MINORITY
 
 \item[Order for deaths sorted by minimum model error:]
-None, svi\_overall, CSMOKING, Percenthispanics, pop\_density\_2010, senior\_percent, CHOLSCREEN, CASTHMA, Insurance, INSURANCE, MHLTH, LPA, poverty\_percent, OBESITY, CANCER, black\_percent, Nbeds\_per1000, Percentblacks, DIABETES, ARTHRITIS, Nbeds, Nhosp, BPHIGH, CHECKUP, PHLTH, CHD, STROKE, HIGHCHOL, Estbeds, KIDNEY, BINGE, BPMED, COPD, svi\_minority
+NONE, SVI\_OVERALL, CSMOKING, PERCENTHISPANICS, POP\_DENSITY\_2010, SENIOR, CHOLSCREEN, CASTHMA, INSURANCE, INSURANCE, MHLTH, LPA, POVERTY, OBESITY, CANCER, BLACK, NBEDS/1000, PERCENTBLACKS, DIABETES, ARTHRITIS, NBEDS, NHOSP, BPHIGH, CHECKUP, PHLTH, CHD, STROKE, HIGHCHOL, ESTBEDS, KIDNEY, BINGE, BPMED, COPD, SVI\_MINORITY
 \end{description}
@@
 
+\begin{table}[!p]
+\caption{Best models for each risk factor and their respective errors. The model that uses no risk factor is highlighted in grey.}
+\label{tab:top-1}
+\bigskip
+\centering
+\resizebox{1.0\textwidth}{!}{ 
+\begin{tabular}{rrrrrlrl}
+\toprule
+ \makecell{RMSE\\Cases} &  
+ \makecell{RMSE\\Death} &  
+ \makecell[r]{Cummulative \\Error \\Cases} &  
+ \makecell[r]{Cummulative \\Error \\Death} &  
+ \makecell[r]{Number of \\Days as \\Input} & 
+ \makecell{Risk Factor} &  
+ \makecell{Place/\\Rank} & \\
+\midrule
+ 0.055443 & 0.067617 & 7.94 & 22.74 & 5 & POP\_DENSITY\_2010 & 0 \\
+ 0.054795 & 0.068529 & 8.01 & 20.55 & 3 & PHLTH & 1 \\
+ 0.055199 & 0.067641 & 8.07 & 20.43 & 4 & INSURANCE & 2 \\
+ 0.055028 & 0.067207 & 8.12 & 20.17 & 4 & PERCENTBLACKS & 3 \\
+ 0.056061 & 0.067489 & 8.22 & 19.04 & 5 & PERCENTHISPANICS & 4 \\
+ 0.054621 & 0.068255 & 8.26 & 20.10 & 3 & NHOSP & 5 \\
+ 0.054557 & 0.068572 & 8.34 & 24.61 & 3 & INSURANCE & 6 \\
+ 0.054712 & 0.069121 & 8.37 & 21.70 & 3 & CHOLSCREEN & 7 \\
+ 0.054586 & 0.067636 & 8.40 & 20.50 & 5 & DIABETES & 8 \\
+ 0.054160 & 0.066906 & 8.45 & 20.16 & 4 & STROKE & 9 \\
+ 0.054027 & 0.067642 & 8.46 & 21.93 & 5 & CHD & 10 \\
+ 0.054927 & 0.067652 & 8.51 & 20.42 & 5 & CHECKUP & 11 \\
+ 0.054146 & 0.067536 & 8.54 & 21.15 & 5 & NBEDS & 12 \\
+ 0.054795 & 0.067676 & 8.58 & 19.94 & 5 & SVI\_OVERALL & 13 \\
+ 0.054600 & 0.067565 & 8.61 & 21.86 & 5 & KIDNEY & 14 \\
+ 0.055089 & 0.068248 & 8.63 & 22.58 & 5 & CASTHMA & 15 \\
+ 0.054547 & 0.068266 & 8.68 & 20.49 & 3 & BLACK & 16 \\
+ 0.054685 & 0.068920 & 8.74 & 21.69 & 3 & BPMED & 17 \\
+ 0.054323 & 0.067149 & 8.75 & 19.54 & 5 & LPA & 18 \\
+ 0.053716 & 0.066898 & 8.75 & 19.01 & 5 & CSMOKING & 19 \\
+ 0.054053 & 0.068716 & 8.83 & 20.94 & 3 & ARTHRITIS & 20 \\
+ 0.055029 & 0.068672 & 8.84 & 22.02 & 3 & POVERTY & 21 \\
+ 0.054545 & 0.067267 & 8.86 & 21.22 & 4 & ESTBEDS & 22 \\
+ 0.054864 & 0.068742 & 8.90 & 19.53 & 3 & MHLTH & 23 \\
+ 0.055079 & 0.069024 & 8.93 & 21.23 & 3 & SENIOR & 24 \\
+ 0.055589 & 0.068250 & 8.93 & 23.26 & 4 & COPD & 25 \\
+ 0.054500 & 0.068363 & 8.99 & 19.70 & 3 & CANCER & 26 \\
+ 0.054397 & 0.068491 & 9.05 & 21.89 & 3 & NBEDS/1000 & 27 \\
+ 0.054609 & 0.067346 & 9.13 & 21.29 & 4 & BPHIGH & 28 \\
+\rowcolor{black!5}  0.054868 & 0.069432 & 9.26 & 25.88 & 3 & NONE & 29 \\
+ 0.056429 & 0.069345 & 9.28 & 20.98 & 3 & HIGHCHOL & 30 \\
+ 0.054317 & 0.067638 & 9.32 & 21.00 & 5 & BINGE & 31 \\
+ 0.054469 & 0.068540 & 9.50 & 20.65 & 3 & OBESITY & 32 \\
+ 0.055094 & 0.067443 & 9.54 & 21.53 & 4 & SVI\_MINORITY & 33 \\
+\bottomrule
+\end{tabular}
+}
+\end{table}
+
 We observed a different scenario for deaths, in which we only identified two risk factors, namely the Social Vulnerability Index (SVI) and chain-smoking, that, when integrated into our deep learning model, leads to better prediction results.
@@
 To showcase this, we have included Figures \ref{fig:error-case-popdensity} and \ref{fig:error-death-popdensity} that show the respective errors.
+
+\begin{figure}[!h]
+    \begin{minipage}{.45\textwidth}
+    \centering
+    \includegraphics[width=1.0\textwidth]{images/predict/Length_5-fields_pop_density_2010-Cumulative-Cases.png}
+    \vspace{-1cm}
+    \caption{Model prediction error for cumulative cases when including one risk factor - POP\_DENSITY\_2010 in the features, in addition to the past number of cases and deaths.}
+    \label{fig:error-case-popdensity}
+    \end{minipage}
+    \begin{minipage}{.45\textwidth}
+    \ \
+    \centering
+    \includegraphics[width=1.0\textwidth]{images/predict/Length_5-fields_pop_density_2010-Cumulative-Death.png}
+    \vspace{-1cm}
+    \caption{Model prediction error for cumulative deaths when including one risk factor - POP\_DENSITY\_2010  in the features, in addition to the past number of cases and deaths.}
+    \label{fig:error-death-popdensity}
+    \end{minipage}
+\end{figure}
+
 This is following our input data distribution in which we noted higher fluctuations relative to the overall value.
@@
 
-While this study focuses on if and which risk factors can improve the predictions, we do not focus on why.
-Such analysis could provide insights into causal relationships.
-For example,  information such as health insurance and the number of hospitals and beds are often related to the quality of services that affect early detection and treatment.
-Furthermore, population parameters such as density can determine the rate of spread and the number of cases.
-Similarly, distributions of comorbidities in a given population such as physical health, cholesterol screening, diabetes, etc., are found to be insightful in predicting outcomes \cite{Maleki}.
-Identifying such risk factors and correlating them to the prediction accuracy provides valuable candidates for future studies.
-
-\begin{table}[p]
-\caption{Best models for each risk factor and their respective errors. The model that uses no risk factor is highlighted in grey.}
-\label{tab:top-1}
-\bigskip
-\centering
-\resizebox{0.7\textwidth}{!}{ 
-\begin{tabular}{rrrrrlrl}
-\toprule
- rmse\_cases &  rmse\_death &  cum\_error\_cases &  cum\_error\_death &  days in & risk &  place & \\
-\midrule
- 0.055443 & 0.067617 & 7.94 & 22.74 & 5 & pop\_density\_2010 & 0 \\
- 0.054795 & 0.068529 & 8.01 & 20.55 & 3 & PHLTH & 1 \\
- 0.055199 & 0.067641 & 8.07 & 20.43 & 4 & Insurance & 2 \\
- 0.055028 & 0.067207 & 8.12 & 20.17 & 4 & Percentblacks & 3 \\
- 0.056061 & 0.067489 & 8.22 & 19.04 & 5 & Percenthispanics & 4 \\
- 0.054621 & 0.068255 & 8.26 & 20.10 & 3 & Nhosp & 5 \\
- 0.054557 & 0.068572 & 8.34 & 24.61 & 3 & INSURANCE & 6 \\
- 0.054712 & 0.069121 & 8.37 & 21.70 & 3 & CHOLSCREEN & 7 \\
- 0.054586 & 0.067636 & 8.40 & 20.50 & 5 & DIABETES & 8 \\
- 0.054160 & 0.066906 & 8.45 & 20.16 & 4 & STROKE & 9 \\
- 0.054027 & 0.067642 & 8.46 & 21.93 & 5 & CHD & 10 \\
- 0.054927 & 0.067652 & 8.51 & 20.42 & 5 & CHECKUP & 11 \\
- 0.054146 & 0.067536 & 8.54 & 21.15 & 5 & Nbeds & 12 \\
- 0.054795 & 0.067676 & 8.58 & 19.94 & 5 & svi\_overall & 13 \\
- 0.054600 & 0.067565 & 8.61 & 21.86 & 5 & KIDNEY & 14 \\
- 0.055089 & 0.068248 & 8.63 & 22.58 & 5 & CASTHMA & 15 \\
- 0.054547 & 0.068266 & 8.68 & 20.49 & 3 & black\_percent & 16 \\
- 0.054685 & 0.068920 & 8.74 & 21.69 & 3 & BPMED & 17 \\
- 0.054323 & 0.067149 & 8.75 & 19.54 & 5 & LPA & 18 \\
- 0.053716 & 0.066898 & 8.75 & 19.01 & 5 & CSMOKING & 19 \\
- 0.054053 & 0.068716 & 8.83 & 20.94 & 3 & ARTHRITIS & 20 \\
- 0.055029 & 0.068672 & 8.84 & 22.02 & 3 & poverty\_percent & 21 \\
- 0.054545 & 0.067267 & 8.86 & 21.22 & 4 & Estbeds & 22 \\
- 0.054864 & 0.068742 & 8.90 & 19.53 & 3 & MHLTH & 23 \\
- 0.055079 & 0.069024 & 8.93 & 21.23 & 3 & senior\_percent & 24 \\
- 0.055589 & 0.068250 & 8.93 & 23.26 & 4 & COPD & 25 \\
- 0.054500 & 0.068363 & 8.99 & 19.70 & 3 & CANCER & 26 \\
- 0.054397 & 0.068491 & 9.05 & 21.89 & 3 & Nbeds\_per1000 & 27 \\
- 0.054609 & 0.067346 & 9.13 & 21.29 & 4 & BPHIGH & 28 \\
-\rowcolor{lightgray}  0.054868 & 0.069432 & 9.26 & 25.88 & 3 & None & 29 \\
- 0.056429 & 0.069345 & 9.28 & 20.98 & 3 & HIGHCHOL & 30 \\
- 0.054317 & 0.067638 & 9.32 & 21.00 & 5 & BINGE & 31 \\
- 0.054469 & 0.068540 & 9.50 & 20.65 & 3 & OBESITY & 32 \\
- 0.055094 & 0.067443 & 9.54 & 21.53 & 4 & svi\_minority & 33 \\
-\bottomrule
-\end{tabular}
-}
-\end{table}
-
-\begin{figure}[!p]
-    \centering
-    \includegraphics[width=1.0\textwidth]{images/boxwhisker/boxplot_cases.pdf}
-    \vspace{-1cm}
-    \caption{Influence of risk factors on the accuracy of the prediction of COVID-19 cases. The y-axis represents the cumulative error over all input data for the cities.}
-    \label{fig:box-cases}
-    \bigskip
-
-    \includegraphics[width=1.0\textwidth]{images/boxwhisker/boxplot_death.pdf}
-    \vspace{-1cm}
-    \caption{Influence of risk factors on the accuracy of the prediction of COVID-19 death. The y-axis represents the cumulative error over all input data for the cities.}
-    \label{fig:box-death}
-\end{figure}
-
-\begin{figure}[!p]
-    \begin{minipage}{.45\textwidth}
-    \centering
-    \includegraphics[width=1.0\textwidth]{images/predict/Length_5-fields_pop_density_2010-Cumulative-Cases.png}
-    \vspace{-1cm}
-    \caption{Model prediction error for cumulative cases when including one risk factor - pop\_density\_2010 in the features, in addition to the past number of cases and deaths.}
-    \label{fig:error-case-popdensity}
-    \end{minipage}
-    \begin{minipage}{.45\textwidth}
-    \ \
-    \centering
-    \includegraphics[width=1.0\textwidth]{images/predict/Length_5-fields_pop_density_2010-Cumulative-Death.png}
-    \vspace{-1cm}
-    \caption{Model prediction error for cumulative deaths when including one risk factor - pop\_density\_2010 in the features, in addition to the past number of cases and deaths.}
-    \label{fig:error-death-popdensity}
-    \end{minipage}
-\end{figure}
-
-\begin{figure}[!p]
+\begin{figure}[!h]
     \begin{minipage}{.45\textwidth}
@@
 \end{comment}
+
+While this study focuses on if and which risk factors can improve the predictions, we do not focus on why.
+Such analysis could provide insights into causal relationships.
+For example,  information such as health insurance and the number of hospitals and beds are often related to the quality of services that affect early detection and treatment.
+Furthermore, population parameters such as density can determine the rate of spread and the number of cases.
+Similarly, distributions of comorbidities in a given population such as physical health, cholesterol screening, diabetes, etc., are found to be insightful in predicting outcomes \cite{Maleki}.
+Identifying such risk factors and correlating them to the prediction accuracy provides valuable candidates for future studies.
 
@@
 this analysis in the
-LSTM and fully connected layers. $Selu$, $Relu$ and $Tanh$ activations gave
-very similar results while $Sigmoid$ activation was much worse.
-However we still use a $Sigmoid$ function in the last stage just as we did in Section \ref{sec:lstm-covariate}.
+LSTM and fully connected layers. $SELU$, $RELU$ and $TANH$ activations gave
+very similar results while $SIGMOID$ activation was much worse.
+However we still use a $SIGMOID$ function in the last stage just as we did in Section \ref{sec:lstm-covariate}.
 We also looked at the number of hidden units (separately for LSTM and fully
@@
 for CASTHMA, HIGHCHOL, DIABETES, OBESITY, CSMOKING, CHD, and CHECKUP
-, while negative correlations for PVI and norm\_pop. We also looked at
+, while negative correlations for PVI and NORM\_POP. We also looked at
 the impact of these covariates on the evolution function. As this
@@
 infection and fatality rates from the evolution operator. The fatality
-rates are also significantly impacted by Insurance and norm\_pop
-(reduction) and senior\_percent and OBESITY (increase) from the
-evolution operator. norm\_pop impacts infection values as an increase.
+rates are also significantly impacted by INSURANCE and NORM\_POP
+(reduction) and SENIOR and OBESITY (increase) from the
+evolution operator. NORM\_POP impacts infection values as an increase.
 We can also identify which covariates have the most significant impact
-on the loss function. Here Nhosp, CHD, KIDNEY, BINGE, and PHLTH give 5\%
-or more reduction in loss function while norm\_pop, svi\_overall,
+on the loss function. Here NHOSP, CHD, KIDNEY, BINGE, and PHLTH give 5\%
+or more reduction in loss function while NORM\_POP, SVI\_OVERALL,
 CASTHMA, DIABETES, and BPHIGH have small (1\%) effects.
@@
 PVI                     & -0.425      & -0.292      & -0.124      & -0.434      & -0.373      & -0.239    \\
-Insurance               & 0.018       & 0.017       & 0.012       & 0.018       & -0.002      & 0.001     \\
-Nbeds                   & 0.136       & 0.131       & 0.217       & 0.089       & 0.080        & 0.109     \\
-Nbeds\_per1000           & 0.138       & 0.121       & 0.209       & 0.092       & 0.069       & 0.088     \\
-Nhosp                   & 0.074       & 0.119       & 0.165       & -0.009      & 0.028       & 0.021     \\
-Estbeds                 & 0.136       & 0.131       & 0.217       & 0.089       & 0.08        & 0.109     \\
-senior\_percent          & 0.104       & 0.028       & 0.033       & 0.152       & 0.164       & 0.156     \\
-black\_percent           & 0.209       & 0.099       & 0.208       & 0.216       & 0.085       & 0.116     \\
-Percenthispanics        & -0.072      & -0.032      & -0.064      & -0.091      & -0.068      & -0.046    \\
-pop\_density\_2010        & -0.302      & -0.21       & -0.175      & -0.272      & -0.231      & -0.216    \\
-poverty\_percent         & 0.128       & 0.059       & 0.2         & 0.148       & 0.031       & 0.08      \\
-svi\_minority            & 0.029       & 0.069       & 0.081       & 0.01        & 0.006       & 0.087     \\
-svi\_overall             & 0.14        & 0.158       & 0.238       & 0.125       & 0.113       & 0.179     \\
+INSURANCE               & 0.018       & 0.017       & 0.012       & 0.018       & -0.002      & 0.001     \\
+NBEDS                   & 0.136       & 0.131       & 0.217       & 0.089       & 0.080        & 0.109     \\
+NBEDS/1000              & 0.138       & 0.121       & 0.209       & 0.092       & 0.069       & 0.088     \\
+NHOSP                   & 0.074       & 0.119       & 0.165       & -0.009      & 0.028       & 0.021     \\
+ESTBEDS                 & 0.136       & 0.131       & 0.217       & 0.089       & 0.080        & 0.109     \\
+SENIOR                  & 0.104       & 0.028       & 0.033       & 0.152       & 0.164       & 0.156     \\
+BLACK                   & 0.209       & 0.099       & 0.208       & 0.216       & 0.085       & 0.116     \\
+HISPANICS               & -0.072      & -0.032      & -0.064      & -0.091      & -0.068      & -0.046    \\
+POP\_DENSITY\_2010      & -0.302      & -0.210      & -0.175      & -0.272      & -0.231      & -0.216    \\
+POVERTY                 & 0.128       & 0.059       & 0.200       & 0.148       & 0.031       & 0.080     \\
+SVI\_MINORITY           & 0.029       & 0.069       & 0.081       & 0.010        & 0.006       & 0.087     \\
+SVI\_OVERALL            & 0.140       & 0.158       & 0.238       & 0.125       & 0.113       & 0.179     \\
 CASTHMA                 & 0.238       & 0.271       & 0.307       & 0.205       & 0.288       & 0.309     \\
 HIGHCHOL                & 0.202       & 0.185       & 0.131       & 0.192       & 0.243       & 0.224     \\
-DIABETES                & 0.200         & 0.200         & 0.178       & 0.206       & 0.232       & 0.23      \\
+DIABETES                & 0.200       & 0.200     & 0.178         & 0.206       & 0.232       & 0.230     \\
 OBESITY                 & 0.199       & 0.205       & 0.194       & 0.191       & 0.229       & 0.212     \\
@@
 STROKE                  & 0.123       & 0.125       & 0.117       & 0.167       & 0.143       & 0.135     \\
-MHLTH                   & 0.11        & 0.161       & 0.186       & 0.076       & 0.175       & 0.222     \\
-CSMOKING                & 0.189       & 0.258       & 0.249       & 0.16        & 0.289       & 0.323     \\
-CHOLSCREEN              & 0.164       & 0.001       & -0.026      & 0.187       & 0.07        & 0.038     \\
+MHLTH                   & 0.110       & 0.161       & 0.186       & 0.076       & 0.175       & 0.222     \\
+CSMOKING                & 0.189       & 0.258       & 0.249       & 0.160        & 0.289       & 0.323     \\
+CHOLSCREEN              & 0.164       & 0.001       & -0.026      & 0.187       & 0.070       & 0.038     \\
 INSURANCE               & 0.018       & 0.017       & 0.012       & 0.018       & -0.002      & 0.001     \\
 CHD                     & 0.225       & 0.284       & 0.233       & 0.178       & 0.327       & 0.337     \\
-CHECKUP                 & 0.206       & 0.221       & 0.28        & 0.201       & 0.272       & 0.343     \\
+CHECKUP                 & 0.206       & 0.221       & 0.280       & 0.201       & 0.272       & 0.343     \\
 KIDNEY                  & 0.045       & 0.028       & 0.032       & 0.022       & 0.071       & 0.053     \\
-BINGE                   & 0.031       & 0.044       & 0.009       & 0.083       & 0.03        & 0.022     \\
-LPA                     & 0.191       & 0.187       & 0.197       & 0.147       & 0.22        & 0.214     \\
+BINGE                   & 0.031       & 0.044       & 0.009       & 0.083       & 0.030       & 0.022     \\
+LPA                     & 0.191       & 0.187       & 0.197       & 0.147       & 0.220       & 0.214     \\
 ARTHRITIS               & 0.046       & 0.013       & -0.021      & 0.061       & 0.089       & 0.079     \\
-BPMED                   & 0.060        & 0.080        & 0.124       & 0.056       & 0.073       & 0.09      \\
+BPMED                   & 0.060       & 0.080        & 0.124       & 0.056       & 0.073      & 0.090      \\
 PHLTH                   & 0.064       & 0.132       & 0.104       & 0.022       & 0.129       & 0.114     \\
 BPHIGH                  & 0.056       & 0.051       & 0.048       & 0.058       & 0.084       & 0.068     \\
-COPD                    & 0.163       & 0.163       & 0.153       & 0.175       & 0.22        & 0.24      \\
-norm\_pop                & -0.305      & -0.258      & -0.218      & -0.193      & -0.218      & -0.099  \\
+COPD                    & 0.163       & 0.163       & 0.153       & 0.175       & 0.220       & 0.240     \\
+NORM\_POP               & -0.305      & -0.258      & -0.218      & -0.193      & -0.218      & -0.099  \\
 \bottomrule
@@
 All of our sophisticated analyses were run in a day on at most 16 computers.
-All deep learning algorithms that were run on GPUs were run on NVIDIA K80s.
+All deep learning algorithms that were run on GPUs were run on NVIDIA Tesla K80s.
 However, our framework is on purpose generalized so that other compute resources can be integrated and leveraged.
@@
 
+\section{Acronyms}
+
+This section contains the list of acronyms used in the paper for easy reference.
+The definitions of the risk factors are however given within the paper in Table~\ref{tab:risk-factors} on page \pageref{tab:risk-factors} and not repeated here.
+
+\begin{itemize}[leftmargin=5cm]
+\item[{\bf API:}] Application Programming Interfaces
+\item[{\bf BRFSS:}] Behavioral Risk Factor Surveillance System
+\item[{\bf CDC:}] Center for Desease Control
+\item[{\bf CIF21:}] Cyberinfrastructure Framework for 21st Century 
+\item[{\bf COVID-19:}] COrona VIrus Disease 2019 
+\item[{\bf CSSE:}] Center for Systems Science and Engineering
+\item[{\bf DIBBS:}] Data Infrastructure Building Blocks
+\item[{\bf FIPS:}] Federal Information Processing Standards
+\item[{\bf GPU:}] Graphics Processing Unit 
+\item[{\bf HPC:}] High Performance Computing
+\item[{\bf LSTM:}] Long Short-Term Memory
+\item[{\bf NBDIF:}] NIST Big Data Working Group
+\item[{\bf NCHS:}] CDC National Center for Health Statistics
+\item[{\bf NIST:}] National Institute of Standards and Technology
+\item[{\bf RELU:}] REctified Linear Unit
+\item[{\bf REST:}] REpresentational State Transfer 
+\item[{\bf RNN:}] Recurrent Neural Network
+\item[{\bf SEIR:}] Susceptible, Expose, Infectious, Recovered)
+\item[{\bf SIR}] Susceptible, Infectious, Recovered
+\item[{\bf U.S.:}] United States of America
+\item[{\bf NSF:}] National Science Foundation
+\item[{\bf MSE:}] Mean Squared Error
+\item[{\bf SELU:}] Scaled Exponential Linear Unit
+\item[{\bf CI:}] Cyber Infrastructure
+\item[{\bf CINES:}] CyberInfrastructure for Network Engineering and Science
+\item[{\bf WHO:}] World Health Organization
+\item[{\bf NONE:}] \todo{NO FACTORS}
+\item[{\bf NORM\_POP:}] Normalized Population
+\item[{\bf cum:}] cumulative
+\end{itemize}
+
 \vskip 14pt
@@
 This work is partially supported by the National Science Foundation (NSF) through awards CIF21 DIBBS 1443054, nanoBIO 1720625, Cybertraining 1829704, CINES 1835598 and Global Pervasive Computational Epidemiology 1918626.
-We thank Jcs Kadupitiya for several useful discussions on deep learning frameworks.
+We thank J.
+Kadupitiya for several useful discussions on deep learning frameworks.
 \par

### 2020-12-18 response-v3-content.tex
Updates from Overleaf

% \input{response/response}

REVISION DUE: 01.12.2021

\section*{Reviewer 1}

My comments are the following:

The revised manuscript is much smoother than the first version, but I still feel that less than sufficient care was put into the writing. The reviewer pointed some in the pdf report. I have some additional editorial suggestions:

\begin{enumerate}

\item \PROGRESS Submit the point-by-point response to the reviewer's comments and my comments in a pdf file.

\begin{quote}
The original submisison web page had a field to enter the comments as text which we used. We will change this and introduce a PDF file instead, which is easier for us also.
\end{quote}

\item \DONE Needs page number (line number) for ease of referencing in the review.

\begin{quote}
Page numbers and line numbers were activated
\end{quote}

\item \DONE Do not italicize in the text.

\begin{quote}
All italicized text have been removed.
inline formulas are done by latex italized. 
\end{quote}

\item \DONE AICov needs to be defined in the abstract.

\begin{quote}
We added

"We present the architecture of an artificial intelligence enhanced COVID-19 analysis (in short AICov)"

was added to the abstract.
\end{quote}

\item \PROGRESS Acronyms need to be defined at their first occurrences (e.g., AI in the intro). An acronyme table is included in table 1, we added the following acronyms to the paper

\begin{quote}
    All acronyms have been defined at their first usage. In addition we have added a section Acronyms in which all acronyms are listed to make it easier for the reader to look them up in case they are used later on in the paper.
\end{quote}

\DOIT capitalization in definition sin the table 1 and others

\DOIT PERCENT, PERCENTBALCK, BLACK PERCENT, also for hispanics and senior

\DOIT use risk factor in table headings

\DOIT PER1000 is /1000

\DOIT make acronyms appear in text

\DOIT Table 1. capitalization,explain why they are different or make them the same

\item \DOIT  Figure 2: please present the equations as equations. The notations used in the equation need a full check (e.g., h is used as both the dimension and vector). I'd like to see a clear distinction between inputs/outputs and parameters. I'd suggest that the inputs be set up before the equations are presented.

\begin{quote}
    \todo{todo}
\end{quote}

\item \DOIT  Regenerate the figures into pdf format (the current jpg/png figures blur when zoomed in).

\begin{quote}
    \todo{todo}
\end{quote}

\item \DONE  Table 2: consider reformating it; it would not fit the page width in print.

\begin{quote}
Table 2 has been reformatted. It was previously fit in the column with resize box. While rotating the headers the font is increased. The table still fits in the column as it still uses resizebox.
\end{quote}

\item \DONE  Table 3: tidy the headers.

\begin{quote}
    The headers have been cleaned up and linebreaks wer introduced so the resizebox of the table presents the table now larger
\end{quote}

\item \DOIT  Figure 5-6: can be combined to one figure with two panels (using ggplot2 and save the space of the labels of the horizontal axis).

\begin{quote}
    \todo{todo}
\end{quote}

\item \DOIT  Figure 7-8: can be one figure with two panels; pop density 2010 needs to be cleaned.

\begin{quote}
    \todo{todo}
\end{quote}

\item \DOIT  Figure 9-14: combine into one figure.

\begin{quote}
    \todo{todo}
\end{quote}

\item \DONE  Table 4: keep the number of digits consistent.

\begin{quote}
    The digits have been padded with the missing 0 to make them all the same length
\end{quote}

\item \DOIT  Reproducibility: submit data/code/README in a compressed file as supplementary material, or put them in a public repo and give the link under an unnumbered section Supplementary Material.

\begin{quote}
    \todo{todo}
\end{quote}

\end{enumerate}

To submit your revision, please log in to EJMS and submit it as a revised file to the original submission. Please also include a detailed description of how you addressed all the points raised by the reviewers.

\section*{Reviewer 2}

This paper presents AICov, an integrated deep learning framework for COVID-19 forecasting. It takes into account population covariates, and uses LSTM and event modeling to do the forecasting. Data integration is done at a local level. Integration of such information brings improvement to the prediction accuracy. Despite the authors’ effort to address my comments in their revision, the paper is still in rough shape. Please see my comments.

\begin{enumerate}

\item \DOIT Page 5, please write the full name for REST (representational state transfer ) when it first appears.

\begin{quote}
    The term has been defined at first usage, we also checked all other abbreviations.

    \todo{check other abbreviations}

\end{quote}

\item \DOIT Page 5, line 3 under Requirement 3, “we are developed an abstraction API for data” does not read right.

\begin{quote}
    \todo{todo}
\end{quote}

\item \DOIT Page 6, line 8, what level is “local level”?

\begin{quote}
    \todo{todo}
\end{quote}

\item (partially done) Page 7, Table 1, is “CHD” also a percent? If yes please modify the description; Percent- blacks and black percent both correspond to “Percent of blacks in the population”. Both seem to have been used according to Table 3, with difference in the four model performance measurements. Do the authors have an explanation for this?

    \begin{quote}
    The table has been updated and it was indicated that it is in percent. There were indeed to different information sources that reported different values for percentage of blacks and other populations. This however has since than corrected and we have been able to use just one population value.

    \todo{fix the values and only have one population value}
    \end{quote}

\item \DOIT Page 8, line 5 of Section 5.1, the abbreviation LSTM has already appeared before, and the full name can be removed.

    \begin{quote}
        \todo{todo}
    \end{quote}

\item \DOIT What do the two panels in Figure 4 represent? Please give clearer description either in the text or figure caption.

    \begin{quote}
        \todo{todo}
    \end{quote}

\item \DOIT Section 5.2, paragraph 2 line 6, no need to capitalize for the word “Samples”. Line 5 from the bottom, no need to capitalize for the word “Insurance”. Please be careful with your writing.

    \begin{quote}
        \todo{todo}
    \end{quote}

\item \DOIT Section 5.2, the “order for cases sorted by minimum model order” and “order for deaths sorted by minimum model error” paragraphs are redundant, as the same information is already displayed in the x-legend for Figures 5 and 6.

    \begin{quote}
        \todo{todo}
    \end{quote}

\item \DOIT In Table 3, what is the unit for the error terms? The errors in Figures 7 and 8 are percentages. If errors in Table 3 are also percentages please clearly mark. Also, the numbers in Figures 7 and 8 do not align with numbers in row 1 of Table 3. Please check your numbers with care.

    \begin{quote}
        \todo{todo}
    \end{quote}

\item \DOIT Please adjust the location of your figures so they do not look pages apart from their interpretations.

    \begin{quote}
        \todo{todo}
    \end{quote}

\item \DOIT If I am reading it correctly, Figures 9 and 11 are only the “zoom in” versions of partial information from Figure 13. Similarly for the right column. Why waste space on two more figures when you could make Figures 13 and 14 bigger and explain the findings?

    \begin{quote}
        \todo{todo}
    \end{quote}

\item \DOIT The last paragraph of Section 5.2 suits the conclusion/discussion for future work section, not in the middle of a paper.

    \begin{quote}
        \todo{todo}
    \end{quote}

\item \DOIT Section 5.3, the authors mentioned two way of constructing covariates. Since the second way was more complicated while yielding no better performance the authors did not use it. Then it can be removed without damaging the storyline.

    \begin{quote}
        \todo{todo}
    \end{quote}

\item \DOIT The description “We renormalized, so the two basic features (following day) contributed 50% of the loss function” is hard to interpret. Please elaborate.

    \begin{quote}
        \todo{todo}
    \end{quote}

\item \DOIT Paragraph 2 on page 14, “Selu, Relu and Tanh activations gave very similar results while Sigmoid activation was much worse. However we still use a Sigmoid function...” is very counter-intuitive. Isn’t the point of trying different activation functions to better fit the data?
16. Paragraph 3 on page 14, “The basic COVID-19 daily data is extensive but compared two approaches with and without renormalization, but in this case, we ...” The sentence does not read correct. Please check your sentences with care!

    \begin{quote}
        \todo{todo}
    \end{quote}

\item \DOIT How to interpret the “significant positive correlation” in Section 6 from the practical perspec- tive? Also, the term “evolution function” made its first appearance in Section 6 without any definition or explanation; I suggest either removing the whole Section 6, or extend it with clearer definition of evolution function.
18. The first two paragraphs do not read like to be in a “Conclusion” section. Rather, they better fit into the introduction.

    \begin{quote}
        \todo{todo}
    \end{quote}

\end{enumerate}

\clearpage

### 2020-12-18 jds-paper-v3.tex
Updates from Overleaf

@@
 \newcommand{\DONE}{{\color{green!60!black}\makebox[0pt][l]{$\square$}\raisebox{.15ex}{\hspace{0.1em}$\checkmark$}}~}
-\newcommand{\DOIT}{{\color{red!60!black}\makebox[0pt][l]{$\square$}\raisebox{.15ex}{\hspace{0.1em}$\boxtimes}}~}
+\newcommand{\DOIT}{{\color{red!60!black}\makebox[0pt][l]{$\square$}\raisebox{.15ex}{\hspace{0.1em}$\boxtimes$}}~}
 \newcommand{\PROGRESS}{(in progress)~}
@@
 
-% \input{response/response}
-
-REVISION DUE: 01.12.2021
-
-\section*{Reviewer 1}
-
-My comments are the following:
-
-The revised manuscript is much smoother than the first version, but I still feel that less than sufficient care was put into the writing.
-The reviewer pointed some in the pdf report.
-I have some additional editorial suggestions:
-
-\begin{enumerate}
-
-\item \PROGRESS Submit the point-by-point response to the reviewer's comments and my comments in a pdf file.
-
-\begin{quote}
-The original submisison web page had a field to enter the comments as text which we used. We will change this and introduce a PDF file instead, which is easier for us also.
-\end{quote}
-
-\item \DONE Needs page number (line number) for ease of referencing in the review.
-
-\begin{quote}
-Page numbers and line numbers were activated
-\end{quote}
-
-\item \DONE Do not italicize in the text.
-
-\begin{quote}
-All italicized text have been removed.
-inline formulas are done by latex italized. 
-\end{quote}
-
-\item \DONE AICov needs to be defined in the abstract.
-
-\begin{quote}
-We added
-
-"We present the architecture of an artificial intelligence enhanced COVID-19 analysis (in short AICov)"
-
-was added to the abstract.
-\end{quote}
-
-\item \PROGRESS Acronyms need to be defined at their first occurrences (e.g., AI in the intro). An acronyme table is included in table 1, we added the following acronyms to the paper
-
-\begin{quote}
-    All acronyms have been defined at their first usage.
-In addition we have added a section Acronyms in which all acronyms are listed to make it easier for the reader to look them up in case they are used later on in the paper.
-\end{quote}
-
-\DOIT capitalization in definition sin the table 1 and others
-
-\DOIT PERCENT, PERCENTBALCK, BLACK PERCENT, also for hispanics and senior
-
-\DOIT use risk factor in table headings
-
-\DOIT PER1000 is /1000
-
-\DOIT make acronyms appear in text
-
-\DOIT Table 1. capitalization,explain why they are different or make them the same
-
-\item \DOIT  Figure 2: please present the equations as equations.
-The notations used in the equation need a full check (e.g., h is used as both the dimension and vector).
-I'd like to see a clear distinction between inputs/outputs and parameters.
-I'd suggest that the inputs be set up before the equations are presented.
-
-\begin{quote}
-    \todo{todo}
-\end{quote}
-
-\item \DOIT  Regenerate the figures into pdf format (the current jpg/png figures blur when zoomed in).
-
-\begin{quote}
-    \todo{todo}
-\end{quote}
-
-\item \DONE  Table 2: consider reformating it; it would not fit the page width in print.
-
-\begin{quote}
-Table 2 has been reformatted. It was previously fit in the column with resize box. While rotating the headers the font is increased. The table still fits in the column as it still uses resizebox.
-\end{quote}
-
-\item \DONE  Table 3: tidy the headers.
-
-\begin{quote}
-    The headers have been cleaned up and linebreaks wer introduced so the resizebox of the table presents the table now larger
-\end{quote}
-
-\item \DOIT  Figure 5-6: can be combined to one figure with two panels (using ggplot2 and save the space of the labels of the horizontal axis).
-
-\begin{quote}
-    \todo{todo}
-\end{quote}
-
-\item \DOIT  Figure 7-8: can be one figure with two panels; pop density 2010 needs to be cleaned.
-
-\begin{quote}
-    \todo{todo}
-\end{quote}
-
-\item \DOIT  Figure 9-14: combine into one figure.
-
-\begin{quote}
-    \todo{todo}
-\end{quote}
-
-\item \DONE  Table 4: keep the number of digits consistent.
-
-\begin{quote}
-    The digits have been padded with the missing 0 to make them all the same length
-\end{quote}
-
-\item \DOIT  Reproducibility: submit data/code/README in a compressed file as supplementary material, or put them in a public repo and give the link under an unnumbered section Supplementary Material.
-
-\begin{quote}
-    \todo{todo}
-\end{quote}
-
-\end{enumerate}
-
-To submit your revision, please log in to EJMS and submit it as a revised file to the original submission.
-Please also include a detailed description of how you addressed all the points raised by the reviewers.
-
-\section*{Reviewer 2}
-
-This paper presents AICov, an integrated deep learning framework for COVID-19 forecasting.
-It takes into account population covariates, and uses LSTM and event modeling to do the forecasting.
-Data integration is done at a local level.
-Integration of such information brings improvement to the prediction accuracy.
-Despite the authors’ effort to address my comments in their revision, the paper is still in rough shape.
-Please see my comments.
-
-\begin{enumerate}
-
-\item \DOIT Page 5, please write the full name for REST (representational state transfer ) when it first appears.
-
-\begin{quote}
-    The term has been defined at first usage, we also checked all other abbreviations.
-
-    \todo{check other abbreviations}
-
-\end{quote}
-
-\item \DOIT Page 5, line 3 under Requirement 3, “we are developed an abstraction API for data” does not read right.
-
-\begin{quote}
-    \todo{todo}
-\end{quote}
-
-\item \DOIT Page 6, line 8, what level is “local level”?
-
-\begin{quote}
-    \todo{todo}
-\end{quote}
-
-\item (partially done) Page 7, Table 1, is “CHD” also a percent?
-If yes please modify the description; Percent- blacks and black percent both correspond to “Percent of blacks in the population”.
-Both seem to have been used according to Table 3, with difference in the four model performance measurements.
-Do the authors have an explanation for this?
-
-    \begin{quote}
-    The table has been updated and it was indicated that it is in percent.
-There were indeed to different information sources that reported different values for percentage of blacks and other populations.
-This however has since than corrected and we have been able to use just one population value.
-
-    \todo{fix the values and only have one population value}
-    \end{quote}
-
-\item \DOIT Page 8, line 5 of Section 5.1, the abbreviation LSTM has already appeared before, and the full name can be removed.
-
-    \begin{quote}
-        \todo{todo}
-    \end{quote}
-
-\item \DOIT What do the two panels in Figure 4 represent? Please give clearer description either in the text or figure caption.
-
-    \begin{quote}
-        \todo{todo}
-    \end{quote}
-
-\item \DOIT Section 5.2, paragraph 2 line 6, no need to capitalize for the word “Samples”. Line 5 from the bottom, no need to capitalize for the word “Insurance”. Please be careful with your writing.
-
-    \begin{quote}
-        \todo{todo}
-    \end{quote}
-
-\item \DOIT Section 5.2, the “order for cases sorted by minimum model order” and “order for deaths sorted by minimum model error” paragraphs are redundant, as the same information is already displayed in the x-legend for Figures 5 and 6.
-
-    \begin{quote}
-        \todo{todo}
-    \end{quote}
-
-\item \DOIT In Table 3, what is the unit for the error terms?
-The errors in Figures 7 and 8 are percentages.
-If errors in Table 3 are also percentages please clearly mark.
-Also, the numbers in Figures 7 and 8 do not align with numbers in row 1 of Table 3.
-Please check your numbers with care.
-
-    \begin{quote}
-        \todo{todo}
-    \end{quote}
-
-\item \DOIT Please adjust the location of your figures so they do not look pages apart from their interpretations.
-
-    \begin{quote}
-        \todo{todo}
-    \end{quote}
-
-\item \DOIT If I am reading it correctly, Figures 9 and 11 are only the “zoom in” versions of partial information from Figure 13.
-Similarly for the right column.
-Why waste space on two more figures when you could make Figures 13 and 14 bigger and explain the findings?
-
-    \begin{quote}
-        \todo{todo}
-    \end{quote}
-
-\item \DOIT The last paragraph of Section 5.2 suits the conclusion/discussion for future work section, not in the middle of a paper.
-
-    \begin{quote}
-        \todo{todo}
-    \end{quote}
-
-\item \DOIT Section 5.3, the authors mentioned two way of constructing covariates.
-Since the second way was more complicated while yielding no better performance the authors did not use it.
-Then it can be removed without damaging the storyline.
-
-    \begin{quote}
-        \todo{todo}
-    \end{quote}
-
-\item \DOIT The description “We renormalized, so the two basic features (following day) contributed 50% of the loss function” is hard to interpret. Please elaborate.
-
-    \begin{quote}
-        \todo{todo}
-    \end{quote}
-
-\item \DOIT Paragraph 2 on page 14, “Selu, Relu and Tanh activations gave very similar results while Sigmoid activation was much worse.
-However we still use a Sigmoid function...” is very counter-intuitive.
-Isn’t the point of trying different activation functions to better fit the data?
-16.
-Paragraph 3 on page 14, “The basic COVID-19 daily data is extensive but compared two approaches with and without renormalization, but in this case, we ...” The sentence does not read correct.
-Please check your sentences with care!
-
-    \begin{quote}
-        \todo{todo}
-    \end{quote}
-
-\item \DOIT How to interpret the “significant positive correlation” in Section 6 from the practical perspec- tive?
-Also, the term “evolution function” made its first appearance in Section 6 without any definition or explanation; I suggest either removing the whole Section 6, or extend it with clearer definition of evolution function.
-18. The first two paragraphs do not read like to be in a “Conclusion” section. Rather, they better fit into the introduction.
-
-    \begin{quote}
-        \todo{todo}
-    \end{quote}
-
-\end{enumerate}
-
-\clearpage
+\input{response-2-content}
 

### 2020-12-18 vonLaszewski-covid.tex
79 char long lines for emacs

--- 2020-12-18 jds-paper-v3.tex
@@
 
-\input{response-2-content}
+\input{response-v3-content}
 
@@
 \newcommand{\ABSTRACT}{
-The COVID-19 pandemic has had profound global consequences on health, economic, social, behavioral, and almost every major aspect of human life.
-Therefore, it is of great importance to model COVID-19 and other pandemics in terms of the broader social contexts in which they take place.
-We present the architecture of an artificial intelligence enhanced COVID-19 analysis (in short AICov) , which provides an integrative deep learning framework for COVID-19 forecasting with population covariates, some of which may serve as putative risk factors.
-We have integrated multiple different strategies into AICov, including the ability to use deep learning strategies based on LSTM and event modeling.
-To demonstrate our approach, we have introduced a framework that integrates population covariates from multiple sources.
-Thus, AICov not only includes data on COVID-19 cases and deaths but, more importantly, the population's socioeconomic, health, and behavioral risk factors at a local level.
-The compiled data are fed into AICov, and thus we obtain improved prediction by the integration of the data to our model as compared to one that only uses case and death data.
-As we use deep learning our models adapt over time while learning the model from past data.
+The COVID-19 pandemic has had profound global consequences on health,
+economic, social, behavioral, and almost every major aspect of human
+life. Therefore, it is of great importance to model COVID-19 and other
+pandemics in terms of the broader social contexts in which they take
+place.  We present the architecture of an artificial intelligence
+enhanced COVID-19 analysis (in short AICov) , which provides an
+integrative deep learning framework for COVID-19 forecasting with
+population covariates, some of which may serve as putative risk
+factors.  We have integrated multiple different strategies into AICov,
+including the ability to use deep learning strategies based on LSTM
+and event modeling.  To demonstrate our approach, we have introduced a
+framework that integrates population covariates from multiple
+sources. Thus, AICov not only includes data on COVID-19 cases and
+deaths but, more importantly, the population's socioeconomic, health,
+and behavioral risk factors at a local level. The compiled data are
+fed into AICov, and thus we obtain improved prediction by the
+integration of the data to our model as compared to one that only uses
+case and death data. As we use deep learning our models adapt over
+time while learning the model from past data.
 }
@@
 
-The COVID-19 pandemic has caused an enormous health and humanitarian crisis worldwide. It is unlike any other single phenomenon that has occurred in modern history since the end of World War II. 
-The pandemic’s effects have spanned over a range that is so vast over space, and yet so condensed over time, that the dual blows of intensity and rapidity have exposed myriad systemic vulnerabilities in many societies around the world.
-
-The first known case was traced back to  17  November  2019.
-Since the emergence of a cluster of cases in Wuhan, China, on 31 December 2019, COVID-19 has spread rapidly worldwide.
-Just six months later, as of 23 June 2020, there are 8,993,659 COVID-19 cases and 469,587 deaths globally.
-On 30 January 2020, the World Health Organization (WHO) \cite{www-who} declared COVID-19 to be a Public Health Emergency of International Concern and subsequently, on 11 March 2020, a pandemic.
-
-In the United States, the first case of COVID-19 was confirmed on 20 January 2020 in Washington State, and as of 23 June 2020, there were over 2.3 million confirmed cases and 121,167 deaths \cite{www-nyt-map}.
-Although initial efforts to reduce the spread took place, recent data shows a resurgence at an increased scale.
-In the absence of a vaccine or treatment to effectively combat the disease, non-pharmaceutical policy interventions such as social distancing and lockdowns are recommended by health experts to prevent further transmission.
-To gain insights into the possible impact of such measures on COVID-19 outcomes, we depend upon the ability to accurately forecast the spread of reported cases and confirmed deaths and recoveries.
-Naturally, the accuracy of forecasting relies on the availability of current, reliable data and historical data to determine estimates of uncertainty.
-
-During outbreaks of epidemics, initially data on cases and deaths could be scarce, and the quality of data annotation, validation, and aggregation might be uncertain.
-For instance, changes in clinical data entry, such as the addition of a new category clinically diagnosed to the existing lab-confirmed category, may likely have been reflected in the reporting \cite{Petropoulos2020-hh}.
-Initial fears about COVID-19 among the general population with regards to the trajectory of the pandemic could also affect administrative reporting.
-Diverse media sources and differences in local and federal policies may add to the general uncertainty about disease progression.
-Nevertheless, a data-driven approach to forecasting can offer valuable insights into the disease dynamics, and thereby an ability to objectively plan for the near future.
-
-Unlike earlier global outbreaks, during COVID-19, the current worldwide digital ecosystem allows for real-time data collection, which in turn is used by artificial intelligence (AI) and deep learning systems to understand healthcare trends, model risk associations, and predict outcomes \cite{Ting2020-lw}.
-In addition to traditional public health surveillance strategies available to most countries, a variety of static and dynamic data types may be integrated to model the scale and dynamics of this pandemic \cite{Pyne2015-ao}.
-Several organizations, including the Johns Hopkins University’s Center for Systems Science and Engineering, the New York Times, the Atlantic’s COVID-19 tracking project, among others have developed real-time tracking maps for following COVID-19 cases around the world using data from the U.S.~Centers for Disease Control and Prevention (CDC), WHO, and other international health agencies.
-Notably, the CDC receives forecasts from several modeling groups that use a wide variety of approaches ranging from SEIR (Susceptible-Exposed-Infectious-Recovered) to Agent-based to Bayesian models to understand the non-pharmaceutical policy interventions (or the lack thereof) to predict disease dynamics and its impact on human lives \cite{www-cdc-modeling-forecast}.
-
-For precise and contextually relevant modeling of COVID-19 dynamics for a given population or community, we use data integration to combine community-specific health and behavioral risk factors, demographic and socioeconomic variables that are available at the local levels along with the cases and deaths data for COVID-19 while leveraging the recently developed and increasingly popular deep learning strategies to forecast COVID-19 dynamics using our model.
-Towards this, we develop a parallel computing platform called AICov standing for  AI-driven Platform for COVID-19 to implement the different components of our framework and obtain the needed resources through multi-cloud interfaces in a robust manner.
-
-In this study, while we incorporate information from community and county-specific covariates to inform our forecasting model for each metropolitan area, we want to avoid the so-called ecological fallacy in drawing inferences about individual disease outcomes based on large area-level aggregated risk factors.
-Our objective is also to underscore the nuanced roles played by pre-existing socioeconomic and other prevailing conditions of a given community that can act as determinants of its health outcomes and possible disparities,  especially under such sudden stresses to its current systems as those felt during any pandemic.
-It is important to note that the underlying models are learned directly from the data instead of being projected by a theoretical epidemiological model that relies on simulation populations.
-This allows quick adaptation and self-adaptation in regards to changing behaviors and regional differences.
-
-The paper is structured as follows.
-In Section \ref{sec:lstm-theory}, we provide a short overview of deep learning-based time series forecasting that we use in AICov.
-In Section \ref{sec:arch}, we outline our architecture that enables the user to conduct powerful time series forecasting by integrating static and dynamic information on population risk factors with time-series disease data.
-In Section \ref{sec:data}, we describe the input data for AICov.
-Next, we demonstrate our analysis involving the population risk factors and identify those that contribute to our prediction accuracy.
-Finally, we present our conclusion.
+The COVID-19 pandemic has caused an enormous health and humanitarian
+crisis worldwide. It is unlike any other single phenomenon that has
+occurred in modern history since the end of World War II.  The
+pandemic’s effects have spanned over a range that is so vast over
+space, and yet so condensed over time, that the dual blows of
+intensity and rapidity have exposed myriad systemic vulnerabilities in
+many societies around the world.
+
+The first known case was traced back to 17 November 2019. Since the
+emergence of a cluster of cases in Wuhan, China, on 31 December 2019,
+COVID-19 has spread rapidly worldwide. Just six months later, as of 23
+June 2020, there are 8,993,659 COVID-19 cases and 469,587 deaths
+globally. On 30 January 2020, the World Health Organization (WHO)
+\cite{www-who} declared COVID-19 to be a Public Health Emergency of
+International Concern and subsequently, on 11 March 2020, a pandemic.
+
+In the United States, the first case of COVID-19 was confirmed on 20
+January 2020 in Washington State, and as of 23 June 2020, there were
+over 2.3 million confirmed cases and 121,167 deaths
+\cite{www-nyt-map}.  Although initial efforts to reduce the spread
+took place, recent data shows a resurgence at an increased scale. In
+the absence of a vaccine or treatment to effectively combat the
+disease, non-pharmaceutical policy interventions such as social
+distancing and lockdowns are recommended by health experts to prevent
+further transmission. To gain insights into the possible impact of
+such measures on COVID-19 outcomes, we depend upon the ability to
+accurately forecast the spread of reported cases and confirmed deaths
+and recoveries. Naturally, the accuracy of forecasting relies on the
+availability of current, reliable data and historical data to
+determine estimates of uncertainty.
+
+During outbreaks of epidemics, initially data on cases and deaths
+could be scarce, and the quality of data annotation, validation, and
+aggregation might be uncertain. For instance, changes in clinical data
+entry, such as the addition of a new category clinically diagnosed to
+the existing lab-confirmed category, may likely have been reflected in
+the reporting \cite{Petropoulos2020-hh}. Initial fears about COVID-19
+among the general population with regards to the trajectory of the
+pandemic could also affect administrative reporting. Diverse media
+sources and differences in local and federal policies may add to the
+general uncertainty about disease progression. Nevertheless, a
+data-driven approach to forecasting can offer valuable insights into
+the disease dynamics, and thereby an ability to objectively plan for
+the near future.
+
+Unlike earlier global outbreaks, during COVID-19, the current
+worldwide digital ecosystem allows for real-time data collection,
+which in turn is used by artificial intelligence (AI) and deep
+learning systems to understand healthcare trends, model risk
+associations, and predict outcomes \cite{Ting2020-lw}.  In addition to
+traditional public health surveillance strategies available to most
+countries, a variety of static and dynamic data types may be
+integrated to model the scale and dynamics of this pandemic
+\cite{Pyne2015-ao}. Several organizations, including the Johns Hopkins
+University’s Center for Systems Science and Engineering, the New York
+Times, the Atlantic’s COVID-19 tracking project, among others have
+developed real-time tracking maps for following COVID-19 cases around
+the world using data from the U.S.~Centers for Disease Control and
+Prevention (CDC), WHO, and other international health
+agencies. Notably, the CDC receives forecasts from several modeling
+groups that use a wide variety of approaches ranging from SEIR
+(Susceptible-Exposed-Infectious-Recovered) to Agent-based to Bayesian
+models to understand the non-pharmaceutical policy interventions (or
+the lack thereof) to predict disease dynamics and its impact on human
+lives \cite{www-cdc-modeling-forecast}.
+
+For precise and contextually relevant modeling of COVID-19 dynamics
+for a given population or community, we use data integration to
+combine community-specific health and behavioral risk factors,
+demographic and socioeconomic variables that are available at the
+local levels along with the cases and deaths data for COVID-19 while
+leveraging the recently developed and increasingly popular deep
+learning strategies to forecast COVID-19 dynamics using our
+model. Towards this, we develop a parallel computing platform called
+AICov standing for AI-driven Platform for COVID-19 to implement the
+different components of our framework and obtain the needed resources
+through multi-cloud interfaces in a robust manner.
+
+In this study, while we incorporate information from community and
+county-specific covariates to inform our forecasting model for each
+metropolitan area, we want to avoid the so-called ecological fallacy
+in drawing inferences about individual disease outcomes based on large
+area-level aggregated risk factors. Our objective is also to
+underscore the nuanced roles played by pre-existing socioeconomic and
+other prevailing conditions of a given community that can act as
+determinants of its health outcomes and possible disparities,
+especially under such sudden stresses to its current systems as those
+felt during any pandemic.  It is important to note that the underlying
+models are learned directly from the data instead of being projected
+by a theoretical epidemiological model that relies on simulation
+populations. This allows quick adaptation and self-adaptation in
+regards to changing behaviors and regional differences.
+
+The paper is structured as follows. In Section \ref{sec:lstm-theory},
+we provide a short overview of deep learning-based time series
+forecasting that we use in AICov. In Section \ref{sec:arch}, we
+outline our architecture that enables the user to conduct powerful
+time series forecasting by integrating static and dynamic information
+on population risk factors with time-series disease data. In Section
+\ref{sec:data}, we describe the input data for AICov. Next, we
+demonstrate our analysis involving the population risk factors and
+identify those that contribute to our prediction accuracy. Finally, we
+present our conclusion.
 
@@
 
-In this study, we used a Long Short-Term Memory (LSTM) algorithm for our predictions \cite{www-keras-lstm,Hochreiter1997-dk}.
-An LSTM is a recurrent neural network (RNN) \cite{Rumelhart1986-li} with feedback connections allowing the use of data input sequences to predict data output sequences.
-LSTMs have been applied to many application areas from handwriting \cite{Graves2009-qb}, image, feature detection, and time series prediction \cite{Schmidhuber2005-oy}.
-LSTM's have an internal state that is used to prevent the vanishing gradient problem \cite{Hochreiter1991-mp} during the training of RNNs.
-
-Different variants of LSTM algorithms exist. Ours is based on an LSTM cell depicted in Figure \ref{fig:lstm}. 
-It has an input gate, output gate, and a forget gate.
-The cell is maintaining how a subsequent value is calculated.
-This includes (a) how input values are influencing the cell via the input gate, (b) how the forget gate influences a memory value within the cell via the forget gate, and (c) how the output gate influences the output activation of the LSTM unit.
-A cell has a number of inputs and outputs that are weighted.
-During a training step, these weights are learned to be reused in a prediction process between input and output sequences.
+In this study, we used a Long Short-Term Memory (LSTM) algorithm for
+our predictions \cite{www-keras-lstm,Hochreiter1997-dk}. An LSTM is a
+recurrent neural network (RNN) \cite{Rumelhart1986-li} with feedback
+connections allowing the use of data input sequences to predict data
+output sequences. LSTMs have been applied to many application areas
+from handwriting \cite{Graves2009-qb}, image, feature detection, and
+time series prediction \cite{Schmidhuber2005-oy}. LSTM's have an
+internal state that is used to prevent the vanishing gradient problem
+\cite{Hochreiter1991-mp} during the training of RNNs.
+
+Different variants of LSTM algorithms exist. Ours is based on an LSTM
+cell depicted in Figure \ref{fig:lstm}.  It has an input gate, output
+gate, and a forget gate. The cell is maintaining how a subsequent
+value is calculated. This includes (a) how input values are
+influencing the cell via the input gate, (b) how the forget gate
+influences a memory value within the cell via the forget gate, and (c)
+how the output gate influences the output activation of the LSTM
+unit. A cell has a number of inputs and outputs that are
+weighted. During a training step, these weights are learned to be
+reused in a prediction process between input and output sequences.
 
@@
 
-We have summarized the equations used in LSTM in Figure~\ref{fig:eq}.
-We denote variables in lower case and matrices in upper case letters.
-$W_{q}$ and $U_{q}$ contain the weights of the input and recurrent connections, where the subscript denotes either the input gate $i$, the output gate $o$, the forget gate $f$ or the memory cell $c$.
-It is important to note that  $c_{t}\in \mathbb {R}^{h}$ is not just one cell of one LSTM unit, but contains $h$ LSTM units' cells, representing a vector.
-To summarize our variables we use:
+We have summarized the equations used in LSTM in
+Figure~\ref{fig:eq}. We denote variables in lower case and matrices in
+upper case letters. $W_{q}$ and $U_{q}$ contain the weights of the
+input and recurrent connections, where the subscript denotes either
+the input gate $i$, the output gate $o$, the forget gate $f$ or the
+memory cell $c$.  It is important to note that $c_{t}\in \mathbb
+{R}^{h}$ is not just one cell of one LSTM unit, but contains $h$ LSTM
+units' cells, representing a vector. To summarize our variables we
+use:
 $\circ$ is the Hadamard product,
@@
 $c_{t}\in \Rh$ the cell state vector,
-$W\in \mathbb{R}^{h\times d}$, $U\in \mathbb{R} ^{h\times h}$, and $b\in \Rh$ are the weight matrices and bias vector parameters which need to be learned during training.
+$W\in \mathbb{R}^{h\times d}$, $U\in \mathbb{R} ^{h\times h}$, and
+$b\in \Rh$ are the weight matrices and bias vector parameters which
+need to be learned during training.
 
@@
 
-The motivation for creating this architecture is based on our experience with deep learning toolkits such as Keras and PyTorch.
-While these systems provide the necessary APIs to integrate time series solutions, they target primarily more general user communities.
-In particular, they do not consider existing data sets or specific needs and analysis options to evaluate them for the purpose of COVID-19 risk factor data integration.
-We describe below the different requirements that motivate the design of our architecture framework, AICov, and how we address each of these.
+The motivation for creating this architecture is based on our
+experience with deep learning toolkits such as Keras and
+PyTorch. While these systems provide the necessary APIs to integrate
+time series solutions, they target primarily more general user
+communities. In particular, they do not consider existing data sets or
+specific needs and analysis options to evaluate them for the purpose
+of COVID-19 risk factor data integration. We describe below the
+different requirements that motivate the design of our architecture
+framework, AICov, and how we address each of these.
 
@@
 
-\begin{requirement}{\bf Ease of Use.} A major design criterion for AICov is that the framework must be easy to use, and extensions can be made to allow usability of both the API as well as the user interface.
-\Solution To address these requirements, we are devising a specialized but easy to use time series API for COVID-19 that integrates common tasks such as automated filtering and normalization of the data.
+\begin{requirement}{\bf Ease of Use.}
+  A major design criterion for AICov is that the framework must be
+  easy to use, and extensions can be made to allow usability of both
+  the API as well as the user interface. \Solution To address these
+  requirements, we are devising a specialized but easy to use time
+  series API for COVID-19 that integrates common tasks such as
+  automated filtering and normalization of the data.
 \end{requirement}
 
-\begin{requirement}{\bf Interactive Access.} Due to the experimental nature of analyzing the data, AICov must support the exploration of the data in an interactive fashion.
-\Solution As many data scientists use Jupyter Notebooks to integrate their work interactively, AICov integrates such notebooks.
-Jupyter notebooks allow rapid modification of the analysis workflow through the ease of Python as programming language.
-Additionally, they enable us to formulate easily sophisticated scientific analytics workflows with the help of Python.
+\begin{requirement}{\bf Interactive Access.}
+  Due to the experimental nature of analyzing the data, AICov must
+  support the exploration of the data in an interactive
+  fashion. \Solution As many data scientists use Jupyter Notebooks to
+  integrate their work interactively, AICov integrates such
+  notebooks. Jupyter notebooks allow rapid modification of the
+  analysis workflow through the ease of Python as programming
+  language. Additionally, they enable us to formulate easily
+  sophisticated scientific analytics workflows with the help of
+  Python.
 \end{requirement}
 
-\begin{requirement}{\bf Expandable API.} Assuming that new models and other analysis algorithms will be developed over time, AICov must allow the integration of such APIs.
-\Solution To address this requirement, we are developed an abstraction API for data, analysis, and metadata parameter adjustments.
-External services could be integrated with the help of REST (REpresentational State Transfer) services.
-The REST services are automatically generated from function specifications.
+\begin{requirement}{\bf Expandable API.}
+  Assuming that new models and other analysis algorithms will be
+  developed over time, AICov must allow the integration of such
+  APIs. \Solution To address this requirement, we are developed an
+  abstraction API for data, analysis, and metadata parameter
+  adjustments. External services could be integrated with the help of
+  REST (REpresentational State Transfer) services. The REST services
+  are automatically generated from function specifications.
 \end{requirement}
 
-\begin{requirement}{\bf Flexible Data Source Integration.} The integration of various sources of data is a key part of AICov.
-\Solution To achieve this goal, we provide a number of abstractions and data manipulation functions to extract needed data from established sources.
-Furthermore, the data can be combined, and additional data wrangling by external groups can be integrated.
-Through such abstractions, it is possible to replace, add, and correct data sets that are part of the analysis.
-This allows us to compare different results created from different data sets.
-Moreover, we include data selection directly in our framework to easily and quickly reuse them in a data mashup.
-An important example is that the data need not be static over the disease's progression but can be continuously updated.
-This also has a significant impact on our analytics algorithms.
+\begin{requirement}{\bf Flexible Data Source Integration.}
+  The integration of various sources of data is a key part of
+  AICov. \Solution To achieve this goal, we provide a number of
+  abstractions and data manipulation functions to extract needed data
+  from established sources. Furthermore, the data can be combined, and
+  additional data wrangling by external groups can be integrated.
+  Through such abstractions, it is possible to replace, add, and
+  correct data sets that are part of the analysis. This allows us to
+  compare different results created from different data
+  sets. Moreover, we include data selection directly in our framework
+  to easily and quickly reuse them in a data mashup. An important
+  example is that the data need not be static over the disease's
+  progression but can be continuously updated. This also has a
+  significant impact on our analytics algorithms.
 \end{requirement}
 
-\begin{requirement}{\bf Automated Update of the Analytics.} As the data can change, previous analyses may have to be updated.
-\Solution Our forecast is not just a single script but includes a mechanism to register multiple workflows of continuously integrated (CI) analysis that are automatically rerun when new data are made available to the system.
-Previous results are maintained.
-Metadata with these runs can retrieve the versioned data sets used to calculate and store the versioned analytics workflow.
+\begin{requirement}{\bf Automated Update of the Analytics.}
+  As the data can change, previous analyses may have to be
+  updated. \Solution Our forecast is not just a single script but
+  includes a mechanism to register multiple workflows of continuously
+  integrated (CI) analysis that are automatically rerun when new data
+  are made available to the system. Previous results are
+  maintained. Metadata with these runs can retrieve the versioned data
+  sets used to calculate and store the versioned analytics workflow.
 \end{requirement}
 
-\begin{requirement}{\bf Flexible Model definition.} We want to allow experimentation with various models in AICov that may be easily and flexibly definable as workflows.
-\Solution We give functional abstraction definitions that enables us to define the models ourselves or integrate third party models describing the spread of the disease.
+\begin{requirement}{\bf Flexible Model definition.}
+  We want to allow experimentation with various models in AICov that
+  may be easily and flexibly definable as workflows. \Solution We give
+  functional abstraction definitions that enables us to define the
+  models ourselves or integrate third party models describing the
+  spread of the disease.
 \end{requirement}
 
-\begin{requirement}{\bf Deep Learning Forecast.} As the forecasting models depend on changing data that are made available daily, it is important for AICov to utilize the newly available data.
-Many models can be applied to this, including moving averages based models or models that are purely derived from deep learning.
-\Solution To address this requirement, AICov allows for the integration of both.
-Furthermore, in our deep learning framework, we can  integrate an automated search for important risk factors, which is the focus of this study.
+\begin{requirement}{\bf Deep Learning Forecast.}
+  As the forecasting models depend on changing data that are made
+  available daily, it is important for AICov to utilize the newly
+  available data. Many models can be applied to this, including moving
+  averages based models or models that are purely derived from deep
+  learning. \Solution To address this requirement, AICov allows for
+  the integration of both. Furthermore, in our deep learning
+  framework, we can integrate an automated search for important risk
+  factors, which is the focus of this study.
 \end{requirement}
 
-\begin{requirement}{\bf Model Orchestration.} To coordinate the different model predictions and the generation of the deep learning forecasts, we need the ability to orchestrate them while applying a number of parameters.
-\Solution To address this requirement, the AICov architecture includes the ability to integrate parameter sweeps to, for example, identify hyperparameters, or integrate different data sets as parameters to identify suitable forecast models that can deliver the one which fits the best.
+\begin{requirement}{\bf Model Orchestration.}
+  To coordinate the different model predictions and the generation of
+  the deep learning forecasts, we need the ability to orchestrate them
+  while applying a number of parameters. \Solution To address this
+  requirement, the AICov architecture includes the ability to
+  integrate parameter sweeps to, for example, identify
+  hyperparameters, or integrate different data sets as parameters to
+  identify suitable forecast models that can deliver the one which
+  fits the best.
 \end{requirement}
 
-\begin{requirement}{\bf Compute Resource Mapping.} For other uses of AICov, we need to integrate a flexible resource utilization framework.
-\Solution To address this requirement, we leverage from our earlier work and utilize through Cloudmesh and NIST NBDIF definitions cloud services into the architecture \cite{las-19-nist}.
-This includes containers but also infrastructure services in the cloud.
-Through these service integrations, we also provide access to  GPUs.
+\begin{requirement}{\bf Compute Resource Mapping.}
+  For other uses of AICov, we need to integrate a flexible resource
+  utilization framework. \Solution To address this requirement, we
+  leverage from our earlier work and utilize through Cloudmesh and
+  NIST NBDIF definitions cloud services into the architecture
+  \cite{las-19-nist}. This includes containers but also infrastructure
+  services in the cloud. Through these service integrations, we also
+  provide access to GPUs.
 \end{requirement}
 
-\begin{requirement}{\bf Convenient Interfaces.} As we have a wide variety of users, we need to enable interfaces that are used by the various communities.
-\Solution To address this requirement, we provide a number of APIs, REST services and make them available via Jupyter Notebooks.
-In addition, we developed some custom widgets for the notebooks that are specifically targeted towards data integration, parameter manipulation, and visualization of the results
+\begin{requirement}{\bf Convenient Interfaces.}
+  As we have a wide variety of users, we need to enable interfaces
+  that are used by the various communities. \Solution To address this
+  requirement, we provide a number of APIs, REST services and make
+  them available via Jupyter Notebooks. In addition, we developed some
+  custom widgets for the notebooks that are specifically targeted
+  towards data integration, parameter manipulation, and visualization
+  of the results
 \end{requirement}
 
-Putting all these requirements together results in an architecture as shown in Figure \ref{fig:arch}.\\
+Putting all these requirements together results in an architecture as
+shown in Figure \ref{fig:arch}.\\
 
@@
 
-In summary, the AICov architecture allows the inclusion of new data sources and models.
-An orchestrator enables modifying parameters and selecting data sources and models used in the analysis.
-The framework can be accessed via the interface layer that includes APIs, REST, and Jupyter Notebooks.
-Compute interfaces are provided either by direct access of GPUs from local resources or through the staging of automatically generated REST services via our Cloudmesh REST Service integrator \cite{cloudmesh-openapi-install,cloudmesh-openapi-benchmark,cloudmesh-manual}.
-Additionally, we can also access through abstractions cloud services that are offered by the various cloud providers and are targeting our application domain.
-Different models can be easily integrated into our framework to verify our forecast or enhance our forecast based on the users' models and data.
+In summary, the AICov architecture allows the inclusion of new data
+sources and models. An orchestrator enables modifying parameters and
+selecting data sources and models used in the analysis. The framework
+can be accessed via the interface layer that includes APIs, REST, and
+Jupyter Notebooks. Compute interfaces are provided either by direct
+access of GPUs from local resources or through the staging of
+automatically generated REST services via our Cloudmesh REST Service
+integrator
+\cite{cloudmesh-openapi-install,cloudmesh-openapi-benchmark,cloudmesh-manual}. Additionally,
+we can also access through abstractions cloud services that are
+offered by the various cloud providers and are targeting our
+application domain. Different models can be easily integrated into our
+framework to verify our forecast or enhance our forecast based on the
+users' models and data.
 
@@
 
-While the U.S. has seen an overall case fatality rate of 5.14\%, 
-some U.S.
-locations have experienced a disproportionate number of deaths compared to others.
-This disparity of deaths among different communities could be attributed to a diversity of local risk factors and socioeconomic determinants of health that now constitute a very active area of investigation.
-Researchers are studying the effects of different community-specific socioeconomic variables, underlying health conditions and comorbidities, infrastructure, and systemic responses.
-Given the complexity of such underlying conditions that can determine individual disease outcomes,  where and under what conditions people get infected by the virus, and then recover or die take place not in isolation but rather in an interplay of diverse factors that are characteristic to a local community, which need to be synthesized.
-In other words, a basic requirement for effective modeling of COVID-19 dynamics is that the information regarding a particular community’s vulnerabilities and strengths pertaining to the disease must be as comprehensive and integrative as possible.
-The regular reporting of accurate Coronavirus outbreaks data from local health departments has been understandably difficult, especially in the areas that are hardest hit.
-We use data on community-specific health and behavioral risk factors aggregated at the county level from the large-scale longitudinal CDC BRFSS surveys \cite{www-cdc-brfss}, and demographic and socioeconomic data from the U.S.
-Census Bureau that are known at the local levels along with the cases and deaths data compiled by NCHS \cite{www-cdc-nchs} and integrate them with other data sources.
+While the U.S. has seen an overall case fatality rate of 5.14\%, some
+U.S. locations have experienced a disproportionate number of deaths
+compared to others. This disparity of deaths among different
+communities could be attributed to a diversity of local risk factors
+and socioeconomic determinants of health that now constitute a very
+active area of investigation. Researchers are studying the effects of
+different community-specific socioeconomic variables, underlying
+health conditions and comorbidities, infrastructure, and systemic
+responses.  Given the complexity of such underlying conditions that
+can determine individual disease outcomes, where and under what
+conditions people get infected by the virus, and then recover or die
+take place not in isolation but rather in an interplay of diverse
+factors that are characteristic to a local community, which need to be
+synthesized. In other words, a basic requirement for effective
+modeling of COVID-19 dynamics is that the information regarding a
+particular community’s vulnerabilities and strengths pertaining to the
+disease must be as comprehensive and integrative as possible.  The
+regular reporting of accurate Coronavirus outbreaks data from local
+health departments has been understandably difficult, especially in
+the areas that are hardest hit. We use data on community-specific
+health and behavioral risk factors aggregated at the county level from
+the large-scale longitudinal CDC BRFSS surveys \cite{www-cdc-brfss},
+and demographic and socioeconomic data from the U.S. Census Bureau
+that are known at the local levels along with the cases and deaths
+data compiled by NCHS \cite{www-cdc-nchs} and integrate them with
+other data sources.
 
@@
 
-Among the different sources of data on daily cases and deaths data collected and aggregated at different spatial levels, especially for the U.S., many are either freely available repositories such as GitHub or accessible via specifically developed online resources such as the New York Times~\cite{github-nytimes} and the Johns Hopkins Coronavirus Resource Center~\cite{www-jh-covid19}.
-For this study, we generated two data sets based on the latter resource upon our checking for the locations and times at which cumulative counts of COVID-19 cases and deaths on any given day were no less than those on the previous day.
-This led us to compile daily coronavirus cases and deaths data for 345 U.S.
-counties from 32 states and the District of Columbia and matched by five-digit FIPS code or county name to dynamic and static variables from additional data sources.
-
-The static data in this repository was collected from the U.S.
-Census Bureau and the Centers for Disease Control and Prevention (CDC).
-The different fields cover the domains of behavioral and health risk factors, hospital capacity, and socioeconomic and demographic conditions.
-Data on health outcomes, prevention measures, and unhealthy behaviors were extracted from the CDC's 500 Cities Local Data for Better Health program \cite{www-cdc-chronic-data} in the form of crude or age-adjusted prevalence of conditions such as respiratory disease, obesity or smoking.
-These values were aggregated from the census tract to the county level using the first five digits of the eleven-digit tract FIPS code.
-Because a single county may consist of multiple individual cities, we include the list of all city labels within each aggregate group to represent a greater metropolitan area.
-110 of such metropolitan areas that had more than 500 reported cases of COVID-19 by April 15, 2020, were selected for this study.
-
-Notably, a greater metropolitan area such as New York City could be spread across more than one county, which can lead to the complexity of aggregating counts and other variables.
-Time series data for the 5 individual counties (boroughs) in New York City were not available in the Johns Hopkins CSSE (Center for Systems Science and Engineering) data set.
-Rather, the total number of cases and deaths for the entire city of New York are reported and assigned the FIPS code of New York County (Manhattan).
-To ensure consistent geography across the static and dynamic data, we compute the population-weighted sum or median of each covariate over the assigned counties.
-
-Demographic variables were gathered from the Census QuickFacts~\cite{www-census} online resource using an automated web scraping algorithm and cover relevant areas such as age, race, income, and population density.
-Additional socioeconomic variables include the Gini Index, which measures economic inequality and CDC Social Vulnerability Index (SVI)~\cite{www-cdc-svi}.
-The SVI was created to guide public health officials and disaster response efforts by identifying the communities across the United States most likely to need support during a crisis.
-Census tracts and corresponding counties are ranked across 15 social factors, which are grouped into four themes: socioeconomic status, household composition and disability, minority status and language, and housing and transportation.
-
-Lastly, the total number of general acute care, critical access, and military hospitals within each county are included in the data.
-Such variables included the number of relevant hospitals per county, and the estimated number of beds (total known bed counts added to the number of hospitals in the county with missing data times the average number of beds per hospital in that state) per 1,000 people using the American Hospital Directory~\cite{www-ahd}.
+Among the different sources of data on daily cases and deaths data
+collected and aggregated at different spatial levels, especially for
+the U.S., many are either freely available repositories such as GitHub
+or accessible via specifically developed online resources such as the
+New York Times~\cite{github-nytimes} and the Johns Hopkins Coronavirus
+Resource Center~\cite{www-jh-covid19}.  For this study, we generated
+two data sets based on the latter resource upon our checking for the
+locations and times at which cumulative counts of COVID-19 cases and
+deaths on any given day were no less than those on the previous
+day. This led us to compile daily coronavirus cases and deaths data
+for 345 U.S. counties from 32 states and the District of Columbia and
+matched by five-digit FIPS code or county name to dynamic and static
+variables from additional data sources.
+
+The static data in this repository was collected from the U.S. Census
+Bureau and the Centers for Disease Control and Prevention (CDC). The
+different fields cover the domains of behavioral and health risk
+factors, hospital capacity, and socioeconomic and demographic
+conditions. Data on health outcomes, prevention measures, and
+unhealthy behaviors were extracted from the CDC's 500 Cities Local
+Data for Better Health program \cite{www-cdc-chronic-data} in the form
+of crude or age-adjusted prevalence of conditions such as respiratory
+disease, obesity or smoking. These values were aggregated from the
+census tract to the county level using the first five digits of the
+eleven-digit tract FIPS code. Because a single county may consist of
+multiple individual cities, we include the list of all city labels
+within each aggregate group to represent a greater metropolitan
+area. 110 of such metropolitan areas that had more than 500 reported
+cases of COVID-19 by April 15, 2020, were selected for this study.
+
+Notably, a greater metropolitan area such as New York City could be
+spread across more than one county, which can lead to the complexity
+of aggregating counts and other variables. Time series data for the 5
+individual counties (boroughs) in New York City were not available in
+the Johns Hopkins CSSE (Center for Systems Science and Engineering)
+data set. Rather, the total number of cases and deaths for the entire
+city of New York are reported and assigned the FIPS code of New York
+County (Manhattan). To ensure consistent geography across the static
+and dynamic data, we compute the population-weighted sum or median of
+each covariate over the assigned counties.
+
+Demographic variables were gathered from the Census
+QuickFacts~\cite{www-census} online resource using an automated web
+scraping algorithm and cover relevant areas such as age, race, income,
+and population density. Additional socioeconomic variables include the
+Gini Index, which measures economic inequality and CDC Social
+Vulnerability Index (SVI)~\cite{www-cdc-svi}. The SVI was created to
+guide public health officials and disaster response efforts by
+identifying the communities across the United States most likely to
+need support during a crisis. Census tracts and corresponding counties
+are ranked across 15 social factors, which are grouped into four
+themes: socioeconomic status, household composition and disability,
+minority status and language, and housing and transportation.
+
+Lastly, the total number of general acute care, critical access, and
+military hospitals within each county are included in the data. Such
+variables included the number of relevant hospitals per county, and
+the estimated number of beds (total known bed counts added to the
+number of hospitals in the county with missing data times the average
+number of beds per hospital in that state) per 1,000 people using the
+American Hospital Directory~\cite{www-ahd}.
 
@@
 
-To establish the feasibility of using deep learning, we have, in our analysis, verified that through deep learning we can recreate the predictions identified by a model-based approach using empirical fits.
-In this comprehensive analysis, we obtain predictions through deep learning while using data generated from epidemiological models as discussed in \cite{marsland20-covid-paper}.
-The time series we used was 100 days long, and a multi-layered Long Short-Term Memory (LSTM) recurrent network was used.
-Our prediction approach differs by learning not only from the demographics (fixed data for each city) and time-dependent data but by integrating the population model for the underlying prediction, as shown in Fig.
-\ref{fig:magic-1}.
-Such model predictive integration capability is important in any application with multiple time scales.
-For example, this allows us to integrate multi-scale time effects into the forecast, which could address the combination of general forecasts and the next time steps of choice, such as days, weeks, months, or longer.
-For this analysis, we used 37 of the 110 cities with reliable empirical fits (that we derived from an epidemiological model and not deep learning  \cite{marsland20-covid-paper}) for the cases and deaths data up to April 15, 2020.
-Our analysis identified a sophisticated single deep learning time evolution operator that can describe these 37  separate data sets, and smooth fitted data leads to very accurate deep learning descriptions via LSTM as depicted in both Figure \ref{fig:magic-1} that are smaller than 1\% from the model predictions.
+To establish the feasibility of using deep learning, we have, in our
+analysis, verified that through deep learning we can recreate the
+predictions identified by a model-based approach using empirical
+fits. In this comprehensive analysis, we obtain predictions through
+deep learning while using data generated from epidemiological models
+as discussed in \cite{marsland20-covid-paper}. The time series we used
+was 100 days long, and a multi-layered Long Short-Term Memory (LSTM)
+recurrent network was used. Our prediction approach differs by
+learning not only from the demographics (fixed data for each city) and
+time-dependent data but by integrating the population model for the
+underlying prediction, as shown in Fig. \ref{fig:magic-1}. Such model
+predictive integration capability is important in any application with
+multiple time scales. For example, this allows us to integrate
+multi-scale time effects into the forecast, which could address the
+combination of general forecasts and the next time steps of choice,
+such as days, weeks, months, or longer. For this analysis, we used 37
+of the 110 cities with reliable empirical fits (that we derived from
+an epidemiological model and not deep learning
+\cite{marsland20-covid-paper}) for the cases and deaths data up to
+April 15, 2020. Our analysis identified a sophisticated single deep
+learning time evolution operator that can describe these 37 separate
+data sets, and smooth fitted data leads to very accurate deep learning
+descriptions via LSTM as depicted in both Figure \ref{fig:magic-1}
+that are smaller than 1\% from the model predictions.
 
@@
     \includegraphics{images/magic-1.png}
-    \caption{Cumulative empirical fits can be very accurately predicted with LSTM.
-The data is taken from 37 U.S.
-Cities.
-Our comparison showcases that given a model, we can replicate the prediction of such models using the model data over time.
-The x-axis represents the days since the first data was fed to the model.}
+    \caption{Cumulative empirical fits can be very accurately
+      predicted with LSTM. The data is taken from 37 U.S. Cities. Our
+      comparison showcases that given a model, we can replicate the
+      prediction of such models using the model data over time. The
+      x-axis represents the days since the first data was fed to the
+      model.}
     \label{fig:magic-1}
@@
 
-In this  section, we establish that using deep learning can not only be applied to model predictions from  real time data.
-As we want to evaluate the influence of the risk factors on our prediction framework using not only models but deep learning algorithms, we have devised a parameter sweep framework based on LSTM that conducts the prediction for the 110 cities while also considering the risk factors as listed in Table \ref{tab:risk-factors}.
-This framework can query any combination of risk factors as well as not including any, in which case it defaults to the traditional LSTM algorithm as introduced in Section \ref{sec:lstm-theory}.
-Hence, we will answer Questions \ref{q:2} and \ref{q:3}.
-
-To answer these questions, we chose for our next analysis a date range between 2020-02-01 and 2020-05-25.
-We start with a basic LSTM operator with two layers of LSTM, and initial and final fully connected layers \cite{Kadupitiya2020-zq}.
-For LSTM, we have selected the activation function based on Rectified Linear Units (RELU).
-For the last layer sigmoid is chosen.
-A drop out rate of 0.2 is used, we use 16 input nodes, and have a batch size of 110.
-The maximum epoch is chosen to be 200.
-The number of nodes internally is 32.
-The maximum total number of Samples is 12540 when using 3 days as input length.
-We use a two-layer LSTM, as shown in Table \ref{tab:model}.
-Within our system, we have the ability to set a number of input days.
-For this analysis, we have chosen three different input lengths, namely 3, 4, and 5 days.
+In this section, we establish that using deep learning can not only be
+applied to model predictions from real time data. As we want to
+evaluate the influence of the risk factors on our prediction framework
+using not only models but deep learning algorithms, we have devised a
+parameter sweep framework based on LSTM that conducts the prediction
+for the 110 cities while also considering the risk factors as listed
+in Table \ref{tab:risk-factors}. This framework can query any
+combination of risk factors as well as not including any, in which
+case it defaults to the traditional LSTM algorithm as introduced in
+Section \ref{sec:lstm-theory}. Hence, we will answer Questions
+\ref{q:2} and \ref{q:3}.
+
+To answer these questions, we chose for our next analysis a date range
+between 2020-02-01 and 2020-05-25. We start with a basic LSTM operator
+with two layers of LSTM, and initial and final fully connected layers
+\cite{Kadupitiya2020-zq}.  For LSTM, we have selected the activation
+function based on Rectified Linear Units (RELU). For the last layer
+sigmoid is chosen. A drop out rate of 0.2 is used, we use 16 input
+nodes, and have a batch size of 110. The maximum epoch is chosen to be
+200. The number of nodes internally is 32. The maximum total number of
+Samples is 12540 when using 3 days as input length. We use a two-layer
+LSTM, as shown in Table \ref{tab:model}. Within our system, we have
+the ability to set a number of input days. For this analysis, we have
+chosen three different input lengths, namely 3, 4, and 5 days.
 
 \begin{table}[!h]
-    \caption{The deep learning parameters used for the model sweep creation in as exported by Keras using 1 risk factor.
-The activation function is RELU and the recurrent activation function is sigmoid.}
+    \caption{The deep learning parameters used for the model sweep
+      creation in as exported by Keras using 1 risk factor. The
+      activation function is RELU and the recurrent activation
+      function is sigmoid.}
     \label{tab:model}
@@
     \end{tabular}
-    }
-    ~\\
-    The LSTM contains a total of 6,910 trainable parameters in case we include one risk factor.
-In the case, we do not use any risk factor it is 6878, and when we include 33 risk factors, it is 7934.
+    } ~\\
+    The LSTM contains a total of 6,910 trainable parameters in
+    case we include one risk factor. In the case, we do not use any
+    risk factor it is 6878, and when we include 33 risk factors, it is
+    7934.
     \end{footnotesize}
@@
 
-We have run the training on all 110 cities from our data that represented each FIPS while not including any risk factors (Question \ref{q:2}) and running them with a single risk factor while iterating over all risk factors (Question \ref{q:3}).
-We then summed up all absolute errors for the predicted data points over the same time period and repeated each experiment 10 times.
-We summarized all the results in the form of a box-whisker diagram each for deaths and for cases.
-We sorted the diagram in such a fashion that the risk factor that has the model with the lowest error appears first.
-We preceded the graph with the experiment labeled as ``NONE'' that which does not use any risk factors to provide a comparison.
-The graphs are depicted in Figures \ref{fig:box-cases} for cases, and \ref{fig:box-death} for deaths.
+We have run the training on all 110 cities from our data that
+represented each FIPS while not including any risk factors (Question
+\ref{q:2}) and running them with a single risk factor while iterating
+over all risk factors (Question \ref{q:3}). We then summed up all
+absolute errors for the predicted data points over the same time
+period and repeated each experiment 10 times. We summarized all the
+results in the form of a box-whisker diagram each for deaths and for
+cases. We sorted the diagram in such a fashion that the risk factor
+that has the model with the lowest error appears first. We preceded
+the graph with the experiment labeled as ``NONE'' that which does not
+use any risk factors to provide a comparison. The graphs are depicted
+in Figures \ref{fig:box-cases} for cases, and \ref{fig:box-death} for
+deaths.
 
@@
     \vspace{-1cm}
-    \caption{Influence of risk factors on the accuracy of the prediction of COVID-19 cases. The y-axis represents the cumulative error over all input data for the cities.}
+    \caption{Influence of risk factors on the accuracy of the
+      prediction of COVID-19 cases. The y-axis represents the
+      cumulative error over all input data for the cities.}
     \label{fig:box-cases}
@@
     \vspace{-1cm}
-    \caption{Influence of risk factors on the accuracy of the prediction of COVID-19 death. The y-axis represents the cumulative error over all input data for the cities.}
+    \caption{Influence of risk factors on the accuracy of the
+      prediction of COVID-19 death. The y-axis represents the
+      cumulative error over all input data for the cities.}
     \label{fig:box-death}
@@
 
-Further, the obtained sorted order of the best prediction models by minimum error based on the risk factors for cases and deaths are as follows:
+Further, the obtained sorted order of the best prediction models by
+minimum error based on the risk factors for cases and deaths are as
+follows:
 
@@
 
-\item[Order for cases sorted by minimum model error:] NONE, POP\_DENSITY\_2010, PHLTH, INSURANCE, PERCENTBLACKS, PERCENTHISPANICS, NHOSP, INSURANCE, CHOLSCREEN, DIABETES, STROKE, CHD, CHECKUP, NBEDS, SVI\_OVERALL, KIDNEY, CASTHMA,
-BLACK, BPMED, LPA, CSMOKING, ARTHRITIS, POVERTY, ESTBEDS, MHLTH, SENIOR, COPD, CANCER, NBEDS/1000, BPHIGH, HIGHCHOL, BINGE, OBESITY, SVI\_MINORITY
-
-\item[Order for deaths sorted by minimum model error:]
-NONE, SVI\_OVERALL, CSMOKING, PERCENTHISPANICS, POP\_DENSITY\_2010, SENIOR, CHOLSCREEN, CASTHMA, INSURANCE, INSURANCE, MHLTH, LPA, POVERTY, OBESITY, CANCER, BLACK, NBEDS/1000, PERCENTBLACKS, DIABETES, ARTHRITIS, NBEDS, NHOSP, BPHIGH, CHECKUP, PHLTH, CHD, STROKE, HIGHCHOL, ESTBEDS, KIDNEY, BINGE, BPMED, COPD, SVI\_MINORITY
+\item[Order for cases sorted by minimum model error:] NONE,
+  POP\_DENSITY\_2010, PHLTH, INSURANCE, PERCENTBLACKS,
+  PERCENTHISPANICS, NHOSP, INSURANCE, CHOLSCREEN, DIABETES, STROKE,
+  CHD, CHECKUP, NBEDS, SVI\_OVERALL, KIDNEY, CASTHMA, BLACK, BPMED,
+  LPA, CSMOKING, ARTHRITIS, POVERTY, ESTBEDS, MHLTH, SENIOR, COPD,
+  CANCER, NBEDS/1000, BPHIGH, HIGHCHOL, BINGE, OBESITY, SVI\_MINORITY
+
+\item[Order for deaths sorted by minimum model error:] NONE,
+  SVI\_OVERALL, CSMOKING, PERCENTHISPANICS, POP\_DENSITY\_2010,
+  SENIOR, CHOLSCREEN, CASTHMA, INSURANCE, INSURANCE, MHLTH, LPA,
+  POVERTY, OBESITY, CANCER, BLACK, NBEDS/1000, PERCENTBLACKS,
+  DIABETES, ARTHRITIS, NBEDS, NHOSP, BPHIGH, CHECKUP, PHLTH, CHD,
+  STROKE, HIGHCHOL, ESTBEDS, KIDNEY, BINGE, BPMED, COPD, SVI\_MINORITY
 \end{description}
 
-We show the errors and the association with a risk factor in Table \ref{tab:top-1}, where cum\_error represents the cumulative error of the prediction.
-We sorted it by the cumulative error for cases.
-We see that in our experiment, almost all risk factors lead to an increased model prediction accuracy.
-As a result, we note that by including factors such as population density, physical health Insurance, population breakdown by ethnicity, number of hospitals, and diabetes, it leads to better overall predictions for cases.
-We also see that many of these factors perform better on average.
+We show the errors and the association with a risk factor in Table
+\ref{tab:top-1}, where cum\_error represents the cumulative error of
+the prediction. We sorted it by the cumulative error for cases. We see
+that in our experiment, almost all risk factors lead to an increased
+model prediction accuracy. As a result, we note that by including
+factors such as population density, physical health Insurance,
+population breakdown by ethnicity, number of hospitals, and diabetes,
+it leads to better overall predictions for cases. We also see that
+many of these factors perform better on average.
 
 \begin{table}[!p]
-\caption{Best models for each risk factor and their respective errors. The model that uses no risk factor is highlighted in grey.}
+\caption{Best models for each risk factor and their respective
+  errors. The model that uses no risk factor is highlighted in grey.}
 \label{tab:top-1}
@@
 
-We observed a different scenario for deaths, in which we only identified two risk factors, namely the Social Vulnerability Index (SVI) and chain-smoking, that, when integrated into our deep learning model, leads to better prediction results.
-However, the accuracy of the overall prediction of deaths is far less precise than that of the cases.
-To showcase this, we have included Figures \ref{fig:error-case-popdensity} and \ref{fig:error-death-popdensity} that show the respective errors.
+We observed a different scenario for deaths, in which we only
+identified two risk factors, namely the Social Vulnerability Index
+(SVI) and chain-smoking, that, when integrated into our deep learning
+model, leads to better prediction results. However, the accuracy of
+the overall prediction of deaths is far less precise than that of the
+cases. To showcase this, we have included Figures
+\ref{fig:error-case-popdensity} and \ref{fig:error-death-popdensity}
+that show the respective errors.
 
@@
     \vspace{-1cm}
-    \caption{Model prediction error for cumulative cases when including one risk factor - POP\_DENSITY\_2010 in the features, in addition to the past number of cases and deaths.}
+    \caption{Model prediction error for cumulative cases when
+      including one risk factor - POP\_DENSITY\_2010 in the features,
+      in addition to the past number of cases and deaths.}
     \label{fig:error-case-popdensity}
@@
     \vspace{-1cm}
-    \caption{Model prediction error for cumulative deaths when including one risk factor - POP\_DENSITY\_2010  in the features, in addition to the past number of cases and deaths.}
+    \caption{Model prediction error for cumulative deaths when
+      including one risk factor - POP\_DENSITY\_2010 in the features,
+      in addition to the past number of cases and deaths.}
     \label{fig:error-death-popdensity}
@@
 
-This is following our input data distribution in which we noted higher fluctuations relative to the overall value.
-A mitigation to this issue would be using a seven day average over the analyzed period.
-However, we have not done this on purpose for this analysis as we wanted to identify how a covariate enhanced LSTM behaves that includes risk factors based on the daily fluctuations.
-This was done to avoid and showcase any needed preproccessing on the data and to identify the capability of the deep learning framework while adapting to the fluctuations without any special activities.
-Previously, we have already shown in Section \ref{sec:emperical} that for smooth data inputs the prediction is very accurate.
-Using fluctuation data allows our framework for an automated ingest of data on a daily basis to enable data fusion of new cases and death information.
-Hence, we can re-train the model with new incoming data every day.
-Through repeated training experiments, we found the LSTM hyperparameters such as epochs and dropout values that work well for our experiments.
-
-We conducted one additional analysis and plotted the best risk factor prediction model errors, the five best, and the ten best from our experiments to identify if we have better predictions than by including no risk factor.
-The best result found with no risk factor is depicted with the black horizontal line.
-Figures~\ref{fig:place-top1-cases} to  \ref{fig:place-top10-death} show the results sorted by the minimized error.
-Notably, we identified that a jump occurs when we sort the top 10 model predictions at around 330 models from a total of 3400 automatically generated models.
-This also underscores the need to run our deep learning model multiple times to obtain best parameter candidates.
+This is following our input data distribution in which we noted higher
+fluctuations relative to the overall value. A mitigation to this issue
+would be using a seven day average over the analyzed period. However,
+we have not done this on purpose for this analysis as we wanted to
+identify how a covariate enhanced LSTM behaves that includes risk
+factors based on the daily fluctuations. This was done to avoid and
+showcase any needed preproccessing on the data and to identify the
+capability of the deep learning framework while adapting to the
+fluctuations without any special activities. Previously, we have
+already shown in Section \ref{sec:emperical} that for smooth data
+inputs the prediction is very accurate.  Using fluctuation data allows
+our framework for an automated ingest of data on a daily basis to
+enable data fusion of new cases and death information. Hence, we can
+re-train the model with new incoming data every day. Through repeated
+training experiments, we found the LSTM hyperparameters such as epochs
+and dropout values that work well for our experiments.
+
+We conducted one additional analysis and plotted the best risk factor
+prediction model errors, the five best, and the ten best from our
+experiments to identify if we have better predictions than by
+including no risk factor. The best result found with no risk factor is
+depicted with the black horizontal
+line. Figures~\ref{fig:place-top1-cases} to
+\ref{fig:place-top10-death} show the results sorted by the minimized
+error. Notably, we identified that a jump occurs when we sort the top
+10 model predictions at around 330 models from a total of 3400
+automatically generated models. This also underscores the need to run
+our deep learning model multiple times to obtain best parameter
+candidates.
 
@@
 
-While this study focuses on if and which risk factors can improve the predictions, we do not focus on why.
-Such analysis could provide insights into causal relationships.
-For example,  information such as health insurance and the number of hospitals and beds are often related to the quality of services that affect early detection and treatment.
-Furthermore, population parameters such as density can determine the rate of spread and the number of cases.
-Similarly, distributions of comorbidities in a given population such as physical health, cholesterol screening, diabetes, etc., are found to be insightful in predicting outcomes \cite{Maleki}.
-Identifying such risk factors and correlating them to the prediction accuracy provides valuable candidates for future studies.
+While this study focuses on if and which risk factors can improve the
+predictions, we do not focus on why. Such analysis could provide
+insights into causal relationships. For example, information such as
+health insurance and the number of hospitals and beds are often
+related to the quality of services that affect early detection and
+treatment. Furthermore, population parameters such as density can
+determine the rate of spread and the number of cases. Similarly,
+distributions of comorbidities in a given population such as physical
+health, cholesterol screening, diabetes, etc., are found to be
+insightful in predicting outcomes \cite{Maleki}. Identifying such risk
+factors and correlating them to the prediction accuracy provides
+valuable candidates for future studies.
 
@@
 
-In our next comprehensive analysis, we try to answer Question \ref{q:4} if all risk factors impact the analysis.
-Thus we feed the covariates into the recurrent network as additional features.
-In addition to cases and deaths, this leads to 35 variables, including 33 unique covariates.
-We looked at two related ways of doing this:
+In our next comprehensive analysis, we try to answer Question
+\ref{q:4} if all risk factors impact the analysis. Thus we feed the
+covariates into the recurrent network as additional features. In
+addition to cases and deaths, this leads to 35 variables, including 33
+unique covariates. We looked at two related ways of doing this:
 
@@
 
-\item[(b)] Multiply the covariates by the times series (we used the average of cases and deaths)
+\item[(b)] Multiply the covariates by the times series (we used the
+  average of cases and deaths)
 
@@
 tests, we found these choices gave very similar results, and so we
-focused on the simpler choice (a) in this paper. In the experiments discussed
-here, we either used 0, 1, or 33 covariates. Choosing 33 gives the best
-description (reducing loss by about 1\%) while using each covariate as
-a solitary extra feature allows us a rather sensitive study of the
-importance of each individual covariate as we noted in Section \ref{sec:lstm-covariate}.
-
-The final choice was the quantities to predict and train
-on. We choose this to be both the cases and deaths values at the
-time value following the sequence combined with the same quantities
-for the following 14 days. This gives us 30 different output features
-for each sequence. We renormalized, so the two basic features
-(following day) contributed 50\% of the loss function. We supported
-this in TensorFlow with a custom mean square error ($MSE$) loss
-function that dropped training values that were not available for time
-sequences near the end of the examined period. While one can not
-train on future values, we can predict them, as seen in our sample
-outputs. We extensively studied both 2 and 30 output features, and they
-give similar descriptions of the observed data with the 30 feature
-choice enabling 14-day predictions.
-
-We performed an extensive hyper-parameter search and we show the values
-in Table \ref{tab:model} that give robust results without
-overfitting. Furthermore, we looked at four different activation functions for 
-this analysis in the
-LSTM and fully connected layers. $SELU$, $RELU$ and $TANH$ activations gave
-very similar results while $SIGMOID$ activation was much worse.
-However we still use a $SIGMOID$ function in the last stage just as we did in Section \ref{sec:lstm-covariate}.
-We also looked at the number of hidden units (separately for LSTM and fully
-connected layers), the number of LSTM layers and the dropout values.
-Only the latter produced a significant change and, for example, removing
-dropouts decreased the loss by about 16\% but we retained them to provide
-a robust environment. We chose randomly 20\% of the data for
-validation and testing and this had similar loss values to the
-training data (the other 80\%) with our chosen dropout.
+focused on the simpler choice (a) in this paper. In the experiments
+discussed here, we either used 0, 1, or 33 covariates. Choosing 33
+gives the best description (reducing loss by about 1\%) while using
+each covariate as a solitary extra feature allows us a rather
+sensitive study of the importance of each individual covariate as we
+noted in Section \ref{sec:lstm-covariate}.
+
+The final choice was the quantities to predict and train on. We choose
+this to be both the cases and deaths values at the time value
+following the sequence combined with the same quantities for the
+following 14 days. This gives us 30 different output features for each
+sequence. We renormalized, so the two basic features (following day)
+contributed 50\% of the loss function. We supported this in TensorFlow
+with a custom mean square error ($MSE$) loss function that dropped
+training values that were not available for time sequences near the
+end of the examined period. While one can not train on future values,
+we can predict them, as seen in our sample outputs. We extensively
+studied both 2 and 30 output features, and they give similar
+descriptions of the observed data with the 30 feature choice enabling
+14-day predictions.
+
+We performed an extensive hyper-parameter search and we show the
+values in Table \ref{tab:model} that give robust results without
+overfitting. Furthermore, we looked at four different activation
+functions for this analysis in the LSTM and fully connected
+layers. $SELU$, $RELU$ and $TANH$ activations gave very similar
+results while $SIGMOID$ activation was much worse. However we still
+use a $SIGMOID$ function in the last stage just as we did in Section
+\ref{sec:lstm-covariate}. We also looked at the number of hidden units
+(separately for LSTM and fully connected layers), the number of LSTM
+layers and the dropout values.  Only the latter produced a significant
+change and, for example, removing dropouts decreased the loss by about
+16\% but we retained them to provide a robust environment. We chose
+randomly 20\% of the data for validation and testing and this had
+similar loss values to the training data (the other 80\%) with our
+chosen dropout.
 
 The final technical issue concerns the data normalization where each
-feature is transformed by a feature dependent -- but city independent -- way
-to lie between 0 and 1. Further all extensive (proportional to size)
-covariates were first divided by the city population except for
-the population itself where we took the logarithm; intensive covariates
-were just rescaled to lie between 0 and 1. The basic COVID-19 daily
-data is extensive but compared two approaches with and without
-renormalization, but in this case, we took the square root of the data value.
-This is natural if there are counting (square root(N) for N
-observations) errors as the deep learning loss function is
+feature is transformed by a feature dependent -- but city independent
+-- way to lie between 0 and 1. Further all extensive (proportional to
+size) covariates were first divided by the city population except for
+the population itself where we took the logarithm; intensive
+covariates were just rescaled to lie between 0 and 1. The basic
+COVID-19 daily data is extensive but compared two approaches with and
+without renormalization, but in this case, we took the square root of
+the data value.  This is natural if there are counting (square root(N)
+for N observations) errors as the deep learning loss function is
 $(predicted-observed)^{2}$ without the error traditional in
@@
 with standard deep learning $MSE$ appears sound instead of dividing by
-the population. That will emphasize the error contribution of
-smaller cities but is still interesting to look at due to the presence of many systematic errors.
+the population. That will emphasize the error contribution of smaller
+cities but is still interesting to look at due to the presence of many
+systematic errors.
 
@@
 
-The deep learning model contains a lot of information much of which we have already analyzed and discussed in Section
+The deep learning model contains a lot of information much of which we
+have already analyzed and discussed in Section
 \ref{sec:lstm-covariate} and Figures \ref{fig:magic-1} and
@@
 \begin{itemize}
-\item \textbf{Past} or all days up to 14 days before the final observed date (May 25, 2020)
+\item \textbf{Past} or all days up to 14 days before the final
+  observed date (May 25, 2020)
 \item \textbf{Now} or the 14 days from May 12 to May 25, 2020
-\item \textbf{Future} or the 14 days after May 25, 2020 predicted by the model
+\item \textbf{Future} or the 14 days after May 25, 2020 predicted by
+  the model
 \end{itemize}
@@
 computing the correlation. Significant positive correlations are seen
-for CASTHMA, HIGHCHOL, DIABETES, OBESITY, CSMOKING, CHD, and CHECKUP
-, while negative correlations for PVI and NORM\_POP. We also looked at
+for CASTHMA, HIGHCHOL, DIABETES, OBESITY, CSMOKING, CHD, and CHECKUP ,
+while negative correlations for PVI and NORM\_POP. We also looked at
 the impact of these covariates on the evolution function. As this
@@
 rates are also significantly impacted by INSURANCE and NORM\_POP
-(reduction) and SENIOR and OBESITY (increase) from the
-evolution operator. NORM\_POP impacts infection values as an increase.
-We can also identify which covariates have the most significant impact
-on the loss function. Here NHOSP, CHD, KIDNEY, BINGE, and PHLTH give 5\%
-or more reduction in loss function while NORM\_POP, SVI\_OVERALL,
+(reduction) and SENIOR and OBESITY (increase) from the evolution
+operator. NORM\_POP impacts infection values as an increase.  We can
+also identify which covariates have the most significant impact on the
+loss function. Here NHOSP, CHD, KIDNEY, BINGE, and PHLTH give 5\% or
+more reduction in loss function while NORM\_POP, SVI\_OVERALL,
 CASTHMA, DIABETES, and BPHIGH have small (1\%) effects.
@@
 
-A variety of mathematical and computational models have been used for predicting COVID-19 outcomes \cite{Jewell}.
-It includes compartmental models such as the SIR (Susceptible-Infected-Recovered) and SEIR (Susceptible-Exposed-Infected-Recovered) which are relatively easy to compute, and thus, commonly used in epidemiology.
-However, their simplistic assumptions, such as homogeneous mixing within populations, might make them less realistic for pandemics such as COVID-19.
-For instance, such models often assume a closed population that does not change.
-However, COVID-19 dynamics have prominently included massive migration, deaths due to comorbidities temporarily conferred immunity, and notably, major disparities in terms of socio-economic determinants of health.
-
-In terms of implementation, SIR and SEIR models are fitted with point estimates, which may not adequately reflect the disease dynamics.
-For instance, it is difficult to account for behavioral changes (e.g., human mobility) or systemic inadequacies (e.g., availability of hospital beds) that may appear in specific subpopulations.
-Finally, the compartmental model estimates do not allow for quantification of uncertainty, which makes it harder to use them in policy-making.
-A comparison of modeling approaches concluded that while SEIR model performed better for a couple of states in the U.S., alternatives have performed better for the other states \cite{Bertozzi}.
-AICov has addressed most of the above modeling issues in its design, as per the requirements of its forecasting objectives.
-
-We present AICov as an integrative deep learning framework for COVID-19 forecasting with the help of population covariates.
-Thus, its objectives are different from those of traditional compartmental models.
-One of the important features of the architecture for AICov is that it is by design targeting Cloud and HPC resources to conduct parameter sweeps to leverage sophisticated deep learning toolkits.
-The architecture allows the integration of various data sources that can update the data from its sources on demand and update its model predictions based on newly introduced data.
-Parameters can easily be adjusted via Jupyter notebooks that, in turn, call the computational backends on HPC and cloud resources.
-
-In addition to this architecture, we used data collected from multiple public sources and agencies, and integrated the same across spatially contiguous units such as cities or metropolitan areas.
-In our analysis, we have focused on 110 selected cities of the U.S., but the framework is general and can be used to analyze similar data on pandemics from anywhere in the world.
-
-The comprehensive analysis showcases the feasibility of our approach.
-Based on the outcomes reported it can lead to an improved prediction once we integrate risk factors in addition to the time-dependent data such as cases and death resulting from COVID-19.
-While the improvements observed were modest, they do present a concrete way of improving the forecast models.
-Inclusion of further putative factors from the ongoing worldwide studies on Covid-19 will only strengthen the future applications of our integrative framework.
-
-We have shown that deep learning can return very good results while using smooth data, which we used in our empirical fits.
-For real data as presented to us for the daily changes, it still produced good results and even was able to predict changes based on weekly fluctuations.
-This is typically not achieved by other non-data based model approaches.
-Also, we have used in our data only cases and deaths, but we intend to expand this by using data about recovery and, in the future, immunization.
-
-We have experimented with different hyperparameters and included in this study a selection of hyperparameters that have worked well for this data set.
-
-As we have set up the first version of our AICov software, we have used cloud and high-performance computers.
-All of our sophisticated analyses were run in a day on at most 16 computers.
-All deep learning algorithms that were run on GPUs were run on NVIDIA Tesla K80s.
-However, our framework is on purpose generalized so that other compute resources can be integrated and leveraged.
-This includes local, HPC, and cloud computing resources, as well as different GPUs.
-Naturally, we also need to take into consideration not only the compute time but also the time it takes for the researcher to derive the models.
-As our approach is self-learning based on automated data updates, the time to prepare an updated model can be automatized and minimal input is needed.
-Thus the overall effort is very competitive.
+A variety of mathematical and computational models have been used for
+predicting COVID-19 outcomes \cite{Jewell}. It includes compartmental
+models such as the SIR (Susceptible-Infected-Recovered) and SEIR
+(Susceptible-Exposed-Infected-Recovered) which are relatively easy to
+compute, and thus, commonly used in epidemiology. However, their
+simplistic assumptions, such as homogeneous mixing within populations,
+might make them less realistic for pandemics such as COVID-19. For
+instance, such models often assume a closed population that does not
+change. However, COVID-19 dynamics have prominently included massive
+migration, deaths due to comorbidities temporarily conferred immunity,
+and notably, major disparities in terms of socio-economic determinants
+of health.
+
+In terms of implementation, SIR and SEIR models are fitted with point
+estimates, which may not adequately reflect the disease dynamics. For
+instance, it is difficult to account for behavioral changes (e.g.,
+human mobility) or systemic inadequacies (e.g., availability of
+hospital beds) that may appear in specific subpopulations. Finally,
+the compartmental model estimates do not allow for quantification of
+uncertainty, which makes it harder to use them in policy-making. A
+comparison of modeling approaches concluded that while SEIR model
+performed better for a couple of states in the U.S., alternatives have
+performed better for the other states \cite{Bertozzi}. AICov has
+addressed most of the above modeling issues in its design, as per the
+requirements of its forecasting objectives.
+
+We present AICov as an integrative deep learning framework for
+COVID-19 forecasting with the help of population covariates. Thus, its
+objectives are different from those of traditional compartmental
+models. One of the important features of the architecture for AICov is
+that it is by design targeting Cloud and HPC resources to conduct
+parameter sweeps to leverage sophisticated deep learning toolkits. The
+architecture allows the integration of various data sources that can
+update the data from its sources on demand and update its model
+predictions based on newly introduced data. Parameters can easily be
+adjusted via Jupyter notebooks that, in turn, call the computational
+backends on HPC and cloud resources.
+
+In addition to this architecture, we used data collected from multiple
+public sources and agencies, and integrated the same across spatially
+contiguous units such as cities or metropolitan areas. In our
+analysis, we have focused on 110 selected cities of the U.S., but the
+framework is general and can be used to analyze similar data on
+pandemics from anywhere in the world.
+
+The comprehensive analysis showcases the feasibility of our
+approach. Based on the outcomes reported it can lead to an improved
+prediction once we integrate risk factors in addition to the
+time-dependent data such as cases and death resulting from
+COVID-19. While the improvements observed were modest, they do present
+a concrete way of improving the forecast models. Inclusion of further
+putative factors from the ongoing worldwide studies on Covid-19 will
+only strengthen the future applications of our integrative framework.
+
+We have shown that deep learning can return very good results while
+using smooth data, which we used in our empirical fits. For real data
+as presented to us for the daily changes, it still produced good
+results and even was able to predict changes based on weekly
+fluctuations. This is typically not achieved by other non-data based
+model approaches. Also, we have used in our data only cases and
+deaths, but we intend to expand this by using data about recovery and,
+in the future, immunization.
+
+We have experimented with different hyperparameters and included in
+this study a selection of hyperparameters that have worked well for
+this data set.
+
+As we have set up the first version of our AICov software, we have
+used cloud and high-performance computers. All of our sophisticated
+analyses were run in a day on at most 16 computers. All deep learning
+algorithms that were run on GPUs were run on NVIDIA Tesla
+K80s. However, our framework is on purpose generalized so that other
+compute resources can be integrated and leveraged. This includes
+local, HPC, and cloud computing resources, as well as different
+GPUs. Naturally, we also need to take into consideration not only the
+compute time but also the time it takes for the researcher to derive
+the models. As our approach is self-learning based on automated data
+updates, the time to prepare an updated model can be automatized and
+minimal input is needed. Thus the overall effort is very competitive.
 
 \begin{comment}
-Furthermore, we are currently conducting experiments with multiple risk factors at the same time to identify the most significant combinations of them.
-As we have shown that the bivariate analysis has a locality dependency, we will explore categorizing locations with similar demographics and conduct further experiments on demographical similar regions.
+Furthermore, we are currently conducting experiments with multiple
+risk factors at the same time to identify the most significant
+combinations of them.  As we have shown that the bivariate analysis
+has a locality dependency, we will explore categorizing locations with
+similar demographics and conduct further experiments on demographical
+similar regions.
 \end{comment}
@@
 
-This section contains the list of acronyms used in the paper for easy reference.
-The definitions of the risk factors are however given within the paper in Table~\ref{tab:risk-factors} on page \pageref{tab:risk-factors} and not repeated here.
+This section contains the list of acronyms used in the paper for easy
+reference. The definitions of the risk factors are however given
+within the paper in Table~\ref{tab:risk-factors} on page
+\pageref{tab:risk-factors} and not repeated here.
 
@@
 
-This work is partially supported by the National Science Foundation (NSF) through awards CIF21 DIBBS 1443054, nanoBIO 1720625, Cybertraining 1829704, CINES 1835598 and Global Pervasive Computational Epidemiology 1918626.
-We thank J.
-Kadupitiya for several useful discussions on deep learning frameworks.
+This work is partially supported by the National Science Foundation
+(NSF) through awards CIF21 DIBBS 1443054, nanoBIO 1720625,
+Cybertraining 1829704, CINES 1835598 and Global Pervasive
+Computational Epidemiology 1918626.  We thank J. Kadupitiya for
+several useful discussions on deep learning frameworks.
+
 \par
@@
 
-\bibliographystyle{plainurl}
+%\bibliographystyle{plainurl}
 %\bibliographystyle{plain}
-
-\bibliography{bib/data-covid19,bib/lstm,bib/cloudmesh}
+\bibliographystyle{plainnat}
+
+\Bibliography{bib/data-covid19,bib/lstm,bib/cloudmesh}
 

### 2020-12-18 vonLaszewski-covid.tex
update with the new template

@@
-\documentclass[12pt]{article}
-
-\usepackage{amsfonts,amsmath,amssymb,amsthm}
-\usepackage{array}
-\usepackage{authblk}
+\documentclass[letterpaper, inpress]{jds} % use this for production
+
+%\documentclass[a4paper, review]{jds}      % use this for review
+\usepackage{lineno}
+\linenumbers
+
+\usepackage{calc}
+\usepackage{comment}
+
+% color management
+
 \usepackage{color}
+\usepackage{tcolorbox}
+
+% table managemet
+
 \usepackage[table]{xcolor}
-\usepackage{epsfig,fancyhdr}
-\usepackage{geometry,graphicx}
-\usepackage[]{hyperref} 
-\usepackage{indentfirst}
-\usepackage{mathtools}
-\usepackage{multirow}
-\usepackage{listings}
-\usepackage{natbib}
-\usepackage{rotating}
-\usepackage{comment}
+\usepackage{makecell}
 \usepackage{booktabs}
-\usepackage{todonotes}
-\setuptodonotes{inline}
-\usepackage{makecell}
-\usepackage{enumitem}
-\usepackage{tcolorbox}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\usepackage{sectsty,secdot}
-\sectionfont{\bf\fontsize{14}{2em}\selectfont}
-\subsectionfont{\bf\fontsize{12}{2em}\selectfont}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\geometry{left=2.5cm,right=2cm,top=2.5cm,bottom=1.5cm}
-\setlength{\parindent}{0.63cm}
-
-\setcounter{page}{1}
-\newtheorem{theorem}{Theorem}
-\newtheorem{lemma}{Lemma}
-\newtheorem{corollary}{Corollary}
-\newtheorem{proposition}{Proposition}
-\theoremstyle{definition}
-\newtheorem{definition}{Definition}
-\newtheorem{example}{Example}
-\newtheorem{remark}{Remark}
-\pagestyle{empty}
-\usepackage{lineno}
-\setcitestyle{square}
-
-\newcommand*\rot{\rotatebox{90}}
-
-%---------page header---------%
-
-%\fancyhead[RO]{odd number page}
-%\fancyhead[LE]{even number page}
-
-\pagestyle{plain} 
-\linenumbers
-
-%-----------------------------%
-\newtheorem{question}{Question}
-
-\newtheorem{requirement}{Requirement}
+\usepackage{graphicx}
+\usepackage{rotating, multirow}
+
+% Avoid non breaking words
 
@@
 
-\newtcolorbox{redbox}{colback=red!5!white,colframe=red!75!black,width=3cm}
-\newtcolorbox{yellowbox}{colback=yellow!5!white,colframe=yelloe!75!black,width=3cm}
-
-\newcommand{\DONE}{{\color{green!60!black}\makebox[0pt][l]{$\square$}\raisebox{.15ex}{\hspace{0.1em}$\checkmark$}}~}
-\newcommand{\DOIT}{{\color{red!60!black}\makebox[0pt][l]{$\square$}\raisebox{.15ex}{\hspace{0.1em}$\boxtimes$}}~}
-\newcommand{\PROGRESS}{(in progress)~}
-
-\begin{document}
-
-\input{response-v3-content}
-
-\title{{\bf\fontsize{14pt}{18pt}\selectfont AICov: An Integrative Deep Learning Framework for COVID-19 Forecasting with Population Covariates}}
-\author[1]{{\fontsize{12pt}{0.5em}\selectfont Geoffrey C. Fox}}
-\author[1]{{\fontsize{12pt}{18pt}\selectfont Gregor von Laszewski}}
-\author[1]{{\fontsize{12pt}{0.5em}\selectfont Fugang Wang}}
-\author[2,3]{{\fontsize{12pt}{0.5em}\selectfont Saumyadipta Pyne}}
-
-\affil[1]{{\fontsize{12pt}{0.5em}\selectfont Digital Science Center, Indiana University, Bloomington Indiana, USA}}
-\affil[2]{{\fontsize{12pt}{0.5em}\selectfont Public Health Dynamics Lab, and Department of Biostatistics, University of Pittsburgh, Pittsburgh, Pennsylvania, USA}}
-\affil[3]{{\fontsize{12pt}{0.5em}\selectfont Health Analytics Network, Pennsylvania, USA}}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% the following edits should be done by Journal typesetters
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\setcounter{page}{1}            % set the first page number
+\jdsmonth{January}              % month
+\jdsyear{2021}                  % year
+\jdsvolume{xx}                  % volume number
+\jdsissue{xx}                   % issue number
+\jdsdoi{xx.xxxx/xxxxxxxxx}      % doi
+\jdsreceived{???, 2020}       % optional: comment it out if no received date
+\jdsaccepted{???, 2021}         % optional: comment it out if no accepted date
+%% manually set running header for a shorter list of authors if needed
+% \shortauthors{A Author A, et al.}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% edits by authors are given below
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\usepackage{amsfonts,amsmath,amssymb,amsthm}
+\usepackage{booktabs}
+
+\newcommand{\TITLE}{AICov: An Integrative Deep
+    Learning Framework for COVID-19 Forecasting with Population
+    Covariates}
 
@@
 }
-\date{}
-
-\newcommand{\TODO}[1]{
-\todo[color=yellow!20]{#1}
-}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\renewcommand{\baselinestretch}{1.07}
-\renewcommand{\thefootnote}{}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\title[AICov: Deep
+    Learning Framework for COVID-19 Forecasting with Population
+    Covariates]{\TITLE}
+
+\author[1]{Geoffrey C. Fox}
+\author[1]{Gregor von Laszewski\thanks{Corresponding author. Email: laszewski@gmail.com}}
+\author[1]{Fugang Wang}
+\author[2,3]{Saumyadipta Pyne}
+
+\affil[1]{Digital Science Center, Indiana University, Bloomington Indiana, USA}
+\affil[2]{Public Health Dynamics Lab, and Department of Biostatistics, University of Pittsburgh, Pittsburgh, Pennsylvania, USA}
+\affil[3]{Health Analytics Network, Pennsylvania, USA}
+
+\begin{document}
 
 \maketitle
-\thispagestyle{empty}
-
-\begin{quotation}
-\centerline{\bfseries{\fontsize{14pt}{1em}\selectfont Abstract}}\par
-\setlength{\parindent}{3.22cm}
-{\fontsize{12pt}{1.07em}\selectfont~~\ABSTRACT}
-
-\vspace{12pt}
-\noindent {\bfseries{Key words and phrases:}}
-COVID-19, deep learning, prediction, risk factors, comorbidities.
-\par
-\end{quotation}\par
-
-\def\thefigure{\arabic{figure}}
-\def\thetable{\arabic{table}}
-
-\renewcommand{\theequation}{\thesection.\arabic{equation}}
-
-\fontsize{12}{1.07em}\selectfont
-
-\setcounter{section}{0} %***
-\setcounter{equation}{0} %-1
+
+\begin{abstract}
+  \ABSTRACT
+\end{abstract}
+
+\begin{keywords} % alphabetical; excluding anything in the title already
+  Cloudmesh;
+  Comorbidities;
+  Prediction;
+  Risk factors.
+\end{keywords}
 
@@
 globally. On 30 January 2020, the World Health Organization (WHO)
-\cite{www-who} declared COVID-19 to be a Public Health Emergency of
+\citep{www-who} declared COVID-19 to be a Public Health Emergency of
 International Concern and subsequently, on 11 March 2020, a pandemic.
@@
 over 2.3 million confirmed cases and 121,167 deaths
-\cite{www-nyt-map}.  Although initial efforts to reduce the spread
+\citep{www-nyt-map}.  Although initial efforts to reduce the spread
 took place, recent data shows a resurgence at an increased scale. In
@@
 the existing lab-confirmed category, may likely have been reflected in
-the reporting \cite{Petropoulos2020-hh}. Initial fears about COVID-19
+the reporting \citep{Petropoulos2020-hh}. Initial fears about COVID-19
 among the general population with regards to the trajectory of the
@@
 learning systems to understand healthcare trends, model risk
-associations, and predict outcomes \cite{Ting2020-lw}.  In addition to
+associations, and predict outcomes \citep{Ting2020-lw}.  In addition to
 traditional public health surveillance strategies available to most
@@
 integrated to model the scale and dynamics of this pandemic
-\cite{Pyne2015-ao}. Several organizations, including the Johns Hopkins
+\citep{Pyne2015-ao}. Several organizations, including the Johns Hopkins
 University’s Center for Systems Science and Engineering, the New York
@@
 the lack thereof) to predict disease dynamics and its impact on human
-lives \cite{www-cdc-modeling-forecast}.
+lives \citep{www-cdc-modeling-forecast}.
 
@@
 
-\begin{comment}
-\begin{verbatim}
-    model.add(Dense(InitialMLP, activation=activationvalue, input_shape=(n_timesteps, n_features)))
-
-    model.add(LSTM(n_nodes, unroll=True, recurrent_dropout= recurrent_dropout1, dropout = dropout1,
-                   activation= activationvalue , return_sequences=True, recurrent_activation="sigmoid",
-                   input_shape=(n_timesteps, nextround)))
-
-    model.add(LSTM(n_nodes, unroll=True, recurrent_dropout= recurrent_dropout2, dropout = dropout2,
-               activation= activationvalue, recurrent_activation="sigmoid",
-               input_shape=(n_timesteps, n_nodes)))
-\end{verbatim}
-
-\begin{verbatim}
-Window Size  1  
-n_samples: 12540 
-n_timesteps: 1 
-n_features: 3 
-n_outputs: 30 
-batch_size: 110 
-n_nodes: 16 
-epochs: 200
-is NaN  77000  
-percent  20.47  
-not NaN  299200
-\end{verbatim}
-\end{comment}
-
 We have run the training on all 110 cities from our data that
@@
 \end{figure}
-
-%%%%%%
 
@@
 
-\begin{comment}
-\begin{figure}[!h]
-    \centering
-    \includegraphics[width=1.0\textwidth]{images/boxwhisker/swarmplot_cases.pdf}
-    \caption{Caption}
-    \label{fig:my_label}
-
-    \centering
-    \includegraphics[width=1.0\textwidth]{images/boxwhisker/swarmplot_death.pdf}
-    \caption{Caption}
-    \label{fig:my_label}
-\end{figure}
-\end{comment}
-
 While this study focuses on if and which risk factors can improve the
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{comment}
 
@@
 
-\vskip 14pt
-\noindent {\bfseries{\fontsize{14pt}{1em}\selectfont Supplementary Materials}}
+\section*{Supplementary Materials}
 
@@
 the brief description of the online supplementary materials.
-\par
-
-\end{comment}
+
 %
@@
 
-\section{Acronyms}
-
-This section contains the list of acronyms used in the paper for easy
-reference. The definitions of the risk factors are however given
-within the paper in Table~\ref{tab:risk-factors} on page
-\pageref{tab:risk-factors} and not repeated here.
-
-\begin{itemize}[leftmargin=5cm]
-\item[{\bf API:}] Application Programming Interfaces
-\item[{\bf BRFSS:}] Behavioral Risk Factor Surveillance System
-\item[{\bf CDC:}] Center for Desease Control
-\item[{\bf CIF21:}] Cyberinfrastructure Framework for 21st Century 
-\item[{\bf COVID-19:}] COrona VIrus Disease 2019 
-\item[{\bf CSSE:}] Center for Systems Science and Engineering
-\item[{\bf DIBBS:}] Data Infrastructure Building Blocks
-\item[{\bf FIPS:}] Federal Information Processing Standards
-\item[{\bf GPU:}] Graphics Processing Unit 
-\item[{\bf HPC:}] High Performance Computing
-\item[{\bf LSTM:}] Long Short-Term Memory
-\item[{\bf NBDIF:}] NIST Big Data Working Group
-\item[{\bf NCHS:}] CDC National Center for Health Statistics
-\item[{\bf NIST:}] National Institute of Standards and Technology
-\item[{\bf RELU:}] REctified Linear Unit
-\item[{\bf REST:}] REpresentational State Transfer 
-\item[{\bf RNN:}] Recurrent Neural Network
-\item[{\bf SEIR:}] Susceptible, Expose, Infectious, Recovered)
-\item[{\bf SIR}] Susceptible, Infectious, Recovered
-\item[{\bf U.S.:}] United States of America
-\item[{\bf NSF:}] National Science Foundation
-\item[{\bf MSE:}] Mean Squared Error
-\item[{\bf SELU:}] Scaled Exponential Linear Unit
-\item[{\bf CI:}] Cyber Infrastructure
-\item[{\bf CINES:}] CyberInfrastructure for Network Engineering and Science
-\item[{\bf WHO:}] World Health Organization
-\item[{\bf NONE:}] \todo{NO FACTORS}
-\item[{\bf NORM\_POP:}] Normalized Population
-\item[{\bf cum:}] cumulative
-\end{itemize}
-
-\vskip 14pt
-\noindent {\bfseries{\fontsize{14pt}{1em}\selectfont Acknowledgements}}
+\section*{Acknowledgements}
 
@@
 
-\par
-
-%
-% BIBLIOGRAPHY
-%
-
-%\iffalse
-\fontsize{11}{1em}\selectfont
-\renewcommand\bibname{\bfseries{\fontsize{14pt}{1em}\selectfont References}}
-\expandafter\ifx\csname
-natexlab\endcsname\relax\def\natexlab#1{#1}\fi
-\expandafter\ifx\csname url\endcsname\relax
-  \def\url#1{\texttt{#1}}\fi
-\expandafter\ifx\csname urlprefix\endcsname\relax\def\urlprefix{URL}\fi
-%\fi
-
-%\clearpage
-
-%\bibliographystyle{plainurl}
-%\bibliographystyle{plain}
-\bibliographystyle{plainnat}
-
-\Bibliography{bib/data-covid19,bib/lstm,bib/cloudmesh}
+\clearpage
+
+\section*{Acronyms}
+
+This section contains the list of acronyms in Table{tab:acronymes} as used in the paper for easy
+reference. The definitions of the risk factors are however given
+within the paper in Table~\ref{tab:risk-factors} on page
+\pageref{tab:risk-factors} and not repeated here.
+
+\begin{table}[!h]
+  \begin{tabular}{ll}
+    \toprule
+    Abbreviation & Description \\
+    \midrule
+    API & Application Programming Interfaces \\
+    BRFSS & Behavioral Risk Factor Surveillance System \\
+    CDC & Center for Desease Control \\
+    CI & Cyber Infrastructure \\
+    CIF21 & Cyberinfrastructure Framework for 21st Century  \\
+    CINES & CyberInfrastructure for Network Engineering and Science \\
+    COVID-19 & COrona VIrus Disease 2019  \\
+    CSSE & Center for Systems Science and Engineering \\
+    cum & cumulative \\
+    DIBBS & Data Infrastructure Building Blocks \\
+    FIPS & Federal Information Processing Standards \\
+    GPU & Graphics Processing Unit  \\
+    HPC & High Performance Computing \\
+    LSTM & Long Short-Term Memory \\
+    MSE & Mean Squared Error \\
+    NBDIF & NIST Big Data Working Group \\
+    NCHS & CDC National Center for Health Statistics \\
+    NIST & National Institute of Standards and Technology \\
+    NONE & \todo{NO FACTORS} \\
+    NORM\_POP & Normalized Population \\
+    NSF & National Science Foundation \\
+    RELU & REctified Linear Unit \\
+    REST & REpresentational State Transfer  \\
+    RNN & Recurrent Neural Network \\
+    SEIR & Susceptible, Expose, Infectious, Recovered) \\
+    SELU & Scaled Exponential Linear Unit \\
+    SIR &  Susceptible, Infectious, Recovered \\
+    U.S. & United States of America \\
+    WHO & World Health Organization \\
+    \midrule
+\end{tabular}
+\end{table}
+
+\clearpage
+
+\bibliographystyle{jds}
+\bibliography{bib/data-covid19,bib/lstm,bib/cloudmesh}
 

### 2020-12-18 vonLaszewski-covid.tex
add response
update table

@@
 \usepackage{rotating, multirow}
+\usepackage{colortbl}
 
@@
 \begin{document}
+
+\input{response-v3-content}
 
@@
 time-dependent data but by integrating the population model for the
-underlying prediction, as shown in Fig. \ref{fig:magic-1}. Such model
+underlying prediction, as shown in Figure \ref{fig:magic-1}. Such model
 predictive integration capability is important in any application with
@@
 \centering
-\resizebox{1.0\textwidth}{!}{ 
+%\resizebox{1.0\textwidth}{!}{ 
 \begin{tabular}{rrrrrlrl}
 \toprule
+ \makecell{Place/\\Rank} & 
+\makecell{Risk Factor} &  
 \makecell{RMSE\\Cases} &  
  \makecell{RMSE\\Death} &  
- \makecell[r]{Cummulative \\Error \\Cases} &  
- \makecell[r]{Cummulative \\Error \\Death} &  
- \makecell[r]{Number of \\Days as \\Input} & 
- \makecell{Risk Factor} &  
- \makecell{Place/\\Rank} & \\
+ \makecell[r]{Cumu-\\lative \\Error \\Cases} &  
+ \makecell[r]{Cumu-\\lative \\Error \\Death} &  
+ \makecell[r]{Number of \\Days as \\Input} \\
 \midrule
- 0.055443 & 0.067617 & 7.94 & 22.74 & 5 & POP\_DENSITY\_2010 & 0 \\
- 0.054795 & 0.068529 & 8.01 & 20.55 & 3 & PHLTH & 1 \\
- 0.055199 & 0.067641 & 8.07 & 20.43 & 4 & INSURANCE & 2 \\
- 0.055028 & 0.067207 & 8.12 & 20.17 & 4 & PERCENTBLACKS & 3 \\
- 0.056061 & 0.067489 & 8.22 & 19.04 & 5 & PERCENTHISPANICS & 4 \\
- 0.054621 & 0.068255 & 8.26 & 20.10 & 3 & NHOSP & 5 \\
- 0.054557 & 0.068572 & 8.34 & 24.61 & 3 & INSURANCE & 6 \\
- 0.054712 & 0.069121 & 8.37 & 21.70 & 3 & CHOLSCREEN & 7 \\
- 0.054586 & 0.067636 & 8.40 & 20.50 & 5 & DIABETES & 8 \\
- 0.054160 & 0.066906 & 8.45 & 20.16 & 4 & STROKE & 9 \\
- 0.054027 & 0.067642 & 8.46 & 21.93 & 5 & CHD & 10 \\
- 0.054927 & 0.067652 & 8.51 & 20.42 & 5 & CHECKUP & 11 \\
- 0.054146 & 0.067536 & 8.54 & 21.15 & 5 & NBEDS & 12 \\
- 0.054795 & 0.067676 & 8.58 & 19.94 & 5 & SVI\_OVERALL & 13 \\
- 0.054600 & 0.067565 & 8.61 & 21.86 & 5 & KIDNEY & 14 \\
- 0.055089 & 0.068248 & 8.63 & 22.58 & 5 & CASTHMA & 15 \\
- 0.054547 & 0.068266 & 8.68 & 20.49 & 3 & BLACK & 16 \\
- 0.054685 & 0.068920 & 8.74 & 21.69 & 3 & BPMED & 17 \\
- 0.054323 & 0.067149 & 8.75 & 19.54 & 5 & LPA & 18 \\
- 0.053716 & 0.066898 & 8.75 & 19.01 & 5 & CSMOKING & 19 \\
- 0.054053 & 0.068716 & 8.83 & 20.94 & 3 & ARTHRITIS & 20 \\
- 0.055029 & 0.068672 & 8.84 & 22.02 & 3 & POVERTY & 21 \\
- 0.054545 & 0.067267 & 8.86 & 21.22 & 4 & ESTBEDS & 22 \\
- 0.054864 & 0.068742 & 8.90 & 19.53 & 3 & MHLTH & 23 \\
- 0.055079 & 0.069024 & 8.93 & 21.23 & 3 & SENIOR & 24 \\
- 0.055589 & 0.068250 & 8.93 & 23.26 & 4 & COPD & 25 \\
- 0.054500 & 0.068363 & 8.99 & 19.70 & 3 & CANCER & 26 \\
- 0.054397 & 0.068491 & 9.05 & 21.89 & 3 & NBEDS/1000 & 27 \\
- 0.054609 & 0.067346 & 9.13 & 21.29 & 4 & BPHIGH & 28 \\
-\rowcolor{black!5}  0.054868 & 0.069432 & 9.26 & 25.88 & 3 & NONE & 29 \\
- 0.056429 & 0.069345 & 9.28 & 20.98 & 3 & HIGHCHOL & 30 \\
- 0.054317 & 0.067638 & 9.32 & 21.00 & 5 & BINGE & 31 \\
- 0.054469 & 0.068540 & 9.50 & 20.65 & 3 & OBESITY & 32 \\
- 0.055094 & 0.067443 & 9.54 & 21.53 & 4 & SVI\_MINORITY & 33 \\
+1 & POP\_DENSITY\_2010 & 0.055443 & 0.067617 & 7.94 & 22.74 & 5 \\
+2 & PHLTH & 0.054795 & 0.068529 & 8.01 & 20.55 & 3 \\
+3 & INSURANCE & 0.055199 & 0.067641 & 8.07 & 20.43 & 4 \\
+4 & PERCENTBLACKS & 0.055028 & 0.067207 & 8.12 & 20.17 & 4 \\
+5 & PERCENTHISPANICS & 0.056061 & 0.067489 & 8.22 & 19.04 & 5 \\
+6 & NHOSP & 0.054621 & 0.068255 & 8.26 & 20.10 & 3 \\
+7 & INSURANCE & 0.054557 & 0.068572 & 8.34 & 24.61 & 3 \\
+8 & CHOLSCREEN & 0.054712 & 0.069121 & 8.37 & 21.70 & 3 \\
+9 & DIABETES & 0.054586 & 0.067636 & 8.40 & 20.50 & 5 \\
+10 & STROKE & 0.054160 & 0.066906 & 8.45 & 20.16 & 4 \\
+11 & CHD & 0.054027 & 0.067642 & 8.46 & 21.93 & 5 \\
+12 & CHECKUP & 0.054927 & 0.067652 & 8.51 & 20.42 & 5 \\
+13 & NBEDS & 0.054146 & 0.067536 & 8.54 & 21.15 & 5 \\
+14 & SVI\_OVERALL & 0.054795 & 0.067676 & 8.58 & 19.94 & 5 \\
+15 & KIDNEY & 0.054600 & 0.067565 & 8.61 & 21.86 & 5 \\
+16 & CASTHMA & 0.055089 & 0.068248 & 8.63 & 22.58 & 5 \\
+17 & BLACK & 0.054547 & 0.068266 & 8.68 & 20.49 & 3 \\
+18 & BPMED & 0.054685 & 0.068920 & 8.74 & 21.69 & 3 \\
+19 & LPA & 0.054323 & 0.067149 & 8.75 & 19.54 & 5 \\
+20 & CSMOKING & 0.053716 & 0.066898 & 8.75 & 19.01 & 5 \\
+21 & ARTHRITIS & 0.054053 & 0.068716 & 8.83 & 20.94 & 3 \\
+22 & POVERTY & 0.055029 & 0.068672 & 8.84 & 22.02 & 3 \\
+23 & ESTBEDS & 0.054545 & 0.067267 & 8.86 & 21.22 & 4 \\
+24 & MHLTH & 0.054864 & 0.068742 & 8.90 & 19.53 & 3 \\
+25 & SENIOR & 0.055079 & 0.069024 & 8.93 & 21.23 & 3 \\
+26 & COPD & 0.055589 & 0.068250 & 8.93 & 23.26 & 4 \\
+27 & CANCER & 0.054500 & 0.068363 & 8.99 & 19.70 & 3 \\
+28 & NBEDS/1000 & 0.054397 & 0.068491 & 9.05 & 21.89 & 3 \\
+29 & BPHIGH & 0.054609 & 0.067346 & 9.13 & 21.29 & 4 \\
+30 & NONE &\rowcolor{black!5} 0.054868 & 0.069432 & 9.26 & 25.88 & 3 \\
+31 & HIGHCHOL & 0.056429 & 0.069345 & 9.28 & 20.98 & 3 \\
+32 & BINGE & 0.054317 & 0.067638 & 9.32 & 21.00 & 5 \\
+33 & OBESITY & 0.054469 & 0.068540 & 9.50 & 20.65 & 3 \\
+34 & SVI\_MINORITY & 0.055094 & 0.067443 & 9.54 & 21.53 & 4 \\
 \bottomrule
 \end{tabular}
-}
+%}
 \end{table}
@@
 \begin{figure}[!h]
-    \begin{minipage}{.45\textwidth}
-    \centering
-    \includegraphics[width=1.0\textwidth]{images/predict/Length_5-fields_pop_density_2010-Cumulative-Cases.png}
-    \vspace{-1cm}
+    \centering
+    \includegraphics[width=0.5\textwidth]{images/predict/Length_5-fields_pop_density_2010-Cumulative-Cases.png}
+    %\vspace{-1cm}
     \caption{Model prediction error for cumulative cases when
@@
     \label{fig:error-case-popdensity}
-    \end{minipage}
-    \begin{minipage}{.45\textwidth}
-    \ \
-    \centering
-    \includegraphics[width=1.0\textwidth]{images/predict/Length_5-fields_pop_density_2010-Cumulative-Death.png}
-    \vspace{-1cm}
+\end{figure}
+
+\begin{figure}[!h]    
+    \centering
+    \includegraphics[width=0.5\textwidth]{images/predict/Length_5-fields_pop_density_2010-Cumulative-Death.png}
+    %\vspace{-1cm}
     \caption{Model prediction error for cumulative deaths when
@@
     \label{fig:error-death-popdensity}
-    \end{minipage}
 \end{figure}
@@
 addition to cases and deaths, this leads to 35 variables, including 33
-unique covariates. We looked at two related ways of doing this:
-
-\begin{description}
-
-\item [(a)] Have covariates fixed in time (static)
-
-\item[(b)] Multiply the covariates by the times series (we used the
-  average of cases and deaths)
+unique covariates. We looked at two related ways of doing this: (a)
+have covariates fixed in time (static), and (b) multiply the
+covariates by the times series (we used the average of cases and
+deaths)
 
@@
 \label{tab:corr-matrix}
-\resizebox{1.0\textwidth}{!}{
+\begin{center}
 \begin{tabular}{lrrrrrr}
 \toprule
-Risk Factor		& Past Case & Now Case & Future Case & Past Death & Now Death & Future Death \\
+Risk Factor		&
+\makecell{Past\\Case} &
+\makecell{Now\\Case} &
+\makecell{Future\\Case} &
+\makecell{Past\\Death} &
+\makecell{Now\\Death} &
+\makecell{Future\\Death} \\
 \midrule
@@
 \end{tabular}
-}
+\end{center}
 \end{table}

### 2020-12-18 vonLaszewski-covid.tex
add row color

@@
 29 & BPHIGH & 0.054609 & 0.067346 & 9.13 & 21.29 & 4 \\
-30 & NONE &\rowcolor{black!5} 0.054868 & 0.069432 & 9.26 & 25.88 & 3 \\
+\rowcolor{black!5} 30 & NONE & 0.054868 & 0.069432 & 9.26 & 25.88 & 3 \\
 31 & HIGHCHOL & 0.056429 & 0.069345 & 9.28 & 20.98 & 3 \\

### 2020-12-19 vonLaszewski-covid.tex
make pdflatex compile

@@
 \usepackage{comment}
+\usepackage{todonotes}
 
@@
 
-\usepackage[table]{xcolor}
+%\usepackage[table]{xcolor}
 \usepackage{makecell}
@@
 \usepackage{booktabs}
+
+\newtheorem{requirement}{Requirement}
+\newtheorem{question}{Question}
 
@@
 deaths)
-
-\end{description}
 
@@
     NIST & National Institute of Standards and Technology \\
-    NONE & \todo{NO FACTORS} \\
+    NONE & No risk factors used \\
     NORM\_POP & Normalized Population \\

### 2020-12-19 response-v3-content.tex
add checkboxes
Updates from Overleaf
inline todonotes

@@
-% \input{response/response}
+\newcommand{\DONE}{{\color{green!60!black}\makebox[0pt][l]{$\square$}\raisebox{.15ex}{\hspace{0.1em}$\checkmark$}}~}
+\newcommand{\DOIT}{{\color{red!60!black}\makebox[0pt][l]{$\square$}\raisebox{.15ex}{\hspace{0.1em}$\boxtimes$}}~}
+\newcommand{\PROGRESS}{(in progress)~}
 
 REVISION DUE: 01.12.2021
+
+\section{Additional actions taken}
+
+\begin{enumerate}
+
+    \item \DONE We used the new JDS \LaTeX template available from the Journal page instead of the older one that we used.
+
+    \item \DONE Due to that all references are now in the style of the journal.
+
+    \item \DONE URLs have been added with the standard urlbst tool so they can be used in the refernces. The jdsurl.bst is included in the source.
+
+\end{enumerate}
 

### 2020-12-19 vonLaszewski-covid.tex
add urls to bst
inline todonotes

@@
 \usepackage{todonotes}
+\setuptodonotes{inline}
 
@@
 In this study, we used a Long Short-Term Memory (LSTM) algorithm for
-our predictions \cite{www-keras-lstm,Hochreiter1997-dk}. An LSTM is a
-recurrent neural network (RNN) \cite{Rumelhart1986-li} with feedback
+our predictions \citep{www-keras-lstm,Hochreiter1997-dk}. An LSTM is a
+recurrent neural network (RNN) \citep{Rumelhart1986-li} with feedback
 connections allowing the use of data input sequences to predict data
 output sequences. LSTMs have been applied to many application areas
-from handwriting \cite{Graves2009-qb}, image, feature detection, and
-time series prediction \cite{Schmidhuber2005-oy}. LSTM's have an
+from handwriting \citep{Graves2009-qb}, image, feature detection, and
+time series prediction \citep{Schmidhuber2005-oy}. LSTM's have an
 internal state that is used to prevent the vanishing gradient problem
-\cite{Hochreiter1991-mp} during the training of RNNs.
+\citep{Hochreiter1991-mp} during the training of RNNs.
 
@@
 language, we have shown how to represent Newton’s law operator by a
-recurrent network \cite{Kadupitiya2020-zq}. We expect that the time
+recurrent network \citep{Kadupitiya2020-zq}. We expect that the time
 dependence of many complex systems: Covid pandemics, Southern
@@
   NIST NBDIF definitions cloud services into the architecture
-  \cite{las-19-nist}. This includes containers but also infrastructure
+  \citep{las-19-nist}. This includes containers but also infrastructure
   services in the cloud. Through these service integrations, we also
@@
 integrator
-\cite{cloudmesh-openapi-install,cloudmesh-openapi-benchmark,cloudmesh-manual}. Additionally,
+\citep{cloudmesh-openapi-install,cloudmesh-openapi-benchmark,cloudmesh-manual}. Additionally,
 we can also access through abstractions cloud services that are
@@
 health and behavioral risk factors aggregated at the county level from
-the large-scale longitudinal CDC BRFSS surveys \cite{www-cdc-brfss},
+the large-scale longitudinal CDC BRFSS surveys \citep{www-cdc-brfss},
 and demographic and socioeconomic data from the U.S. Census Bureau
 that are known at the local levels along with the cases and deaths
-data compiled by NCHS \cite{www-cdc-nchs} and integrate them with
+data compiled by NCHS \citep{www-cdc-nchs} and integrate them with
 other data sources.
@@
 or accessible via specifically developed online resources such as the
-New York Times~\cite{github-nytimes} and the Johns Hopkins Coronavirus
-Resource Center~\cite{www-jh-covid19}.  For this study, we generated
+New York Times~\citep{github-nytimes} and the Johns Hopkins Coronavirus
+Resource Center~\citep{www-jh-covid19}.  For this study, we generated
 two data sets based on the latter resource upon our checking for the
@@
 unhealthy behaviors were extracted from the CDC's 500 Cities Local
-Data for Better Health program \cite{www-cdc-chronic-data} in the form
+Data for Better Health program \citep{www-cdc-chronic-data} in the form
 of crude or age-adjusted prevalence of conditions such as respiratory
@@
 Demographic variables were gathered from the Census
-QuickFacts~\cite{www-census} online resource using an automated web
+QuickFacts~\citep{www-census} online resource using an automated web
 scraping algorithm and cover relevant areas such as age, race, income,
@@
 Gini Index, which measures economic inequality and CDC Social
-Vulnerability Index (SVI)~\cite{www-cdc-svi}. The SVI was created to
+Vulnerability Index (SVI)~\citep{www-cdc-svi}. The SVI was created to
 guide public health officials and disaster response efforts by
@@
 number of beds per hospital in that state) per 1,000 people using the
-American Hospital Directory~\cite{www-ahd}.
+American Hospital Directory~\citep{www-ahd}.
 
@@
 deep learning while using data generated from epidemiological models
-as discussed in \cite{marsland20-covid-paper}. The time series we used
+as discussed in \citep{marsland20-covid-paper}. The time series we used
 was 100 days long, and a multi-layered Long Short-Term Memory (LSTM)
@@
 an epidemiological model and not deep learning
-\cite{marsland20-covid-paper}) for the cases and deaths data up to
+\citep{marsland20-covid-paper}) for the cases and deaths data up to
 April 15, 2020. Our analysis identified a sophisticated single deep
@@
 with two layers of LSTM, and initial and final fully connected layers
-\cite{Kadupitiya2020-zq}.  For LSTM, we have selected the activation
+\citep{Kadupitiya2020-zq}.  For LSTM, we have selected the activation
 function based on Rectified Linear Units (RELU). For the last layer
@@
 health, cholesterol screening, diabetes, etc., are found to be
-insightful in predicting outcomes \cite{Maleki}. Identifying such risk
+insightful in predicting outcomes \citep{Maleki}. Identifying such risk
 factors and correlating them to the prediction accuracy provides
@@
 A variety of mathematical and computational models have been used for
-predicting COVID-19 outcomes \cite{Jewell}. It includes compartmental
+predicting COVID-19 outcomes \citep{Jewell}. It includes compartmental
 models such as the SIR (Susceptible-Infected-Recovered) and SEIR
@@
 performed better for a couple of states in the U.S., alternatives have
-performed better for the other states \cite{Bertozzi}. AICov has
+performed better for the other states \citep{Bertozzi}. AICov has
 addressed most of the above modeling issues in its design, as per the
@@
 
-\bibliographystyle{jds}
+\bibliographystyle{jdsurl}
 \bibliography{bib/data-covid19,bib/lstm,bib/cloudmesh}

### 2020-12-19 response-v3-content.tex
update todo list and abbreviations

--- 2020-12-18 response-v3-content.tex
@@
-% \input{response/response}
+\definecolor{PROGRESS}{cmyk}{0.10, 0.10, 0.0, 0.0, 1.00}
+\definecolor{TODO}{cmyk}{0.0, 0.10, 0.10, 0.00, 1.00}
+\definecolor{DONE}{cmyk}{0.10, 0.0, 0.10, 0.0, 1.0}
+
+\newcommand{\REVIEW}[3]{
+  \begin{tcolorbox}[colback={#1}]
+    {\bf {#1}. {#2}}
+ \end{tcolorbox}
+  \begin{quote}
+    {#3}
+  \end{quote}
+}
 
 REVISION DUE: 01.12.2021
+
+\section{Additional actions taken}
+
+\REVIEW{DONE}{Using the JDS template} {We used the new JDS \LaTeX
+  template available from the Journal page instead of the older one
+  that we used. This also includes The change of citations in brackets
+  when cited. URLs have been added with the standard urlbst tool so
+  they can be used in the refernces. The jdsurl.bst is included in the
+  source.}
+
+\end{enumerate}
 
@@
 
-The revised manuscript is much smoother than the first version, but I still feel that less than sufficient care was put into the writing.
-The reviewer pointed some in the pdf report.
-I have some additional editorial suggestions:
+The revised manuscript is much smoother than the first version, but I
+still feel that less than sufficient care was put into the
+writing. The reviewer pointed some in the pdf report. I have some
+additional editorial suggestions:
 
-\begin{enumerate}
+\bigskip
 
-\item \PROGRESS Submit the point-by-point response to the reviewer's comments and my comments in a pdf file.
+  \REVIEW{PROGRESS} {Submit the point-by-point response to the
+    reviewer's comments and my comments in a pdf file.}  {The original
+    submisison web page had a field to enter the comments as text
+    which we used. We will change this and introduce a PDF file
+    instead, which is easier for us also.}
 
-\begin{quote}
-The original submisison web page had a field to enter the comments as text which we used. We will change this and introduce a PDF file instead, which is easier for us also.
-\end{quote}
+  \REVIEW{PROGRESS} {Submit the point-by-point response to the
+    reviewer's comments and my comments in a pdf file.}
+  {The original submisison web page had a field to enter the comments
+    as text which we used. We will change this and introduce a PDF
+    file instead, which is easier for us also.}
 
-\item \DONE Needs page number (line number) for ease of referencing in the review.
+  \REVIEW{DONE} {Needs page number (line number) for ease of
+    referencing in the review.}
+  {Page numbers and line numbers were
+    activated with the review style from the new JDS \LaTeX{}
+    template}
 
-\begin{quote}
-Page numbers and line numbers were activated
-\end{quote}
+  \REVIEW{DONE} {Do not italicize in the text.}
+  {All italicized text
+    have been removed.  Formulas enclosed in dollar signs are in
+    italized text as is default for the JDS document style.}
 
-\item \DONE Do not italicize in the text.
+  \REVIEW{DONE} {AICov needs to be defined in the abstract.}
+  {We added ``We present the architecture of an artificial
+    intelligence enhanced COVID-19 analysis (in short AICov)''
+  in the abstract. We also defined COVID-10 in the abstract.}
 
-\begin{quote}
-All italicized text have been removed.
-inline formulas are done by latex italized. 
-\end{quote}
+\REVIEW{DONE}{Acronyms need to be defined at their first occurrences
+  (e.g., AI in the intro). An acronyme table is included in table 1,
+  we added the following acronyms to the paper}
+  {All acronyms have been defined at their first usage. In addition
+    we have added a section Acronyms in which all acronyms are listed
+    to make it easier for the reader to look them up in case they are
+    used later on in the paper.
 
-\item \DONE AICov needs to be defined in the abstract.
+    All riskfactors have been capitalized.
 
-\begin{quote}
-We added
+TODO. PERCENT, PERCENTBALCK, BLACK PERCENT, also for hispanics and senior
 
-"We present the architecture of an artificial intelligence enhanced COVID-19 analysis (in short AICov)"
+TODO. use risk factor in table headings
 
-was added to the abstract.
-\end{quote}
+TODO. PER1000 is /1000
 
-\item \PROGRESS Acronyms need to be defined at their first occurrences (e.g., AI in the intro). An acronyme table is included in table 1, we added the following acronyms to the paper
+TODO. make acronyms appear in text
 
-\begin{quote}
-    All acronyms have been defined at their first usage.
-In addition we have added a section Acronyms in which all acronyms are listed to make it easier for the reader to look them up in case they are used later on in the paper.
-\end{quote}
+TODO. Table 1. capitalization,explain why they are different or make them the same
+  }
 
-\DOIT capitalization in definition sin the table 1 and others
+\REVIEW{TODO}{Figure 2: please present the equations as equations. The
+  notations used in the equation need a full check (e.g., h is used as
+  both the dimension and vector). I'd like to see a clear distinction
+  between inputs/outputs and parameters. I'd suggest that the inputs
+  be set up before the equations are presented.}
+  { todo}
 
-\DOIT PERCENT, PERCENTBALCK, BLACK PERCENT, also for hispanics and senior
+\REVIEW{TODO}{Regenerate the figures into pdf format (the current
+  jpg/png figures blur when zoomed in).}
+  { todo}
 
-\DOIT use risk factor in table headings
+\REVIEW{DONE}{Table 2: consider reformating it; it would not fit the
+  page width in print.}
+  { Table 2 has been reformatted. It was previously fit in the column
+    with resize box. While rotating the headers the font is
+    increased. The table still fits in the column as it still uses
+    resizebox.}
 
-\DOIT PER1000 is /1000
+\REVIEW{DONE}{Table 3: tidy the headers.}
+  { The headers have been cleaned up and linebreaks wer introduced so
+    the resizebox of the table presents the table now larger} }
 
-\DOIT make acronyms appear in text
+\REVIEW{TODO}{Figure 5-6: can be combined to one figure with two
+  panels (using ggplot2 and save the space of the labels of the
+  horizontal axis).  }
+  { todo}
 
-\DOIT Table 1. capitalization,explain why they are different or make them the same
+\REVIEW{TODO}{Figure 7-8: can be one figure with two panels; pop
+  density 2010 needs to be cleaned.}
+  { todo}
 
-\item \DOIT  Figure 2: please present the equations as equations.
-The notations used in the equation need a full check (e.g., h is used as both the dimension and vector).
-I'd like to see a clear distinction between inputs/outputs and parameters.
-I'd suggest that the inputs be set up before the equations are presented.
+\REVIEW{TODO}{Figure 9-14: combine into one figure.}
+  { todo}
 
-\begin{quote}
-    \todo{todo}
-\end{quote}
+\REVIEW{DONE}{Table 4: keep the number of digits consistent.
+   The digits have been padded with the missing 0 to make them all
+    the same length}
+  { todo}
 
-\item \DOIT  Regenerate the figures into pdf format (the current jpg/png figures blur when zoomed in).
+\REVIEW{TODO}{Reproducibility: submit data/code/README in a
+  compressed file as supplementary material, or put them in a public
+  repo and give the link under an unnumbered section Supplementary
+  Material.}
+  { todo}
 
-\begin{quote}
-    \todo{todo}
-\end{quote}
-
-\item \DONE  Table 2: consider reformating it; it would not fit the page width in print.
-
-\begin{quote}
-Table 2 has been reformatted. It was previously fit in the column with resize box. While rotating the headers the font is increased. The table still fits in the column as it still uses resizebox.
-\end{quote}
-
-\item \DONE  Table 3: tidy the headers.
-
-\begin{quote}
-    The headers have been cleaned up and linebreaks wer introduced so the resizebox of the table presents the table now larger
-\end{quote}
-
-\item \DOIT  Figure 5-6: can be combined to one figure with two panels (using ggplot2 and save the space of the labels of the horizontal axis).
-
-\begin{quote}
-    \todo{todo}
-\end{quote}
-
-\item \DOIT  Figure 7-8: can be one figure with two panels; pop density 2010 needs to be cleaned.
-
-\begin{quote}
-    \todo{todo}
-\end{quote}
-
-\item \DOIT  Figure 9-14: combine into one figure.
-
-\begin{quote}
-    \todo{todo}
-\end{quote}
-
-\item \DONE  Table 4: keep the number of digits consistent.
-
-\begin{quote}
-    The digits have been padded with the missing 0 to make them all the same length
-\end{quote}
-
-\item \DOIT  Reproducibility: submit data/code/README in a compressed file as supplementary material, or put them in a public repo and give the link under an unnumbered section Supplementary Material.
-
-\begin{quote}
-    \todo{todo}
-\end{quote}
-
-\end{enumerate}
-
-To submit your revision, please log in to EJMS and submit it as a revised file to the original submission.
-Please also include a detailed description of how you addressed all the points raised by the reviewers.
+To submit your revision, please log in to EJMS and submit it as a
+revised file to the original submission. Please also include a
+detailed description of how you addressed all the points raised by the
+reviewers.
 
@@
 
-This paper presents AICov, an integrated deep learning framework for COVID-19 forecasting.
-It takes into account population covariates, and uses LSTM and event modeling to do the forecasting.
-Data integration is done at a local level.
-Integration of such information brings improvement to the prediction accuracy.
-Despite the authors’ effort to address my comments in their revision, the paper is still in rough shape.
-Please see my comments.
+This paper presents AICov, an integrated deep learning framework for
+COVID-19 forecasting. It takes into account population covariates, and
+uses LSTM and event modeling to do the forecasting. Data integration
+is done at a local level. Integration of such information brings
+improvement to the prediction accuracy. Despite the authors’ effort to
+address my comments in their revision, the paper is still in rough
+shape. Please see my comments.
 
-\begin{enumerate}
+\bigskip
 
-\item \DOIT Page 5, please write the full name for REST (representational state transfer ) when it first appears.
+\REVIEW{DONE}{Page 5, please write the full name for REST
+  (representational state transfer) when it first appears.}
+  { The term has been defined at first usage, we also checked all
+    other abbreviations. }
 
-\begin{quote}
-    The term has been defined at first usage, we also checked all other abbreviations.
+\REVIEW{TODO}{Page 5, line 3 under Requirement 3, “we are developed an
+  abstraction API for data” does not read right.
+}
+  { todo}
 
-    \todo{check other abbreviations}
+\REVIEW{TODO}{Page 6, line 8, what level is “local level”?
+}
+  { todo}
 
-\end{quote}
+\REVIEW{PROGRESS} Page 7, Table 1, is “CHD” also a percent? If
+  yes please modify the description; Percent- blacks and black percent
+  both correspond to “Percent of blacks in the population”. Both seem
+  to have been used according to Table 3, with difference in the four
+  model performance measurements. Do the authors have an explanation
+  for this?
+}
+  { The table has been updated and it was indicated that it is in
+    percent. There were indeed to different information sources that
+    reported different values for percentage of blacks and other
+    populations. This however has since than corrected and we have
+    been able to use just one population value.
 
-\item \DOIT Page 5, line 3 under Requirement 3, “we are developed an abstraction API for data” does not read right.
+    \todo{fix the values and only have one population value}}
 
-\begin{quote}
-    \todo{todo}
-\end{quote}
+\REVIEW{TODO}{Page 8, line 5 of Section 5.1, the abbreviation LSTM has
+  already appeared before, and the full name can be removed.
+}
+  { todo}
 
-\item \DOIT Page 6, line 8, what level is “local level”?
+\REVIEW{TODO}{What do the two panels in Figure 4 represent? Please give
+  clearer description either in the text or figure caption.
+}
+  { todo}
 
-\begin{quote}
-    \todo{todo}
-\end{quote}
+\REVIEW{TODO}{Section 5.2, paragraph 2 line 6, no need to capitalize for
+  the word “Samples”. Line 5 from the bottom, no need to capitalize
+  for the word “Insurance”. Please be careful with your writing.
+}
+  { todo}
 
-\item (partially done) Page 7, Table 1, is “CHD” also a percent?
-If yes please modify the description; Percent- blacks and black percent both correspond to “Percent of blacks in the population”.
-Both seem to have been used according to Table 3, with difference in the four model performance measurements.
-Do the authors have an explanation for this?
+\REVIEW{TODO}{Section 5.2, the “order for cases sorted by minimum model
+  order” and “order for deaths sorted by minimum model error”
+  paragraphs are redundant, as the same information is already
+  displayed in the x-legend for Figures 5 and 6.
+}
+  { todo}
 
-    \begin{quote}
-    The table has been updated and it was indicated that it is in percent.
-There were indeed to different information sources that reported different values for percentage of blacks and other populations.
-This however has since than corrected and we have been able to use just one population value.
+\REVIEW{TODO}{In Table 3, what is the unit for the error terms? The
+  errors in Figures 7 and 8 are percentages. If errors in Table 3 are
+  also percentages please clearly mark. Also, the numbers in Figures 7
+  and 8 do not align with numbers in row 1 of Table 3. Please check
+  your numbers with care.
+}
+  { todo}
 
-    \todo{fix the values and only have one population value}
-    \end{quote}
+\REVIEW{TODO}{Please adjust the location of your figures so they do not
+  look pages apart from their interpretations.
+}
+  { todo}
 
-\item \DOIT Page 8, line 5 of Section 5.1, the abbreviation LSTM has already appeared before, and the full name can be removed.
+\REVIEW{TODO}{If I am reading it correctly, Figures 9 and 11 are only
+  the “zoom in” versions of partial information from Figure
+  13. Similarly for the right column. Why waste space on two more
+  figures when you could make Figures 13 and 14 bigger and explain the
+  findings?
+}
+  { todo}
 
-    \begin{quote}
-        \todo{todo}
-    \end{quote}
+\REVIEW{TODO}{The last paragraph of Section 5.2 suits the
+  conclusion/discussion for future work section, not in the middle of
+  a paper.
+}
+  { todo}
 
-\item \DOIT What do the two panels in Figure 4 represent? Please give clearer description either in the text or figure caption.
+\REVIEW{TODO}{Section 5.3, the authors mentioned two way of constructing
+  covariates. Since the second way was more complicated while yielding
+  no better performance the authors did not use it. Then it can be
+  removed without damaging the storyline.
+}
+  { todo}
 
-    \begin{quote}
-        \todo{todo}
-    \end{quote}
+\REVIEW{TODO}{The description “We renormalized, so the two basic
+  features (following day) contributed 50\% of the loss function” is
+  hard to interpret. Please elaborate.
+}
+  { todo}
 
-\item \DOIT Section 5.2, paragraph 2 line 6, no need to capitalize for the word “Samples”. Line 5 from the bottom, no need to capitalize for the word “Insurance”. Please be careful with your writing.
+\REVIEW{TODO}{Paragraph 2 on page 14, “Selu, Relu and Tanh activations
+  gave very similar results while Sigmoid activation was much
+  worse. However we still use a Sigmoid function...” is very
+  counter-intuitive. Isn’t the point of trying different activation
+  functions to better fit the data?  16. Paragraph 3 on page 14, “The
+  basic COVID-19 daily data is extensive but compared two approaches
+  with and without renormalization, but in this case, we ...” The
+  sentence does not read correct. Please check your sentences with
+  care!
+}
+  { todo}
 
-    \begin{quote}
-        \todo{todo}
-    \end{quote}
-
-\item \DOIT Section 5.2, the “order for cases sorted by minimum model order” and “order for deaths sorted by minimum model error” paragraphs are redundant, as the same information is already displayed in the x-legend for Figures 5 and 6.
-
-    \begin{quote}
-        \todo{todo}
-    \end{quote}
-
-\item \DOIT In Table 3, what is the unit for the error terms?
-The errors in Figures 7 and 8 are percentages.
-If errors in Table 3 are also percentages please clearly mark.
-Also, the numbers in Figures 7 and 8 do not align with numbers in row 1 of Table 3.
-Please check your numbers with care.
-
-    \begin{quote}
-        \todo{todo}
-    \end{quote}
-
-\item \DOIT Please adjust the location of your figures so they do not look pages apart from their interpretations.
-
-    \begin{quote}
-        \todo{todo}
-    \end{quote}
-
-\item \DOIT If I am reading it correctly, Figures 9 and 11 are only the “zoom in” versions of partial information from Figure 13.
-Similarly for the right column.
-Why waste space on two more figures when you could make Figures 13 and 14 bigger and explain the findings?
-
-    \begin{quote}
-        \todo{todo}
-    \end{quote}
-
-\item \DOIT The last paragraph of Section 5.2 suits the conclusion/discussion for future work section, not in the middle of a paper.
-
-    \begin{quote}
-        \todo{todo}
-    \end{quote}
-
-\item \DOIT Section 5.3, the authors mentioned two way of constructing covariates.
-Since the second way was more complicated while yielding no better performance the authors did not use it.
-Then it can be removed without damaging the storyline.
-
-    \begin{quote}
-        \todo{todo}
-    \end{quote}
-
-\item \DOIT The description “We renormalized, so the two basic features (following day) contributed 50% of the loss function” is hard to interpret. Please elaborate.
-
-    \begin{quote}
-        \todo{todo}
-    \end{quote}
-
-\item \DOIT Paragraph 2 on page 14, “Selu, Relu and Tanh activations gave very similar results while Sigmoid activation was much worse.
-However we still use a Sigmoid function...” is very counter-intuitive.
-Isn’t the point of trying different activation functions to better fit the data?
-16.
-Paragraph 3 on page 14, “The basic COVID-19 daily data is extensive but compared two approaches with and without renormalization, but in this case, we ...” The sentence does not read correct.
-Please check your sentences with care!
-
-    \begin{quote}
-        \todo{todo}
-    \end{quote}
-
-\item \DOIT How to interpret the “significant positive correlation” in Section 6 from the practical perspec- tive?
-Also, the term “evolution function” made its first appearance in Section 6 without any definition or explanation; I suggest either removing the whole Section 6, or extend it with clearer definition of evolution function.
-18. The first two paragraphs do not read like to be in a “Conclusion” section. Rather, they better fit into the introduction.
-
-    \begin{quote}
-        \todo{todo}
-    \end{quote}
+\REVIEW{TODO}{How to interpret the “significant positive correlation” in
+  Section 6 from the practical perspective? Also, the term
+  “evolution function” made its first appearance in Section 6 without
+  any definition or explanation; I suggest either removing the whole
+  Section 6, or extend it with clearer definition of evolution
+  function.  18. The first two paragraphs do not read like to be in a
+  “Conclusion” section. Rather, they better fit into the introduction.
+}
+  { todo}
 

### 2020-12-19 vonLaszewski-covid.tex
update todo list and abbreviations

@@
 
-\newcommand{\ABSTRACT}{
-The COVID-19 pandemic has had profound global consequences on health,
-economic, social, behavioral, and almost every major aspect of human
-life. Therefore, it is of great importance to model COVID-19 and other
-pandemics in terms of the broader social contexts in which they take
-place.  We present the architecture of an artificial intelligence
-enhanced COVID-19 analysis (in short AICov) , which provides an
-integrative deep learning framework for COVID-19 forecasting with
-population covariates, some of which may serve as putative risk
-factors.  We have integrated multiple different strategies into AICov,
-including the ability to use deep learning strategies based on LSTM
-and event modeling.  To demonstrate our approach, we have introduced a
+\newcommand{\ABSTRACT}{ The COVID-19 (COrona VIrus Disease 2019)
+  pandemic has had profound global consequences on health, economic,
+  social, behavioral, and almost every major aspect of human
+  life. Therefore, it is of great importance to model COVID-19 and
+  other pandemics in terms of the broader social contexts in which
+  they take place.  We present the architecture of an artificial
+  intelligence enhanced COVID-19 analysis (in short AICov) , which
+  provides an integrative deep learning framework for COVID-19
+  forecasting with population covariates, some of which may serve as
+  putative risk factors.  We have integrated multiple different
+  strategies into AICov, including the ability to use deep learning
+  strategies based on Long Short-Term Memory (LSTM) and event
+  modeling.  To demonstrate our approach, we have introduced a
   framework that integrates population covariates from multiple
   sources. Thus, AICov not only includes data on COVID-19 cases and
-deaths but, more importantly, the population's socioeconomic, health,
-and behavioral risk factors at a local level. The compiled data are
-fed into AICov, and thus we obtain improved prediction by the
-integration of the data to our model as compared to one that only uses
-case and death data. As we use deep learning our models adapt over
-time while learning the model from past data.
-}
+  deaths but, more importantly, the population's socioeconomic,
+  health, and behavioral risk factors at a local level. The compiled
+  data are fed into AICov, and thus we obtain improved prediction by
+  the integration of the data to our model as compared to one that
+  only uses case and death data. As we use deep learning our models
+  adapt over time while learning the model from past data.}
 
@@
 \affil[1]{Digital Science Center, Indiana University, Bloomington Indiana, USA}
-\affil[2]{Public Health Dynamics Lab, and Department of Biostatistics, University of Pittsburgh, Pittsburgh, Pennsylvania, USA}
+\affil[2]{Public Health Dynamics Lab, and Department of Biostatistics,
+  University of Pittsburgh, Pittsburgh, Pennsylvania, USA}
 \affil[3]{Health Analytics Network, Pennsylvania, USA}
@@
 our predictions \citep{www-keras-lstm,Hochreiter1997-dk}. An LSTM is a
-recurrent neural network (RNN) \citep{Rumelhart1986-li} with feedback
+Recurrent Neural Network (RNN) \citep{Rumelhart1986-li} with feedback
 connections allowing the use of data input sequences to predict data
@@
   easy to use, and extensions can be made to allow usability of both
-  the API as well as the user interface. \Solution To address these
-  requirements, we are devising a specialized but easy to use time
-  series API for COVID-19 that integrates common tasks such as
-  automated filtering and normalization of the data.
+  the Application Programming Interfaces (API) as well as the user
+  interface. \Solution To address these requirements, we are devising
+  a specialized but easy to use time series API for COVID-19 that
+  integrates common tasks such as automated filtering and
+  normalization of the data.
 \end{requirement}
@@
   leverage from our earlier work and utilize through Cloudmesh and
-  NIST NBDIF definitions cloud services into the architecture
-  \citep{las-19-nist}. This includes containers but also infrastructure
-  services in the cloud. Through these service integrations, we also
-  provide access to GPUs.
+  National Institute of Standards and Technology (NIST) Big Data
+  Working Group (NBDIF) definitions cloud services into the
+  architecture \citep{las-19-nist}. This includes containers but also
+  infrastructure services in the cloud. Through these service
+  integrations, we also provide access to Graphics Processing Units
+  (GPUs).
 \end{requirement}
@@
 health and behavioral risk factors aggregated at the county level from
-the large-scale longitudinal CDC BRFSS surveys \citep{www-cdc-brfss},
-and demographic and socioeconomic data from the U.S. Census Bureau
-that are known at the local levels along with the cases and deaths
-data compiled by NCHS \citep{www-cdc-nchs} and integrate them with
-other data sources.
+the large-scale longitudinal CDC Behavioral Risk Factor Surveillance
+System (BRFSS) surveys \citep{www-cdc-brfss}, and demographic and
+socioeconomic data from the U.S. Census Bureau that are known at the
+local levels along with the cases and deaths data compiled by the CDC
+National Center for Health Statistics (NCHS) \citep{www-cdc-nchs} and
+integrate them with other data sources.
 
@@
 or accessible via specifically developed online resources such as the
-New York Times~\citep{github-nytimes} and the Johns Hopkins Coronavirus
-Resource Center~\citep{www-jh-covid19}.  For this study, we generated
-two data sets based on the latter resource upon our checking for the
-locations and times at which cumulative counts of COVID-19 cases and
-deaths on any given day were no less than those on the previous
-day. This led us to compile daily coronavirus cases and deaths data
-for 345 U.S. counties from 32 states and the District of Columbia and
-matched by five-digit FIPS code or county name to dynamic and static
-variables from additional data sources.
+New York Times~\citep{github-nytimes} and the Johns Hopkins
+Coronavirus Resource Center~\citep{www-jh-covid19}.  For this study,
+we generated two data sets based on the latter resource upon our
+checking for the locations and times at which cumulative counts of
+COVID-19 cases and deaths on any given day were no less than those on
+the previous day. This led us to compile daily coronavirus cases and
+deaths data for 345 U.S. counties from 32 states and the District of
+Columbia and matched by five-digit Federal Information Processing
+Standards (FIPS) code or county name to dynamic and static variables
+from additional data sources.
 
@@
 contributed 50\% of the loss function. We supported this in TensorFlow
-with a custom mean square error ($MSE$) loss function that dropped
+with a custom mean square error (MSE) loss function that dropped
 training values that were not available for time sequences near the
@@
 functions for this analysis in the LSTM and fully connected
-layers. $SELU$, $RELU$ and $TANH$ activations gave very similar
-results while $SIGMOID$ activation was much worse. However we still
-use a $SIGMOID$ function in the last stage just as we did in Section
-\ref{sec:lstm-covariate}. We also looked at the number of hidden units
-(separately for LSTM and fully connected layers), the number of LSTM
-layers and the dropout values.  Only the latter produced a significant
-change and, for example, removing dropouts decreased the loss by about
-16\% but we retained them to provide a robust environment. We chose
-randomly 20\% of the data for validation and testing and this had
-similar loss values to the training data (the other 80\%) with our
-chosen dropout.
+layers. Scaled Exponential Linear Unit (SELU), REctified Linear Unit
+(RELU) and TANH activations gave very similar results while SIGMOID
+activation was much worse. However we still use a SIGMOID function in
+the last stage just as we did in Section \ref{sec:lstm-covariate}. We
+also looked at the number of hidden units (separately for LSTM and
+fully connected layers), the number of LSTM layers and the dropout
+values.  Only the latter produced a significant change and, for
+example, removing dropouts decreased the loss by about 16\% but we
+retained them to provide a robust environment. We chose randomly 20\%
+of the data for validation and testing and this had similar loss
+values to the training data (the other 80\%) with our chosen dropout.
 
@@
 $\chi^{2}$, namely $((predicted-observed)/error)^{2}$. This approach
-with standard deep learning $MSE$ appears sound instead of dividing by
+with standard deep learning MSE appears sound instead of dividing by
 the population. That will emphasize the error contribution of smaller
@@
 models. One of the important features of the architecture for AICov is
-that it is by design targeting Cloud and HPC resources to conduct
-parameter sweeps to leverage sophisticated deep learning toolkits. The
-architecture allows the integration of various data sources that can
-update the data from its sources on demand and update its model
-predictions based on newly introduced data. Parameters can easily be
-adjusted via Jupyter notebooks that, in turn, call the computational
-backends on HPC and cloud resources.
+that it is by design targeting Cloud and High Performance Computing
+(HPC) resources to conduct parameter sweeps to leverage sophisticated
+deep learning toolkits. The architecture allows the integration of
+various data sources that can update the data from its sources on
+demand and update its model predictions based on newly introduced
+data. Parameters can easily be adjusted via Jupyter notebooks that, in
+turn, call the computational backends on HPC and cloud resources.
 
@@
 This work is partially supported by the National Science Foundation
-(NSF) through awards CIF21 DIBBS 1443054, nanoBIO 1720625,
-Cybertraining 1829704, CINES 1835598 and Global Pervasive
-Computational Epidemiology 1918626.  We thank J. Kadupitiya for
-several useful discussions on deep learning frameworks.
+(NSF) through awards Cyberinfrastructure Framework for 21st Century
+Data Infrastructure Building Blocks (1443054), Network for
+Computational Nanotechnology Engineered nanoBIO Node (1720625),
+Cybertraining (1829704), CyberInfrastructure for Network Engineering
+and Science (1835598) and Global Pervasive Computational Epidemiology
+(1918626).  We thank J. Kadupitiya for several useful discussions on
+deep learning frameworks.
 
@@
 
-This section contains the list of acronyms in Table{tab:acronymes} as used in the paper for easy
-reference. The definitions of the risk factors are however given
-within the paper in Table~\ref{tab:risk-factors} on page
-\pageref{tab:risk-factors} and not repeated here.
+This section contains the list of acronyms in Table \ref{tab:acronyms}
+as used in the paper for easy reference. The definitions of the risk
+factors are however given within the paper in
+Table~\ref{tab:risk-factors} on page \pageref{tab:risk-factors} and
+not repeated here.
+
+\todo{NORM\_POP Normalized population}
 
 \begin{table}[!h]
+  \caption{Non-risk factor acronyms used in this paper.}
+  \label{tab:acronyms}
   \begin{tabular}{ll}
@@
     BRFSS & Behavioral Risk Factor Surveillance System \\
-    CDC & Center for Desease Control \\
-    CI & Cyber Infrastructure \\
-    CIF21 & Cyberinfrastructure Framework for 21st Century  \\
-    CINES & CyberInfrastructure for Network Engineering and Science \\
+    CDC & U.S.~Centers for Disease Control and Prevention \\
+    CI & Continious INtegration \\
     COVID-19 & COrona VIrus Disease 2019  \\
@@
     cum & cumulative \\
-    DIBBS & Data Infrastructure Building Blocks \\
     FIPS & Federal Information Processing Standards \\
-    GPU & Graphics Processing Unit  \\
+    GPUs & Graphics Processing Units  \\
     HPC & High Performance Computing \\
@@
     REST & REpresentational State Transfer  \\
+    RMSE & Root Mean Squared Error \\
     RNN & Recurrent Neural Network \\

### 2020-12-19 response-v3-content.tex
update progress

@@
 
-My comments are the following:
-
 The revised manuscript is much smoother than the first version, but I
@@
   \REVIEW{PROGRESS} {Submit the point-by-point response to the
-    reviewer's comments and my comments in a pdf file.}  {The original
+    reviewer's comments and my comments in a pdf file.}{The original
     submisison web page had a field to enter the comments as text
-    which we used. We will change this and introduce a PDF file
-    instead, which is easier for us also.}
-
-  \REVIEW{PROGRESS} {Submit the point-by-point response to the
-    reviewer's comments and my comments in a pdf file.}
-  {The original submisison web page had a field to enter the comments
-    as text which we used. We will change this and introduce a PDF
-    file instead, which is easier for us also.}
+    which we used. We just added ``see attached PDF'' to make it
+    easier for reviewers as suggested by the editor.}
 
@@
 
-\REVIEW{DONE}{Acronyms need to be defined at their first occurrences
+\REVIEW{PROGRESS}{Acronyms need to be defined at their first occurrences
   (e.g., AI in the intro). An acronyme table is included in table 1,

### 2020-12-21 response-v3-content.tex
add local level fixes

@@
   source.}
-
-\end{enumerate}
 
@@
     to make it easier for the reader to look them up in case they are
-    used later on in the paper.
-
-    All riskfactors have been capitalized.
-
-TODO. PERCENT, PERCENTBALCK, BLACK PERCENT, also for hispanics and senior
-
-TODO. use risk factor in table headings
-
-TODO. PER1000 is /1000
-
-TODO. make acronyms appear in text
-
-TODO. Table 1. capitalization,explain why they are different or make them the same
+    used later on in the paper. All riskfactors have been capitalized.
+    The duplicated Risk factors were removed.
   }
@@
 \REVIEW{DONE}{Table 3: tidy the headers.}
-  { The headers have been cleaned up and linebreaks wer introduced so
-    the resizebox of the table presents the table now larger} }
+  { The headers have been cleaned up and linebreaks were introduced so
+    the resizebox of the table presents the table now larger}
 
@@
 
-\REVIEW{DONE}{Table 4: keep the number of digits consistent.
+\REVIEW{DONE}{Table 4: keep the number of digits consistent.}{
    The digits have been padded with the missing 0 to make them all
     the same length}
-  { todo}
 
@@
 
-\REVIEW{TODO}{Page 5, line 3 under Requirement 3, “we are developed an
+\REVIEW{DONE}{Page 5, line 3 under Requirement 3, “we are developed an
   abstraction API for data” does not read right.
 }
-  { todo}
+  {The sentence has been corrected to: ``To address this requirement, we are developing an
+  abstract API for data, analysis, and metadata parameter
+  adjustments.''}
 
@@
 }
-  { todo}
+  {PYNE: todo
 
-\REVIEW{PROGRESS} Page 7, Table 1, is “CHD” also a percent? If
+  (a) Absract: the term local level has been removed from the abstract and
+  replaced by at their ``specific locations''.
+
+  (b) Introduction: The term ``local level'' has been replaced with local which
+  according to an editor is suffiecient to refer to data that is
+  available for a locality.
+
+  (c) Line \linelabel{comment:level}  ``local level'' has been cahnged as follows
+
+We use data on community-specific health and behavioral risk factors
+aggregated at the county level from the large-scale longitudinal CDC
+Behavioral Risk Factor Surveillance System (BRFSS) surveys
+\citep{www-cdc-brfss} and demographic and socioeconomic data from the
+\citet{www-sensus}, i.e., both at the county and
+census tract, along with the cases and deaths data compiled by
+the CDC National Center for Health Statistics (NCHS)
+\citep{www-cdc-nchs} and integrate them with other data sources. These
+other data sources include
+
+\todo{Pyne}.
+
+}
+
+\REVIEW{PROGRESS}{ Page 7, Table 1, is “CHD” also a percent? If
   yes please modify the description; Percent- blacks and black percent
@@
 
-    \todo{fix the values and only have one population value}}
+    TODO. fix the values and only have one population value}
 
-\REVIEW{TODO}{Page 8, line 5 of Section 5.1, the abbreviation LSTM has
-  already appeared before, and the full name can be removed.
-}
-  { todo}
+\REVIEW{DONE}{Page 8, line 5 of Section 5.1, the abbreviation LSTM has
+  already appeared before, and the full name can be removed.}
+{All acronyms have been defined at first usage. A table has been added
+  at the end to make it easier for the reader to look them up.}
 
@@
 
-\REVIEW{TODO}{Section 5.2, paragraph 2 line 6, no need to capitalize for
+\REVIEW{DONE}{Section 5.2, paragraph 2 line 6, no need to capitalize for
   the word “Samples”. Line 5 from the bottom, no need to capitalize
@@
 }
-  { todo}
+{The capitaization has been corrected. The paper has been reviewed by
+  2 native English speakers. In addition 2 grammar checking programs
+  were used on the entire paper. One of the reviewers serves also as
+  editor in other journals.}
 
@@
 
-\end{enumerate}
-
 \clearpage

### 2020-12-21 vonLaszewski-covid.tex
add local level fixes

@@
 \documentclass[letterpaper, inpress]{jds} % use this for production
-
 %\documentclass[a4paper, review]{jds}      % use this for review
@@
 
+\usepackage{subcaption}
 \usepackage{calc}
@@
   deaths but, more importantly, the population's socioeconomic,
-  health, and behavioral risk factors at a local level. The compiled
-  data are fed into AICov, and thus we obtain improved prediction by
-  the integration of the data to our model as compared to one that
-  only uses case and death data. As we use deep learning our models
-  adapt over time while learning the model from past data.}
+  health, and behavioral risk factors at their specific locations.
+  The compiled data are fed into AICov, and thus we obtain improved
+  prediction by the integration of the data to our model as compared
+  to one that only uses case and death data. As we use deep learning
+  our models adapt over time while learning the model from past data.}
 
@@
 demographic and socioeconomic variables that are available at the
-local levels along with the cases and deaths data for COVID-19 while
+local along with the cases and deaths data for COVID-19 while
 leveraging the recently developed and increasingly popular deep
@@
 
-In this study, we used a Long Short-Term Memory (LSTM) algorithm for
+In this study, we used a LSTM algorithm for
 our predictions \citep{www-keras-lstm,Hochreiter1997-dk}. An LSTM is a
@@
   sophisticated scientific analytics workflows with the help of
-  Python.
+  Python and Jupyter Notebooks.
 \end{requirement}
@@
   developed over time, AICov must allow the integration of such
-  APIs. \Solution To address this requirement, we are developed an
-  abstraction API for data, analysis, and metadata parameter
-  adjustments. External services could be integrated with the help of
+  APIs. \Solution To address this requirement, we are developing an
+  abstract API for data, analysis, and metadata parameter
+  adjustments. External services can be integrated with the help of
   REST (REpresentational State Transfer) services. The REST services
@@
 integrator
-\citep{cloudmesh-openapi-install,cloudmesh-openapi-benchmark,cloudmesh-manual}. Additionally,
-we can also access through abstractions cloud services that are
-offered by the various cloud providers and are targeting our
+\citep{cloudmesh-openapi-install,cloudmesh-openapi-benchmark,cloudmesh-manual}.
+Additionally, we can also access through abstractions cloud services
+that are offered by the various cloud providers and are targeting our
 application domain. Different models can be easily integrated into our
@@
 health departments has been understandably difficult, especially in
-the areas that are hardest hit. We use data on community-specific
-health and behavioral risk factors aggregated at the county level from
-the large-scale longitudinal CDC Behavioral Risk Factor Surveillance
-System (BRFSS) surveys \citep{www-cdc-brfss}, and demographic and
-socioeconomic data from the U.S. Census Bureau that are known at the
-local levels along with the cases and deaths data compiled by the CDC
-National Center for Health Statistics (NCHS) \citep{www-cdc-nchs} and
-integrate them with other data sources.
+the areas that are hardest hit.
+%
+\linelabel{comment:level} We use data on community-specific health and
+behavioral risk factors aggregated at the county level from the
+large-scale longitudinal CDC Behavioral Risk Factor Surveillance
+System (BRFSS) surveys \citep{www-cdc-brfss} and demographic and
+socioeconomic data from the U.S. Census Bureau \cite{www-sensus},
+i.e., both at the county and census tract, along with the cases and
+deaths data compiled by the CDC National Center for Health Statistics
+(NCHS) \citep{www-cdc-nchs} and integrate them with other data
+sources. These other data sources include
+
+\todo{Pyne}.
 
@@
  PVI 		  & Pandemic vulnerability index \\
- PERCENTBLACKS 	  & Percent of blacks in the population\\
- PERCENTHISPANICS & Percent of hispanics in the population\\
  STROKE 	  & Percent reported with stroke\\
@@
 as discussed in \citep{marsland20-covid-paper}. The time series we used
-was 100 days long, and a multi-layered Long Short-Term Memory (LSTM)
+was 100 days long, and a multi-layered LSTM
 recurrent network was used. Our prediction approach differs by
@@
     \centering
-    \includegraphics{images/magic-1.png}
+    \includegraphics[width=1.0\textwidth]{images/magic-1.png}
+
+    \begin{subfigure}{.4\textwidth}
+      \centering
+      % include first image
+      % \includegraphics[width=.8\linewidth]{image_file_name}  
+      \caption{Prediction acuracy for deaths.}
+      \label{fig:magic-a}
+    \end{subfigure}
+    \begin{subfigure}{.4\textwidth}
+      \centering
+      % include second image
+      %\includegraphics[width=.8\linewidth]{image_file_name}  
+      \caption{Prediction acuracy for cases.}
+      \label{fig:magic-b}
+    \end{subfigure}
+
     \caption{Cumulative empirical fits can be very accurately
@@
 200. The number of nodes internally is 32. The maximum total number of
-Samples is 12540 when using 3 days as input length. We use a two-layer
+samples is 12540 when using 3 days as input length. We use a two-layer
 LSTM, as shown in Table \ref{tab:model}. Within our system, we have
 the ability to set a number of input days. For this analysis, we have
-chosen three different input lengths, namely 3, 4, and 5 days.
+chosen three different input lengths, namely 3, 4, and 5 days. The
+term ``None'' in the output shape means this dimension is
+variable. For our model it is the batch size.  \todo{What is the batch
+size.}
 
@@
 \item[Order for cases sorted by minimum model error:] NONE,
-  POP\_DENSITY\_2010, PHLTH, INSURANCE, PERCENTBLACKS,
-  PERCENTHISPANICS, NHOSP, INSURANCE, CHOLSCREEN, DIABETES, STROKE,
-  CHD, CHECKUP, NBEDS, SVI\_OVERALL, KIDNEY, CASTHMA, BLACK, BPMED,
+  POP\_DENSITY\_2010, PHLTH, INSURANCE, BLACK,
+  HISPANICS, NHOSP, CHOLSCREEN, DIABETES, STROKE,
+  CHD, CHECKUP, NBEDS, SVI\_OVERALL, KIDNEY, CASTHMA, BPMED,
   LPA, CSMOKING, ARTHRITIS, POVERTY, ESTBEDS, MHLTH, SENIOR, COPD,
@@
 \item[Order for deaths sorted by minimum model error:] NONE,
-  SVI\_OVERALL, CSMOKING, PERCENTHISPANICS, POP\_DENSITY\_2010,
-  SENIOR, CHOLSCREEN, CASTHMA, INSURANCE, INSURANCE, MHLTH, LPA,
-  POVERTY, OBESITY, CANCER, BLACK, NBEDS/1000, PERCENTBLACKS,
-  DIABETES, ARTHRITIS, NBEDS, NHOSP, BPHIGH, CHECKUP, PHLTH, CHD,
-  STROKE, HIGHCHOL, ESTBEDS, KIDNEY, BINGE, BPMED, COPD, SVI\_MINORITY
+  SVI\_OVERALL, CSMOKING, HISPANICS, POP\_DENSITY\_2010,
+  SENIOR, CHOLSCREEN, CASTHMA, INSURANCE, MHLTH, LPA,
+  POVERTY, OBESITY, CANCER, BLACK, NBEDS/1000, DIABETES, ARTHRITIS,
+  NBEDS, NHOSP, BPHIGH, CHECKUP, PHLTH, CHD, STROKE, HIGHCHOL,
+  ESTBEDS, KIDNEY, BINGE, BPMED, COPD, SVI\_MINORITY
 \end{description}
@@
 3 & INSURANCE & 0.055199 & 0.067641 & 8.07 & 20.43 & 4 \\
-4 & PERCENTBLACKS & 0.055028 & 0.067207 & 8.12 & 20.17 & 4 \\
-5 & PERCENTHISPANICS & 0.056061 & 0.067489 & 8.22 & 19.04 & 5 \\
+4 & BLACK & 0.055028 & 0.067207 & 8.12 & 20.17 & 4 \\
+5 & HISPANIC & 0.056061 & 0.067489 & 8.22 & 19.04 & 5 \\
 6 & NHOSP & 0.054621 & 0.068255 & 8.26 & 20.10 & 3 \\
-7 & INSURANCE & 0.054557 & 0.068572 & 8.34 & 24.61 & 3 \\
-8 & CHOLSCREEN & 0.054712 & 0.069121 & 8.37 & 21.70 & 3 \\
-9 & DIABETES & 0.054586 & 0.067636 & 8.40 & 20.50 & 5 \\
-10 & STROKE & 0.054160 & 0.066906 & 8.45 & 20.16 & 4 \\
-11 & CHD & 0.054027 & 0.067642 & 8.46 & 21.93 & 5 \\
-12 & CHECKUP & 0.054927 & 0.067652 & 8.51 & 20.42 & 5 \\
-13 & NBEDS & 0.054146 & 0.067536 & 8.54 & 21.15 & 5 \\
-14 & SVI\_OVERALL & 0.054795 & 0.067676 & 8.58 & 19.94 & 5 \\
-15 & KIDNEY & 0.054600 & 0.067565 & 8.61 & 21.86 & 5 \\
-16 & CASTHMA & 0.055089 & 0.068248 & 8.63 & 22.58 & 5 \\
-17 & BLACK & 0.054547 & 0.068266 & 8.68 & 20.49 & 3 \\
-18 & BPMED & 0.054685 & 0.068920 & 8.74 & 21.69 & 3 \\
-19 & LPA & 0.054323 & 0.067149 & 8.75 & 19.54 & 5 \\
-20 & CSMOKING & 0.053716 & 0.066898 & 8.75 & 19.01 & 5 \\
-21 & ARTHRITIS & 0.054053 & 0.068716 & 8.83 & 20.94 & 3 \\
-22 & POVERTY & 0.055029 & 0.068672 & 8.84 & 22.02 & 3 \\
-23 & ESTBEDS & 0.054545 & 0.067267 & 8.86 & 21.22 & 4 \\
-24 & MHLTH & 0.054864 & 0.068742 & 8.90 & 19.53 & 3 \\
-25 & SENIOR & 0.055079 & 0.069024 & 8.93 & 21.23 & 3 \\
-26 & COPD & 0.055589 & 0.068250 & 8.93 & 23.26 & 4 \\
-27 & CANCER & 0.054500 & 0.068363 & 8.99 & 19.70 & 3 \\
-28 & NBEDS/1000 & 0.054397 & 0.068491 & 9.05 & 21.89 & 3 \\
-29 & BPHIGH & 0.054609 & 0.067346 & 9.13 & 21.29 & 4 \\
-\rowcolor{black!5} 30 & NONE & 0.054868 & 0.069432 & 9.26 & 25.88 & 3 \\
-31 & HIGHCHOL & 0.056429 & 0.069345 & 9.28 & 20.98 & 3 \\
-32 & BINGE & 0.054317 & 0.067638 & 9.32 & 21.00 & 5 \\
-33 & OBESITY & 0.054469 & 0.068540 & 9.50 & 20.65 & 3 \\
-34 & SVI\_MINORITY & 0.055094 & 0.067443 & 9.54 & 21.53 & 4 \\
+7 & CHOLSCREEN & 0.054712 & 0.069121 & 8.37 & 21.70 & 3 \\
+8 & DIABETES & 0.054586 & 0.067636 & 8.40 & 20.50 & 5 \\
+9 & STROKE & 0.054160 & 0.066906 & 8.45 & 20.16 & 4 \\
+10 & CHD & 0.054027 & 0.067642 & 8.46 & 21.93 & 5 \\
+11 & CHECKUP & 0.054927 & 0.067652 & 8.51 & 20.42 & 5 \\
+12 & NBEDS & 0.054146 & 0.067536 & 8.54 & 21.15 & 5 \\
+13 & SVI\_OVERALL & 0.054795 & 0.067676 & 8.58 & 19.94 & 5 \\
+14 & KIDNEY & 0.054600 & 0.067565 & 8.61 & 21.86 & 5 \\
+15 & CASTHMA & 0.055089 & 0.068248 & 8.63 & 22.58 & 5 \\
+16 & BPMED & 0.054685 & 0.068920 & 8.74 & 21.69 & 3 \\
+17 & LPA & 0.054323 & 0.067149 & 8.75 & 19.54 & 5 \\
+18 & CSMOKING & 0.053716 & 0.066898 & 8.75 & 19.01 & 5 \\
+19 & ARTHRITIS & 0.054053 & 0.068716 & 8.83 & 20.94 & 3 \\
+20 & POVERTY & 0.055029 & 0.068672 & 8.84 & 22.02 & 3 \\
+21 & ESTBEDS & 0.054545 & 0.067267 & 8.86 & 21.22 & 4 \\
+22 & MHLTH & 0.054864 & 0.068742 & 8.90 & 19.53 & 3 \\
+23 & SENIOR & 0.055079 & 0.069024 & 8.93 & 21.23 & 3 \\
+24 & COPD & 0.055589 & 0.068250 & 8.93 & 23.26 & 4 \\
+25 & CANCER & 0.054500 & 0.068363 & 8.99 & 19.70 & 3 \\
+26 & NBEDS/1000 & 0.054397 & 0.068491 & 9.05 & 21.89 & 3 \\
+27 & BPHIGH & 0.054609 & 0.067346 & 9.13 & 21.29 & 4 \\
+\rowcolor{black!5} 28 & NONE & 0.054868 & 0.069432 & 9.26 & 25.88 & 3 \\
+29 & HIGHCHOL & 0.056429 & 0.069345 & 9.28 & 20.98 & 3 \\
+30 & BINGE & 0.054317 & 0.067638 & 9.32 & 21.00 & 5 \\
+31 & OBESITY & 0.054469 & 0.068540 & 9.50 & 20.65 & 3 \\
+32 & SVI\_MINORITY & 0.055094 & 0.067443 & 9.54 & 21.53 & 4 \\
 \bottomrule
@@
 BLACK                   & 0.209       & 0.099       & 0.208       & 0.216       & 0.085       & 0.116     \\
-HISPANICS               & -0.072      & -0.032      & -0.064      & -0.091      & -0.068      & -0.046    \\
+HISPANIC                & -0.072      & -0.032      & -0.064      & -0.091      & -0.068      & -0.046    \\
 POP\_DENSITY\_2010      & -0.302      & -0.210      & -0.175      & -0.272      & -0.231      & -0.216    \\

### 2020-12-21 response-v3-content.tex
update top10

@@
 
-\REVIEW{PROGRESS}{ Page 7, Table 1, is “CHD” also a percent? If
-  yes please modify the description; Percent- blacks and black percent
+\REVIEW{PROGRESS}{ Page 7, Table 1, is “CHD” also a percent? If yes
+  please modify the description; Percent- blacks and black percent
   both correspond to “Percent of blacks in the population”. Both seem
@@
   model performance measurements. Do the authors have an explanation
-  for this?
+  for this?  }
+%
+{ (a) The CHD and other values in the table have been updated wether
+  they are given in percent.
+
+  (b) There were two different information sources that we originally
+  used indicated by the capitalization. However since the submission
+  we have improved this data and have now merged all data and
+  corrected information for these duplicates, while only presenting
+  the one that is more accurate.
+
+  (c) \todo{TODO. fix the values and only have one population value}
+
 }
-  { The table has been updated and it was indicated that it is in
-    percent. There were indeed to different information sources that
-    reported different values for percentage of blacks and other
-    populations. This however has since than corrected and we have
-    been able to use just one population value.
-
-    TODO. fix the values and only have one population value}
 

### 2020-12-21 vonLaszewski-covid.tex
update top10

@@
 reused in a prediction process between input and output sequences.
+
+\todo{USE NEW REF: \cite{lstm-explained}}
+\todo{arrow on tanh is wrong}
 
@@
     \centering
-    \includegraphics[width=1.0\textwidth]{images/boxwhisker/boxplot_cases.pdf}
+    \includegraphics[width=1.0\textwidth]{images/boxwhisker/boxplot_casesNEW.pdf}
     \vspace{-1cm}
@@
 
-    \includegraphics[width=1.0\textwidth]{images/boxwhisker/boxplot_death.pdf}
+    \includegraphics[width=1.0\textwidth]{images/boxwhisker/boxplot_deathNEW.pdf}
     \vspace{-1cm}
@@
 \begin{figure}[!h]
+  \begin{subfigure}{.49\textwidth}
     \centering
-    \includegraphics[width=0.5\textwidth]{images/predict/Length_5-fields_pop_density_2010-Cumulative-Cases.png}
+    \includegraphics[width=1.0\textwidth]{images/predict/Length_5-fields_pop_density_2010-Cumulative-Cases.png}
     %\vspace{-1cm}
-    \caption{Model prediction error for cumulative cases when
-      including one risk factor - POP\_DENSITY\_2010 in the features,
-      in addition to the past number of cases and deaths.}
+    \caption{Model prediction error cases.}
     \label{fig:error-case-popdensity}
-\end{figure}
-
-\begin{figure}[!h]    
+  \end{subfigure}
+  \begin{subfigure}{.49\textwidth}
     \centering
-    \includegraphics[width=0.5\textwidth]{images/predict/Length_5-fields_pop_density_2010-Cumulative-Death.png}
+    \includegraphics[width=1.0\textwidth]{images/predict/Length_5-fields_pop_density_2010-Cumulative-Death.png}
     %\vspace{-1cm}
+    \caption{Model prediction error deaths.}
+    \label{fig:error-death-popdensity}
+    \end{subfigure}
+
     \caption{Model prediction error for cumulative deaths when
-      including one risk factor - POP\_DENSITY\_2010 in the features,
+      including one risk factor POP\_DENSITY\_2010
       in addition to the past number of cases and deaths.}
     \label{fig:error-death-popdensity}
+
 \end{figure}
@@
 
-We conducted one additional analysis and plotted the best risk factor
-prediction model errors, the five best, and the ten best from our
-experiments to identify if we have better predictions than by
-including no risk factor. The best result found with no risk factor is
-depicted with the black horizontal
-line. Figures~\ref{fig:place-top1-cases} to
-\ref{fig:place-top10-death} show the results sorted by the minimized
-error. Notably, we identified that a jump occurs when we sort the top
-10 model predictions at around 330 models from a total of 3400
+We conducted one additional analysis and plotted the top 10 best
+cummulative prediction model errors for each risk factor in
+Figure \label{fig:place-top10} for cases and deaths.  The horizontal
+black line shows the first occurance of no risk factor being used.
+
+Notably, we identified that a jump occurs when we sort the top 10
+model predictions at around 330 models from a total of 3400
 automatically generated models. This also underscores the need to run
 our deep learning model multiple times to obtain best parameter
-candidates.
+candidates.  Furthermore, we see that the cummulative error for cases
+is about half the size as for death. This is caused by cases having a
+larger number of sample.
+
+% In case of cases we also see a drop in the better
+% model predictions.
 
 \begin{figure}[!h]
-    \begin{minipage}{.45\textwidth}
+%    \begin{minipage}{.45\textwidth}
+%        \centering
+%        \includegraphics[width=1.0\textwidth]{images/predict/PlaceTop1_Cases.pdf}
+%        \vspace{-1cm}
+%        \caption{Best predictions for cases over all risk factors.}
+%        \label{fig:place-top1-cases}
+%
+%    \end{minipage}
+%    \ \
+%    \begin{minipage}{0.45\textwidth}
+%        \centering
+%        \includegraphics[width=1.0\textwidth]{images/predict/PlaceTop1_Death.pdf}
+%        \vspace{-1cm}
+%        \caption{Best predictions for deaths over all risk factors.}
+%        \label{fig:place-top1-death}
+%    \end{minipage}
+
+%    \begin{minipage}{.45\textwidth}
+%        \centering
+%        \includegraphics[width=1.0\textwidth]{images/predict/PlaceTop5_Cases.pdf}
+%        \vspace{-1cm}
+%        \caption{Top 5 predictions from all cases over all risk factors.}
+%        \label{fig:place-top5-cases}
+%    \end{minipage}
+%    \ \
+%    \begin{minipage}{.45\textwidth}
+%        \centering
+%        \includegraphics[width=1.0\textwidth]{images/predict/PlaceTop5_Death.pdf}
+%        \vspace{-1cm}
+%        \caption{Top 5 predictions from all deaths  over all risk factors.}
+%        \label{fig:place-top5-death}
+%    \end{minipage}
+    \begin{subfigure}{.45\textwidth}
         \centering
-        \includegraphics[width=1.0\textwidth]{images/predict/PlaceTop1_Cases.pdf}
-        \vspace{-1cm}
-        \caption{Best predictions for cases over all risk factors.}
-        \label{fig:place-top1-cases}
-
-    \end{minipage}
-    \ \
-    \begin{minipage}{0.45\textwidth}
+        \includegraphics[width=1.0\textwidth]{images/predict/PlaceTop10_CasesNEW.pdf}
+        \caption{Cummulative error for cases}
+        \label{fig:place-top10-cases}
+    \end{subfigure}
+    \begin{subfigure}{.45\textwidth}
         \centering
-        \includegraphics[width=1.0\textwidth]{images/predict/PlaceTop1_Death.pdf}
-        \vspace{-1cm}
-        \caption{Best predictions for deaths over all risk factors.}
-        \label{fig:place-top1-death}
-    \end{minipage}
-
-    \begin{minipage}{.45\textwidth}
-        \centering
-        \includegraphics[width=1.0\textwidth]{images/predict/PlaceTop5_Cases.pdf}
-        \vspace{-1cm}
-        \caption{Top 5 predictions from all cases over all risk factors.}
-        \label{fig:place-top5-cases}
-    \end{minipage}
-    \ \
-    \begin{minipage}{.45\textwidth}
-        \centering
-        \includegraphics[width=1.0\textwidth]{images/predict/PlaceTop5_Death.pdf}
-        \vspace{-1cm}
-        \caption{Top 5 predictions from all deaths  over all risk factors.}
-        \label{fig:place-top5-death}
-    \end{minipage}
-
-    \begin{minipage}{.45\textwidth}
-
-        \centering
-        \includegraphics[width=1.0\textwidth]{images/predict/PlaceTop10_Cases.pdf}
-        \vspace{-1cm}
-        \caption{Top 10 predictions from all deaths  over all risk factors.}
-        \label{fig:place-top10-cases}
-    \end{minipage}
-    \ \
-    \begin{minipage}{.45\textwidth}
-
-        \centering
-        \includegraphics[width=1.0\textwidth]{images/predict/PlaceTop10_Death.pdf}
-        \vspace{-1cm}
-        \caption{Top 10 predictions from all deaths over all risk factors.}
+        \includegraphics[width=1.0\textwidth]{images/predict/PlaceTop10_DeathNEW.pdf}
+        \caption{Cumulatice error for deaths.}
         \label{fig:place-top10-death}
-    \end{minipage}
+    \end{subfigure}
+
+    \caption{Top 10 predictions from all differnt risk factors sorted by cummulative error for cases and deaths.}
+    \label{fig:place-top10}
+
 \end{figure}

### 2020-12-22 response-v3-content.tex
Updates from Overleaf
move the top 10 to the reviews

@@
 
-\REVIEW{TODO}{Figure 9-14: combine into one figure.}
-  { todo}
+\REVIEW{PROGRESS}{Figure 9-14: combine into one figure.}
+  {
+
+\todo{the lines are missing in the images showing the NONE line}
+
+First we decided to only disaly the top 10 results of each prediction
+with a factor. Second we have included the combined picture of it for
+deth and cases in the {\bf Figure Combined}. We can include this
+figure in the paper if the editor thinks this figure is better or
+sufficient. At this tiem we still have included them in a subfigure
+(a) and (b) in the paper as the scales on the y axis are
+different. Please let us know what you think and we can adapt this.
+
+\begin{figure}[!h]
+        \centering
+        \includegraphics[width=1.0\textwidth]{images/predict/PlaceTop10_CasesAndDeath.pdf}
+
+        {\bf Figure Combined} Best predictions for cases over all risk factors.
+
+        %\label{fig:place-top10-joint}
+
+\end{figure}
+
+}
 

### 2020-12-31 response-v3-content.tex
update response

@@
 REVISION DUE: 01.12.2021
+
+\begin{itemize}
+\item Manuscript ID JDS2007-001 to Journal of Data Science.
+
+\item \url{http://jds.ruc.edu.cn/EN/column/item8.shtml}
+
+\item \url{http://jds.ruc.edu.cn/EN/column/column3.shtml}
+
+\item Progress: \url{https://www.e-publications.org/ruc/sbs/JDS/}
+
+\end{itemize}
+
+\bigskip
+
+\parindent 0pt We are happy to resubmit our paper and are very thankful for the great reviews we received.
+
+If possible we would like to acknowledge the reviewers by name in the
+acknowledgement section, but this depends on them and you.
+
+``AICov: A Deep Learning Framework for COVID-19 Forecastingwith
+CovariatesAICov: An Integrative Deep Learning Framework for
+COVID-19Forecasting with Covariates''
+
+Manuscript ID JDS2007-001 to Journal of Data Science.
+
+Progress: \url{https://www.e-publications.org/ruc/sbs/JDS/}
 
@@
 
-\REVIEW{PROGRESS}{Acronyms need to be defined at their first occurrences
+\REVIEW{DONE}{Acronyms need to be defined at their first occurrences
   (e.g., AI in the intro). An acronyme table is included in table 1,
@@
 
-\REVIEW{TODO}{Regenerate the figures into pdf format (the current
+\REVIEW{DONE}{Regenerate the figures into pdf format (the current
   jpg/png figures blur when zoomed in).}
-  { todo}
+  {All images have been included as PDF images}
 
@@
 
-\REVIEW{TODO}{Figure 5-6: can be combined to one figure with two
+\REVIEW{DONE}{Figure 5-6: can be combined to one figure with two
   panels (using ggplot2 and save the space of the labels of the
   horizontal axis).  }
-  { todo}
+  {The figures were combined as subfigures into Figure \ref{fig:magic-1} (a) and (b)}
 
-\REVIEW{TODO}{Figure 7-8: can be one figure with two panels; pop
+\REVIEW{DONE}{Figure 7-8: can be one figure with two panels; pop
   density 2010 needs to be cleaned.}
-  { todo}
+    {The figures were combined as subfigures into Figure \ref{fig:error-death-popdensity} (a) and (b)}
 
@@
 
-\begin{figure}[!h]
+%\begin{figure}[!h]
+{
         \centering
-        \includegraphics[width=1.0\textwidth]{images/predict/PlaceTop10_CasesAndDeath.pdf}
+        \includegraphics[width=.9\textwidth]{images/predict/PlaceTop10_CasesAndDeathNEW2.pdf}
 
         {\bf Figure Combined} Best predictions for cases over all risk factors.
-
+The two vertical lines represent the experiment in which no additional Risk factor is used for training. The top one for deaths and the bottom one for cases.}        
         %\label{fig:place-top10-joint}
 
-\end{figure}
+%\end{figure}
 
@@
 
-\REVIEW{TODO}{Please adjust the location of your figures so they do not
+\REVIEW{DONE}{Please adjust the location of your figures so they do not
   look pages apart from their interpretations.
 }
-  { todo}
+  {The location have een adjusted, but LaTeX tries to optimize the
+  placement. We have used the !h flag to force placement. If this
+  placement is still not sufficient please let us know. We also
+  removed images in order to facilitatie better placement. The
+  document is submitted in drafts mode which also could influence the
+  placement. We have tried the document in final mode with the LaTeX
+  style and it seems the placement is now fixed as we removed images.}
 
-\REVIEW{TODO}{If I am reading it correctly, Figures 9 and 11 are only
+\REVIEW{DONE}{If I am reading it correctly, Figures 9 and 11 are only
   the “zoom in” versions of partial information from Figure
@@
 }
-  { todo}
+  {We have provided in an earlier comment that we now only include the
+  Top 10 instead of Top 1, and Top 5 to fulfull the desire of the
+  reviewers to have these images combined. However we still separated
+  the deaths from teh cases view to ease the readers ability to see
+  more details. HOwever we can easily replace the image with the one
+  provided earlier if the reviewers desire. This is a small change and
+  will appreciate the revieers feedback on this further.}
 

### 2020-12-31 vonLaszewski-covid.tex
update response

@@
 \jdsdoi{xx.xxxx/xxxxxxxxx}      % doi
-\jdsreceived{???, 2020}       % optional: comment it out if no received date
+\jdsreceived{Jun 30, 2020}       % optional: comment it out if no received date
 \jdsaccepted{???, 2021}         % optional: comment it out if no accepted date
@@
 
-This section presents a short overview of the recurrent neural network algorithms used in this work.
+This section presents a short overview of the recurrent neural network
+algorithms used in this work.
 
@@
 
-In this study, we used a LSTM algorithm for
-our predictions \citep{www-keras-lstm,Hochreiter1997-dk}. An LSTM is a
-Recurrent Neural Network (RNN) \citep{Rumelhart1986-li} with feedback
+In this study, we used a LSTM algorithm for our predictions
+\citep{www-keras-lstm,Hochreiter1997-dk}. An LSTM is a Recurrent
+Neural Network (RNN) \citep{Rumelhart1986-li} with feedback
 connections allowing the use of data input sequences to predict data
@@
 
-\begin{requirement}{\bf Ease of Use.}
+\begin{requirement}{\bf Ease of Use.} \normalfont
   A major design criterion for AICov is that the framework must be
@@
 
-\begin{requirement}{\bf Interactive Access.}
+\begin{requirement}{\bf Interactive Access.} \normalfont
   Due to the experimental nature of analyzing the data, AICov must
@@
 
-\begin{requirement}{\bf Expandable API.}
+\begin{requirement}{\bf Expandable API.} \normalfont
   Assuming that new models and other analysis algorithms will be
@@
 
-\begin{requirement}{\bf Flexible Data Source Integration.}
+\begin{requirement}{\bf Flexible Data Source Integration.} \normalfont
   The integration of various sources of data is a key part of
@@
 
-\begin{requirement}{\bf Automated Update of the Analytics.}
+\begin{requirement}{\bf Automated Update of the Analytics.} \normalfont
   As the data can change, previous analyses may have to be
@@
 
-\begin{requirement}{\bf Flexible Model definition.}
+\begin{requirement}{\bf Flexible Model definition.} \normalfont
   We want to allow experimentation with various models in AICov that
@@
 
-\begin{requirement}{\bf Deep Learning Forecast.}
+\begin{requirement}{\bf Deep Learning Forecast.} \normalfont
   As the forecasting models depend on changing data that are made
@@
 
-\begin{requirement}{\bf Model Orchestration.}
+\begin{requirement}{\bf Model Orchestration.} \normalfont
   To coordinate the different model predictions and the generation of
@@
 
-\begin{requirement}{\bf Compute Resource Mapping.}
+\begin{requirement}{\bf Compute Resource Mapping.} \normalfont
   For other uses of AICov, we need to integrate a flexible resource
@@
 
-\begin{requirement}{\bf Convenient Interfaces.}
+\begin{requirement}{\bf Convenient Interfaces.} \normalfont
   As we have a wide variety of users, we need to enable interfaces
@@
 
-\begin{question} Can we apply deep learning strategies that self-learn the behavior of a model in order to provide future predictions? See Section \ref{sec:emperical}.
+\begin{question} \normalfont Can we apply deep learning strategies that self-learn the behavior of a model in order to provide future predictions? See Section \ref{sec:emperical}.
 \label{q:1}
@@
 
-\begin{question} Can we use real time data as input and apply deep learning strategies instead of models that self learn and provide future predictions? See Section \ref{sec:lstm-covariate}.
+\begin{question} \normalfont Can we use real time data as input and apply deep learning strategies instead of models that self learn and provide future predictions?
+See Section \ref{sec:lstm-covariate}.
 \label{q:2}
@@
 
-\begin{question} Does the inclusion of covariates improve the prediction quality while including geospatial and socio-economic risk factors? See Section \ref{sec:lstm-covariate}.
+\begin{question} \normalfont Does the inclusion of covariates improve the prediction quality while including geospatial and socio-economic risk factors? See Section \ref{sec:lstm-covariate}.
 \label{q:3}
@@
 
-\begin{question} What impact do we see when we introduce many risk factors? See Section \ref{sec:lstm-many}.
+\begin{question} \normalfont What impact do we see when we introduce many risk factors? See Section \ref{sec:lstm-many}.
 \label{q:4}
@@
     \centering
-    \includegraphics[width=1.0\textwidth]{images/magic-1.png}
-
-    \begin{subfigure}{.4\textwidth}
+
+    \begin{subfigure}{.49\textwidth}
       \centering
-      % include first image
-      % \includegraphics[width=.8\linewidth]{image_file_name}  
+      \includegraphics[width=1.0\linewidth]{images/predict/Fig4a.pdf}  
       \caption{Prediction acuracy for deaths.}
@@
     \end{subfigure}
-    \begin{subfigure}{.4\textwidth}
+    \begin{subfigure}{0.49\textwidth}
       \centering
-      % include second image
-      %\includegraphics[width=.8\linewidth]{image_file_name}  
+      \includegraphics[width=1.0\linewidth]{images/predict/Fig4b.pdf}        
       \caption{Prediction acuracy for cases.}
@@
     \centering
-    \includegraphics[width=1.0\textwidth]{images/predict/Length_5-fields_pop_density_2010-Cumulative-Cases.png}
+    \includegraphics[width=1.0\textwidth]{images/predict/Length_5-fields_pop_density_2010-Cumulative-Cases.pdf}
     %\vspace{-1cm}
@@
     \centering
-    \includegraphics[width=1.0\textwidth]{images/predict/Length_5-fields_pop_density_2010-Cumulative-Death.png}
+    \includegraphics[width=1.0\textwidth]{images/predict/Length_5-fields_pop_density_2010-Cumulative-Death.pdf}
     %\vspace{-1cm}
@@
         \centering
-        \includegraphics[width=1.0\textwidth]{images/predict/PlaceTop10_CasesNEW.pdf}
+        \includegraphics[width=1.0\textwidth]{images/predict/PlaceTop10_Cases.pdf}
         \caption{Cummulative error for cases}
@@
         \centering
-        \includegraphics[width=1.0\textwidth]{images/predict/PlaceTop10_DeathNEW.pdf}
+        \includegraphics[width=1.0\textwidth]{images/predict/PlaceTop10_Death.pdf}
         \caption{Cumulatice error for deaths.}

### 2020-12-31 vonLaszewski-covid.tex
update top10

@@
 %    \end{minipage}
-    \begin{subfigure}{.45\textwidth}
+    \begin{subfigure}{.95\textwidth}
+        \centering
+        \includegraphics[width=1.0\textwidth]{images/predict/PlaceTop10_Death.pdf}
+        \caption{Cumulatice error for deaths.}
+        \label{fig:place-top10-death}
+    \end{subfigure}
+    \begin{subfigure}{.95\textwidth}
         \centering
@@
     \end{subfigure}
-    \begin{subfigure}{.45\textwidth}
-        \centering
-        \includegraphics[width=1.0\textwidth]{images/predict/PlaceTop10_Death.pdf}
-        \caption{Cumulatice error for deaths.}
-        \label{fig:place-top10-death}
-    \end{subfigure}
-
-    \caption{Top 10 predictions from all differnt risk factors sorted by cummulative error for cases and deaths.}
+
+    \caption{Top 10 predictions from all differnt risk factors sorted by cummulative error for deaths (\ref{fig:place-top10-death}) and cases (\ref{fig:place-top10-cases}).
+The horizontal line in each figure represents the model cummulative error with no risk factor as input for deaths and cases respectively.}
     \label{fig:place-top10}

### 2020-12-31 vonLaszewski-covid.tex
Updates from Overleaf

--- 2020-12-21 vonLaszewski-covid.tex
@@
 %    \end{minipage}
-    \begin{subfigure}{.45\textwidth}
+    \begin{subfigure}{.90\textwidth}
         \centering
-        \includegraphics[width=1.0\textwidth]{images/predict/PlaceTop10_CasesNEW.pdf}
+        \includegraphics[width=1.0\textwidth]{images/predict/PlaceTop10_CasesNEW2.pdf}
         \caption{Cummulative error for cases}
@@
     \end{subfigure}
-    \begin{subfigure}{.45\textwidth}
+    \begin{subfigure}{.90\textwidth}
         \centering
-        \includegraphics[width=1.0\textwidth]{images/predict/PlaceTop10_DeathNEW.pdf}
+        \includegraphics[width=1.0\textwidth]{images/predict/PlaceTop10_DeathNEW2.pdf}
         \caption{Cumulatice error for deaths.}

### 2020-12-31 response-v3-content.tex
Updates from Overleaf

@@
 
-\REVIEW{TODO}{Reproducibility: submit data/code/README in a
+\REVIEW{DONE}{Reproducibility: submit data/code/README in a
   compressed file as supplementary material, or put them in a public
@@
   Material.}
-  { todo}
+  {A supplementary section was added with pointers to all available code and data. Usage of the data is restricted prior to publication of this paper.}
 

### 2020-12-31 vonLaszewski-covid.tex
Updates from Overleaf

--- 2020-12-31 vonLaszewski-covid.tex
@@
 
+The code and paper document represented to implement AICov are contained in several repositories:
+
+\begin{enumerate}
+    \item The entire cloudmesh code on which the cloud based implementation of the AICov framework is based and contains over 70 contributors is available publicly at \url{https://github.com/cloudmesh}.
+Cloudmesh contains a number of modules that dependent on the users access to cloud resources can be customized.
+A detailed manual about the configuration is available at \url{https://cloudmesh.github.io/cloudmesh-manual/}
+
+    \item The entire COVID-19 analysis is based leverages cloudmesh and uses Jupyter notebooks to coordinate its workflow as discussed in the architecture Figure \ref{fig:arch}.
+The code and data for the results presented in this paper are located in the repository at \url{https://github.com/cloudmesh/cloudmesh-covid}.
+
+    The data was analysed on a variety of supercomputing resources including an allocation of 20 compute nodes that were utilized to execute the  repeated model creation to assure reproducible results.
+
+    However, as a publication is pending, at this time, the use of the data is copyrighted and can not be used for other publications without contacting the authors.
+The data gathering and analysis is a significant intellectual contribution and we like to avoid that the data is taken before we have not secured a publication.
+
+    \item The entire paper is located in \LaTeX source in the github repository \url{https://github.com/cyberaide/paper-covid}.
+This repository will be open sourced after acceptance of publication.
+If desireed the authors can grant acess to this repository prior to publication.
+Please contact the corresponding author.
+
+\end{enumerate}
+
 Contain

### 2021-01-04 response-v3-content.tex
Updates from Overleaf

@@
 \REVIEW{DONE}{Acronyms need to be defined at their first occurrences
-  (e.g., AI in the intro). An acronyme table is included in table 1,
+  (e.g., AI in the intro). A table of acronyms is included in table 1,
   we added the following acronyms to the paper}
@@
     to make it easier for the reader to look them up in case they are
-    used later on in the paper. All riskfactors have been capitalized.
+    used later on in the paper. All risk factors have been capitalized.
     The duplicated Risk factors were removed.
@@
     increased. The table still fits in the column as it still uses
-    resizebox.}
+    resizebox.Thus it will fit in the print.}
 
@@
   { The headers have been cleaned up and linebreaks were introduced so
-    the resizebox of the table presents the table now larger}
+    the resizebox of the table presents the table now larger.}
 
@@
 
-First we decided to only disaly the top 10 results of each prediction
+First we decided to only display the top 10 results of each prediction
 with a factor. Second we have included the combined picture of it for
-deth and cases in the {\bf Figure Combined}. We can include this
-figure in the paper if the editor thinks this figure is better or
-sufficient. At this tiem we still have included them in a subfigure
+deaths and cases in the {\bf Figure Combined}. We can include this
+figure in the paper if the editor thinks this figure is acceptable. Presently, we have still included these in  subfigure
 (a) and (b) in the paper as the scales on the y axis are
-different. Please let us know what you think and we can adapt this.
+clearly distinct. Please let us know what if we need to change this.
 
@@
         {\bf Figure Combined} Best predictions for cases over all risk factors.
-The two vertical lines represent the experiment in which no additional Risk factor is used for training. The top one for deaths and the bottom one for cases.}        
+The two horizontal lines represent the experiment in which no additional Risk factor is used for training. The top one for deaths and the bottom one for cases.}        
         %\label{fig:place-top10-joint}
@@
 
-\REVIEW{TODO}{Page 6, line 8, what level is “local level”?
+\REVIEW{DONE}{Page 6, line 8, what level is “local level”?
 }
-  {PYNE: todo
+  {
 
-  (a) Absract: the term local level has been removed from the abstract and
-  replaced by at their ``specific locations''.
+  (a) Abstract: the phrase ``local level'' has been removed from the abstract and replaced by ``specific locations''.
 
-  (b) Introduction: The term ``local level'' has been replaced with local which
-  according to an editor is suffiecient to refer to data that is
-  available for a locality.
+  (b) Introduction: The phrase ``at the local level'' has been replaced with ``for the corresponding county'' as a U.S.~county is a well-defined geographical unit of area.
 
-  (c) Line \linelabel{comment:level}  ``local level'' has been cahnged as follows
+  (c) In Section 4, the sentence containing \linelabel{comment:level}  ``local level'' has been modified as follows:
 
@@
 \citep{www-cdc-brfss} and demographic and socioeconomic data from the
-\citet{www-sensus}, i.e., both at the county and
-census tract, along with the cases and deaths data compiled by
+\citet{www-sensus} for each county, along with the cases and deaths data compiled by
 the CDC National Center for Health Statistics (NCHS)
-\citep{www-cdc-nchs} and integrate them with other data sources. These
-other data sources include
-
-\todo{Pyne}.
+\citep{www-cdc-nchs} and integrate them with other data sources such as the American Hospital Directory \citep{www-ahd}.
 
@@
 %
-{ (a) The CHD and other values in the table have been updated wether
+{ (a) The CHD and other values in the table have been updated wherever
   they are given in percent.
@@
 }
-{The capitaization has been corrected. The paper has been reviewed by
+{The capitalization has been corrected. The paper has been reviewed by
   2 native English speakers. In addition 2 grammar checking programs
@@
 }
-  {The location have een adjusted, but LaTeX tries to optimize the
+  {The locations of the figures have been adjusted, although LaTeX tries to automatically optimize their
   placement. We have used the !h flag to force placement. If this
-  placement is still not sufficient please let us know. We also
-  removed images in order to facilitatie better placement. The
-  document is submitted in drafts mode which also could influence the
+  placement is still not appropriate, then please let us know. We have also
+  removed images in order to facilitate better placement. The
+  document is submitted in draft mode which might also influence the
   placement. We have tried the document in final mode with the LaTeX
-  style and it seems the placement is now fixed as we removed images.}
+  style and it seems that the placement is now fixed as we removed images.}
 
@@
 }
-  {We have provided in an earlier comment that we now only include the
-  Top 10 instead of Top 1, and Top 5 to fulfull the desire of the
-  reviewers to have these images combined. However we still separated
-  the deaths from teh cases view to ease the readers ability to see
-  more details. HOwever we can easily replace the image with the one
-  provided earlier if the reviewers desire. This is a small change and
-  will appreciate the revieers feedback on this further.}
+  {As we noted in an earlier comment, now we only include the
+  Top 10 instead of Top 1, and Top 5, to ensure -- as per the suggestion of the
+  reviewers -- that these images are combined. However we still separated
+  the deaths from the cases in the figure to ease the readers' ability to view
+  more details. However, we can easily replace the image with the ones
+  provided earlier if required. This is a minor change and we welcome any further feedback on this from the reviewers.}
 

### 2021-01-04 vonLaszewski-covid.tex
Updates from Overleaf

@@
 \affil[1]{Digital Science Center, Indiana University, Bloomington Indiana, USA}
-\affil[2]{Public Health Dynamics Lab, and Department of Biostatistics,
+\affil[2]{Public Health Dynamics Laboratory, and Department of Biostatistics,
   University of Pittsburgh, Pittsburgh, Pennsylvania, USA}
@@
 disease, non-pharmaceutical policy interventions such as social
-distancing and lockdowns are recommended by health experts to prevent
-further transmission. To gain insights into the possible impact of
+distancing and lockdowns are recommended by health experts to prevent further transmission. To gain insights into the possible impact of
 such measures on COVID-19 outcomes, we depend upon the ability to
@@
 integrated to model the scale and dynamics of this pandemic
-\citep{Pyne2015-ao}. Several organizations, including the Johns Hopkins
-University’s Center for Systems Science and Engineering, the New York
+\citep{Pyne2015-ao}. Several organizations, including the Johns Hopkins University’s Center for Systems Science and Engineering, the New York
 Times, the Atlantic’s COVID-19 tracking project, among others have
@@
 (Susceptible-Exposed-Infectious-Recovered) to Agent-based to Bayesian
-models to understand the non-pharmaceutical policy interventions (or
-the lack thereof) to predict disease dynamics and its impact on human
+models to understand the non-pharmaceutical policy interventions (or the lack thereof) to predict disease dynamics and its impact on human
 lives \citep{www-cdc-modeling-forecast}.
@@
 combine community-specific health and behavioral risk factors,
-demographic and socioeconomic variables that are available at the
-local along with the cases and deaths data for COVID-19 while
+demographic and socioeconomic variables that are available for the corresponding county along with the cases and deaths data for COVID-19 while
 leveraging the recently developed and increasingly popular deep
@@
 System (BRFSS) surveys \citep{www-cdc-brfss} and demographic and
-socioeconomic data from the U.S. Census Bureau \cite{www-sensus},
-i.e., both at the county and census tract, along with the cases and
+socioeconomic data from the U.S. Census Bureau \cite{www-sensus} for each county, along with the cases and
 deaths data compiled by the CDC National Center for Health Statistics
 (NCHS) \citep{www-cdc-nchs} and integrate them with other data
-sources. These other data sources include
-
-\todo{Pyne}.
+sources such as the American Hospital Directory \citep{www-ahd}.
 
@@
 We conducted one additional analysis and plotted the top 10 best
-cummulative prediction model errors for each risk factor in
+cumulative prediction model errors for each risk factor in
 Figure \label{fig:place-top10} for cases and deaths.  The horizontal
-black line shows the first occurance of no risk factor being used.
+black line shows the first occurrence of no risk factor being used.
 
@@
 our deep learning model multiple times to obtain best parameter
-candidates.  Furthermore, we see that the cummulative error for cases
+candidates.  Furthermore, we see that the cumulative error for cases
 is about half the size as for death. This is caused by cases having a
@@
         \includegraphics[width=1.0\textwidth]{images/predict/PlaceTop10_Death.pdf}
-        \caption{Cumulatice error for deaths.}
+        \caption{Cumulative error for deaths.}
         \label{fig:place-top10-death}
@@
         \includegraphics[width=1.0\textwidth]{images/predict/PlaceTop10_Cases.pdf}
-        \caption{Cummulative error for cases}
+        \caption{Cumulative error for cases}
         \label{fig:place-top10-cases}
@@
 
-    \caption{Top 10 predictions from all differnt risk factors sorted by cummulative error for deaths (\ref{fig:place-top10-death}) and cases (\ref{fig:place-top10-cases}).
-The horizontal line in each figure represents the model cummulative error with no risk factor as input for deaths and cases respectively.}
+    \caption{Top 10 predictions from all different risk factors sorted by cumulative error for deaths (\ref{fig:place-top10-death}) and cases (\ref{fig:place-top10-cases}).
+The horizontal line in each figure represents the model cumulative error with no risk factor as input for deaths and cases respectively.}
     \label{fig:place-top10}

### 2021-01-11 vonLaszewski-covid.tex
Updates from Overleaf

@@
 the areas that are hardest hit.
-%
-\linelabel{comment:level} We use data on community-specific health and
+
+%\linelabel{comment:level}
+
+We use data on community-specific health and
 behavioral risk factors aggregated at the county level from the
@@
       in addition to the past number of cases and deaths.}
-    \label{fig:error-death-popdensity}
+    \label{fig:error-death-popdensity-both}
 
@@
 
-\begin{figure}[!h]
+\begin{figure}[!htbp]
 %    \begin{minipage}{.45\textwidth}
@@
 The horizontal line in each figure represents the model cumulative error with no risk factor as input for deaths and cases respectively.}
-    \label{fig:place-top10}
+    \label{fig:place-top10-both}
 
@@
 
-Contain
-the brief description of the online supplementary materials.
-
 %

### 2021-01-12 response-v3-content.tex
Updates from Overleaf

@@
 
-REVISION DUE: 01.12.2021
+\bigskip
 
@@
 \item Manuscript ID JDS2007-001 to Journal of Data Science.
+
+\item REVISION DUE: 01.12.2021
 
@@
 
-``AICov: A Deep Learning Framework for COVID-19 Forecastingwith
-CovariatesAICov: An Integrative Deep Learning Framework for
-COVID-19Forecasting with Covariates''
+We include the following comments
 
-Manuscript ID JDS2007-001 to Journal of Data Science.
-
-Progress: \url{https://www.e-publications.org/ruc/sbs/JDS/}
-
-\section{Additional actions taken}
+\section*{Additional actions taken}
 
@@
 
-  \REVIEW{PROGRESS} {Submit the point-by-point response to the
+  \REVIEW{DONE}{Submit the point-by-point response to the
     reviewer's comments and my comments in a pdf file.}{The original
@@
     have been removed.  Formulas enclosed in dollar signs are in
-    italized text as is default for the JDS document style.}
+    italicized text as is default for the JDS document style.}
 
@@
 
-\REVIEW{TODO}{Figure 2: please present the equations as equations. The
+\REVIEW{DONE}{Figure 2: please present the equations as equations. The
   notations used in the equation need a full check (e.g., h is used as
@@
   be set up before the equations are presented.}
-  { todo}
+  { We have replaced the h with a capital H, We have started our explanation of the equations with the input and output vectors, we have removed the equations from the image and displayed them as separate equations using the LaTeX equation environment with numbers.
+}
 
@@
 
-\REVIEW{DONE}{Table 2: consider reformating it; it would not fit the
+\REVIEW{DONE}{Table 2: consider reformatting it; it would not fit the
   page width in print.}
@@
     increased. The table still fits in the column as it still uses
-    resizebox.Thus it will fit in the print.}
+    resizebox. Thus it will fit in the print.}
 
@@
 
-\REVIEW{PROGRESS}{Figure 9-14: combine into one figure.}
+\REVIEW{DONE}{Figure 9-14: combine into one figure.}
   {
 
-\todo{the lines are missing in the images showing the NONE line}
+% \todo{the lines are missing in the images showing the NONE line}
 
@@
 
-\REVIEW{PROGRESS}{ Page 7, Table 1, is “CHD” also a percent? If yes
+\REVIEW{DONE}{ Page 7, Table 1, is “CHD” also a percent? If yes
   please modify the description; Percent- blacks and black percent
@@
 
-  (c) \todo{TODO. fix the values and only have one population value}
-
+  (c) We have fixed all values.
+Furthermore we have split the table with the values in two tables fro deaths and for cases to relate them more easily for the reader to the box diagrams.
+This makes the display of the data consistent and the tables and figures match.
 }
@@
 
-\REVIEW{TODO}{What do the two panels in Figure 4 represent? Please give
+\REVIEW{DONE}{What do the two panels in Figure 4 represent? Please give
   clearer description either in the text or figure caption.
 }
-  { todo}
+  { Better descriptions have been added. }
 
 \REVIEW{DONE}{Section 5.2, paragraph 2 line 6, no need to capitalize for
-  the word “Samples”. Line 5 from the bottom, no need to capitalize
+  the word ``Samples''. Line 5 from the bottom, no need to capitalize
   for the word “Insurance”. Please be careful with your writing.
@@
 
-\REVIEW{TODO}{Section 5.2, the “order for cases sorted by minimum model
-  order” and “order for deaths sorted by minimum model error”
+\REVIEW{DONE}{Section 5.2, the ``order for cases sorted by minimum model
+  order'' and ``order for deaths sorted by minimum model error''
   paragraphs are redundant, as the same information is already
@@
 }
-  { todo}
+  { The figures have been redone as subfigures based on Reviewer ones comment. }
 
-\REVIEW{TODO}{In Table 3, what is the unit for the error terms? The
+\REVIEW{DONE}{In Table 3, what is the unit for the error terms? The
   errors in Figures 7 and 8 are percentages. If errors in Table 3 are
@@
 }
-  { todo}
+  {The table has been corrected. See details Reviewer 1}
 
@@
   placement. We have used the !h flag to force placement. If this
-  placement is still not appropriate, then please let us know. We have also
+  placement is still not appropriate, then please let us know. 
+   We have also
   removed images in order to facilitate better placement. The
@@
 
-\REVIEW{TODO}{The last paragraph of Section 5.2 suits the
+\REVIEW{DONE}{The last paragraph of Section 5.2 suits the
   conclusion/discussion for future work section, not in the middle of
@@
 }
-  { todo}
+  { This paragraph has been moved to the conclusion.}
 
-\REVIEW{TODO}{Section 5.3, the authors mentioned two way of constructing
+\REVIEW{DONE}{Section 5.3, the authors mentioned two way of constructing
   covariates. Since the second way was more complicated while yielding
@@
 }
-  { todo}
+  { Based on the reviewers suggestion we removed section 5.3 so we could focus on the previous data analysis only. }
 
-\REVIEW{TODO}{The description “We renormalized, so the two basic
+\REVIEW{DONE}{The description “We renormalized, so the two basic
   features (following day) contributed 50\% of the loss function” is
@@
 }
-  { todo}
+    { Based on the reviewers suggestion we removed section 5.3 so we could focus on the previous data analysis only. }
 
-\REVIEW{TODO}{Paragraph 2 on page 14, “Selu, Relu and Tanh activations
+\REVIEW{DONE}{Paragraph 2 on page 14, “Selu, Relu and Tanh activations
   gave very similar results while Sigmoid activation was much
@@
 }
-  { todo}
+  { Based on the reviewers suggestion we removed section 5.3 so we could focus on the previous data analysis only. }
 
-\REVIEW{TODO}{How to interpret the “significant positive correlation” in
+\REVIEW{DONE}{How to interpret the “significant positive correlation” in
   Section 6 from the practical perspective? Also, the term
@@
 }
-  { todo}
+  { Based on the suggestion of the reviewer we deleted Section 6}
 

### 2021-01-12 vonLaszewski-covid.tex
Updates from Overleaf

@@
-\documentclass[letterpaper, inpress]{jds} % use this for production
+%\documentclass[letterpaper, inpress]{jds} % use this for production
+\documentclass[a4paper, inpress]{jds} % use this for production
 %\documentclass[a4paper, review]{jds}      % use this for review
+
+% spacing
+% outcomment this for production
+\usepackage{setspace}
 \usepackage{lineno}
 \linenumbers
+\doublespace
+% above needs to be commented out for production
 
@@
 
-% table managemet
+% table management
 
@@
 
-\input{response-v3-content}
-
 \maketitle
@@
 
-In this study, we used a LSTM algorithm for our predictions
+In this study, we base our work on a traditional LSTM algorithm for our predictions
 \citep{www-keras-lstm,Hochreiter1997-dk}. An LSTM is a Recurrent
@@
 Different variants of LSTM algorithms exist. Ours is based on an LSTM
-cell depicted in Figure \ref{fig:lstm}.  It has an input gate, output
+cell depicted in Figure \ref{fig:lstm}.
+
+\begin{figure}[!htb]
+    \centering
+    \includegraphics[width=0.6\columnwidth]{images/lstm.pdf}
+    \caption{LSTM Unit Diagram}
+    \label{fig:lstm}
+\end{figure}
+
+It has an input gate, output
 gate, and a forget gate. The cell is maintaining how a subsequent
@@
 reused in a prediction process between input and output sequences.
-
-\todo{USE NEW REF: \cite{lstm-explained}}
-\todo{arrow on tanh is wrong}
-
-\begin{figure}[h!]
-\begin{minipage}[b]{0.6\textwidth}
-    \centering
-    \includegraphics[width=0.6\columnwidth]{images/lstm.pdf}
-    \caption{LSTM Unit Diagram}
-    \label{fig:lstm}
-\end{minipage}
-\ \
-\begin{minipage}[b]{0.3\textwidth}
-\vspace{-3cm}
-\begin{equation*}
-\begin{split}
-    f_t &= \sigma(W_f x_t + U_f h_{t-1} + b_f) \\
-    i_t &= \sigma(W_i x_t + U_i h_{t-1} + b_i) \\
-    o_t &= \sigma(W_o x_t + U_o h_{t-1} + b_o)\\
-    \tilde{c}_t &= tanh(W_c x_t + U_c h_{t-1} + b_c)\\
+The LSTM network will have a number of input values represented as  vector x and a number of output values in the last layer of the network predicting a number of future values \citep{lstm-explained}.
+For our experiments this includes the numbers of days we use as input to our prediction and the number of days we use as output of our prediction for a single cell.
+In our implementation the number for inputs and outputs can be adjusted.
+We repeat this prediction over all input sequences to create the output forecast sequences.
+Next we list the equations (\ref{eq:first}) - (\ref{eq:last}) used in LSTM.
+
+\newcommand{\RH}{\mathbb{R}^{H}}
+\newcommand{\RD}{\mathbb{R}^{D}}
+\begin{align}
+    f_t &= \sigma_g(W_f x_t + U_f h_{t-1} + b_f) \label{eq:first} \\
+    i_t &= \sigma_g(W_i x_t + U_i h_{t-1} + b_i) \\
+    o_t &= \sigma_g(W_o x_t + U_o h_{t-1} + b_o) \\
+    \tilde{c}_t &= \sigma_c(W_c x_t + U_c h_{t-1} + b_c) \\
     c_t &= f_t \circ c_{t-1} +i_t \circ \tilde{c}_t  \\
-    h_t &= o_t \circ tanh(c_t) 
-    \label{eq:lstm_equations}
-\end{split}
-\end{equation*}
-\caption{LSTM Equations}
-\label{fig:eq}
-\end{minipage}
-\end{figure}
-
-\newcommand{\Rh}{\mathbb{R}^{h}}
-
-We have summarized the equations used in LSTM in
-Figure~\ref{fig:eq}. We denote variables in lower case and matrices in
-upper case letters. $W_{q}$ and $U_{q}$ contain the weights of the
+    h_t &= o_t \circ \sigma_h(c_t) \label{eq:last}
+\end{align}
+
+We use in the euqations $\circ$ as the Hadamard product,
+$\sigma_g$ is the sigmoid function for the gate,
+$D$ the number of days as input features, and
+$H$ the number of hidden layers.
+Inputs and outputs are defined by 
+$x_{t} \in \RD$ the input vector to the LSTM unit, and 
+$h_{t}\in \RH$ the hidden state vector e.g., the last of which is the output vector of the LSTM.
+In our case the output contains the days we predict based on the input features.
+Other variables include
+$f_t \in \RH$ is the forget gate's activation vector, 
+$i_t \in \RH$ is the input/update gates activation vector, 
+$o_{t}\in \RH$ output gate's activation vector, 
+$\tilde{c}_{t}\in \RH$ the cell input activation vector
+$c_{t}\in \RH$ the cell state vector,  
+$\sigma_c$ is the cell activation function, 
+$W\in \mathbb {R} ^{H\times D}$, $U\in \mathbb {R} ^{H\times H}$ and $b\in \RH$ being the weight matrices and bias vector parameters which are learned during training.
+In particular we denote $W_{q}$ and $U_{q}$ contain the weights of the
 input and recurrent connections, where the subscript denotes either
 the input gate $i$, the output gate $o$, the forget gate $f$ or the
-memory cell $c$.  It is important to note that $c_{t}\in \mathbb
-{R}^{h}$ is not just one cell of one LSTM unit, but contains $h$ LSTM
-units' cells, representing a vector. To summarize our variables we
-use:
-$\circ$ is the Hadamard product,
-$\sigma$ is the sigmoid function,
-$d$ the number of input features,
-$h$ the number of hidden units,
-$x_{t}\in \mathbb R^d$ the input vector to the LSTM unit,
-$f_t\in \Rh$ the forget gate's activation vector,
-$i_{t}\in \Rh$ the input/update gate's activation vector,
-$o_{t}\in \Rh$ the output gate's activation vector,
-$h_{t}\in \Rh$ the hidden state vector (e.g., the output vector of the LSTM unit),
-$\tilde{c}_t\in \Rh$ the cell input activation vector,
-$c_{t}\in \Rh$ the cell state vector,
-$W\in \mathbb{R}^{h\times d}$, $U\in \mathbb{R} ^{h\times h}$, and
-$b\in \Rh$ are the weight matrices and bias vector parameters which
-need to be learned during training.
+memory cell $c$. For a more detailed discussion about LSTM's we refer to \cite{lstm-explained}.
 
@@
 for COVID-19 data that naturally depends on the covariates for each
-city. Thus instead, we feed the covariates into the recurrent network.
+city. Thus instead, we feed the covariates into the recurrent network as input in addition to the days.
 
@@
   developed over time, AICov must allow the integration of such
-  APIs. \Solution To address this requirement, we are developing an
+  models through programs and APIs. \Solution To address this 
+  requirement, we are developing an
   abstract API for data, analysis, and metadata parameter
@@
   REST (REpresentational State Transfer) services. The REST services
-  are automatically generated from function specifications.
+  are automatically generated from function specifications \citep{las-19-openapi-function}.
 \end{requirement}
@@
   architecture \citep{las-19-nist}. This includes containers but also
-  infrastructure services in the cloud. Through these service
-  integrations, we also provide access to Graphics Processing Units
+  infrastructure services in the cloud. Through the service
+  integration, we also provide access to Graphics Processing Units
   (GPUs).
@@
   towards data integration, parameter manipulation, and visualization
-  of the results
+  of the results.
 \end{requirement}
@@
 American Hospital Directory~\citep{www-ahd}.
-
 The list of covariates is presented in Table \ref{tab:risk-factors}.
@@
 
+\begin{question} \normalfont Are there some impact factors that influence the prediction more than others? See Section \ref{sec:lstm-covariate}.
+\label{q:4}
+\end{question}
+
+\begin{comment}
 \begin{question} \normalfont What impact do we see when we introduce many risk factors? See Section \ref{sec:lstm-many}.
@@
 \end{question}
+\end{comment}
 
@@
 
-To establish the feasibility of using deep learning, we have, in our
+To establish the feasibility of using LSTM, we have, in our
 analysis, verified that through deep learning we can recreate the
@@
 data sets, and smooth fitted data leads to very accurate deep learning
-descriptions via LSTM as depicted in both Figure \ref{fig:magic-1}
+descriptions via LSTM as depicted in Figure \ref{fig:magic-1}
 that are smaller than 1\% from the model predictions.
@@
 
-\subsection{Real time-Data Covariate LSTM Prediction}
+\subsection{Realtime-Data Covariate LSTM Prediction}
 \label{sec:lstm-covariate}
@@
 In this section, we establish that using deep learning can not only be
-applied to model predictions from real time data. As we want to
+applied to model predictions but also from real time data. As we want to
 evaluate the influence of the risk factors on our prediction framework
@@
 parameter sweep framework based on LSTM that conducts the prediction
-for the 110 cities while also considering the risk factors as listed
+for 110 cities while also considering the risk factors as listed
 in Table \ref{tab:risk-factors}. This framework can query any
@@
 sigmoid is chosen. A drop out rate of 0.2 is used, we use 16 input
-nodes, and have a batch size of 110. The maximum epoch is chosen to be
-200. The number of nodes internally is 32. The maximum total number of
+nodes, and have a batch size of 110. The  
+batch size is the subset size of the training sample (e.g. 110 out of 12540 for a 3 day input length) that is used to train the LSTM during its learning process. The maximum epoch is chosen to be
+200. This is the number of steps after we interrupt the learning process. The number of nodes internally is 32. The maximum total number of
 samples is 12540 when using 3 days as input length. We use a two-layer
@@
 term ``None'' in the output shape means this dimension is
-variable. For our model it is the batch size.  \todo{What is the batch
-size.}
+variable.
 
@@
     \centering
-    \includegraphics[width=1.0\textwidth]{images/boxwhisker/boxplot_casesNEW.pdf}
+    \includegraphics[width=1.0\textwidth]{images/boxwhisker/boxplot_casesNEW2.pdf}
     \vspace{-1cm}
@@
 
-    \includegraphics[width=1.0\textwidth]{images/boxwhisker/boxplot_deathNEW.pdf}
+    \includegraphics[width=1.0\textwidth]{images/boxwhisker/boxplot_deathNEW2.pdf}
     \vspace{-1cm}
@@
 
-We show the errors and the association with a risk factor in Table
-\ref{tab:top-1}, where cum\_error represents the cumulative error of
-the prediction. We sorted it by the cumulative error for cases. We see
-that in our experiment, almost all risk factors lead to an increased
-model prediction accuracy. As a result, we note that by including
+We show the errors and the association with a risk factors in Table
+ \ref{tab:top-1-cases} and \ref{tab:top-1-deaths}.
+In the tables we include the following columns (a) the Risk Factor (b) the root mean square error of all models for this risk factor (c) the minimum cumulative error by risk factor and (d) the number of days that wer used as input and identified to relate to the model with the minimum cumulative error.
+The table has in the first column the rank this particular model comperes to all other models with minimal risk factor.
+The rank was obtained by sorting the best model for each risk factor. 
+We display two tables. One for cases (Table \ref{tab:top-1-cases}) and one for deaths (Table \ref{tab:top-1-deaths}).
+
+We see that in our experiment, almost all risk factors lead to an increased
+model prediction accuracy for cases.  We note that by including
 factors such as population density, physical health Insurance,
 population breakdown by ethnicity, number of hospitals, and diabetes,
-it leads to better overall predictions for cases. We also see that
+lead to better overall predictions for cases. 
+We also see that
 many of these factors perform better on average.
@@
 \begin{table}[!p]
-\caption{Best models for each risk factor and their respective
-  errors. The model that uses no risk factor is highlighted in grey.}
-\label{tab:top-1}
+\caption{Best models for cases for each risk factor and their respective
+  errors.
+The model that uses no risk factor is highlighted in grey.
+We present the Root Mean Square Error (RMSE), the cumulative error and how many days as input were used for this corresponding model.}
+\label{tab:top-1-cases}
 \bigskip
@@
 \makecell{RMSE\\Cases} &  
- \makecell{RMSE\\Death} &  
- \makecell[r]{Cumu-\\lative \\Error \\Cases} &  
- \makecell[r]{Cumu-\\lative \\Error \\Death} &  
+  \makecell[r]{Minimum\\Cumu-\\lative \\Error for \\Cases by\\Risk Factor} & 
  \makecell[r]{Number of \\Days as \\Input} \\
 \midrule
-1 & POP\_DENSITY\_2010 & 0.055443 & 0.067617 & 7.94 & 22.74 & 5 \\
-2 & PHLTH & 0.054795 & 0.068529 & 8.01 & 20.55 & 3 \\
-3 & INSURANCE & 0.055199 & 0.067641 & 8.07 & 20.43 & 4 \\
-4 & BLACK & 0.055028 & 0.067207 & 8.12 & 20.17 & 4 \\
-5 & HISPANIC & 0.056061 & 0.067489 & 8.22 & 19.04 & 5 \\
-6 & NHOSP & 0.054621 & 0.068255 & 8.26 & 20.10 & 3 \\
-7 & CHOLSCREEN & 0.054712 & 0.069121 & 8.37 & 21.70 & 3 \\
-8 & DIABETES & 0.054586 & 0.067636 & 8.40 & 20.50 & 5 \\
-9 & STROKE & 0.054160 & 0.066906 & 8.45 & 20.16 & 4 \\
-10 & CHD & 0.054027 & 0.067642 & 8.46 & 21.93 & 5 \\
-11 & CHECKUP & 0.054927 & 0.067652 & 8.51 & 20.42 & 5 \\
-12 & NBEDS & 0.054146 & 0.067536 & 8.54 & 21.15 & 5 \\
-13 & SVI\_OVERALL & 0.054795 & 0.067676 & 8.58 & 19.94 & 5 \\
-14 & KIDNEY & 0.054600 & 0.067565 & 8.61 & 21.86 & 5 \\
-15 & CASTHMA & 0.055089 & 0.068248 & 8.63 & 22.58 & 5 \\
-16 & BPMED & 0.054685 & 0.068920 & 8.74 & 21.69 & 3 \\
-17 & LPA & 0.054323 & 0.067149 & 8.75 & 19.54 & 5 \\
-18 & CSMOKING & 0.053716 & 0.066898 & 8.75 & 19.01 & 5 \\
-19 & ARTHRITIS & 0.054053 & 0.068716 & 8.83 & 20.94 & 3 \\
-20 & POVERTY & 0.055029 & 0.068672 & 8.84 & 22.02 & 3 \\
-21 & ESTBEDS & 0.054545 & 0.067267 & 8.86 & 21.22 & 4 \\
-22 & MHLTH & 0.054864 & 0.068742 & 8.90 & 19.53 & 3 \\
-23 & SENIOR & 0.055079 & 0.069024 & 8.93 & 21.23 & 3 \\
-24 & COPD & 0.055589 & 0.068250 & 8.93 & 23.26 & 4 \\
-25 & CANCER & 0.054500 & 0.068363 & 8.99 & 19.70 & 3 \\
-26 & NBEDS/1000 & 0.054397 & 0.068491 & 9.05 & 21.89 & 3 \\
-27 & BPHIGH & 0.054609 & 0.067346 & 9.13 & 21.29 & 4 \\
-\rowcolor{black!5} 28 & NONE & 0.054868 & 0.069432 & 9.26 & 25.88 & 3 \\
-29 & HIGHCHOL & 0.056429 & 0.069345 & 9.28 & 20.98 & 3 \\
-30 & BINGE & 0.054317 & 0.067638 & 9.32 & 21.00 & 5 \\
-31 & OBESITY & 0.054469 & 0.068540 & 9.50 & 20.65 & 3 \\
-32 & SVI\_MINORITY & 0.055094 & 0.067443 & 9.54 & 21.53 & 4 \\
+
+1	&	POP\_DENSITY\_2010	&	0.055443	&	7.94	&	5	 \\
+2	&	PHLTH	&	0.054795	&	8.01	&	3	 \\
+3	&	BLACK	&	0.055028	&	8.12	&	4	 \\
+4	&	HISPANIC	&	0.056061	&	8.22	&	5	 \\
+5	&	NHOSP	&	0.054621	&	8.26	&	3	 \\
+6	&	INSURANCE	&	0.054557	&	8.34	&	3	 \\
+7	&	CHOLSCREEN	&	0.054712	&	8.37	&	3	 \\
+8	&	DIABETES	&	0.054586	&	8.4	&	5	 \\
+9	&	STROKE	&	0.05416	&	8.45	&	4	 \\
+10	&	CHD	&	0.054027	&	8.46	&	5	 \\
+11	&	CHECKUP	&	0.054927	&	8.51	&	5	 \\
+12	&	NBEDS	&	0.054146	&	8.54	&	5	 \\
+13	&	SVI\_OVERALL	&	0.054795	&	8.58	&	5	 \\
+14	&	KIDNEY	&	0.0546	&	8.61	&	5	 \\
+15	&	CASTHMA	&	0.055089	&	8.63	&	5	 \\
+16	&	BPMED	&	0.054685	&	8.74	&	3	 \\
+17	&	LPA	&	0.054323	&	8.75	&	5	 \\
+18	&	CSMOKING	&	0.053716	&	8.75	&	5	 \\
+19	&	ARTHRITIS	&	0.054053	&	8.83	&	3	 \\
+20	&	POVERTY	&	0.055029	&	8.84	&	3	 \\
+21	&	ESTBEDS	&	0.054545	&	8.86	&	4	 \\
+22	&	MHLTH	&	0.054864	&	8.9	&	3	 \\
+23	&	COPD	&	0.055589	&	8.93	&	4	 \\
+24	&	SENIOR	&	0.055079	&	8.93	&	3	 \\
+25	&	CANCER	&	0.0545	&	8.99	&	3	 \\
+26	&	NBEDS/1000	&	0.054397	&	9.05	&	3	 \\
+27	&	BPHIGH	&	0.054609	&	9.13	&	4	 \\
+\rowcolor{black!5} 28	&	NONE	&	0.054868	&	9.26	&	3	 \\
+29	&	HIGHCHOL	&	0.056429	&	9.28	&	3	 \\
+30	&	BINGE	&	0.054317	&	9.32	&	5	 \\
+31	&	OBESITY	&	0.054469	&	9.5	&	3	 \\
+32	&	SVI\_MINORITY	&	0.055094	&	9.54	&	4	 \\
+\bottomrule
+\end{tabular}
+%}
+\end{table}
+
+\begin{table}[!p]
+\caption{Best models for deaths for each risk factor and their respective
+  errors. The model that uses no risk factor is highlighted in grey.
+  We present the Root Mean Square Error (RMSE), the cumulative error and how many days as input were used for this corresponding model.
+  }
+\label{tab:top-1-deaths}
+\bigskip
+\centering
+%\resizebox{1.0\textwidth}{!}{ 
+\begin{tabular}{rrrrrlrl}
+\toprule
+ \makecell{Place/\\Rank} & 
+\makecell{Risk Factor} &  
+\makecell{RMSE\\Deaths} &  
+\makecell[r]{Minimum\\Cumu-\\lative \\Error for \\Deaths by\\Risk Factor} &  
+ \makecell[r]{Number of \\Days as \\Input} \\
+\midrule
+1	&	SVI\_OVERALL	&	0.067457	&	18.75	&	5	 \\
+2	&	CSMOKING	&	0.066898	&	19.01	&	5	 \\
+3	&	HISPANIC	&	0.067489	&	19.04	&	5	 \\
+\rowcolor{black!5} 4	&	NONE	&	0.06696	&	19.09	&	4	 \\
+5	&	POP\_DENSITY\_2010	&	0.066968	&	19.13	&	4	 \\
+6	&	SENIOR	&	0.068498	&	19.27	&	3	 \\
+7	&	CHOLSCREEN	&	0.067859	&	19.43	&	5	 \\
+8	&	CASTHMA	&	0.066952	&	19.46	&	5	 \\
+9	&	INSURANCE	&	0.067127	&	19.48	&	5	 \\
+10	&	MHLTH	&	0.068742	&	19.53	&	3	 \\
+11	&	LPA	&	0.067149	&	19.54	&	5	 \\
+12	&	POVERTY	&	0.067267	&	19.57	&	5	 \\
+13	&	OBESITY	&	0.067123	&	19.68	&	5	 \\
+14	&	CANCER	&	0.068363	&	19.70	&	3	 \\
+15	&	NBEDS/1000	&	0.06752	&	19.73	&	5	 \\
+16	&	BLACK	&	0.068352	&	19.83	&	3	 \\
+17	&	DIABETES	&	0.067636	&	19.84	&	5	 \\
+18	&	ARTHRITIS	&	0.066927	&	19.94	&	4	 \\
+19	&	NBEDS	&	0.066879	&	19.95	&	4	 \\
+20	&	NHOSP	&	0.067391	&	19.95	&	5	 \\
+21	&	BPHIGH	&	0.067799	&	19.99	&	5	 \\
+22	&	PHLTH	&	0.068737	&	20.02	&	3	 \\
+23	&	CHECKUP	&	0.067543	&	20.02	&	5	 \\
+24	&	CHD	&	0.06874	&	20.15	&	3	 \\
+25	&	STROKE	&	0.066906	&	20.16	&	4	 \\
+26	&	HIGHCHOL	&	0.068557	&	20.35	&	3	 \\
+27	&	ESTBEDS	&	0.067052	&	20.42	&	4	 \\
+28	&	KIDNEY	&	0.068948	&	20.49	&	3	 \\
+29	&	BINGE	&	0.067447	&	20.56	&	5	 \\
+30	&	BPMED	&	0.067399	&	20.69	&	5	 \\
+31	&	COPD	&	0.06745	&	20.98	&	5	 \\
+32	&	SVI\_MINORITY	&	0.067443	&	21.53	&	4	 \\
 \bottomrule
@@
 
-This is following our input data distribution in which we noted higher
-fluctuations relative to the overall value. A mitigation to this issue
+This can be explained by our input data distribution between cases and deaths in which we noted higher
+fluctuations relative to the overall value in cases of deaths. 
+A mitigation to this issue
 would be using a seven day average over the analyzed period. However,
@@
 
-We conducted one additional analysis and plotted the top 10 best
+We conducted an additional analysis and plotted the top 10 best
 cumulative prediction model errors for each risk factor in
@@
 is about half the size as for death. This is caused by cases having a
-larger number of sample.
+larger number of samples than those from deaths.
 
@@
         \centering
-        \includegraphics[width=1.0\textwidth]{images/predict/PlaceTop10_Death.pdf}
+        \includegraphics[width=1.0\textwidth]{images/predict/PlaceTop10_DeathNEW3.pdf}
         \caption{Cumulative error for deaths.}
@@
         \centering
-        \includegraphics[width=1.0\textwidth]{images/predict/PlaceTop10_Cases.pdf}
+        \includegraphics[width=1.0\textwidth]{images/predict/PlaceTop10_CasesNEW3.pdf}
         \caption{Cumulative error for cases}
@@
 
-While this study focuses on if and which risk factors can improve the
-predictions, we do not focus on why. Such analysis could provide
-insights into causal relationships. For example, information such as
-health insurance and the number of hospitals and beds are often
-related to the quality of services that affect early detection and
-treatment. Furthermore, population parameters such as density can
-determine the rate of spread and the number of cases. Similarly,
-distributions of comorbidities in a given population such as physical
-health, cholesterol screening, diabetes, etc., are found to be
-insightful in predicting outcomes \citep{Maleki}. Identifying such risk
-factors and correlating them to the prediction accuracy provides
-valuable candidates for future studies.
+\begin{comment}
 
@@
 
+\end{comment}
+
 \section{Conclusion}
+
+While this study focuses on if and which risk factors can improve the
+predictions, we do not focus on why. Such analysis could provide
+insights into causal relationships. For example, information such as
+health insurance and the number of hospitals and beds are often
+related to the quality of services that affect early detection and
+treatment. Furthermore, population parameters such as population density can
+determine the rate of spread and the number of cases. Similarly,
+distributions of comorbidities in a given population such as physical
+health, cholesterol screening, diabetes, etc., are found to be
+insightful in predicting outcomes \citep{Maleki}. Identifying such risk
+factors and correlating them to the prediction accuracy provides
+valuable candidates for future studies.
 
@@
 Cloudmesh contains a number of modules that dependent on the users access to cloud resources can be customized.
-A detailed manual about the configuration is available at \url{https://cloudmesh.github.io/cloudmesh-manual/}
-
-    \item The entire COVID-19 analysis is based leverages cloudmesh and uses Jupyter notebooks to coordinate its workflow as discussed in the architecture Figure \ref{fig:arch}.
+A detailed manual about the configuration is available at \url{https://cloudmesh.github.io/cloudmesh-manual/}.
+
+    \item The entire COVID-19 analysis leverages cloudmesh and uses Jupyter notebooks to coordinate its workflow as discussed in the architecture Figure \ref{fig:arch}.
 The code and data for the results presented in this paper are located in the repository at \url{https://github.com/cloudmesh/cloudmesh-covid}.
@@
     \item The entire paper is located in \LaTeX source in the github repository \url{https://github.com/cyberaide/paper-covid}.
-This repository will be open sourced after acceptance of publication.
-If desireed the authors can grant acess to this repository prior to publication.
+This repository will be open sourced after acceptance of publication to not violate any publisher restrictions.
+If desired the authors can grant access to this repository prior to publication.
 Please contact the corresponding author.
@@
 not repeated here.
-
-\todo{NORM\_POP Normalized population}
 
@@
 
+\clearpage
+
+\singlespace
+
+\maketitle
+
+\input{response-v3-content}
+
 \end{document}

### 2021-01-12 vonLaszewski-covid.tex
Updates from Overleaf

@@
 \linenumbers
-\doublespace
+%\doublespace
 % above needs to be commented out for production
@@
 \midrule
- NONE & No risk factor used\\
  ARTHRITIS 	  & Percent reported with arthritis \\
  BINGE 		  & Percent reported with binge drinking \\
+ BLACK	  & Percent of Blacks in the population\\
  BPHIGH 	  & Percent reported with  high blood pressure\\
@@
  HIGHCHOL 	  & Percent reported with high cholesterol\\
+ HISPANIC	  & Percent of Hispanic in the population\\
  INSURANCE 	  & Percent reported with insurance\\
@@
  NHOSP 		  & Number of hospitals\\
+ NONE & No risk factor used\\
  OBESITY 	  & Percent reported with  population that is obese\\
  PHLTH        & Percent reported with physical health issues\\
- PVI 		  & Pandemic vulnerability index \\
+ POP\_DENSITY\_2010 & Population Density from the 2010 Census\\
+ POVERTY & Percent of poverty in the population\\
+ % PVI 		  & Pandemic vulnerability index \\
+ SENIOR  & Percent of seniors in the population\\
  STROKE 	  & Percent reported with stroke\\
- BLACK	  & Percent of Blacks in the population\\
- POVERTY & Percent of poverty in the population\\
- SENIOR  & Percent of seniors in the population\\
  SVI\_MINORITY 	  & Percent of social vulnerability index in the minority population\\
@@
       \includegraphics[width=1.0\linewidth]{images/predict/Fig4a.pdf}  
-      \caption{Prediction acuracy for deaths.}
+      \caption{Prediction accuracy for deaths.}
       \label{fig:magic-a}
@@
       \includegraphics[width=1.0\linewidth]{images/predict/Fig4b.pdf}        
-      \caption{Prediction acuracy for cases.}
+      \caption{Prediction accuracy for cases.}
       \label{fig:magic-b}
@@
     \caption{Influence of risk factors on the accuracy of the
-      prediction of COVID-19 death. The y-axis represents the
+      prediction of COVID-19 deaths. The y-axis represents the
       cumulative error over all input data for the cities.}
@@
 
-\item[Order for cases sorted by minimum model error:] NONE,
-  POP\_DENSITY\_2010, PHLTH, INSURANCE, BLACK,
-  HISPANICS, NHOSP, CHOLSCREEN, DIABETES, STROKE,
-  CHD, CHECKUP, NBEDS, SVI\_OVERALL, KIDNEY, CASTHMA, BPMED,
-  LPA, CSMOKING, ARTHRITIS, POVERTY, ESTBEDS, MHLTH, SENIOR, COPD,
-  CANCER, NBEDS/1000, BPHIGH, HIGHCHOL, BINGE, OBESITY, SVI\_MINORITY
-
-\item[Order for deaths sorted by minimum model error:] NONE,
-  SVI\_OVERALL, CSMOKING, HISPANICS, POP\_DENSITY\_2010,
+\item[Order for cases sorted by minimum model error:] 
+  POP\_DENSITY\_2010, PHLTH, BLACK, HISPANIC, NHOSP, INSURANCE, 
+  CHOLSCREEN, DIABETES, STROKE, CHD, CHECKUP, NBEDS, SVI\_OVERALL, 
+  KIDNEY, CASTHMA, BPMED, LPA, CSMOKING, ARTHRITIS, POVERTY, ESTBEDS, 
+  MHLTH, COPD, SENIOR, CANCER, NBEDS/1000, BPHIGH, NONE, HIGHCHOL, 
+  BINGE, OBESITY, SVI\_MINORITY
+
+\item[Order for deaths sorted by minimum model error:] 
+  SVI\_OVERALL, CSMOKING, HISPANIC, NONE, POP\_DENSITY\_2010, 
   SENIOR, CHOLSCREEN, CASTHMA, INSURANCE, MHLTH, LPA, 
-  POVERTY, OBESITY, CANCER, BLACK, NBEDS/1000, DIABETES, ARTHRITIS,
-  NBEDS, NHOSP, BPHIGH, CHECKUP, PHLTH, CHD, STROKE, HIGHCHOL,
+  POVERTY, OBESITY, CANCER, NBEDS/1000, BLACK, DIABETES, ARTHRITIS, 
+  NBEDS, NHOSP, BPHIGH, PHLTH, CHECKUP, CHD, STROKE, HIGHCHOL, 
   ESTBEDS, KIDNEY, BINGE, BPMED, COPD, SVI\_MINORITY
+
 \end{description}
@@
 We observed a different scenario for deaths, in which we only
-identified two risk factors, namely the Social Vulnerability Index
-(SVI) and chain-smoking, that, when integrated into our deep learning
+identified three risk factors, namely the Social Vulnerability Index
+(SVI), chain-smoking, and Hispanic that, when integrated into our deep learning
 model, leads to better prediction results. However, the accuracy of
 the overall prediction of deaths is far less precise than that of the
-cases. To showcase this, we have included Figures
+cases. 
+To showcase this, we have included Figures
 \ref{fig:error-case-popdensity} and \ref{fig:error-death-popdensity}
-that show the respective errors.
+that show the respective errors for example for the population density.
 
@@
 our framework for an automated ingest of data on a daily basis to
-enable data fusion of new cases and death information. Hence, we can
+enable data fusion of new cases and deaths information. Hence, we can
 re-train the model with new incoming data every day. Through repeated
@@
 Notably, we identified that a jump occurs when we sort the top 10
-model predictions at around 330 models from a total of 3400
+model predictions at around 320 models from a total of 3200
 automatically generated models. This also underscores the need to run
@@
 candidates.  Furthermore, we see that the cumulative error for cases
-is about half the size as for death. This is caused by cases having a
+is about half the size as for deaths. This is caused by cases having a
 larger number of samples than those from deaths.
@@
         \includegraphics[width=1.0\textwidth]{images/predict/PlaceTop10_CasesNEW3.pdf}
-        \caption{Cumulative error for cases}
+        \caption{Cumulative error for cases.}
         \label{fig:place-top10-cases}
@@
 prediction once we integrate risk factors in addition to the
-time-dependent data such as cases and death resulting from
+time-dependent data such as cases and deaths resulting from
 COVID-19. While the improvements observed were modest, they do present

### 2021-01-27 vonLaszewski-covid.tex
Updates from Overleaf

@@
 \linenumbers
-%\doublespace
+\doublespace
 % above needs to be commented out for production
@@
 
+\newcommand{\RH}{\mathbb{R}^{H}}
+\newcommand{\RD}{\mathbb{R}^{D}}
+
 It has an input gate, output
@@
 Next we list the equations (\ref{eq:first}) - (\ref{eq:last}) used in LSTM.
-
-\newcommand{\RH}{\mathbb{R}^{H}}
-\newcommand{\RD}{\mathbb{R}^{D}}
+We use in the equations $\circ$ as the Hadamard product,
+$\sigma_g$ is the sigmoid function for the gate.
+With 
+$D$ being the number of days as input features, and
+$H$ being the number of hidden layers. 
+The inputs and outputs are defined by  
+$x_{t} \in \RD$ the input vector to the LSTM unit, and 
+$h_{t} \in \RH$ the hidden state vector e.g., the last of which is the output vector of the LSTM. In our case the output contains the days we predict based on the input features. The equations are
+
 \begin{align}
-    f_t &= \sigma_g(W_f x_t + U_f h_{t-1} + b_f) \label{eq:first} \\
-    i_t &= \sigma_g(W_i x_t + U_i h_{t-1} + b_i) \\
-    o_t &= \sigma_g(W_o x_t + U_o h_{t-1} + b_o) \\
-    \tilde{c}_t &= \sigma_c(W_c x_t + U_c h_{t-1} + b_c) \\
-    c_t &= f_t \circ c_{t-1} +i_t \circ \tilde{c}_t  \\
-    h_t &= o_t \circ \sigma_h(c_t) \label{eq:last}
+    f_t &= \sigma_g(W_f x_t + U_f h_{t-1} + b_f), \label{eq:first} \\
+    i_t &= \sigma_g(W_i x_t + U_i h_{t-1} + b_i), \\
+    o_t &= \sigma_g(W_o x_t + U_o h_{t-1} + b_o), \\
+    \tilde{c}_t &= \sigma_c(W_c x_t + U_c h_{t-1} + b_c), \\
+    c_t &= f_t \circ c_{t-1} +i_t \circ \tilde{c}_t,  \\
+    h_t &= o_t \circ \sigma_h(c_t) \label{eq:last},
 \end{align}
 
-We use in the euqations $\circ$ as the Hadamard product,
-$\sigma_g$ is the sigmoid function for the gate,
-$D$ the number of days as input features, and
-$H$ the number of hidden layers.
-Inputs and outputs are defined by 
-$x_{t} \in \RD$ the input vector to the LSTM unit, and 
-$h_{t}\in \RH$ the hidden state vector e.g., the last of which is the output vector of the LSTM.
-In our case the output contains the days we predict based on the input features.
-Other variables include
+where 
 $f_t \in \RH$ is the forget gate's activation vector, 
@@
     \resizebox{1.0\textwidth}{!}{ 
-    \begin{tabular}{|ll|lll|ll|ll|}
+    %\begin{tabular}{|ll|lll|ll|ll|}
+    \begin{tabular}{lllllllll}
     \toprule 
@@
       prediction of COVID-19 cases. The y-axis represents the
-      cumulative error over all input data for the cities.}
+      cumulative error over all input data for the cities. The x-axis labels correspond to the order of models for minimum model error.}
     \label{fig:box-cases}
@@
       prediction of COVID-19 deaths. The y-axis represents the
-      cumulative error over all input data for the cities.}
+      cumulative error over all input data for the cities. The x-axis labels correspond to the order of models for minimum model error.}
     \label{fig:box-death}
@@
 
+\begin{comment}
 Further, the obtained sorted order of the best prediction models by
@@
 \end{description}
+\end{comment}
 
@@
 
-1	&	POP\_DENSITY\_2010	&	0.055443	&	7.94	&	5	 \\
-2	&	PHLTH	&	0.054795	&	8.01	&	3	 \\
-3	&	BLACK	&	0.055028	&	8.12	&	4	 \\
-4	&	HISPANIC	&	0.056061	&	8.22	&	5	 \\
-5	&	NHOSP	&	0.054621	&	8.26	&	3	 \\
-6	&	INSURANCE	&	0.054557	&	8.34	&	3	 \\
-7	&	CHOLSCREEN	&	0.054712	&	8.37	&	3	 \\
-8	&	DIABETES	&	0.054586	&	8.4	&	5	 \\
-9	&	STROKE	&	0.05416	&	8.45	&	4	 \\
-10	&	CHD	&	0.054027	&	8.46	&	5	 \\
-11	&	CHECKUP	&	0.054927	&	8.51	&	5	 \\
-12	&	NBEDS	&	0.054146	&	8.54	&	5	 \\
-13	&	SVI\_OVERALL	&	0.054795	&	8.58	&	5	 \\
+1	&	POP\_DENSITY\_2010	&	0.0554	&	7.94	&	5	 \\
+2	&	PHLTH	            &	0.0548	&	8.01	&	3	 \\
+3	&	BLACK	            &	0.0550	&	8.12	&	4	 \\
+4	&	HISPANIC	        &	0.0561	&	8.22	&	5	 \\
+5	&	NHOSP	            &	0.0546	&	8.26	&	3	 \\
+6	&	INSURANCE	        &	0.0546	&	8.34	&	3	 \\
+7	&	CHOLSCREEN	        &	0.0547	&	8.37	&	3	 \\
+8	&	DIABETES	        &	0.0546	&	8.4	&	5	 \\
+9	&	STROKE	            &	0.0542	&	8.45	&	4	 \\
+10	&	CHD	                &	0.0540	&	8.46	&	5	 \\
+11	&	CHECKUP	            &	0.0549	&	8.51	&	5	 \\
+12	&	NBEDS	            &	0.0541	&	8.54	&	5	 \\
+13	&	SVI\_OVERALL	    &	0.0548	&	8.58	&	5	 \\
 14	&	KIDNEY	            &	0.0546	&	8.61	&	5	 \\
-15	&	CASTHMA	&	0.055089	&	8.63	&	5	 \\
-16	&	BPMED	&	0.054685	&	8.74	&	3	 \\
-17	&	LPA	&	0.054323	&	8.75	&	5	 \\
-18	&	CSMOKING	&	0.053716	&	8.75	&	5	 \\
-19	&	ARTHRITIS	&	0.054053	&	8.83	&	3	 \\
-20	&	POVERTY	&	0.055029	&	8.84	&	3	 \\
-21	&	ESTBEDS	&	0.054545	&	8.86	&	4	 \\
-22	&	MHLTH	&	0.054864	&	8.9	&	3	 \\
-23	&	COPD	&	0.055589	&	8.93	&	4	 \\
-24	&	SENIOR	&	0.055079	&	8.93	&	3	 \\
+15	&	CASTHMA	            &	0.0551	&	8.63	&	5	 \\
+16	&	BPMED	            &	0.0547	&	8.74	&	3	 \\
+17	&	LPA	                &	0.0543	&	8.75	&	5	 \\
+18	&	CSMOKING	        &	0.0537	&	8.75	&	5	 \\
+19	&	ARTHRITIS	        &	0.0541	&	8.83	&	3	 \\
+20	&	POVERTY	            &	0.0550	&	8.84	&	3	 \\
+21	&	ESTBEDS	            &	0.0545	&	8.86	&	4	 \\
+22	&	MHLTH	            &	0.0549	&	8.9	&	3	 \\
+23	&	COPD	            &	0.0556	&	8.93	&	4	 \\
+24	&	SENIOR	            &	0.0551	&	8.93	&	3	 \\
 25	&	CANCER	            &	0.0545	&	8.99	&	3	 \\
-26	&	NBEDS/1000	&	0.054397	&	9.05	&	3	 \\
-27	&	BPHIGH	&	0.054609	&	9.13	&	4	 \\
-\rowcolor{black!5} 28	&	NONE	&	0.054868	&	9.26	&	3	 \\
-29	&	HIGHCHOL	&	0.056429	&	9.28	&	3	 \\
-30	&	BINGE	&	0.054317	&	9.32	&	5	 \\
-31	&	OBESITY	&	0.054469	&	9.5	&	3	 \\
-32	&	SVI\_MINORITY	&	0.055094	&	9.54	&	4	 \\
+26	&	NBEDS/1000	        &	0.0544	&	9.05	&	3	 \\
+27	&	BPHIGH	            &	0.0546	&	9.13	&	4	 \\
+\rowcolor{black!5} 28 &	NONE	&	0.0549	&	9.26	&	3	 \\
+29	&	HIGHCHOL	        &	0.0564	&	9.28	&	3	 \\
+30	&	BINGE	            &	0.0543	&	9.32	&	5	 \\
+31	&	OBESITY	            &	0.0545	&	9.5	&	3	 \\
+32	&	SVI\_MINORITY	    &	0.0551	&	9.54	&	4	 \\
 \bottomrule
@@
 \midrule
-1	&	SVI\_OVERALL	&	0.067457	&	18.75	&	5	 \\
-2	&	CSMOKING	&	0.066898	&	19.01	&	5	 \\
-3	&	HISPANIC	&	0.067489	&	19.04	&	5	 \\
+1	&	SVI\_OVERALL	&	0.0675	&	18.75	&	5	 \\
+2	&	CSMOKING	    &	0.0669	&	19.01	&	5	 \\
+3	&	HISPANIC	    &	0.0676	&	19.04	&	5	 \\
 \rowcolor{black!5} 4	&	NONE	&	0.06696	&	19.09	&	4	 \\
-5	&	POP\_DENSITY\_2010	&	0.066968	&	19.13	&	4	 \\
-6	&	SENIOR	&	0.068498	&	19.27	&	3	 \\
-7	&	CHOLSCREEN	&	0.067859	&	19.43	&	5	 \\
-8	&	CASTHMA	&	0.066952	&	19.46	&	5	 \\
-9	&	INSURANCE	&	0.067127	&	19.48	&	5	 \\
-10	&	MHLTH	&	0.068742	&	19.53	&	3	 \\
-11	&	LPA	&	0.067149	&	19.54	&	5	 \\
-12	&	POVERTY	&	0.067267	&	19.57	&	5	 \\
-13	&	OBESITY	&	0.067123	&	19.68	&	5	 \\
-14	&	CANCER	&	0.068363	&	19.70	&	3	 \\
-15	&	NBEDS/1000	&	0.06752	&	19.73	&	5	 \\
-16	&	BLACK	&	0.068352	&	19.83	&	3	 \\
-17	&	DIABETES	&	0.067636	&	19.84	&	5	 \\
-18	&	ARTHRITIS	&	0.066927	&	19.94	&	4	 \\
-19	&	NBEDS	&	0.066879	&	19.95	&	4	 \\
-20	&	NHOSP	&	0.067391	&	19.95	&	5	 \\
-21	&	BPHIGH	&	0.067799	&	19.99	&	5	 \\
-22	&	PHLTH	&	0.068737	&	20.02	&	3	 \\
-23	&	CHECKUP	&	0.067543	&	20.02	&	5	 \\
-24	&	CHD	&	0.06874	&	20.15	&	3	 \\
-25	&	STROKE	&	0.066906	&	20.16	&	4	 \\
-26	&	HIGHCHOL	&	0.068557	&	20.35	&	3	 \\
-27	&	ESTBEDS	&	0.067052	&	20.42	&	4	 \\
-28	&	KIDNEY	&	0.068948	&	20.49	&	3	 \\
-29	&	BINGE	&	0.067447	&	20.56	&	5	 \\
-30	&	BPMED	&	0.067399	&	20.69	&	5	 \\
-31	&	COPD	&	0.06745	&	20.98	&	5	 \\
-32	&	SVI\_MINORITY	&	0.067443	&	21.53	&	4	 \\
+5	&	POP\_DENSITY\_2010	&	0.0670	&	19.13	&	4	 \\
+6	&	SENIOR	        &	0.0685	&	19.27	&	3	 \\
+7	&	CHOLSCREEN	    &	0.0679	&	19.43	&	5	 \\
+8	&	CASTHMA	        &	0.0670	&	19.46	&	5	 \\
+9	&	INSURANCE	    &	0.0671	&	19.48	&	5	 \\
+10	&	MHLTH	        &	0.0687	&	19.53	&	3	 \\
+11	&	LPA	            &	0.0671	&	19.54	&	5	 \\
+12	&	POVERTY	        &	0.0673	&	19.57	&	5	 \\
+13	&	OBESITY	        &	0.0671	&	19.68	&	5	 \\
+14	&	CANCER	        &	0.0687	&	19.70	&	3	 \\
+15	&	NBEDS/1000	    &	0.0675	&	19.73	&	5	 \\
+16	&	BLACK	        &	0.0684	&	19.83	&	3	 \\
+17	&	DIABETES	    &	0.0676	&	19.84	&	5	 \\
+18	&	ARTHRITIS	    &	0.0669	&	19.94	&	4	 \\
+19	&	NBEDS	        &	0.0669	&	19.95	&	4	 \\
+20	&	NHOSP	        &	0.0674	&	19.95	&	5	 \\
+21	&	BPHIGH	        &	0.0678	&	19.99	&	5	 \\
+22	&	PHLTH	        &	0.0687	&	20.02	&	3	 \\
+23	&	CHECKUP	        &	0.0675	&	20.02	&	5	 \\
+24	&	CHD	            &	0.0687	&	20.15	&	3	 \\
+25	&	STROKE	        &	0.0669	&	20.16	&	4	 \\
+26	&	HIGHCHOL	    &	0.0686	&	20.35	&	3	 \\
+27	&	ESTBEDS	        &	0.0671	&	20.42	&	4	 \\
+28	&	KIDNEY	        &	0.0689	&	20.49	&	3	 \\
+29	&	BINGE	        &	0.0674	&	20.56	&	5	 \\
+30	&	BPMED	        &	0.0674	&	20.69	&	5	 \\
+31	&	COPD	        &	0.0675	&	20.98	&	5	 \\
+32	&	SVI\_MINORITY	&	0.0674	&	21.53	&	4	 \\
 \bottomrule
@@
 
+%\clearpage
+%\singlespace
+%\maketitle
+%\input{response-v3-content}
+
 \clearpage
-
-\singlespace
-
-\maketitle
-
-\input{response-v3-content}
+\input{response-v4-content}
 

### 2021-01-27 draft/response-3.tex
prepare separation betweein final and draft

\documentclass[letterpaper, inpress]{jds} % use this for production
%\documentclass[a4paper, review]{jds}      % use this for review
\usepackage{lineno}
\linenumbers

\usepackage{subcaption}
\usepackage{calc}
\usepackage{comment}
\usepackage{todonotes}
\setuptodonotes{inline}

% color management

\usepackage{color}
\usepackage{tcolorbox}

% table management

%\usepackage[table]{xcolor}
\usepackage{makecell}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{rotating, multirow}
\usepackage{colortbl}

% Avoid non breaking words

\renewcommand{\_}{%
    \textunderscore\hspace{0pt}%
}
\setlength{\emergencystretch}{3em}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% the following edits should be done by Journal typesetters
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\setcounter{page}{1}            % set the first page number
\jdsmonth{January}              % month
\jdsyear{2021}                  % year
\jdsvolume{xx}                  % volume number
\jdsissue{xx}                   % issue number
\jdsdoi{xx.xxxx/xxxxxxxxx}      % doi
\jdsreceived{Jun 30, 2020}       % optional: comment it out if no received date
\jdsaccepted{???, 2021}         % optional: comment it out if no accepted date
%% manually set running header for a shorter list of authors if needed
% \shortauthors{A Author A, et al.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% edits by authors are given below
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{amsfonts,amsmath,amssymb,amsthm}
\usepackage{booktabs}

\newtheorem{requirement}{Requirement}
\newtheorem{question}{Question}

\newcommand{\TITLE}{AICov: An Integrative Deep
    Learning Framework for COVID-19 Forecasting with Population
    Covariates}

\newcommand{\ABSTRACT}{ The COVID-19 (COrona VIrus Disease 2019)
  pandemic has had profound global consequences on health, economic,
  social, behavioral, and almost every major aspect of human
  life. Therefore, it is of great importance to model COVID-19 and
  other pandemics in terms of the broader social contexts in which
  they take place.  We present the architecture of an artificial
  intelligence enhanced COVID-19 analysis (in short AICov) , which
  provides an integrative deep learning framework for COVID-19
  forecasting with population covariates, some of which may serve as
  putative risk factors.  We have integrated multiple different
  strategies into AICov, including the ability to use deep learning
  strategies based on Long Short-Term Memory (LSTM) and event
  modeling.  To demonstrate our approach, we have introduced a
  framework that integrates population covariates from multiple
  sources. Thus, AICov not only includes data on COVID-19 cases and
  deaths but, more importantly, the population's socioeconomic,
  health, and behavioral risk factors at their specific locations.
  The compiled data are fed into AICov, and thus we obtain improved
  prediction by the integration of the data to our model as compared
  to one that only uses case and death data. As we use deep learning
  our models adapt over time while learning the model from past data.}

\title[AICov: Deep
    Learning Framework for COVID-19 Forecasting with Population
    Covariates]{\TITLE}

\author[1]{Geoffrey C. Fox}
\author[1]{Gregor von Laszewski\thanks{Corresponding author. Email: laszewski@gmail.com}}
\author[1]{Fugang Wang}
\author[2,3]{Saumyadipta Pyne}

\affil[1]{Digital Science Center, Indiana University, Bloomington Indiana, USA}
\affil[2]{Public Health Dynamics Laboratory, and Department of Biostatistics,
  University of Pittsburgh, Pittsburgh, Pennsylvania, USA}
\affil[3]{Health Analytics Network, Pennsylvania, USA}

\begin{document}

\input{response-v3-content}

\maketitle

\input{response-v3-content}

\begin{abstract}
  \ABSTRACT
\end{abstract}

\begin{keywords} % alphabetical; excluding anything in the title already
  Cloudmesh;
  Comorbidities;
  Prediction;
  Risk factors.
\end{keywords}

\section{Introduction}

The COVID-19 pandemic has caused an enormous health and humanitarian
crisis worldwide. It is unlike any other single phenomenon that has
occurred in modern history since the end of World War II.  The
pandemic’s effects have spanned over a range that is so vast over
space, and yet so condensed over time, that the dual blows of
intensity and rapidity have exposed myriad systemic vulnerabilities in
many societies around the world.

The first known case was traced back to 17 November 2019. Since the
emergence of a cluster of cases in Wuhan, China, on 31 December 2019,
COVID-19 has spread rapidly worldwide. Just six months later, as of 23
June 2020, there are 8,993,659 COVID-19 cases and 469,587 deaths
globally. On 30 January 2020, the World Health Organization (WHO)
\citep{www-who} declared COVID-19 to be a Public Health Emergency of
International Concern and subsequently, on 11 March 2020, a pandemic.

In the United States, the first case of COVID-19 was confirmed on 20
January 2020 in Washington State, and as of 23 June 2020, there were
over 2.3 million confirmed cases and 121,167 deaths
\citep{www-nyt-map}.  Although initial efforts to reduce the spread
took place, recent data shows a resurgence at an increased scale. In
the absence of a vaccine or treatment to effectively combat the
disease, non-pharmaceutical policy interventions such as social
distancing and lockdowns are recommended by health experts to prevent further transmission. To gain insights into the possible impact of
such measures on COVID-19 outcomes, we depend upon the ability to
accurately forecast the spread of reported cases and confirmed deaths
and recoveries. Naturally, the accuracy of forecasting relies on the
availability of current, reliable data and historical data to
determine estimates of uncertainty.

During outbreaks of epidemics, initially data on cases and deaths
could be scarce, and the quality of data annotation, validation, and
aggregation might be uncertain. For instance, changes in clinical data
entry, such as the addition of a new category clinically diagnosed to
the existing lab-confirmed category, may likely have been reflected in
the reporting \citep{Petropoulos2020-hh}. Initial fears about COVID-19
among the general population with regards to the trajectory of the
pandemic could also affect administrative reporting. Diverse media
sources and differences in local and federal policies may add to the
general uncertainty about disease progression. Nevertheless, a
data-driven approach to forecasting can offer valuable insights into
the disease dynamics, and thereby an ability to objectively plan for
the near future.

Unlike earlier global outbreaks, during COVID-19, the current
worldwide digital ecosystem allows for real-time data collection,
which in turn is used by artificial intelligence (AI) and deep
learning systems to understand healthcare trends, model risk
associations, and predict outcomes \citep{Ting2020-lw}.  In addition to
traditional public health surveillance strategies available to most
countries, a variety of static and dynamic data types may be
integrated to model the scale and dynamics of this pandemic
\citep{Pyne2015-ao}. Several organizations, including the Johns Hopkins University’s Center for Systems Science and Engineering, the New York
Times, the Atlantic’s COVID-19 tracking project, among others have
developed real-time tracking maps for following COVID-19 cases around
the world using data from the U.S.~Centers for Disease Control and
Prevention (CDC), WHO, and other international health
agencies. Notably, the CDC receives forecasts from several modeling
groups that use a wide variety of approaches ranging from SEIR
(Susceptible-Exposed-Infectious-Recovered) to Agent-based to Bayesian
models to understand the non-pharmaceutical policy interventions (or the lack thereof) to predict disease dynamics and its impact on human
lives \citep{www-cdc-modeling-forecast}.

For precise and contextually relevant modeling of COVID-19 dynamics
for a given population or community, we use data integration to
combine community-specific health and behavioral risk factors,
demographic and socioeconomic variables that are available for the corresponding county along with the cases and deaths data for COVID-19 while
leveraging the recently developed and increasingly popular deep
learning strategies to forecast COVID-19 dynamics using our
model. Towards this, we develop a parallel computing platform called
AICov standing for AI-driven Platform for COVID-19 to implement the
different components of our framework and obtain the needed resources
through multi-cloud interfaces in a robust manner.

In this study, while we incorporate information from community and
county-specific covariates to inform our forecasting model for each
metropolitan area, we want to avoid the so-called ecological fallacy
in drawing inferences about individual disease outcomes based on large
area-level aggregated risk factors. Our objective is also to
underscore the nuanced roles played by pre-existing socioeconomic and
other prevailing conditions of a given community that can act as
determinants of its health outcomes and possible disparities,
especially under such sudden stresses to its current systems as those
felt during any pandemic.  It is important to note that the underlying
models are learned directly from the data instead of being projected
by a theoretical epidemiological model that relies on simulation
populations. This allows quick adaptation and self-adaptation in
regards to changing behaviors and regional differences.

The paper is structured as follows. In Section \ref{sec:lstm-theory},
we provide a short overview of deep learning-based time series
forecasting that we use in AICov. In Section \ref{sec:arch}, we
outline our architecture that enables the user to conduct powerful
time series forecasting by integrating static and dynamic information
on population risk factors with time-series disease data. In Section
\ref{sec:data}, we describe the input data for AICov. Next, we
demonstrate our analysis involving the population risk factors and
identify those that contribute to our prediction accuracy. Finally, we
present our conclusion.

\section{LSTM}
\label{sec:lstm-theory}

This section presents a short overview of the recurrent neural network
algorithms used in this work.

% ------------------------------
\subsection{LSTM Background}
\label{sec:lstm-background}
% ------------------------------

In this study, we base our work on a traditional LSTM algorithm for our predictions
\citep{www-keras-lstm,Hochreiter1997-dk}. An LSTM is a Recurrent
Neural Network (RNN) \citep{Rumelhart1986-li} with feedback
connections allowing the use of data input sequences to predict data
output sequences. LSTMs have been applied to many application areas
from handwriting \citep{Graves2009-qb}, image, feature detection, and
time series prediction \citep{Schmidhuber2005-oy}. LSTM's have an
internal state that is used to prevent the vanishing gradient problem
\citep{Hochreiter1991-mp} during the training of RNNs.

Different variants of LSTM algorithms exist. Ours is based on an LSTM
cell depicted in Figure \ref{fig:lstm}.

\begin{figure}[!htb]
    \centering
    \includegraphics[width=0.6\columnwidth]{images/lstm.pdf}
    \caption{LSTM Unit Diagram}
    \label{fig:lstm}
\end{figure}

It has an input gate, output
gate, and a forget gate. The cell is maintaining how a subsequent
value is calculated. This includes (a) how input values are
influencing the cell via the input gate, (b) how the forget gate
influences a memory value within the cell via the forget gate, and (c)
how the output gate influences the output activation of the LSTM
unit. A cell has a number of inputs and outputs that are
weighted. During a training step, these weights are learned to be
reused in a prediction process between input and output sequences.
The LSTM network will have a number of input values represented as  vector x and a number of output values in the last layer of the network predicting a number of future values \citep{lstm-explained}. For our experiments this includes the numbers of days we use as input to our prediction and the number of days we use as output of our prediction for a single cell. In our implementation the number for inputs and outputs can be adjusted. We repeat this prediction over all input sequences to create the output forecast sequences. Next we list the equations (\ref{eq:first}) - (\ref{eq:last}) used in LSTM.

\newcommand{\RH}{\mathbb{R}^{H}}
\newcommand{\RD}{\mathbb{R}^{D}}
\begin{align}
    f_t &= \sigma_g(W_f x_t + U_f h_{t-1} + b_f) \label{eq:first} \\
    i_t &= \sigma_g(W_i x_t + U_i h_{t-1} + b_i) \\
    o_t &= \sigma_g(W_o x_t + U_o h_{t-1} + b_o) \\
    \tilde{c}_t &= \sigma_c(W_c x_t + U_c h_{t-1} + b_c) \\
    c_t &= f_t \circ c_{t-1} +i_t \circ \tilde{c}_t  \\
    h_t &= o_t \circ \sigma_h(c_t) \label{eq:last}
\end{align}

We use in the euqations $\circ$ as the Hadamard product,
$\sigma_g$ is the sigmoid function for the gate,
$D$ the number of days as input features, and
$H$ the number of hidden layers.
Inputs and outputs are defined by 
$x_{t} \in \RD$ the input vector to the LSTM unit, and 
$h_{t}\in \RH$ the hidden state vector e.g., the last of which is the output vector of the LSTM. In our case the output contains the days we predict based on the input features. Other variables include
$f_t \in \RH$ is the forget gate's activation vector, 
$i_t \in \RH$ is the input/update gates activation vector, 
$o_{t}\in \RH$ output gate's activation vector, 
$\tilde{c}_{t}\in \RH$ the cell input activation vector
$c_{t}\in \RH$ the cell state vector,  
$\sigma_c$ is the cell activation function, 
$W\in \mathbb {R} ^{H\times D}$, $U\in \mathbb {R} ^{H\times H}$ and $b\in \RH$ being the weight matrices and bias vector parameters which are learned during training.
In particular we denote $W_{q}$ and $U_{q}$ contain the weights of the
input and recurrent connections, where the subscript denotes either
the input gate $i$, the output gate $o$, the forget gate $f$ or the
memory cell $c$. For a more detailed discussion about LSTM's we refer to \cite{lstm-explained}.

\subsection{Covariate Enhanced LSTM}

We often think of the laws of physics described by operators that
evolve the system given sufficient initial conditions and in this
language, we have shown how to represent Newton’s law operator by a
recurrent network \citep{Kadupitiya2020-zq}. We expect that the time
dependence of many complex systems: Covid pandemics, Southern
California earthquakes, traffic flow, security events can be described
by deep learning operators that both capture the dynamics and allow
predictions.

In this paper, we adopt this approach to describe the
time dependence of COVID-19 infection and fatality counts in different
regions. 
We intend to
process the two time-series (infections, fatalities/deaths) associated with
each city plus the covariates (fixed in time but dependent on region)
to build a model for the time evolution of the data. There is a
nontrivial architecture design choice as to the method of handling the
time-independent covariates. We could combine a recurrent network for
the time series with some fixed network for the covariates. However,
that does not fit with our concept of deriving an evolution operator
for COVID-19 data that naturally depends on the covariates for each
city. Thus instead, we feed the covariates into the recurrent network as input in addition to the days.

\section{AICov Architecture}
\label{sec:arch}

The motivation for creating this architecture is based on our
experience with deep learning toolkits such as Keras and
PyTorch. While these systems provide the necessary APIs to integrate
time series solutions, they target primarily more general user
communities. In particular, they do not consider existing data sets or
specific needs and analysis options to evaluate them for the purpose
of COVID-19 risk factor data integration. We describe below the
different requirements that motivate the design of our architecture
framework, AICov, and how we address each of these.

%\newcommand{\Solution}{{\bf Solution: }}
\newcommand{\Solution}{}

\begin{requirement}{\bf Ease of Use.} \normalfont
  A major design criterion for AICov is that the framework must be
  easy to use, and extensions can be made to allow usability of both
  the Application Programming Interfaces (API) as well as the user
  interface. \Solution To address these requirements, we are devising
  a specialized but easy to use time series API for COVID-19 that
  integrates common tasks such as automated filtering and
  normalization of the data.
\end{requirement}

\begin{requirement}{\bf Interactive Access.} \normalfont
  Due to the experimental nature of analyzing the data, AICov must
  support the exploration of the data in an interactive
  fashion. \Solution As many data scientists use Jupyter Notebooks to
  integrate their work interactively, AICov integrates such
  notebooks. Jupyter notebooks allow rapid modification of the
  analysis workflow through the ease of Python as programming
  language. Additionally, they enable us to formulate easily
  sophisticated scientific analytics workflows with the help of
  Python and Jupyter Notebooks.
\end{requirement}

\begin{requirement}{\bf Expandable API.} \normalfont
  Assuming that new models and other analysis algorithms will be
  developed over time, AICov must allow the integration of such
  models through programs and APIs. \Solution To address this 
  requirement, we are developing an
  abstract API for data, analysis, and metadata parameter
  adjustments. External services can be integrated with the help of
  REST (REpresentational State Transfer) services. The REST services
  are automatically generated from function specifications \citep{las-19-openapi-function}.
\end{requirement}

\begin{requirement}{\bf Flexible Data Source Integration.} \normalfont
  The integration of various sources of data is a key part of
  AICov. \Solution To achieve this goal, we provide a number of
  abstractions and data manipulation functions to extract needed data
  from established sources. Furthermore, the data can be combined, and
  additional data wrangling by external groups can be integrated.
  Through such abstractions, it is possible to replace, add, and
  correct data sets that are part of the analysis. This allows us to
  compare different results created from different data
  sets. Moreover, we include data selection directly in our framework
  to easily and quickly reuse them in a data mashup. An important
  example is that the data need not be static over the disease's
  progression but can be continuously updated. This also has a
  significant impact on our analytics algorithms.
\end{requirement}

\begin{requirement}{\bf Automated Update of the Analytics.} \normalfont
  As the data can change, previous analyses may have to be
  updated. \Solution Our forecast is not just a single script but
  includes a mechanism to register multiple workflows of continuously
  integrated (CI) analysis that are automatically rerun when new data
  are made available to the system. Previous results are
  maintained. Metadata with these runs can retrieve the versioned data
  sets used to calculate and store the versioned analytics workflow.
\end{requirement}

\begin{requirement}{\bf Flexible Model definition.} \normalfont
  We want to allow experimentation with various models in AICov that
  may be easily and flexibly definable as workflows. \Solution We give
  functional abstraction definitions that enables us to define the
  models ourselves or integrate third party models describing the
  spread of the disease.
\end{requirement}

\begin{requirement}{\bf Deep Learning Forecast.} \normalfont
  As the forecasting models depend on changing data that are made
  available daily, it is important for AICov to utilize the newly
  available data. Many models can be applied to this, including moving
  averages based models or models that are purely derived from deep
  learning. \Solution To address this requirement, AICov allows for
  the integration of both. Furthermore, in our deep learning
  framework, we can integrate an automated search for important risk
  factors, which is the focus of this study.
\end{requirement}

\begin{requirement}{\bf Model Orchestration.} \normalfont
  To coordinate the different model predictions and the generation of
  the deep learning forecasts, we need the ability to orchestrate them
  while applying a number of parameters. \Solution To address this
  requirement, the AICov architecture includes the ability to
  integrate parameter sweeps to, for example, identify
  hyperparameters, or integrate different data sets as parameters to
  identify suitable forecast models that can deliver the one which
  fits the best.
\end{requirement}

\begin{requirement}{\bf Compute Resource Mapping.} \normalfont
  For other uses of AICov, we need to integrate a flexible resource
  utilization framework. \Solution To address this requirement, we
  leverage from our earlier work and utilize through Cloudmesh and
  National Institute of Standards and Technology (NIST) Big Data
  Working Group (NBDIF) definitions cloud services into the
  architecture \citep{las-19-nist}. This includes containers but also
  infrastructure services in the cloud. Through the service
  integration, we also provide access to Graphics Processing Units
  (GPUs).
\end{requirement}

\begin{requirement}{\bf Convenient Interfaces.} \normalfont
  As we have a wide variety of users, we need to enable interfaces
  that are used by the various communities. \Solution To address this
  requirement, we provide a number of APIs, REST services and make
  them available via Jupyter Notebooks. In addition, we developed some
  custom widgets for the notebooks that are specifically targeted
  towards data integration, parameter manipulation, and visualization
  of the results.
\end{requirement}

Putting all these requirements together results in an architecture as
shown in Figure \ref{fig:arch}.\\

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.6\columnwidth]{images/arch.pdf}
    \caption{AICov Architecture}
    \label{fig:arch}
\end{figure}

In summary, the AICov architecture allows the inclusion of new data
sources and models. An orchestrator enables modifying parameters and
selecting data sources and models used in the analysis. The framework
can be accessed via the interface layer that includes APIs, REST, and
Jupyter Notebooks. Compute interfaces are provided either by direct
access of GPUs from local resources or through the staging of
automatically generated REST services via our Cloudmesh REST Service
integrator
\citep{cloudmesh-openapi-install,cloudmesh-openapi-benchmark,cloudmesh-manual}.
Additionally, we can also access through abstractions cloud services
that are offered by the various cloud providers and are targeting our
application domain. Different models can be easily integrated into our
framework to verify our forecast or enhance our forecast based on the
users' models and data.

\section{Data for the COVID-19 Analysis}
\label{sec:data}

While the U.S. has seen an overall case fatality rate of 5.14\%, some
U.S. locations have experienced a disproportionate number of deaths
compared to others. This disparity of deaths among different
communities could be attributed to a diversity of local risk factors
and socioeconomic determinants of health that now constitute a very
active area of investigation. Researchers are studying the effects of
different community-specific socioeconomic variables, underlying
health conditions and comorbidities, infrastructure, and systemic
responses.  Given the complexity of such underlying conditions that
can determine individual disease outcomes, where and under what
conditions people get infected by the virus, and then recover or die
take place not in isolation but rather in an interplay of diverse
factors that are characteristic to a local community, which need to be
synthesized. In other words, a basic requirement for effective
modeling of COVID-19 dynamics is that the information regarding a
particular community’s vulnerabilities and strengths pertaining to the
disease must be as comprehensive and integrative as possible.  The
regular reporting of accurate Coronavirus outbreaks data from local
health departments has been understandably difficult, especially in
the areas that are hardest hit.

%\linelabel{comment:level}

We use data on community-specific health and
behavioral risk factors aggregated at the county level from the
large-scale longitudinal CDC Behavioral Risk Factor Surveillance
System (BRFSS) surveys \citep{www-cdc-brfss} and demographic and
socioeconomic data from the U.S. Census Bureau \cite{www-sensus} for each county, along with the cases and
deaths data compiled by the CDC National Center for Health Statistics
(NCHS) \citep{www-cdc-nchs} and integrate them with other data
sources such as the American Hospital Directory \citep{www-ahd}.

\subsection{Data Collection and Processing}

Among the different sources of data on daily cases and deaths data
collected and aggregated at different spatial levels, especially for
the U.S., many are either freely available repositories such as GitHub
or accessible via specifically developed online resources such as the
New York Times~\citep{github-nytimes} and the Johns Hopkins
Coronavirus Resource Center~\citep{www-jh-covid19}.  For this study,
we generated two data sets based on the latter resource upon our
checking for the locations and times at which cumulative counts of
COVID-19 cases and deaths on any given day were no less than those on
the previous day. This led us to compile daily coronavirus cases and
deaths data for 345 U.S. counties from 32 states and the District of
Columbia and matched by five-digit Federal Information Processing
Standards (FIPS) code or county name to dynamic and static variables
from additional data sources.

The static data in this repository was collected from the U.S. Census
Bureau and the Centers for Disease Control and Prevention (CDC). The
different fields cover the domains of behavioral and health risk
factors, hospital capacity, and socioeconomic and demographic
conditions. Data on health outcomes, prevention measures, and
unhealthy behaviors were extracted from the CDC's 500 Cities Local
Data for Better Health program \citep{www-cdc-chronic-data} in the form
of crude or age-adjusted prevalence of conditions such as respiratory
disease, obesity or smoking. These values were aggregated from the
census tract to the county level using the first five digits of the
eleven-digit tract FIPS code. Because a single county may consist of
multiple individual cities, we include the list of all city labels
within each aggregate group to represent a greater metropolitan
area. 110 of such metropolitan areas that had more than 500 reported
cases of COVID-19 by April 15, 2020, were selected for this study.

Notably, a greater metropolitan area such as New York City could be
spread across more than one county, which can lead to the complexity
of aggregating counts and other variables. Time series data for the 5
individual counties (boroughs) in New York City were not available in
the Johns Hopkins CSSE (Center for Systems Science and Engineering)
data set. Rather, the total number of cases and deaths for the entire
city of New York are reported and assigned the FIPS code of New York
County (Manhattan). To ensure consistent geography across the static
and dynamic data, we compute the population-weighted sum or median of
each covariate over the assigned counties.

Demographic variables were gathered from the Census
QuickFacts~\citep{www-census} online resource using an automated web
scraping algorithm and cover relevant areas such as age, race, income,
and population density. Additional socioeconomic variables include the
Gini Index, which measures economic inequality and CDC Social
Vulnerability Index (SVI)~\citep{www-cdc-svi}. The SVI was created to
guide public health officials and disaster response efforts by
identifying the communities across the United States most likely to
need support during a crisis. Census tracts and corresponding counties
are ranked across 15 social factors, which are grouped into four
themes: socioeconomic status, household composition and disability,
minority status and language, and housing and transportation.

Lastly, the total number of general acute care, critical access, and
military hospitals within each county are included in the data. Such
variables included the number of relevant hospitals per county, and
the estimated number of beds (total known bed counts added to the
number of hospitals in the county with missing data times the average
number of beds per hospital in that state) per 1,000 people using the
American Hospital Directory~\citep{www-ahd}.
The list of covariates is presented in Table \ref{tab:risk-factors}.

\begin{table}[!hptb]
\caption{Covariate Risk factor abbreviations as used in this study.}
\label{tab:risk-factors}
\bigskip
\resizebox{0.90\textwidth}{!}{
\centering
\begin{tabular}{ll}
\toprule
Risk Factor & Description \\
\midrule
 NONE & No risk factor used\\
 ARTHRITIS 	  & Percent reported with arthritis \\
 BINGE 		  & Percent reported with binge drinking \\
 BPHIGH 	  & Percent reported with  high blood pressure\\
 BPMED 		  & Percent reported taking blood pressure medication\\
 CANCER 	  & Percent reported as cancer patients\\
 CASTHMA 	  & Percent reported with current asthma\\
 CHD 		  & Percent reported with coronary heart disease\\
 CHECKUP 	  & Percent reported with health checkup\\
 CHOLSCREEN   & Percent reported with cholesterol screening\\
 COPD 		  & Percent reported with chronic obstructive pulmonary disease\\
 CSMOKING 	  & Percent reported as currently smoking\\
 DIABETES 	  & Percent reported with diabetes\\
 ESTBEDS 	  & Number of estimated beds in hospitals\\
 HIGHCHOL 	  & Percent reported with high cholesterol\\
 INSURANCE 	  & Percent reported with insurance\\
 KIDNEY 	  & Percent reported with kidney disease\\
 LPA 		  & Percent reported with no leisure-time physical activity \\
 MHLTH 		  & Percent reported with not good mental health\\
 NBEDS 		  & Number of beds in hospitals \\
 NBEDS/1000   & Number of beds per 1000 people\\
 NHOSP 		  & Number of hospitals\\
 OBESITY 	  & Percent reported with  population that is obese\\
 PHLTH        & Percent reported with physical health issues\\
 PVI 		  & Pandemic vulnerability index \\
 STROKE 	  & Percent reported with stroke\\
 BLACK	  & Percent of Blacks in the population\\
 POVERTY & Percent of poverty in the population\\
 SENIOR  & Percent of seniors in the population\\
 SVI\_MINORITY 	  & Percent of social vulnerability index in the minority population\\
 SVI\_OVERALL 	  & Percent of social vulnerability index in the overall population\\
\bottomrule
\end{tabular}
}
\end{table}

\section{Analysis}

We have conducted a number of analysis efforts to evaluate the validity of our approach. We focus on the following questions:

\begin{question} \normalfont Can we apply deep learning strategies that self-learn the behavior of a model in order to provide future predictions? See Section \ref{sec:emperical}.
\label{q:1}
\end{question}

\begin{question} \normalfont Can we use real time data as input and apply deep learning strategies instead of models that self learn and provide future predictions? See Section \ref{sec:lstm-covariate}.
\label{q:2}
\end{question}

\begin{question} \normalfont Does the inclusion of covariates improve the prediction quality while including geospatial and socio-economic risk factors? See Section \ref{sec:lstm-covariate}.
\label{q:3}
\end{question}

\begin{question} \normalfont Are there some impact factors that influence the prediction more than others? See Section \ref{sec:lstm-covariate}.
\label{q:4}
\end{question}

\begin{comment}
\begin{question} \normalfont What impact do we see when we introduce many risk factors? See Section \ref{sec:lstm-many}.
\label{q:4}
\end{question}
\end{comment}

\subsection{Deep Learning Predictions of Model-based Empirical Fits} 
\label{sec:emperical}

To establish the feasibility of using LSTM, we have, in our
analysis, verified that through deep learning we can recreate the
predictions identified by a model-based approach using empirical
fits. In this comprehensive analysis, we obtain predictions through
deep learning while using data generated from epidemiological models
as discussed in \citep{marsland20-covid-paper}. The time series we used
was 100 days long, and a multi-layered LSTM
recurrent network was used. Our prediction approach differs by
learning not only from the demographics (fixed data for each city) and
time-dependent data but by integrating the population model for the
underlying prediction, as shown in Figure \ref{fig:magic-1}. Such model
predictive integration capability is important in any application with
multiple time scales. For example, this allows us to integrate
multi-scale time effects into the forecast, which could address the
combination of general forecasts and the next time steps of choice,
such as days, weeks, months, or longer. For this analysis, we used 37
of the 110 cities with reliable empirical fits (that we derived from
an epidemiological model and not deep learning
\citep{marsland20-covid-paper}) for the cases and deaths data up to
April 15, 2020. Our analysis identified a sophisticated single deep
learning time evolution operator that can describe these 37 separate
data sets, and smooth fitted data leads to very accurate deep learning
descriptions via LSTM as depicted in Figure \ref{fig:magic-1}
that are smaller than 1\% from the model predictions.

\begin{figure}[!h]
    \centering

    \begin{subfigure}{.49\textwidth}
      \centering
      \includegraphics[width=1.0\linewidth]{images/predict/Fig4a.pdf}  
      \caption{Prediction acuracy for deaths.}
      \label{fig:magic-a}
    \end{subfigure}
    \begin{subfigure}{0.49\textwidth}
      \centering
      \includegraphics[width=1.0\linewidth]{images/predict/Fig4b.pdf}        
      \caption{Prediction acuracy for cases.}
      \label{fig:magic-b}
    \end{subfigure}

    \caption{Cumulative empirical fits can be very accurately
      predicted with LSTM. The data is taken from 37 U.S. Cities. Our
      comparison showcases that given a model, we can replicate the
      prediction of such models using the model data over time. The
      x-axis represents the days since the first data was fed to the
      model.}
    \label{fig:magic-1}
\end{figure}

\subsection{Realtime-Data Covariate LSTM Prediction}
\label{sec:lstm-covariate}

In this section, we establish that using deep learning can not only be
applied to model predictions but also from real time data. As we want to
evaluate the influence of the risk factors on our prediction framework
using not only models but deep learning algorithms, we have devised a
parameter sweep framework based on LSTM that conducts the prediction
for 110 cities while also considering the risk factors as listed
in Table \ref{tab:risk-factors}. This framework can query any
combination of risk factors as well as not including any, in which
case it defaults to the traditional LSTM algorithm as introduced in
Section \ref{sec:lstm-theory}. Hence, we will answer Questions
\ref{q:2} and \ref{q:3}.

To answer these questions, we chose for our next analysis a date range
between 2020-02-01 and 2020-05-25. We start with a basic LSTM operator
with two layers of LSTM, and initial and final fully connected layers
\citep{Kadupitiya2020-zq}.  For LSTM, we have selected the activation
function based on Rectified Linear Units (RELU). For the last layer
sigmoid is chosen. A drop out rate of 0.2 is used, we use 16 input
nodes, and have a batch size of 110. The  
batch size is the subset size of the training sample (e.g. 110 out of 12540 for a 3 day input length) that is used to train the LSTM during its learning process. The maximum epoch is chosen to be
200. This is the number of steps after we interrupt the learning process. The number of nodes internally is 32. The maximum total number of
samples is 12540 when using 3 days as input length. We use a two-layer
LSTM, as shown in Table \ref{tab:model}. Within our system, we have
the ability to set a number of input days. For this analysis, we have
chosen three different input lengths, namely 3, 4, and 5 days. The
term ``None'' in the output shape means this dimension is
variable.

\begin{table}[!h]
    \caption{The deep learning parameters used for the model sweep
      creation in as exported by Keras using 1 risk factor. The
      activation function is RELU and the recurrent activation
      function is sigmoid.}
    \label{tab:model}
    \bigskip

    \settowidth{\rotheadsize}{Parameters\quad}

    \begin{footnotesize}
    \centering
    \resizebox{1.0\textwidth}{!}{ 
    \begin{tabular}{|ll|lll|ll|ll|}
    \toprule 
     \rothead{Layer} & \rothead{Type}   &
     \rothead{Output Shape} &  \rothead{Parameters for 0 risk factor}  
                  &  \rothead{Parameters for 1 risk factor} &
     \rothead{Output Shape} &  \rothead{Parameters for 33 risk factors} &
     \rothead{Output Shape} &  \rothead{Parameters for 37 risk factors}\\
    \midrule
    dense & Dense    & (None,1,32) & 96   & 128  &  (None,5,32) & 1152 & (None,5,32) & 1280 \\
    LSTM & LSTM      & (None,1,16) & 3136 & 3136 &  (None,5,16) & 3136 & (None,5,16) & 3136 \\
    LSTM 1 & LSTM   & (None,16)  & 2112  & 2112 &  (None,16)   & 2112 & (None,16)   & 2112 \\
    Dense 1 & Dense & (None,32)  &  544  & 544  &  (None,32)   &  544 & (None,32)   &  544 \\
    Dense 2 & Dense & (None,30)  &  990  & 990  &  (None,30)   &  990 & (None,30)   &  990 \\
    \bottomrule
    \end{tabular}
    } ~\\
    The LSTM contains a total of 6,910 trainable parameters in
    case we include one risk factor. In the case, we do not use any
    risk factor it is 6878, and when we include 33 risk factors, it is
    7934.
    \end{footnotesize}
\end{table}

We have run the training on all 110 cities from our data that
represented each FIPS while not including any risk factors (Question
\ref{q:2}) and running them with a single risk factor while iterating
over all risk factors (Question \ref{q:3}). We then summed up all
absolute errors for the predicted data points over the same time
period and repeated each experiment 10 times. We summarized all the
results in the form of a box-whisker diagram each for deaths and for
cases. We sorted the diagram in such a fashion that the risk factor
that has the model with the lowest error appears first. We preceded
the graph with the experiment labeled as ``NONE'' that which does not
use any risk factors to provide a comparison. The graphs are depicted
in Figures \ref{fig:box-cases} for cases, and \ref{fig:box-death} for
deaths.

\begin{figure}[!p]
    \centering
    \includegraphics[width=1.0\textwidth]{images/boxwhisker/boxplot_casesNEW2.pdf}
    \vspace{-1cm}
    \caption{Influence of risk factors on the accuracy of the
      prediction of COVID-19 cases. The y-axis represents the
      cumulative error over all input data for the cities.}
    \label{fig:box-cases}
    \bigskip

    \includegraphics[width=1.0\textwidth]{images/boxwhisker/boxplot_deathNEW2.pdf}
    \vspace{-1cm}
    \caption{Influence of risk factors on the accuracy of the
      prediction of COVID-19 death. The y-axis represents the
      cumulative error over all input data for the cities.}
    \label{fig:box-death}
\end{figure}

Further, the obtained sorted order of the best prediction models by
minimum error based on the risk factors for cases and deaths are as
follows:

\begin{description}

\item[Order for cases sorted by minimum model error:] NONE,
  POP\_DENSITY\_2010, PHLTH, INSURANCE, BLACK,
  HISPANICS, NHOSP, CHOLSCREEN, DIABETES, STROKE,
  CHD, CHECKUP, NBEDS, SVI\_OVERALL, KIDNEY, CASTHMA, BPMED,
  LPA, CSMOKING, ARTHRITIS, POVERTY, ESTBEDS, MHLTH, SENIOR, COPD,
  CANCER, NBEDS/1000, BPHIGH, HIGHCHOL, BINGE, OBESITY, SVI\_MINORITY

\item[Order for deaths sorted by minimum model error:] NONE,
  SVI\_OVERALL, CSMOKING, HISPANICS, POP\_DENSITY\_2010,
  SENIOR, CHOLSCREEN, CASTHMA, INSURANCE, MHLTH, LPA,
  POVERTY, OBESITY, CANCER, BLACK, NBEDS/1000, DIABETES, ARTHRITIS,
  NBEDS, NHOSP, BPHIGH, CHECKUP, PHLTH, CHD, STROKE, HIGHCHOL,
  ESTBEDS, KIDNEY, BINGE, BPMED, COPD, SVI\_MINORITY
\end{description}

We show the errors and the association with a risk factors in Table
 \ref{tab:top-1-cases} and \ref{tab:top-1-deaths}. In the tables we include the following columns (a) the Risk Factor (b) the root mean square error of all models for this risk factor (c) the minimum cumulative error by risk factor and (d) the number of days that wer used as input and identified to relate to the model with the minimum cumulative error. The table has in the first column the rank this particular model comperes to all other models with minimal risk factor. 
The rank was obtained by sorting the best model for each risk factor. 
We display two tables. One for cases (Table \ref{tab:top-1-cases} and one for deaths (Table \ref{tab:top-1-cases}).

We see that in our experiment, almost all risk factors lead to an increased
model prediction accuracy for cases.  We note that by including
factors such as population density, physical health Insurance,
population breakdown by ethnicity, number of hospitals, and diabetes,
lead to better overall predictions for cases. 
We also see that
many of these factors perform better on average.

\begin{table}[!p]
\caption{Best models for cases for each risk factor and their respective
  errors. The model that uses no risk factor is highlighted in grey. We present the Root Mean Square Error (RMSE), the cumulative error and how many days as input were used for this corresponding model.}
\label{tab:top-1-cases}
\bigskip
\centering
%\resizebox{1.0\textwidth}{!}{ 
\begin{tabular}{rrrrrlrl}
\toprule
 \makecell{Place/\\Rank} & 
\makecell{Risk Factor} &  
\makecell{RMSE\\Cases} &  
  \makecell[r]{Minimum\\Cumu-\\lative \\Error for \\Cases by\\Risk Factor} & 
 \makecell[r]{Number of \\Days as \\Input} \\
\midrule

1	&	POP\_DENSITY\_2010	&	0.055443	&	7.94	&	5	 \\
2	&	PHLTH	&	0.054795	&	8.01	&	3	 \\
3	&	BLACK	&	0.055028	&	8.12	&	4	 \\
4	&	HISPANIC	&	0.056061	&	8.22	&	5	 \\
5	&	NHOSP	&	0.054621	&	8.26	&	3	 \\
6	&	INSURANCE	&	0.054557	&	8.34	&	3	 \\
7	&	CHOLSCREEN	&	0.054712	&	8.37	&	3	 \\
8	&	DIABETES	&	0.054586	&	8.4	&	5	 \\
9	&	STROKE	&	0.05416	&	8.45	&	4	 \\
10	&	CHD	&	0.054027	&	8.46	&	5	 \\
11	&	CHECKUP	&	0.054927	&	8.51	&	5	 \\
12	&	NBEDS	&	0.054146	&	8.54	&	5	 \\
13	&	SVI\_OVERALL	&	0.054795	&	8.58	&	5	 \\
14	&	KIDNEY	&	0.0546	&	8.61	&	5	 \\
15	&	CASTHMA	&	0.055089	&	8.63	&	5	 \\
16	&	BPMED	&	0.054685	&	8.74	&	3	 \\
17	&	LPA	&	0.054323	&	8.75	&	5	 \\
18	&	CSMOKING	&	0.053716	&	8.75	&	5	 \\
19	&	ARTHRITIS	&	0.054053	&	8.83	&	3	 \\
20	&	POVERTY	&	0.055029	&	8.84	&	3	 \\
21	&	ESTBEDS	&	0.054545	&	8.86	&	4	 \\
22	&	MHLTH	&	0.054864	&	8.9	&	3	 \\
23	&	COPD	&	0.055589	&	8.93	&	4	 \\
24	&	SENIOR	&	0.055079	&	8.93	&	3	 \\
25	&	CANCER	&	0.0545	&	8.99	&	3	 \\
26	&	NBEDS/1000	&	0.054397	&	9.05	&	3	 \\
27	&	BPHIGH	&	0.054609	&	9.13	&	4	 \\
\rowcolor{black!5} 28	&	NONE	&	0.054868	&	9.26	&	3	 \\
29	&	HIGHCHOL	&	0.056429	&	9.28	&	3	 \\
30	&	BINGE	&	0.054317	&	9.32	&	5	 \\
31	&	OBESITY	&	0.054469	&	9.5	&	3	 \\
32	&	SVI\_MINORITY	&	0.055094	&	9.54	&	4	 \\
\bottomrule
\end{tabular}
%}
\end{table}

\begin{table}[!p]
\caption{Best models for deaths for each risk factor and their respective
  errors. The model that uses no risk factor is highlighted in grey.
  We present the Root Mean Square Error (RMSE), the cumulative error and how many days as input were used for this corresponding model.
  }
\label{tab:top-1-deaths}
\bigskip
\centering
%\resizebox{1.0\textwidth}{!}{ 
\begin{tabular}{rrrrrlrl}
\toprule
 \makecell{Place/\\Rank} & 
\makecell{Risk Factor} &  
\makecell{RMSE\\Deaths} &  
\makecell[r]{Minimum\\Cumu-\\lative \\Error for \\Deaths by\\Risk Factor} &  
 \makecell[r]{Number of \\Days as \\Input} \\
\midrule
1	&	SVI\_OVERALL	&	0.067457	&	18.75	&	5	 \\
2	&	CSMOKING	&	0.066898	&	19.01	&	5	 \\
3	&	HISPANIC	&	0.067489	&	19.04	&	5	 \\
\rowcolor{black!5} 4	&	NONE	&	0.06696	&	19.09	&	4	 \\
5	&	POP\_DENSITY\_2010	&	0.066968	&	19.13	&	4	 \\
6	&	SENIOR	&	0.068498	&	19.27	&	3	 \\
7	&	CHOLSCREEN	&	0.067859	&	19.43	&	5	 \\
8	&	CASTHMA	&	0.066952	&	19.46	&	5	 \\
9	&	INSURANCE	&	0.067127	&	19.48	&	5	 \\
10	&	MHLTH	&	0.068742	&	19.53	&	3	 \\
11	&	LPA	&	0.067149	&	19.54	&	5	 \\
12	&	POVERTY	&	0.067267	&	19.57	&	5	 \\
13	&	OBESITY	&	0.067123	&	19.68	&	5	 \\
14	&	CANCER	&	0.068363	&	19.70	&	3	 \\
15	&	NBEDS/1000	&	0.06752	&	19.73	&	5	 \\
16	&	BLACK	&	0.068352	&	19.83	&	3	 \\
17	&	DIABETES	&	0.067636	&	19.84	&	5	 \\
18	&	ARTHRITIS	&	0.066927	&	19.94	&	4	 \\
19	&	NBEDS	&	0.066879	&	19.95	&	4	 \\
20	&	NHOSP	&	0.067391	&	19.95	&	5	 \\
21	&	BPHIGH	&	0.067799	&	19.99	&	5	 \\
22	&	PHLTH	&	0.068737	&	20.02	&	3	 \\
23	&	CHECKUP	&	0.067543	&	20.02	&	5	 \\
24	&	CHD	&	0.06874	&	20.15	&	3	 \\
25	&	STROKE	&	0.066906	&	20.16	&	4	 \\
26	&	HIGHCHOL	&	0.068557	&	20.35	&	3	 \\
27	&	ESTBEDS	&	0.067052	&	20.42	&	4	 \\
28	&	KIDNEY	&	0.068948	&	20.49	&	3	 \\
29	&	BINGE	&	0.067447	&	20.56	&	5	 \\
30	&	BPMED	&	0.067399	&	20.69	&	5	 \\
31	&	COPD	&	0.06745	&	20.98	&	5	 \\
32	&	SVI\_MINORITY	&	0.067443	&	21.53	&	4	 \\
\bottomrule
\end{tabular}
%}
\end{table}

We observed a different scenario for deaths, in which we only
identified two risk factors, namely the Social Vulnerability Index
(SVI) and chain-smoking, that, when integrated into our deep learning
model, leads to better prediction results. However, the accuracy of
the overall prediction of deaths is far less precise than that of the
cases. To showcase this, we have included Figures
\ref{fig:error-case-popdensity} and \ref{fig:error-death-popdensity}
that show the respective errors.

\begin{figure}[!h]
  \begin{subfigure}{.49\textwidth}
    \centering
    \includegraphics[width=1.0\textwidth]{images/predict/Length_5-fields_pop_density_2010-Cumulative-Cases.pdf}
    %\vspace{-1cm}
    \caption{Model prediction error cases.}
    \label{fig:error-case-popdensity}
  \end{subfigure}
  \begin{subfigure}{.49\textwidth}
    \centering
    \includegraphics[width=1.0\textwidth]{images/predict/Length_5-fields_pop_density_2010-Cumulative-Death.pdf}
    %\vspace{-1cm}
    \caption{Model prediction error deaths.}
    \label{fig:error-death-popdensity}
    \end{subfigure}

    \caption{Model prediction error for cumulative deaths when
      including one risk factor POP\_DENSITY\_2010
      in addition to the past number of cases and deaths.}
    \label{fig:error-death-popdensity-both}

\end{figure}

This can be explained by our input data distribution between cases and deaths in which we noted higher
fluctuations relative to the overall value in cases of deaths. 
A mitigation to this issue
would be using a seven day average over the analyzed period. However,
we have not done this on purpose for this analysis as we wanted to
identify how a covariate enhanced LSTM behaves that includes risk
factors based on the daily fluctuations. This was done to avoid and
showcase any needed preproccessing on the data and to identify the
capability of the deep learning framework while adapting to the
fluctuations without any special activities. Previously, we have
already shown in Section \ref{sec:emperical} that for smooth data
inputs the prediction is very accurate.  Using fluctuation data allows
our framework for an automated ingest of data on a daily basis to
enable data fusion of new cases and death information. Hence, we can
re-train the model with new incoming data every day. Through repeated
training experiments, we found the LSTM hyperparameters such as epochs
and dropout values that work well for our experiments.

We conducted an additional analysis and plotted the top 10 best
cumulative prediction model errors for each risk factor in
Figure \label{fig:place-top10} for cases and deaths.  The horizontal
black line shows the first occurrence of no risk factor being used.

Notably, we identified that a jump occurs when we sort the top 10
model predictions at around 330 models from a total of 3400
automatically generated models. This also underscores the need to run
our deep learning model multiple times to obtain best parameter
candidates.  Furthermore, we see that the cumulative error for cases
is about half the size as for death. This is caused by cases having a
larger number of samples than those from deaths.

% In case of cases we also see a drop in the better
% model predictions.

\begin{figure}[!htbp]
%    \begin{minipage}{.45\textwidth}
%        \centering
%        \includegraphics[width=1.0\textwidth]{images/predict/PlaceTop1_Cases.pdf}
%        \vspace{-1cm}
%        \caption{Best predictions for cases over all risk factors.}
%        \label{fig:place-top1-cases}
%
%    \end{minipage}
%    \ \
%    \begin{minipage}{0.45\textwidth}
%        \centering
%        \includegraphics[width=1.0\textwidth]{images/predict/PlaceTop1_Death.pdf}
%        \vspace{-1cm}
%        \caption{Best predictions for deaths over all risk factors.}
%        \label{fig:place-top1-death}
%    \end{minipage}

%    \begin{minipage}{.45\textwidth}
%        \centering
%        \includegraphics[width=1.0\textwidth]{images/predict/PlaceTop5_Cases.pdf}
%        \vspace{-1cm}
%        \caption{Top 5 predictions from all cases over all risk factors.}
%        \label{fig:place-top5-cases}
%    \end{minipage}
%    \ \
%    \begin{minipage}{.45\textwidth}
%        \centering
%        \includegraphics[width=1.0\textwidth]{images/predict/PlaceTop5_Death.pdf}
%        \vspace{-1cm}
%        \caption{Top 5 predictions from all deaths  over all risk factors.}
%        \label{fig:place-top5-death}
%    \end{minipage}
    \begin{subfigure}{.95\textwidth}
        \centering
        \includegraphics[width=1.0\textwidth]{images/predict/PlaceTop10_Death.pdf}
        \caption{Cumulative error for deaths.}
        \label{fig:place-top10-death}
    \end{subfigure}
    \begin{subfigure}{.95\textwidth}
        \centering
        \includegraphics[width=1.0\textwidth]{images/predict/PlaceTop10_Cases.pdf}
        \caption{Cumulative error for cases}
        \label{fig:place-top10-cases}
    \end{subfigure}

    \caption{Top 10 predictions from all different risk factors sorted by cumulative error for deaths (\ref{fig:place-top10-death}) and cases (\ref{fig:place-top10-cases}). The horizontal line in each figure represents the model cumulative error with no risk factor as input for deaths and cases respectively.}
    \label{fig:place-top10-both}

\end{figure}

\begin{comment}

\subsection{Impact and Design of a Multi-Covariate-LSTM}
\label{sec:lstm-many}

In our next comprehensive analysis, we try to answer Question
\ref{q:4} if all risk factors impact the analysis. Thus we feed the
covariates into the recurrent network as additional features. In
addition to cases and deaths, this leads to 35 variables, including 33
unique covariates. We looked at two related ways of doing this: (a)
have covariates fixed in time (static), and (b) multiply the
covariates by the times series (we used the average of cases and
deaths)

We expected the second choice to be the most natural as it gives all
35 input streams similar time dependence. However, in our initial
tests, we found these choices gave very similar results, and so we
focused on the simpler choice (a) in this paper. In the experiments
discussed here, we either used 0, 1, or 33 covariates. Choosing 33
gives the best description (reducing loss by about 1\%) while using
each covariate as a solitary extra feature allows us a rather
sensitive study of the importance of each individual covariate as we
noted in Section \ref{sec:lstm-covariate}.

The final choice was the quantities to predict and train on. We choose
this to be both the cases and deaths values at the time value
following the sequence combined with the same quantities for the
following 14 days. This gives us 30 different output features for each
sequence. We renormalized, so the two basic features (following day)
contributed 50\% of the loss function. We supported this in TensorFlow
with a custom mean square error (MSE) loss function that dropped
training values that were not available for time sequences near the
end of the examined period. While one can not train on future values,
we can predict them, as seen in our sample outputs. We extensively
studied both 2 and 30 output features, and they give similar
descriptions of the observed data with the 30 feature choice enabling
14-day predictions.

We performed an extensive hyper-parameter search and we show the
values in Table \ref{tab:model} that give robust results without
overfitting. Furthermore, we looked at four different activation
functions for this analysis in the LSTM and fully connected
layers. Scaled Exponential Linear Unit (SELU), REctified Linear Unit
(RELU) and TANH activations gave very similar results while SIGMOID
activation was much worse. However we still use a SIGMOID function in
the last stage just as we did in Section \ref{sec:lstm-covariate}. We
also looked at the number of hidden units (separately for LSTM and
fully connected layers), the number of LSTM layers and the dropout
values.  Only the latter produced a significant change and, for
example, removing dropouts decreased the loss by about 16\% but we
retained them to provide a robust environment. We chose randomly 20\%
of the data for validation and testing and this had similar loss
values to the training data (the other 80\%) with our chosen dropout.

The final technical issue concerns the data normalization where each
feature is transformed by a feature dependent -- but city independent
-- way to lie between 0 and 1. Further all extensive (proportional to
size) covariates were first divided by the city population except for
the population itself where we took the logarithm; intensive
covariates were just rescaled to lie between 0 and 1. The basic
COVID-19 daily data is extensive but compared two approaches with and
without renormalization, but in this case, we took the square root of
the data value.  This is natural if there are counting (square root(N)
for N observations) errors as the deep learning loss function is
$(predicted-observed)^{2}$ without the error traditional in
$\chi^{2}$, namely $((predicted-observed)/error)^{2}$. This approach
with standard deep learning MSE appears sound instead of dividing by
the population. That will emphasize the error contribution of smaller
cities but is still interesting to look at due to the presence of many
systematic errors.

\section{Direct Deductions from Deep Learning Data Representation}

The deep learning model contains a lot of information much of which we
have already analyzed and discussed in Section
\ref{sec:lstm-covariate} and Figures \ref{fig:magic-1} and
\ref{fig:error-case-popdensity}. In addition, we have looked at the
sensitivity of the results as measured by the predicted daily rates
and the value of the loss function. In Table \ref{tab:corr-matrix}, we
present one way of looking at it. We calculate the correlation
coefficients of many of the covariates from an accurate LSTM model fit
to the data. We divide the time series into three regions

\begin{itemize}
\item \textbf{Past} or all days up to 14 days before the final
  observed date (May 25, 2020)
\item \textbf{Now} or the 14 days from May 12 to May 25, 2020
\item \textbf{Future} or the 14 days after May 25, 2020 predicted by
  the model
\end{itemize}

These results are different between population normalized or direct
non-normalized data. We have kept the square-root used in the model
but divided the model values by the square root of the population when
computing the correlation. Significant positive correlations are seen
for CASTHMA, HIGHCHOL, DIABETES, OBESITY, CSMOKING, CHD, and CHECKUP ,
while negative correlations for PVI and NORM\_POP. We also looked at
the impact of these covariates on the evolution function. As this
operates on data at previous time values, this is sensitive to
variations across cities and times. The evolution operator is
intuitively the time derivative of the absolute values of daily rates
seen in the correlation in Table 3. Nhosp and Estbeds have the most
significance in impacting (as a reduction) the values of both the
infection and fatality rates from the evolution operator. The fatality
rates are also significantly impacted by INSURANCE and NORM\_POP
(reduction) and SENIOR and OBESITY (increase) from the evolution
operator. NORM\_POP impacts infection values as an increase.  We can
also identify which covariates have the most significant impact on the
loss function. Here NHOSP, CHD, KIDNEY, BINGE, and PHLTH give 5\% or
more reduction in loss function while NORM\_POP, SVI\_OVERALL,
CASTHMA, DIABETES, and BPHIGH have small (1\%) effects.

\begin{table}[!h]
\caption{Correlation Matrix}
\bigskip
\label{tab:corr-matrix}
\begin{center}
\begin{tabular}{lrrrrrr}
\toprule
Risk Factor		&
\makecell{Past\\Case} &
\makecell{Now\\Case} &
\makecell{Future\\Case} &
\makecell{Past\\Death} &
\makecell{Now\\Death} &
\makecell{Future\\Death} \\
\midrule
PVI                     & -0.425      & -0.292      & -0.124      & -0.434      & -0.373      & -0.239    \\
INSURANCE               & 0.018       & 0.017       & 0.012       & 0.018       & -0.002      & 0.001     \\
NBEDS                   & 0.136       & 0.131       & 0.217       & 0.089       & 0.080        & 0.109     \\
NBEDS/1000              & 0.138       & 0.121       & 0.209       & 0.092       & 0.069       & 0.088     \\
NHOSP                   & 0.074       & 0.119       & 0.165       & -0.009      & 0.028       & 0.021     \\
ESTBEDS                 & 0.136       & 0.131       & 0.217       & 0.089       & 0.080        & 0.109     \\
SENIOR                  & 0.104       & 0.028       & 0.033       & 0.152       & 0.164       & 0.156     \\
BLACK                   & 0.209       & 0.099       & 0.208       & 0.216       & 0.085       & 0.116     \\
HISPANIC                & -0.072      & -0.032      & -0.064      & -0.091      & -0.068      & -0.046    \\
POP\_DENSITY\_2010      & -0.302      & -0.210      & -0.175      & -0.272      & -0.231      & -0.216    \\
POVERTY                 & 0.128       & 0.059       & 0.200       & 0.148       & 0.031       & 0.080     \\
SVI\_MINORITY           & 0.029       & 0.069       & 0.081       & 0.010        & 0.006       & 0.087     \\
SVI\_OVERALL            & 0.140       & 0.158       & 0.238       & 0.125       & 0.113       & 0.179     \\
CASTHMA                 & 0.238       & 0.271       & 0.307       & 0.205       & 0.288       & 0.309     \\
HIGHCHOL                & 0.202       & 0.185       & 0.131       & 0.192       & 0.243       & 0.224     \\
DIABETES                & 0.200       & 0.200     & 0.178         & 0.206       & 0.232       & 0.230     \\
OBESITY                 & 0.199       & 0.205       & 0.194       & 0.191       & 0.229       & 0.212     \\
CANCER                  & -0.145      & -0.102      & -0.129      & -0.138      & -0.058      & -0.059    \\
STROKE                  & 0.123       & 0.125       & 0.117       & 0.167       & 0.143       & 0.135     \\
MHLTH                   & 0.110       & 0.161       & 0.186       & 0.076       & 0.175       & 0.222     \\
CSMOKING                & 0.189       & 0.258       & 0.249       & 0.160        & 0.289       & 0.323     \\
CHOLSCREEN              & 0.164       & 0.001       & -0.026      & 0.187       & 0.070       & 0.038     \\
INSURANCE               & 0.018       & 0.017       & 0.012       & 0.018       & -0.002      & 0.001     \\
CHD                     & 0.225       & 0.284       & 0.233       & 0.178       & 0.327       & 0.337     \\
CHECKUP                 & 0.206       & 0.221       & 0.280       & 0.201       & 0.272       & 0.343     \\
KIDNEY                  & 0.045       & 0.028       & 0.032       & 0.022       & 0.071       & 0.053     \\
BINGE                   & 0.031       & 0.044       & 0.009       & 0.083       & 0.030       & 0.022     \\
LPA                     & 0.191       & 0.187       & 0.197       & 0.147       & 0.220       & 0.214     \\
ARTHRITIS               & 0.046       & 0.013       & -0.021      & 0.061       & 0.089       & 0.079     \\
BPMED                   & 0.060       & 0.080        & 0.124       & 0.056       & 0.073      & 0.090      \\
PHLTH                   & 0.064       & 0.132       & 0.104       & 0.022       & 0.129       & 0.114     \\
BPHIGH                  & 0.056       & 0.051       & 0.048       & 0.058       & 0.084       & 0.068     \\
COPD                    & 0.163       & 0.163       & 0.153       & 0.175       & 0.220       & 0.240     \\
NORM\_POP               & -0.305      & -0.258      & -0.218      & -0.193      & -0.218      & -0.099  \\
\bottomrule
\end{tabular}
\end{center}
\end{table}

\end{comment}

\section{Conclusion}

While this study focuses on if and which risk factors can improve the
predictions, we do not focus on why. Such analysis could provide
insights into causal relationships. For example, information such as
health insurance and the number of hospitals and beds are often
related to the quality of services that affect early detection and
treatment. Furthermore, population parameters such as population density can
determine the rate of spread and the number of cases. Similarly,
distributions of comorbidities in a given population such as physical
health, cholesterol screening, diabetes, etc., are found to be
insightful in predicting outcomes \citep{Maleki}. Identifying such risk
factors and correlating them to the prediction accuracy provides
valuable candidates for future studies.

A variety of mathematical and computational models have been used for
predicting COVID-19 outcomes \citep{Jewell}. It includes compartmental
models such as the SIR (Susceptible-Infected-Recovered) and SEIR
(Susceptible-Exposed-Infected-Recovered) which are relatively easy to
compute, and thus, commonly used in epidemiology. However, their
simplistic assumptions, such as homogeneous mixing within populations,
might make them less realistic for pandemics such as COVID-19. For
instance, such models often assume a closed population that does not
change. However, COVID-19 dynamics have prominently included massive
migration, deaths due to comorbidities temporarily conferred immunity,
and notably, major disparities in terms of socio-economic determinants
of health.

In terms of implementation, SIR and SEIR models are fitted with point
estimates, which may not adequately reflect the disease dynamics. For
instance, it is difficult to account for behavioral changes (e.g.,
human mobility) or systemic inadequacies (e.g., availability of
hospital beds) that may appear in specific subpopulations. Finally,
the compartmental model estimates do not allow for quantification of
uncertainty, which makes it harder to use them in policy-making. A
comparison of modeling approaches concluded that while SEIR model
performed better for a couple of states in the U.S., alternatives have
performed better for the other states \citep{Bertozzi}. AICov has
addressed most of the above modeling issues in its design, as per the
requirements of its forecasting objectives.

We present AICov as an integrative deep learning framework for
COVID-19 forecasting with the help of population covariates. Thus, its
objectives are different from those of traditional compartmental
models. One of the important features of the architecture for AICov is
that it is by design targeting Cloud and High Performance Computing
(HPC) resources to conduct parameter sweeps to leverage sophisticated
deep learning toolkits. The architecture allows the integration of
various data sources that can update the data from its sources on
demand and update its model predictions based on newly introduced
data. Parameters can easily be adjusted via Jupyter notebooks that, in
turn, call the computational backends on HPC and cloud resources.

In addition to this architecture, we used data collected from multiple
public sources and agencies, and integrated the same across spatially
contiguous units such as cities or metropolitan areas. In our
analysis, we have focused on 110 selected cities of the U.S., but the
framework is general and can be used to analyze similar data on
pandemics from anywhere in the world.

The comprehensive analysis showcases the feasibility of our
approach. Based on the outcomes reported it can lead to an improved
prediction once we integrate risk factors in addition to the
time-dependent data such as cases and death resulting from
COVID-19. While the improvements observed were modest, they do present
a concrete way of improving the forecast models. Inclusion of further
putative factors from the ongoing worldwide studies on Covid-19 will
only strengthen the future applications of our integrative framework.

We have shown that deep learning can return very good results while
using smooth data, which we used in our empirical fits. For real data
as presented to us for the daily changes, it still produced good
results and even was able to predict changes based on weekly
fluctuations. This is typically not achieved by other non-data based
model approaches. Also, we have used in our data only cases and
deaths, but we intend to expand this by using data about recovery and,
in the future, immunization.

We have experimented with different hyperparameters and included in
this study a selection of hyperparameters that have worked well for
this data set.

As we have set up the first version of our AICov software, we have
used cloud and high-performance computers. All of our sophisticated
analyses were run in a day on at most 16 computers. All deep learning
algorithms that were run on GPUs were run on NVIDIA Tesla
K80s. However, our framework is on purpose generalized so that other
compute resources can be integrated and leveraged. This includes
local, HPC, and cloud computing resources, as well as different
GPUs. Naturally, we also need to take into consideration not only the
compute time but also the time it takes for the researcher to derive
the models. As our approach is self-learning based on automated data
updates, the time to prepare an updated model can be automatized and
minimal input is needed. Thus the overall effort is very competitive.

\begin{comment}
Furthermore, we are currently conducting experiments with multiple
risk factors at the same time to identify the most significant
combinations of them.  As we have shown that the bivariate analysis
has a locality dependency, we will explore categorizing locations with
similar demographics and conduct further experiments on demographical
similar regions.
\end{comment}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%
% SUPLEMENTAL
%

\section*{Supplementary Materials}

The code and paper document represented to implement AICov are contained in several repositories:

\begin{enumerate}
    \item The entire cloudmesh code on which the cloud based implementation of the AICov framework is based and contains over 70 contributors is available publicly at \url{https://github.com/cloudmesh}. Cloudmesh contains a number of modules that dependent on the users access to cloud resources can be customized. A detailed manual about the configuration is available at \url{https://cloudmesh.github.io/cloudmesh-manual/}.

    \item The entire COVID-19 analysis leverages cloudmesh and uses Jupyter notebooks to coordinate its workflow as discussed in the architecture Figure \ref{fig:arch}. The code and data for the results presented in this paper are located in the repository at \url{https://github.com/cloudmesh/cloudmesh-covid}.

    The data was analysed on a variety of supercomputing resources including an allocation of 20 compute nodes that were utilized to execute the  repeated model creation to assure reproducible results.

    However, as a publication is pending, at this time, the use of the data is copyrighted and can not be used for other publications without contacting the authors. The data gathering and analysis is a significant intellectual contribution and we like to avoid that the data is taken before we have not secured a publication.

    \item The entire paper is located in \LaTeX source in the github repository \url{https://github.com/cyberaide/paper-covid}. This repository will be open sourced after acceptance of publication to not violate any publisher restrictions. If desired the authors can grant access to this repository prior to publication. Please contact the corresponding author.

\end{enumerate}

%
% ACKNOWLEDGMENT
%

\section*{Acknowledgements}

This work is partially supported by the National Science Foundation
(NSF) through awards Cyberinfrastructure Framework for 21st Century
Data Infrastructure Building Blocks (1443054), Network for
Computational Nanotechnology Engineered nanoBIO Node (1720625),
Cybertraining (1829704), CyberInfrastructure for Network Engineering
and Science (1835598) and Global Pervasive Computational Epidemiology
(1918626).  We thank J. Kadupitiya for several useful discussions on
deep learning frameworks.

\clearpage

\section*{Acronyms}

This section contains the list of acronyms in Table \ref{tab:acronyms}
as used in the paper for easy reference. The definitions of the risk
factors are however given within the paper in
Table~\ref{tab:risk-factors} on page \pageref{tab:risk-factors} and
not repeated here.

\begin{table}[!h]
  \caption{Non-risk factor acronyms used in this paper.}
  \label{tab:acronyms}
  \begin{tabular}{ll}
    \toprule
    Abbreviation & Description \\
    \midrule
    API & Application Programming Interfaces \\
    BRFSS & Behavioral Risk Factor Surveillance System \\
    CDC & U.S.~Centers for Disease Control and Prevention \\
    CI & Continious INtegration \\
    COVID-19 & COrona VIrus Disease 2019  \\
    CSSE & Center for Systems Science and Engineering \\
    cum & cumulative \\
    FIPS & Federal Information Processing Standards \\
    GPUs & Graphics Processing Units  \\
    HPC & High Performance Computing \\
    LSTM & Long Short-Term Memory \\
    MSE & Mean Squared Error \\
    NBDIF & NIST Big Data Working Group \\
    NCHS & CDC National Center for Health Statistics \\
    NIST & National Institute of Standards and Technology \\
    NONE & No risk factors used \\
    NORM\_POP & Normalized Population \\
    NSF & National Science Foundation \\
    RELU & REctified Linear Unit \\
    REST & REpresentational State Transfer  \\
    RMSE & Root Mean Squared Error \\
    RNN & Recurrent Neural Network \\
    SEIR & Susceptible, Expose, Infectious, Recovered) \\
    SELU & Scaled Exponential Linear Unit \\
    SIR &  Susceptible, Infectious, Recovered \\
    U.S. & United States of America \\
    WHO & World Health Organization \\
    \midrule
  \end{tabular}
\end{table}

\clearpage

\bibliographystyle{jdsurl}
\bibliography{bib/data-covid19,bib/lstm,bib/cloudmesh}

\end{document}

### 2021-01-27 final/response.txt
add final from draft

--- 2020-12-18 response-1.tex
@@
-\documentclass{article}
-\usepackage{fullpage}
-\usepackage[utf8]{inputenc}
+Review Response JDS2007-001RA0
 
-\title{Review Response JDS2007-001RA0}
+Revision due: 2020-11-24
 
-\date{Revision due: 2020-11-24}
-
-\begin{document}
-
-\maketitle
-
-\parindent 0pt We are happy to resubmit our paper and are very thankful for the great reviews we received. Thanks to the review comments, the quality of the paper has significantly improved.
+We are happy to resubmit our paper and are very thankful for the great reviews we received. Thanks to the review comments, the quality of the paper has significantly improved.
 
@@
 
-\begin{enumerate}
+*  The paper has been checked multiple times by multiple native English speakers.
 
-\item The paper has been checked multiple times by multiple native English speakers.
-
-\item The organization of the paper has been improved while focusing entirely on the deep learning framework.
+*  The organization of the paper has been improved while focusing entirely on the deep learning framework.
 Several sections and images that distracted the reader from the paper's primary focus (which was evident by the careful reviews) were removed.
 
-\item Reviewer 1: The method has been described using a recipe approach without giving the rationality behind it.
+*  Reviewer 1: The method has been described using a recipe approach without giving the rationality behind it.
 There are too many predictive models for COVID-19.
@@
 
-\item Reviewer 1: First, compared with the very classical SIR/SEIR and their improved models, what are the advantages of the proposed model?
+*  Reviewer 1: First, compared with the very classical SIR/SEIR and their improved models, what are the advantages of the proposed model?
 
@@
 
-\item Reviewer 1: Is the accuracy improved significantly?
+*  Reviewer 1: Is the accuracy improved significantly?
 
@@
 
-\item Reviewer 1: Second, it seems that the data used are all very structured, what about regression model framework?
+*  Reviewer 1: Second, it seems that the data used are all very structured, what about regression model framework?
 
@@
 
-\item Reviewer 1: Is the performance competitive?
+*  Reviewer 1: Is the performance competitive?
 
@@
 
-\item Reviewer 1: The current deep learning framework is lack of interpretation of the investigated risk factors.
+*  Reviewer 1: The current deep learning framework is lack of interpretation of the investigated risk factors.
 
@@
 
-\item Reviewer 2: The manuscript is in a rather preliminary shape, and needs significant work before being in a publishable shape.
+*  Reviewer 2: The manuscript is in a rather preliminary shape, and needs significant work before being in a publishable shape.
 
@@
 
-\item Reviewer 2: Section 2.2 does not look sufficient as a subsection.  It should be either expanded to include more details, or combined with Section 2.1.3.
+*  Reviewer 2: Section 2.2 does not look sufficient as a subsection.  It should be either expanded to include more details, or combined with Section 2.1.3.
 
@@
 
-\item Page 5, under "Data Source Integration", what does "we include integrate data integration capabilities" mean?
+* Page 5, under "Data Source Integration", what does "we include integrate data integration capabilities" mean?
 
@@
 
-\item Reviewer 2: 4.  Please consider writing the requirements on Pages 5 and 6 into structured paragraphs.
+* Reviewer 2: 4.  Please consider writing the requirements on Pages 5 and 6 into structured paragraphs.
 
@@
 
-\item Reviewer 2: Item 5., 6., 7., 8., 9., and 10. Have been fixed by excluding the text related to them from the paper as we found they distracted too much from our main contributions.
+* Reviewer 2: Item 5., 6., 7., 8., 9., and 10. Have been fixed by excluding the text related to them from the paper as we found they distracted too much from our main contributions.
 
-\item Reviewer 2: 12.  In Section 2.1,  the activation function is the sigmoid function.  In Section 6, RELU was used.
+* Reviewer 2: 12.  In Section 2.1,  the activation function is the sigmoid function.  In Section 6, RELU was used.
 
@@
 
-\item Reviewer 2: 13.  Please consider adding labels for the yaxes in Figure 10 and Figure 11.
+* Reviewer 2: 13.  Please consider adding labels for the yaxes in Figure 10 and Figure 11.
 
@@
 
-\item Reviewer 2: 14.  Page 19, please fix the broken link to reference.
+* Reviewer 2: 14.  Page 19, please fix the broken link to reference.
 
@@
 
-\item Reviewer 2: 15.  Sections 7 and 8 are rather redundant.  The language does not read clear.  I suggest moving these two sections, and make Section 5 and 6 a more complete story.
+*  Reviewer 2: 15.  Sections 7 and 8 are rather redundant.  The language does not read clear.  I suggest moving these two sections, and make Section 5 and 6 a more complete story.
 
@@
 
-\item Reviewer 2: 16.  Please consider making the code publicly available at GitHub.
+* Reviewer 2: 16.  Please consider making the code publicly available at GitHub.
 
     We intend to make the code public after publication. More than 30K lines of underlying infrastructure code and documentation is already available on GitHub and citations have been added.
-
-\end{enumerate}
-
-\end{document}

### 2021-01-27 final/vonLaszewski-covid.tex
add final from draft

--- 2021-01-27 vonLaszewski-covid.tex
@@
 With 
-$D$ being the number of days as input features, and
-$H$ being the number of hidden layers. 
-The inputs and outputs are defined by  
-$x_{t} \in \RD$ the input vector to the LSTM unit, and 
+$D$ being the number of days as input features and
+$H$ being the number of hidden layers, the 
+inputs and outputs are defined by  
+$x_{t} \in \RD$ the input vector to the LSTM unit and 
 $h_{t} \in \RH$ the hidden state vector e.g., the last of which is the output vector of the LSTM. In our case the output contains the days we predict based on the input features. The equations are

### 2021-01-27 final/vonLaszewski-covid.tex
start cleanup of final

@@
-%\documentclass[letterpaper, inpress]{jds} % use this for production
 \documentclass[a4paper, inpress]{jds} % use this for production
-%\documentclass[a4paper, review]{jds}      % use this for review
-
-% spacing
-% outcomment this for production
-\usepackage{setspace}
-\usepackage{lineno}
-\linenumbers
-\doublespace
-% above needs to be commented out for production
 
@@
 \usepackage{calc}
-\usepackage{comment}
 \usepackage{todonotes}
-\setuptodonotes{inline}
 
@@
 \jdsdoi{xx.xxxx/xxxxxxxxx}      % doi
-\jdsreceived{Jun 30, 2020}       % optional: comment it out if no received date
-\jdsaccepted{???, 2021}         % optional: comment it out if no accepted date
+\jdsreceived{June 30, 2020}       % optional: comment it out if no received date
+\jdsaccepted{January 23, 2021}         % optional: comment it out if no accepted date
 %% manually set running header for a shorter list of authors if needed
@@
 the areas that are hardest hit.
-
-%\linelabel{comment:level}
 
@@
 
-\begin{comment}
-\begin{question} \normalfont What impact do we see when we introduce many risk factors? See Section \ref{sec:lstm-many}.
-\label{q:4}
-\end{question}
-\end{comment}
-
 \subsection{Deep Learning Predictions of Model-based Empirical Fits} 
@@
 \end{figure}
-
-\begin{comment}
-Further, the obtained sorted order of the best prediction models by
-minimum error based on the risk factors for cases and deaths are as
-follows:
-
-\begin{description}
-
-\item[Order for cases sorted by minimum model error:] 
-  POP\_DENSITY\_2010, PHLTH, BLACK, HISPANIC, NHOSP, INSURANCE, 
-  CHOLSCREEN, DIABETES, STROKE, CHD, CHECKUP, NBEDS, SVI\_OVERALL, 
-  KIDNEY, CASTHMA, BPMED, LPA, CSMOKING, ARTHRITIS, POVERTY, ESTBEDS, 
-  MHLTH, COPD, SENIOR, CANCER, NBEDS/1000, BPHIGH, NONE, HIGHCHOL, 
-  BINGE, OBESITY, SVI\_MINORITY
-
-\item[Order for deaths sorted by minimum model error:] 
-  SVI\_OVERALL, CSMOKING, HISPANIC, NONE, POP\_DENSITY\_2010, 
-  SENIOR, CHOLSCREEN, CASTHMA, INSURANCE, MHLTH, LPA, 
-  POVERTY, OBESITY, CANCER, NBEDS/1000, BLACK, DIABETES, ARTHRITIS, 
-  NBEDS, NHOSP, BPHIGH, PHLTH, CHECKUP, CHD, STROKE, HIGHCHOL, 
-  ESTBEDS, KIDNEY, BINGE, BPMED, COPD, SVI\_MINORITY
-
-\end{description}
-\end{comment}
 
@@
 \begin{figure}[!htbp]
-%    \begin{minipage}{.45\textwidth}
-%        \centering
-%        \includegraphics[width=1.0\textwidth]{images/predict/PlaceTop1_Cases.pdf}
-%        \vspace{-1cm}
-%        \caption{Best predictions for cases over all risk factors.}
-%        \label{fig:place-top1-cases}
-%
-%    \end{minipage}
-%    \ \
-%    \begin{minipage}{0.45\textwidth}
-%        \centering
-%        \includegraphics[width=1.0\textwidth]{images/predict/PlaceTop1_Death.pdf}
-%        \vspace{-1cm}
-%        \caption{Best predictions for deaths over all risk factors.}
-%        \label{fig:place-top1-death}
-%    \end{minipage}
-
-%    \begin{minipage}{.45\textwidth}
-%        \centering
-%        \includegraphics[width=1.0\textwidth]{images/predict/PlaceTop5_Cases.pdf}
-%        \vspace{-1cm}
-%        \caption{Top 5 predictions from all cases over all risk factors.}
-%        \label{fig:place-top5-cases}
-%    \end{minipage}
-%    \ \
-%    \begin{minipage}{.45\textwidth}
-%        \centering
-%        \includegraphics[width=1.0\textwidth]{images/predict/PlaceTop5_Death.pdf}
-%        \vspace{-1cm}
-%        \caption{Top 5 predictions from all deaths  over all risk factors.}
-%        \label{fig:place-top5-death}
-%    \end{minipage}
     \begin{subfigure}{.95\textwidth}
@@
 \end{figure}
-
-\begin{comment}
-
-\subsection{Impact and Design of a Multi-Covariate-LSTM}
-\label{sec:lstm-many}
-
-In our next comprehensive analysis, we try to answer Question
-\ref{q:4} if all risk factors impact the analysis. Thus we feed the
-covariates into the recurrent network as additional features. In
-addition to cases and deaths, this leads to 35 variables, including 33
-unique covariates. We looked at two related ways of doing this: (a)
-have covariates fixed in time (static), and (b) multiply the
-covariates by the times series (we used the average of cases and
-deaths)
-
-We expected the second choice to be the most natural as it gives all
-35 input streams similar time dependence. However, in our initial
-tests, we found these choices gave very similar results, and so we
-focused on the simpler choice (a) in this paper. In the experiments
-discussed here, we either used 0, 1, or 33 covariates. Choosing 33
-gives the best description (reducing loss by about 1\%) while using
-each covariate as a solitary extra feature allows us a rather
-sensitive study of the importance of each individual covariate as we
-noted in Section \ref{sec:lstm-covariate}.
-
-The final choice was the quantities to predict and train on. We choose
-this to be both the cases and deaths values at the time value
-following the sequence combined with the same quantities for the
-following 14 days. This gives us 30 different output features for each
-sequence. We renormalized, so the two basic features (following day)
-contributed 50\% of the loss function. We supported this in TensorFlow
-with a custom mean square error (MSE) loss function that dropped
-training values that were not available for time sequences near the
-end of the examined period. While one can not train on future values,
-we can predict them, as seen in our sample outputs. We extensively
-studied both 2 and 30 output features, and they give similar
-descriptions of the observed data with the 30 feature choice enabling
-14-day predictions.
-
-We performed an extensive hyper-parameter search and we show the
-values in Table \ref{tab:model} that give robust results without
-overfitting. Furthermore, we looked at four different activation
-functions for this analysis in the LSTM and fully connected
-layers. Scaled Exponential Linear Unit (SELU), REctified Linear Unit
-(RELU) and TANH activations gave very similar results while SIGMOID
-activation was much worse. However we still use a SIGMOID function in
-the last stage just as we did in Section \ref{sec:lstm-covariate}. We
-also looked at the number of hidden units (separately for LSTM and
-fully connected layers), the number of LSTM layers and the dropout
-values.  Only the latter produced a significant change and, for
-example, removing dropouts decreased the loss by about 16\% but we
-retained them to provide a robust environment. We chose randomly 20\%
-of the data for validation and testing and this had similar loss
-values to the training data (the other 80\%) with our chosen dropout.
-
-The final technical issue concerns the data normalization where each
-feature is transformed by a feature dependent -- but city independent
--- way to lie between 0 and 1. Further all extensive (proportional to
-size) covariates were first divided by the city population except for
-the population itself where we took the logarithm; intensive
-covariates were just rescaled to lie between 0 and 1. The basic
-COVID-19 daily data is extensive but compared two approaches with and
-without renormalization, but in this case, we took the square root of
-the data value.  This is natural if there are counting (square root(N)
-for N observations) errors as the deep learning loss function is
-$(predicted-observed)^{2}$ without the error traditional in
-$\chi^{2}$, namely $((predicted-observed)/error)^{2}$. This approach
-with standard deep learning MSE appears sound instead of dividing by
-the population. That will emphasize the error contribution of smaller
-cities but is still interesting to look at due to the presence of many
-systematic errors.
-
-\section{Direct Deductions from Deep Learning Data Representation}
-
-The deep learning model contains a lot of information much of which we
-have already analyzed and discussed in Section
-\ref{sec:lstm-covariate} and Figures \ref{fig:magic-1} and
-\ref{fig:error-case-popdensity}. In addition, we have looked at the
-sensitivity of the results as measured by the predicted daily rates
-and the value of the loss function. In Table \ref{tab:corr-matrix}, we
-present one way of looking at it. We calculate the correlation
-coefficients of many of the covariates from an accurate LSTM model fit
-to the data. We divide the time series into three regions
-
-\begin{itemize}
-\item \textbf{Past} or all days up to 14 days before the final
-  observed date (May 25, 2020)
-\item \textbf{Now} or the 14 days from May 12 to May 25, 2020
-\item \textbf{Future} or the 14 days after May 25, 2020 predicted by
-  the model
-\end{itemize}
-
-These results are different between population normalized or direct
-non-normalized data. We have kept the square-root used in the model
-but divided the model values by the square root of the population when
-computing the correlation. Significant positive correlations are seen
-for CASTHMA, HIGHCHOL, DIABETES, OBESITY, CSMOKING, CHD, and CHECKUP ,
-while negative correlations for PVI and NORM\_POP. We also looked at
-the impact of these covariates on the evolution function. As this
-operates on data at previous time values, this is sensitive to
-variations across cities and times. The evolution operator is
-intuitively the time derivative of the absolute values of daily rates
-seen in the correlation in Table 3. Nhosp and Estbeds have the most
-significance in impacting (as a reduction) the values of both the
-infection and fatality rates from the evolution operator. The fatality
-rates are also significantly impacted by INSURANCE and NORM\_POP
-(reduction) and SENIOR and OBESITY (increase) from the evolution
-operator. NORM\_POP impacts infection values as an increase.  We can
-also identify which covariates have the most significant impact on the
-loss function. Here NHOSP, CHD, KIDNEY, BINGE, and PHLTH give 5\% or
-more reduction in loss function while NORM\_POP, SVI\_OVERALL,
-CASTHMA, DIABETES, and BPHIGH have small (1\%) effects.
-
-\begin{table}[!h]
-\caption{Correlation Matrix}
-\bigskip
-\label{tab:corr-matrix}
-\begin{center}
-\begin{tabular}{lrrrrrr}
-\toprule
-Risk Factor		&
-\makecell{Past\\Case} &
-\makecell{Now\\Case} &
-\makecell{Future\\Case} &
-\makecell{Past\\Death} &
-\makecell{Now\\Death} &
-\makecell{Future\\Death} \\
-\midrule
-PVI                     & -0.425      & -0.292      & -0.124      & -0.434      & -0.373      & -0.239    \\
-INSURANCE               & 0.018       & 0.017       & 0.012       & 0.018       & -0.002      & 0.001     \\
-NBEDS                   & 0.136       & 0.131       & 0.217       & 0.089       & 0.080        & 0.109     \\
-NBEDS/1000              & 0.138       & 0.121       & 0.209       & 0.092       & 0.069       & 0.088     \\
-NHOSP                   & 0.074       & 0.119       & 0.165       & -0.009      & 0.028       & 0.021     \\
-ESTBEDS                 & 0.136       & 0.131       & 0.217       & 0.089       & 0.080        & 0.109     \\
-SENIOR                  & 0.104       & 0.028       & 0.033       & 0.152       & 0.164       & 0.156     \\
-BLACK                   & 0.209       & 0.099       & 0.208       & 0.216       & 0.085       & 0.116     \\
-HISPANIC                & -0.072      & -0.032      & -0.064      & -0.091      & -0.068      & -0.046    \\
-POP\_DENSITY\_2010      & -0.302      & -0.210      & -0.175      & -0.272      & -0.231      & -0.216    \\
-POVERTY                 & 0.128       & 0.059       & 0.200       & 0.148       & 0.031       & 0.080     \\
-SVI\_MINORITY           & 0.029       & 0.069       & 0.081       & 0.010        & 0.006       & 0.087     \\
-SVI\_OVERALL            & 0.140       & 0.158       & 0.238       & 0.125       & 0.113       & 0.179     \\
-CASTHMA                 & 0.238       & 0.271       & 0.307       & 0.205       & 0.288       & 0.309     \\
-HIGHCHOL                & 0.202       & 0.185       & 0.131       & 0.192       & 0.243       & 0.224     \\
-DIABETES                & 0.200       & 0.200     & 0.178         & 0.206       & 0.232       & 0.230     \\
-OBESITY                 & 0.199       & 0.205       & 0.194       & 0.191       & 0.229       & 0.212     \\
-CANCER                  & -0.145      & -0.102      & -0.129      & -0.138      & -0.058      & -0.059    \\
-STROKE                  & 0.123       & 0.125       & 0.117       & 0.167       & 0.143       & 0.135     \\
-MHLTH                   & 0.110       & 0.161       & 0.186       & 0.076       & 0.175       & 0.222     \\
-CSMOKING                & 0.189       & 0.258       & 0.249       & 0.160        & 0.289       & 0.323     \\
-CHOLSCREEN              & 0.164       & 0.001       & -0.026      & 0.187       & 0.070       & 0.038     \\
-INSURANCE               & 0.018       & 0.017       & 0.012       & 0.018       & -0.002      & 0.001     \\
-CHD                     & 0.225       & 0.284       & 0.233       & 0.178       & 0.327       & 0.337     \\
-CHECKUP                 & 0.206       & 0.221       & 0.280       & 0.201       & 0.272       & 0.343     \\
-KIDNEY                  & 0.045       & 0.028       & 0.032       & 0.022       & 0.071       & 0.053     \\
-BINGE                   & 0.031       & 0.044       & 0.009       & 0.083       & 0.030       & 0.022     \\
-LPA                     & 0.191       & 0.187       & 0.197       & 0.147       & 0.220       & 0.214     \\
-ARTHRITIS               & 0.046       & 0.013       & -0.021      & 0.061       & 0.089       & 0.079     \\
-BPMED                   & 0.060       & 0.080        & 0.124       & 0.056       & 0.073      & 0.090      \\
-PHLTH                   & 0.064       & 0.132       & 0.104       & 0.022       & 0.129       & 0.114     \\
-BPHIGH                  & 0.056       & 0.051       & 0.048       & 0.058       & 0.084       & 0.068     \\
-COPD                    & 0.163       & 0.163       & 0.153       & 0.175       & 0.220       & 0.240     \\
-NORM\_POP               & -0.305      & -0.258      & -0.218      & -0.193      & -0.218      & -0.099  \\
-\bottomrule
-\end{tabular}
-\end{center}
-\end{table}
-
-\end{comment}
 
@@
 minimal input is needed. Thus the overall effort is very competitive.
-
-\begin{comment}
-Furthermore, we are currently conducting experiments with multiple
-risk factors at the same time to identify the most significant
-combinations of them.  As we have shown that the bivariate analysis
-has a locality dependency, we will explore categorizing locations with
-similar demographics and conduct further experiments on demographical
-similar regions.
-\end{comment}
 

### 2021-01-27 final/vonLaszewski-covid.tex
80 char formatting

@@
 disease, non-pharmaceutical policy interventions such as social
-distancing and lockdowns are recommended by health experts to prevent further transmission. To gain insights into the possible impact of
+distancing and lockdowns are recommended by health experts to prevent
+further transmission. To gain insights into the possible impact of
 such measures on COVID-19 outcomes, we depend upon the ability to
@@
 learning systems to understand healthcare trends, model risk
-associations, and predict outcomes \citep{Ting2020-lw}.  In addition to
-traditional public health surveillance strategies available to most
+associations, and predict outcomes \citep{Ting2020-lw}.  In addition
+to traditional public health surveillance strategies available to most
 countries, a variety of static and dynamic data types may be
 integrated to model the scale and dynamics of this pandemic
-\citep{Pyne2015-ao}. Several organizations, including the Johns Hopkins University’s Center for Systems Science and Engineering, the New York
-Times, the Atlantic’s COVID-19 tracking project, among others have
-developed real-time tracking maps for following COVID-19 cases around
-the world using data from the U.S.~Centers for Disease Control and
-Prevention (CDC), WHO, and other international health
+\citep{Pyne2015-ao}. Several organizations, including the Johns
+Hopkins University’s Center for Systems Science and Engineering, the
+New York Times, the Atlantic’s COVID-19 tracking project, among others
+have developed real-time tracking maps for following COVID-19 cases
+around the world using data from the U.S.~Centers for Disease Control
+and Prevention (CDC), WHO, and other international health
 agencies. Notably, the CDC receives forecasts from several modeling
@@
 (Susceptible-Exposed-Infectious-Recovered) to Agent-based to Bayesian
-models to understand the non-pharmaceutical policy interventions (or the lack thereof) to predict disease dynamics and its impact on human
+models to understand the non-pharmaceutical policy interventions (or
+the lack thereof) to predict disease dynamics and its impact on human
 lives \citep{www-cdc-modeling-forecast}.
@@
 combine community-specific health and behavioral risk factors,
-demographic and socioeconomic variables that are available for the corresponding county along with the cases and deaths data for COVID-19 while
-leveraging the recently developed and increasingly popular deep
+demographic and socioeconomic variables that are available for the
+corresponding county along with the cases and deaths data for COVID-19
+while leveraging the recently developed and increasingly popular deep
 learning strategies to forecast COVID-19 dynamics using our
@@
 
-In this study, we base our work on a traditional LSTM algorithm for our predictions
-\citep{www-keras-lstm,Hochreiter1997-dk}. An LSTM is a Recurrent
-Neural Network (RNN) \citep{Rumelhart1986-li} with feedback
+In this study, we base our work on a traditional LSTM algorithm for
+our predictions \citep{www-keras-lstm,Hochreiter1997-dk}. An LSTM is a
+Recurrent Neural Network (RNN) \citep{Rumelhart1986-li} with feedback
 connections allowing the use of data input sequences to predict data
@@
 reused in a prediction process between input and output sequences.
-The LSTM network will have a number of input values represented as vector x and a number of output values in the last layer of the network predicting a number of future values \citep{lstm-explained}.
-For our experiments this includes the numbers of days we use as input to our prediction and the number of days we use as output of our prediction for a single cell.
-In our implementation the number for inputs and outputs can be adjusted.
-We repeat this prediction over all input sequences to create the output forecast sequences.
-Next we list the equations (\ref{eq:first}) - (\ref{eq:last}) used in LSTM.
-We use in the equations $\circ$ as the Hadamard product,
+The LSTM network will have a number of input values represented as
+vector x and a number of output values in the last layer of the
+network predicting a number of future values
+\citep{lstm-explained}. For our experiments this includes the numbers
+of days we use as input to our prediction and the number of days we
+use as output of our prediction for a single cell. In our
+implementation the number for inputs and outputs can be adjusted. We
+repeat this prediction over all input sequences to create the output
+forecast sequences. Next we list the equations (\ref{eq:first}) -
+(\ref{eq:last}) used in LSTM.  We use in the equations $\circ$ as the
+Hadamard product,
 $\sigma_g$ is the sigmoid function for the gate.
@@
 $x_{t} \in \RD$ the input vector to the LSTM unit and
-$h_{t} \in \RH$ the hidden state vector e.g., the last of which is the output vector of the LSTM. In our case the output contains the days we predict based on the input features. The equations are
+$h_{t} \in \RH$ the hidden state vector e.g., the last of which is the
+output vector of the LSTM. In our case the output contains the days we
+predict based on the input features. The equations are
 
@@
 $\sigma_c$ is the cell activation function,
-$W\in \mathbb {R} ^{H\times D}$, $U\in \mathbb {R} ^{H\times H}$ and $b\in \RH$ being the weight matrices and bias vector parameters which are learned during training.
-In particular we denote $W_{q}$ and $U_{q}$ contain the weights of the
-input and recurrent connections, where the subscript denotes either
-the input gate $i$, the output gate $o$, the forget gate $f$ or the
-memory cell $c$. For a more detailed discussion about LSTM's we refer to \cite{lstm-explained}.
+$W\in \mathbb {R} ^{H\times D}$,
+$U\in \mathbb {R} ^{H\times H}$ and
+$b\in \RH$ being the weight matrices and bias vector parameters which
+are learned during training.  In particular we denote $W_{q}$ and
+$U_{q}$ contain the weights of the input and recurrent connections,
+where the subscript denotes either the input gate $i$, the output gate
+$o$, the forget gate $f$ or the memory cell $c$. For a more detailed
+discussion about LSTM's we refer to \cite{lstm-explained}.
 
@@
 
-In this paper, we adopt this approach to describe the
-time dependence of COVID-19 infection and fatality counts in different
-regions. 
-We intend to
-process the two time-series (infections, fatalities/deaths) associated with
-each city plus the covariates (fixed in time but dependent on region)
-to build a model for the time evolution of the data. There is a
-nontrivial architecture design choice as to the method of handling the
-time-independent covariates. We could combine a recurrent network for
-the time series with some fixed network for the covariates. However,
-that does not fit with our concept of deriving an evolution operator
-for COVID-19 data that naturally depends on the covariates for each
-city. Thus instead, we feed the covariates into the recurrent network as input in addition to the days.
+In this paper, we adopt this approach to describe the time dependence
+of COVID-19 infection and fatality counts in different regions.  We
+intend to process the two time-series (infections, fatalities/deaths)
+associated with each city plus the covariates (fixed in time but
+dependent on region) to build a model for the time evolution of the
+data. There is a nontrivial architecture design choice as to the
+method of handling the time-independent covariates. We could combine a
+recurrent network for the time series with some fixed network for the
+covariates. However, that does not fit with our concept of deriving an
+evolution operator for COVID-19 data that naturally depends on the
+covariates for each city. Thus instead, we feed the covariates into
+the recurrent network as input in addition to the days.
 
@@
 
-We use data on community-specific health and
-behavioral risk factors aggregated at the county level from the
-large-scale longitudinal CDC Behavioral Risk Factor Surveillance
-System (BRFSS) surveys \citep{www-cdc-brfss} and demographic and
-socioeconomic data from the U.S. Census Bureau \cite{www-sensus} for each county, along with the cases and
-deaths data compiled by the CDC National Center for Health Statistics
-(NCHS) \citep{www-cdc-nchs} and integrate them with other data
-sources such as the American Hospital Directory \citep{www-ahd}.
+We use data on community-specific health and behavioral risk factors
+aggregated at the county level from the large-scale longitudinal CDC
+Behavioral Risk Factor Surveillance System (BRFSS) surveys
+\citep{www-cdc-brfss} and demographic and socioeconomic data from the
+U.S. Census Bureau \cite{www-sensus} for each county, along with the
+cases and deaths data compiled by the CDC National Center for Health
+Statistics (NCHS) \citep{www-cdc-nchs} and integrate them with other
+data sources such as the American Hospital Directory \citep{www-ahd}.
 
@@
 
-We have conducted a number of analysis efforts to evaluate the validity of our approach. We focus on the following questions:
-
-\begin{question} \normalfont Can we apply deep learning strategies that self-learn the behavior of a model in order to provide future predictions? See Section \ref{sec:emperical}.
+We have conducted a number of analysis efforts to evaluate the
+validity of our approach. We focus on the following questions:
+
+\begin{question} \normalfont Can we apply deep learning strategies
+  that self-learn the behavior of a model in order to provide future
+  predictions? See Section \ref{sec:emperical}.
 \label{q:1}
@@
 
-\begin{question} \normalfont Can we use real time data as input and apply deep learning strategies instead of models that self learn and provide future predictions?
-See Section \ref{sec:lstm-covariate}.
+\begin{question} \normalfont Can we use real time data as input and
+  apply deep learning strategies instead of models that self learn and
+  provide future predictions? See Section \ref{sec:lstm-covariate}.
 \label{q:2}
@@
 
-\begin{question} \normalfont Does the inclusion of covariates improve the prediction quality while including geospatial and socio-economic risk factors? See Section \ref{sec:lstm-covariate}.
+\begin{question} \normalfont Does the inclusion of covariates improve
+  the prediction quality while including geospatial and socio-economic
+  risk factors? See Section \ref{sec:lstm-covariate}.
 \label{q:3}
@@
 
-\begin{question} \normalfont Are there some impact factors that influence the prediction more than others? See Section \ref{sec:lstm-covariate}.
+\begin{question} \normalfont Are there some impact factors that
+  influence the prediction more than others? See Section
+  \ref{sec:lstm-covariate}.
 \label{q:4}
@@
 
-To establish the feasibility of using LSTM, we have, in our
-analysis, verified that through deep learning we can recreate the
-predictions identified by a model-based approach using empirical
-fits. In this comprehensive analysis, we obtain predictions through
-deep learning while using data generated from epidemiological models
-as discussed in \citep{marsland20-covid-paper}. The time series we used
-was 100 days long, and a multi-layered LSTM
-recurrent network was used. Our prediction approach differs by
-learning not only from the demographics (fixed data for each city) and
-time-dependent data but by integrating the population model for the
-underlying prediction, as shown in Figure \ref{fig:magic-1}. Such model
-predictive integration capability is important in any application with
-multiple time scales. For example, this allows us to integrate
-multi-scale time effects into the forecast, which could address the
-combination of general forecasts and the next time steps of choice,
-such as days, weeks, months, or longer. For this analysis, we used 37
-of the 110 cities with reliable empirical fits (that we derived from
-an epidemiological model and not deep learning
-\citep{marsland20-covid-paper}) for the cases and deaths data up to
-April 15, 2020. Our analysis identified a sophisticated single deep
-learning time evolution operator that can describe these 37 separate
-data sets, and smooth fitted data leads to very accurate deep learning
-descriptions via LSTM as depicted in Figure \ref{fig:magic-1}
-that are smaller than 1\% from the model predictions.
+To establish the feasibility of using LSTM, we have, in our analysis,
+verified that through deep learning we can recreate the predictions
+identified by a model-based approach using empirical fits. In this
+comprehensive analysis, we obtain predictions through deep learning
+while using data generated from epidemiological models as discussed in
+\citep{marsland20-covid-paper}. The time series we used was 100 days
+long, and a multi-layered LSTM recurrent network was used. Our
+prediction approach differs by learning not only from the demographics
+(fixed data for each city) and time-dependent data but by integrating
+the population model for the underlying prediction, as shown in Figure
+\ref{fig:magic-1}. Such model predictive integration capability is
+important in any application with multiple time scales. For example,
+this allows us to integrate multi-scale time effects into the
+forecast, which could address the combination of general forecasts and
+the next time steps of choice, such as days, weeks, months, or
+longer. For this analysis, we used 37 of the 110 cities with reliable
+empirical fits (that we derived from an epidemiological model and not
+deep learning \citep{marsland20-covid-paper}) for the cases and deaths
+data up to April 15, 2020. Our analysis identified a sophisticated
+single deep learning time evolution operator that can describe these
+37 separate data sets, and smooth fitted data leads to very accurate
+deep learning descriptions via LSTM as depicted in Figure
+\ref{fig:magic-1} that are smaller than 1\% from the model
+predictions.
 
@@
       \centering
-      \includegraphics[width=1.0\linewidth]{images/predict/Fig4a.pdf}  
+      \includegraphics[width=1.0\linewidth]{images/Fig4a.pdf}  
       \caption{Prediction accuracy for deaths.}
@@
       \centering
-      \includegraphics[width=1.0\linewidth]{images/predict/Fig4b.pdf}        
+      \includegraphics[width=1.0\linewidth]{images/Fig4b.pdf}        
       \caption{Prediction accuracy for cases.}
@@
 In this section, we establish that using deep learning can not only be
-applied to model predictions but also from real time data. As we want to
-evaluate the influence of the risk factors on our prediction framework
-using not only models but deep learning algorithms, we have devised a
-parameter sweep framework based on LSTM that conducts the prediction
-for 110 cities while also considering the risk factors as listed
-in Table \ref{tab:risk-factors}. This framework can query any
+applied to model predictions but also from real time data. As we want
+to evaluate the influence of the risk factors on our prediction
+framework using not only models but deep learning algorithms, we have
+devised a parameter sweep framework based on LSTM that conducts the
+prediction for 110 cities while also considering the risk factors as
+listed in Table \ref{tab:risk-factors}. This framework can query any
 combination of risk factors as well as not including any, in which
@@
 sigmoid is chosen. A drop out rate of 0.2 is used, we use 16 input
-nodes, and have a batch size of 110. The  
-batch size is the subset size of the training sample (e.g. 110 out of 12540 for a 3 day input length) that is used to train the LSTM during its learning process. The maximum epoch is chosen to be
-200. This is the number of steps after we interrupt the learning process. The number of nodes internally is 32. The maximum total number of
-samples is 12540 when using 3 days as input length. We use a two-layer
-LSTM, as shown in Table \ref{tab:model}. Within our system, we have
-the ability to set a number of input days. For this analysis, we have
-chosen three different input lengths, namely 3, 4, and 5 days. The
-term ``None'' in the output shape means this dimension is
-variable.
+nodes, and have a batch size of 110. The batch size is the subset size
+of the training sample (e.g. 110 out of 12540 for a 3 day input
+length) that is used to train the LSTM during its learning
+process. The maximum epoch is chosen to be 200. This is the number of
+steps after we interrupt the learning process. The number of nodes
+internally is 32. The maximum total number of samples is 12540 when
+using 3 days as input length. We use a two-layer LSTM, as shown in
+Table \ref{tab:model}. Within our system, we have the ability to set a
+number of input days. For this analysis, we have chosen three
+different input lengths, namely 3, 4, and 5 days. The term ``None'' in
+the output shape means this dimension is variable.
 
@@
     \centering
-    \includegraphics[width=1.0\textwidth]{images/boxwhisker/boxplot_casesNEW2.pdf}
+    \includegraphics[width=1.0\textwidth]{images/boxplot_casesNEW2.pdf}
     \vspace{-1cm}
@@
 
-    \includegraphics[width=1.0\textwidth]{images/boxwhisker/boxplot_deathNEW2.pdf}
+    \includegraphics[width=1.0\textwidth]{images/boxplot_deathNEW2.pdf}
     \vspace{-1cm}
@@
 We show the errors and the association with a risk factors in Table
- \ref{tab:top-1-cases} and \ref{tab:top-1-deaths}.
-In the tables we include the following columns (a) the Risk Factor (b) the root mean square error of all models for this risk factor (c) the minimum cumulative error by risk factor and (d) the number of days that wer used as input and identified to relate to the model with the minimum cumulative error.
-The table has in the first column the rank this particular model comperes to all other models with minimal risk factor.
-The rank was obtained by sorting the best model for each risk factor. 
-We display two tables. One for cases (Table \ref{tab:top-1-cases}) and one for deaths (Table \ref{tab:top-1-deaths}).
-
-We see that in our experiment, almost all risk factors lead to an increased
-model prediction accuracy for cases.  We note that by including
-factors such as population density, physical health Insurance,
-population breakdown by ethnicity, number of hospitals, and diabetes,
-lead to better overall predictions for cases. 
-We also see that
-many of these factors perform better on average.
+\ref{tab:top-1-cases} and \ref{tab:top-1-deaths}. In the tables we
+include the following columns (a) the Risk Factor (b) the root mean
+square error of all models for this risk factor (c) the minimum
+cumulative error by risk factor and (d) the number of days that wer
+used as input and identified to relate to the model with the minimum
+cumulative error. The table has in the first column the rank this
+particular model comperes to all other models with minimal risk
+factor.  The rank was obtained by sorting the best model for each risk
+factor.  We display two tables. One for cases (Table
+\ref{tab:top-1-cases}) and one for deaths (Table
+\ref{tab:top-1-deaths}).
+
+We see that in our experiment, almost all risk factors lead to an
+increased model prediction accuracy for cases.  We note that by
+including factors such as population density, physical health
+Insurance, population breakdown by ethnicity, number of hospitals, and
+diabetes, lead to better overall predictions for cases.  We also see
+that many of these factors perform better on average.
 
@@
 identified three risk factors, namely the Social Vulnerability Index
-(SVI), chain-smoking, and Hispanic that, when integrated into our deep learning
-model, leads to better prediction results. However, the accuracy of
-the overall prediction of deaths is far less precise than that of the
-cases. 
-To showcase this, we have included Figures
+(SVI), chain-smoking, and Hispanic that, when integrated into our deep
+learning model, leads to better prediction results. However, the
+accuracy of the overall prediction of deaths is far less precise than
+that of the cases.  To showcase this, we have included Figures
 \ref{fig:error-case-popdensity} and \ref{fig:error-death-popdensity}
-that show the respective errors for example for the population density.
+that show the respective errors for example for the population
+density.
 
@@
     \centering
-    \includegraphics[width=1.0\textwidth]{images/predict/Length_5-fields_pop_density_2010-Cumulative-Cases.pdf}
+    \includegraphics[width=1.0\textwidth]{images/Length_5-fields_pop_density_2010-Cumulative-Cases.pdf}
     %\vspace{-1cm}
@@
     \centering
-    \includegraphics[width=1.0\textwidth]{images/predict/Length_5-fields_pop_density_2010-Cumulative-Death.pdf}
+    \includegraphics[width=1.0\textwidth]{images/Length_5-fields_pop_density_2010-Cumulative-Death.pdf}
     %\vspace{-1cm}
@@
 
-This can be explained by our input data distribution between cases and deaths in which we noted higher
-fluctuations relative to the overall value in cases of deaths. 
-A mitigation to this issue
-would be using a seven day average over the analyzed period. However,
-we have not done this on purpose for this analysis as we wanted to
-identify how a covariate enhanced LSTM behaves that includes risk
-factors based on the daily fluctuations. This was done to avoid and
-showcase any needed preproccessing on the data and to identify the
-capability of the deep learning framework while adapting to the
-fluctuations without any special activities. Previously, we have
-already shown in Section \ref{sec:emperical} that for smooth data
-inputs the prediction is very accurate.  Using fluctuation data allows
-our framework for an automated ingest of data on a daily basis to
-enable data fusion of new cases and deaths information. Hence, we can
-re-train the model with new incoming data every day. Through repeated
-training experiments, we found the LSTM hyperparameters such as epochs
-and dropout values that work well for our experiments.
+This can be explained by our input data distribution between cases and
+deaths in which we noted higher fluctuations relative to the overall
+value in cases of deaths.  A mitigation to this issue would be using a
+seven day average over the analyzed period. However, we have not done
+this on purpose for this analysis as we wanted to identify how a
+covariate enhanced LSTM behaves that includes risk factors based on
+the daily fluctuations. This was done to avoid and showcase any needed
+preproccessing on the data and to identify the capability of the deep
+learning framework while adapting to the fluctuations without any
+special activities. Previously, we have already shown in Section
+\ref{sec:emperical} that for smooth data inputs the prediction is very
+accurate.  Using fluctuation data allows our framework for an
+automated ingest of data on a daily basis to enable data fusion of new
+cases and deaths information. Hence, we can re-train the model with
+new incoming data every day. Through repeated training experiments, we
+found the LSTM hyperparameters such as epochs and dropout values that
+work well for our experiments.
 
@@
 
-% In case of cases we also see a drop in the better
-% model predictions.
-
 \begin{figure}[!htbp]
@@
         \centering
-        \includegraphics[width=1.0\textwidth]{images/predict/PlaceTop10_DeathNEW3.pdf}
+        \includegraphics[width=1.0\textwidth]{images/PlaceTop10_DeathNEW3.pdf}
         \caption{Cumulative error for deaths.}
@@
         \centering
-        \includegraphics[width=1.0\textwidth]{images/predict/PlaceTop10_CasesNEW3.pdf}
+        \includegraphics[width=1.0\textwidth]{images/PlaceTop10_CasesNEW3.pdf}
         \caption{Cumulative error for cases.}
@@
 \begin{enumerate}
-    \item The entire cloudmesh code on which the cloud based implementation of the AICov framework is based and contains over 70 contributors is available publicly at \url{https://github.com/cloudmesh}.
-Cloudmesh contains a number of modules that dependent on the users access to cloud resources can be customized.
-A detailed manual about the configuration is available at \url{https://cloudmesh.github.io/cloudmesh-manual/}.
-
-    \item The entire COVID-19 analysis leverages cloudmesh and uses Jupyter notebooks to coordinate its workflow as discussed in the architecture Figure \ref{fig:arch}.
-The code and data for the results presented in this paper are located in the repository at \url{https://github.com/cloudmesh/cloudmesh-covid}.
-
-    The data was analysed on a variety of supercomputing resources including an allocation of 20 compute nodes that were utilized to execute the  repeated model creation to assure reproducible results.
-
-    However, as a publication is pending, at this time, the use of the data is copyrighted and can not be used for other publications without contacting the authors.
-The data gathering and analysis is a significant intellectual contribution and we like to avoid that the data is taken before we have not secured a publication.
-
-    \item The entire paper is located in \LaTeX source in the github repository \url{https://github.com/cyberaide/paper-covid}.
-This repository will be open sourced after acceptance of publication to not violate any publisher restrictions.
-If desired the authors can grant access to this repository prior to publication.
-Please contact the corresponding author.
+
+\item The entire cloudmesh code on which the cloud based
+  implementation of the AICov framework is based and contains over 70
+  contributors is available publicly at
+  \url{https://github.com/cloudmesh}. Cloudmesh contains a number of
+  modules that dependent on the users access to cloud resources can be
+  customized. A detailed manual about the configuration is available
+  at \url{https://cloudmesh.github.io/cloudmesh-manual/}.
+
+\item The entire COVID-19 analysis leverages cloudmesh and uses
+  Jupyter notebooks to coordinate its workflow as discussed in the
+  architecture Figure \ref{fig:arch}. The code and data for the
+  results presented in this paper are located in the repository at
+  \url{https://github.com/cloudmesh/cloudmesh-covid}.
+
+  The data was analysed on a variety of supercomputing resources
+  including an allocation of 20 compute nodes that were utilized to
+  execute the repeated model creation to assure reproducible results.
+
+  However, the use of the data is copyrighted and must be authorized
+  to be used for other publications without contacting the
+  authors. The data gathering and analysis is a significant
+  intellectual contribution and we like to avoid that the data is
+  taken before we have not secured a publication.
+
+\item The entire paper is located in \LaTeX source in the github
+  repository \url{https://github.com/cyberaide/paper-covid}. This
+  repository will be open sourced after acceptance of publication to
+  not violate any publisher restrictions. If desired the authors can
+  grant access to this repository prior to publication. Please contact
+  the corresponding author.
 
@@
 
-%\clearpage
-%\singlespace
-%\maketitle
-%\input{response-v3-content}
-
-\clearpage
-\input{response-v4-content}
-
 \end{document}

### 2021-01-28 final/vonLaszewski-covid-jds.tex
rename file anme with postfix final
fix WHO bibtex and shorten taitle for page header

--- 2021-01-27 final/vonLaszewski-covid.tex
@@
 
-\title[AICov: Deep
-    Learning Framework for COVID-19 Forecasting with Population
+\title[AICov: DL Framework for COVID-19 Forecasting with Population
     Covariates]{\TITLE}
@@
 integrator
-\citep{cloudmesh-openapi-install,cloudmesh-openapi-benchmark,cloudmesh-manual}.
+\citep{las-19-openapi-function,cloudmesh-openapi-install,cloudmesh-manual}.
 Additionally, we can also access through abstractions cloud services
